### 2018-01-29 abstract.tex
29Jan

% ==============================================================================
% DOCUMENT HEADER
% ==============================================================================

\documentclass[11t, a4paper, twocolumn]{article} 
\input{structure.tex}

\title{Estimating real-time highstreet footfall from the Wi-Fi probe requests}
\author{
	\authorstyle{
		Balamurugan Soundararaj\textsuperscript{1}, 
		James Cheshire\textsuperscript{1} and 
		Paul Longley\textsuperscript{1}}
	\newline\newline
	\textsuperscript{1}\institution{
		Department of Geography, 
		University College London, 
		United Kingdom}
}
\date{\today}

% ==============================================================================
% DOCUMENT BODY
% ==============================================================================

\begin{document}

	% ---------------------------------------------------------------------------
	% We talk about wifi and probe requests. Possibility of 
	% counting people without revealing identity. There is uncertainity. We look
	% at resolving these. We have a setup to collect data from sensor and manual
	% We propose methodology to clean sensor count. compare with manual to see
	% it corresponds properly.
	% ---------------------------------------------------------------------------

	\maketitle
	\thispagestyle{firstpage}

	\textbf{Abstract}

	Wi-Fi has become an ubiquitous technology used to provide internet access in
	public and private spaces to people's mobile devices such as smartphones, 
	tablets and laptops. Any point in a dense built environment such as cities 
	has multiple Wi-Fi networks. Anticipating and remembering these networks
	mobile devices, to be able to switch seamlessly between them,
	broadcast a special type of management packets known as probe requests.
	This is a low level package specified in IEEE 802.1 specification
	which relays information about the source mobile device to
	any Access Points (AP) listening to them.
	This often the first steo in establishing a handshake between these devices,
	These probe requests are constinuous, passive and wireless stream of
	infromation available at any urban location which can act as a proxy to
	understanding the number of people present in that area in real-time.
	In this paper, utilising a set of probe requests collected
	at a highstreet location, along with manually collected data, we demonstrate
	that highstreet footfall can be estimated with reasonable accuracy without
	infringing on people's privacy.

	Use of wireless technology to measure movement of people in built
	environment has been done previously. It is either done at mobile
	level or network level.
	At network level it can be done with recording mobile devices coming in
	and out of network or measuing mobile devices recording aps.
	These can be done with various technologies such as gps, mobile network
	where some of them are more accurate than others.
	In this Wi-Fi provides us a good middle ground - interms of cost,
	scalability,  etc. The major disadvantage is that the unique identifier
	in wifi based counting is the MAC address which is global and can 
	cause privacy risk. This can be solved by hashing the personal data
	collected. Recently mobile devices have also started randomising their
	mac to avoid detection and tracking.
	Though this can be over come by using various techniques, cite papers,
	these methods are usually infringing on privacy.
	We need a method to count people using this data without tracking them
	which is the major challange in which we address here.

	The data collection is done through tshark, data collected were - mac,
	vendor, length, ssid, tags, signal, duration. - manual count was done in
	parallel through mobile phone app. The survey was carried out at oxford st.
	from 12:30 to 13:00. The total numbers. how many people. How many probes?
	how many unique macs? what are the uncertainities in the

	internet access to mobile devices such as smartphones,tablets and laptops.
	In addition to internet these devices use these networks as a way to quick 
	geo location without waiting for GPS.
	To produce a list of WiFi networks available, mobile devices constantly 
	broadcast signals called probe requests.
	These signals have various information regarding the device and its 
	capabilities which are identifiable such as MAC addresses, non identifiable 
	such as signal strength and partly identifiable such as randomised mac 
	addresses.
	There is a significant use in uniquely identify the number of mobile devices
	at a specific location wihout actually affecting the privacy of the mobile
	devices.
	Such enumeration can be done by looking at identifiable information after 
	anonymising them whenever they are available and from the patterns in the 
	other information when they are not.
	We collect data on probe requests sent at set of locations in 
	different times and device a methodology for enumerating number of people
	at these locations from the data.
	We compare the enumerated counts with the counts collected manually at
	these locations to find that we are able to estimate real life with a 
	confidence of XX\%.

	\citep{vanhoef2016}
	\citep{matte2016}
	\citep{martin2017}
	\citep{vo2016}
	\citep{konto2017}

	\printbibliography[title={References}]

\end{document}

### 2018-01-29 abstract.tex
first draft

@@
 
-	Wi-Fi has become an ubiquitous technology used to provide internet access in
+	Wi-Fi has become an ubiquitous technology used in provinding internet access in
 	public and private spaces to people's mobile devices such as smartphones, 
-	tablets and laptops. Any point in a dense built environment such as cities 
-	has multiple Wi-Fi networks. Anticipating and remembering these networks
-	mobile devices, to be able to switch seamlessly between them,
+	tablets and laptops. This has resulted in a situation
+	where any point in a dense urban built environment 
+	has multiple Wi-Fi networks available. Modern mobile devices anticipate and 
+	remember these networks and to be able to switch seamlessly between them,
 	broadcast a special type of management packets known as probe requests.
-	This is a low level package specified in IEEE 802.1 specification
+	This is a low level signal (probe request) specified in IEEE 802.11b/g specification
 	which relays information about the source mobile device to
-	any Access Points (AP) listening to them.
-	This often the first steo in establishing a handshake between these devices,
-	These probe requests are constinuous, passive and wireless stream of
-	infromation available at any urban location which can act as a proxy to
-	understanding the number of people present in that area in real-time.
-	In this paper, utilising a set of probe requests collected
-	at a highstreet location, along with manually collected data, we demonstrate
-	that highstreet footfall can be estimated with reasonable accuracy without
-	infringing on people's privacy.
+	any Access Points (AP) listening around them.
+	This is the first step in establishing a connection between these devices and
+	is universal to any device which has a Wi-Fi radio to communicate.
+	This makes these probe requests a continuous, passive and wireless source of
+	data available at any urban location which can be a clue in
+	understanding the number of people present in that area in real-time (\citep{freud2015},\citep{konto2017}).
+	In this paper, from a set of probe requests collected
+	at a highstreet location in London, along with manually collected data, we demonstrate
+	that pedestrian footfall can be estimated with reasonable accuracy without
+	infringing on the privacy of the mobile users.
 
-	Use of wireless technology to measure movement of people in built
-	environment has been done previously. It is either done at mobile
-	level or network level.
-	At network level it can be done with recording mobile devices coming in
-	and out of network or measuing mobile devices recording aps.
-	These can be done with various technologies such as gps, mobile network
-	where some of them are more accurate than others.
-	In this Wi-Fi provides us a good middle ground - interms of cost,
-	scalability,  etc. The major disadvantage is that the unique identifier
-	in wifi based counting is the MAC address which is global and can 
-	cause privacy risk. This can be solved by hashing the personal data
-	collected. Recently mobile devices have also started randomising their
-	mac to avoid detection and tracking.
-	Though this can be over come by using various techniques, cite papers,
-	these methods are usually infringing on privacy.
-	We need a method to count people using this data without tracking them
-	which is the major challange in which we address here.
+	\begin{figure*}
+		\begin{center}
+	  \includegraphics[width=0.8\textwidth]{outputs/clustering_1.png}
+	  \caption{Clustering probe requests based on increasing sequence numbers generated over time.}
+		\end{center}
+	\end{figure*}
 
-	The data collection is done through tshark, data collected were - mac,
-	vendor, length, ssid, tags, signal, duration. - manual count was done in
-	parallel through mobile phone app. The survey was carried out at oxford st.
-	from 12:30 to 13:00. The total numbers. how many people. How many probes?
-	how many unique macs? what are the uncertainities in the
+	There have been numerous attempts at using Wi-Fi
+	to measure the volume and movement of people in built environment for various
+	applications (\citep{zarim2006},\citep{sap2015},\citep{reki2007}).
+	Though favorable results have been observed, one of the major 
+	challanges been the changes in the MAC address randomisation process which 
+	aims to protect the users' privacy by anonymising the only globally identifiable
+	portion of the probe requests \citep{green2008}. There have been various 
+	successful research in breaking this
+	randomisation process to extract real MAC addresses \citep{martin2017}
+	but this usually results in serious infringement on the user's
+	privacy. There is a clear gap in the research for
+	exploring methodologies with which we can estimate the
+	number of unique mobile devices at a location without the need to fingerprint
+	them using their MAC address which we try to address in this paper.
 
-	internet access to mobile devices such as smartphones,tablets and laptops.
-	In addition to internet these devices use these networks as a way to quick 
-	geo location without waiting for GPS.
-	To produce a list of WiFi networks available, mobile devices constantly 
-	broadcast signals called probe requests.
-	These signals have various information regarding the device and its 
-	capabilities which are identifiable such as MAC addresses, non identifiable 
-	such as signal strength and partly identifiable such as randomised mac 
-	addresses.
-	There is a significant use in uniquely identify the number of mobile devices
-	at a specific location wihout actually affecting the privacy of the mobile
-	devices.
-	Such enumeration can be done by looking at identifiable information after 
-	anonymising them whenever they are available and from the patterns in the 
-	other information when they are not.
-	We collect data on probe requests sent at set of locations in 
-	different times and device a methodology for enumerating number of people
-	at these locations from the data.
-	We compare the enumerated counts with the counts collected manually at
-	these locations to find that we are able to estimate real life with a 
-	confidence of XX\%.
+	The pilot survey was conducted on Oxford st. in London on 20 December 2017 from 12:30 to
+	13:00. The data was collected parallely through two methods - Wi-Fi sensor and Android app.
+	The Wi-Fi sensor collected all the probe requests broadcated around the area
+	and recorded the timestamp at which they were colleted, MAC address of the source
+	device anonymised using a hashing algorithm, organisationally unique identifier (OUI) of the
+	vendor who manufactured the device, total length of the singal in bits, signal
+	strength reported by the device in dBm, duration for which the signal was tranmmitted, 
+	the service set identifier (SSID) to which the probe request has been sent to and the
+	length of the extra tags in the packets. The android app recorded the timestamp
+	of every touch on the screen by the surveyor which corresponds to a pedestrian 
+	crossing a cordon on the sidewalk.
 
-	\citep{vanhoef2016}
-	\citep{matte2016}
-	\citep{martin2017}
-	\citep{vo2016}
-	\citep{konto2017}
+	Approximately 60,000 probe requests were collected out of which   \% had
+	randomised MAC addresses thus effectively anonymised.
+	When we include these anonymised probe requests and count the number of unique
+	devices present in dataset, we find that the difference between manual counting and sensor based counting
+	is 425\% on average thus showing the need for using a different fingerprint other than MAC address for the
+	anonymised probe requests. We first filter the data by removing all the probe requests reporting
+	low signal strength. This classification is done using a kmeans - class interval algorithm.
+	This reduces the difference from the manual count to 30\%.
+
+	Then we turn our attention to other fields which can provide us information
+	on the device which generated the probe request and hence enabling us to 
+	finger print them uniquely.
+	From a preliminary investigation, we find that the data fields - SSID and length of
+	the tags are very sparse and not useful in identifying the unique device while 
+	OUIs, length of the packet and sequence numbers can be crucial in doing the same.
+	A pair of probes with the same OUI shows the device manufacturer and same length of the packet signifies that 
+	there is a high probability that it was generated by a similar devcie.
+
+	We first split the anonymised probe requests by the OUI and length then run a 
+	graph based partition algorithm on them based on the sequence numbers. This 
+	algortithm links probes with increasing sequence numbers in increasing time within a
+	specific distance. It also enforces the rule that a probe request has only one 
+	incoming and outgoing links. It then returns the membership of the probe requests in
+	distinct connected clusters in the resulting graph thus classifying probe requests
+	with sequentially increasing numbers as ones generated from the same device as shown in Figure 1. Finally
+	all the individual subsets are joined and cleaned for repeating probe requests with
+	unique clusters. This filtering in-turn reduces the difference of the sensor count to the manual count to 
+	-18\%.
+
+	The important detail to notice is that we have clustered and estimated unique device
+	without the knowledge of the MAC address of the devices hence delivering output 
+	without compromising the users' privacy. The methodology can be applications in 
+	numerous footfall counting projects in retail, urban planning, facilities management etc.
 

### 2018-01-30 abstract.tex
second draft

@@
 
-\title{Estimating real-time highstreet footfall from the Wi-Fi probe requests}
+\title{Estimating real-time high street footfall from Wi-Fi probe requests}
 \author{
@@
 
-	% ---------------------------------------------------------------------------
-	% We talk about wifi and probe requests. Possibility of 
-	% counting people without revealing identity. There is uncertainity. We look
-	% at resolving these. We have a setup to collect data from sensor and manual
-	% We propose methodology to clean sensor count. compare with manual to see
-	% it corresponds properly.
-	% ---------------------------------------------------------------------------
-
 	\maketitle
@@
 
-	\textbf{Abstract}
+	\section{Abstract}
 
-	Wi-Fi has become an ubiquitous technology used in provinding internet access in
-	public and private spaces to people's mobile devices such as smartphones, 
-	tablets and laptops. This has resulted in a situation
-	where any point in a dense urban built environment 
-	has multiple Wi-Fi networks available. Modern mobile devices anticipate and 
-	remember these networks and to be able to switch seamlessly between them,
-	broadcast a special type of management packets known as probe requests.
-	This is a low level signal (probe request) specified in IEEE 802.11b/g specification
-	which relays information about the source mobile device to
-	any Access Points (AP) listening around them.
-	This is the first step in establishing a connection between these devices and
-	is universal to any device which has a Wi-Fi radio to communicate.
-	This makes these probe requests a continuous, passive and wireless source of
-	data available at any urban location which can be a clue in
-	understanding the number of people present in that area in real-time (\citep{freud2015},\citep{konto2017}).
+	In the past decade Wi-Fi has emerged as the most
+	commonly used technology in providing 
+	internet access to mobile devices such as smartphones, tablets
+	and laptops in public and private spaces.
+	This has resulted in multiple Wi-Fi networks being available at
+	almost every location in an urban environment.
+	Traversing through this overlapping mesh of Wi-Fi networks,
+	modern mobile devices with Wi-Fi antennae regularly
+	broadcast special type of signals known as probe requests to
+	discover available Wi-Fi networks and switch seamlessly
+	between them.
+	This is a hardware level signal which was standardised by IEEE 802.11b/g
+	specification relays information about the source mobile device to
+	any Access Points (AP) available around it.
+	Since this is the first step in establishing
+	a connection between any two devices, it is
+	is universal to any device which uses a Wi-Fi radio to communicate.
+	This makes these probe requests
+	an open, passive, continuous, and wireless source of
+	data available at any urban location which can provide us with clues in
+	understanding the number of people present in the immediate surroundings
+	in real-time with high granularity \citep{freud2015,konto2017}.
 	In this paper, from a set of probe requests collected
-	at a highstreet location in London, along with manually collected data, we demonstrate
-	that pedestrian footfall can be estimated with reasonable accuracy without
-	infringing on the privacy of the mobile users.
+	at a high street location in London,
+	along with manually counted data, we demonstrate
+	that pedestrian footfall can be estimated with considerable accuracy without
+	infringing on the privacy of the mobile users involved.
 
@@
 			\includegraphics[width=0.8\textwidth]{outputs/clustering_1.png}
-	  \caption{Clustering probe requests based on increasing sequence numbers generated over time.}
+			\caption{Clustering probe requests based on increasing sequence
+			numbers present in them.}
 		\end{center}
@@
 	to measure the volume and movement of people in built environment for various
-	applications (\citep{zarim2006},\citep{sap2015},\citep{reki2007}).
-	Though favorable results have been observed, one of the major 
-	challanges been the changes in the MAC address randomisation process which 
-	aims to protect the users' privacy by anonymising the only globally identifiable
-	portion of the probe requests \citep{green2008}. There have been various 
-	successful research in breaking this
+	applications \citep{zarim2006,sap2015,reki2007}.
+	Though most research observe feasible and favorable results,
+	in recent years, one of the major challenges faced in such attempts
+	has been the MAC address randomisation process.
+	This process aims to protect the users' privacy by anonymising
+	the only globally identifiable portion of the probe requests
+	resulting in a set of probe requests generated by the same device
+	with different random MAC addresses \citep{green2008}.
+	There have been various successful research in breaking this
 	randomisation process to extract real MAC addresses \citep{martin2017}
-	but this usually results in serious infringement on the user's
-	privacy. There is a clear gap in the research for
-	exploring methodologies with which we can estimate the
-	number of unique mobile devices at a location without the need to fingerprint
-	them using their MAC address which we try to address in this paper.
+	but this usually results in serious risk in infringement of
+	privacy of the users of the mobile devices.
+	There is a clear gap in the research for exploring methodologies
+	which enables us to estimate the number of unique mobile
+	devices from a set of anonymised probe requests
+	without the need to reveal their original MAC addresses.
 
-	The pilot survey was conducted on Oxford st. in London on 20 December 2017 from 12:30 to
-	13:00. The data was collected parallely through two methods - Wi-Fi sensor and Android app.
-	The Wi-Fi sensor collected all the probe requests broadcated around the area
-	and recorded the timestamp at which they were colleted, MAC address of the source
-	device anonymised using a hashing algorithm, organisationally unique identifier (OUI) of the
-	vendor who manufactured the device, total length of the singal in bits, signal
-	strength reported by the device in dBm, duration for which the signal was tranmmitted, 
-	the service set identifier (SSID) to which the probe request has been sent to and the
-	length of the extra tags in the packets. The android app recorded the timestamp
-	of every touch on the screen by the surveyor which corresponds to a pedestrian 
-	crossing a cordon on the sidewalk.
+	A pilot survey was conducted at Oxford street in London
+	on 20 December 2017 from 12:30 to 13:00 where two sets of data
+	were collected on pedestrian footfall through Wi-Fi sensing
+	and manual counting in parallel.
+	The Wi-Fi sensor collected all the probe requests that were 
+	broadcast around the area and recorded
+	the time-stamp at which they were collected,
+	MAC address of the source mobile device
+	(anonymised using a hashing algorithm),
+	organisationally unique identifier (OUI) of the manufacturer of the device,
+	total length of the signal in bits,
+	the strength of the signal reported by the mobile device in dBm,
+	the sequence number of the signal,
+	duration for which the signal was transmitted, 
+	the service set identifier (SSID) of the access point targeted
+	by the probe request and 
+	the length of the extra information (tags) embedded in the packets.
+	The manual count was done by an android application on a mobile phone
+	which records the time-stamp every time the surveyor touches the screen
+	which corresponds to one pedestrian footfall on the sidewalk.
 
-	Approximately 60,000 probe requests were collected out of which   \% had
-	randomised MAC addresses thus effectively anonymised.
-	When we include these anonymised probe requests and count the number of unique
-	devices present in dataset, we find that the difference between manual counting and sensor based counting
-	is 425\% on average thus showing the need for using a different fingerprint other than MAC address for the
-	anonymised probe requests. We first filter the data by removing all the probe requests reporting
-	low signal strength. This classification is done using a kmeans - class interval algorithm.
-	This reduces the difference from the manual count to 30\%.
+	Being one of the busiest location in the United Kingdom, the location
+	generated approximately 60,000 probe requests in the 30 minutes
+	interval through the Wi-Fi sensor and 3,722 people were counted manually.
+	When we aggregate the probe requests by their MAC address for every minute,
+	the difference between the sensor counts and the manual counts
+	is observed to be on average 425\%. 
+	This shows that there is a large amount of noise
+	in the data which might include signals from devices outside the area where
+	the manual count was conducted and the anonymised probe requests from the
+	same devices with different MAC addresses.
+	Before we go look at the data
+	in detail, an initial analysis shows that the fields - SSID and tags are
+	very sparse and doesn't provide much information for our cleaning process 
+	and the duration field is closely related to the length of the 
+	probe request and provides no new information.
+	So we remove these fields from further analysis.
+
+	We remove the noise from devices outside the area of interest 
+	by removing all the probe requests which report a "low" signal strength.
+	This classification of "high" vs "low" is done through "k-means"
+	classification algorithm.
+	The cutoff point for the collected data is -71dBm.
+	This process of filtering is highly effective and reduces the difference
+	between the sensor counts and manual counts to 30\%. We observe that
+	around 55\% of all probe requests collected are anonymised.
+	We assign the hashed MAC address the unique identifier for the rest of
+	the 45\% and investigate the anonymised probe requests further.
 

### 2018-01-30 abstract.tex
update

@@
 
+	We then use the fields - OUI, lengths and sequence number,
+	to tackle the noise from devices which anonymising the probe requests.
+	OUI and length are used to split the dataset into
+	groups probe requests from similar
+	devices and each subset is classified further
+	based on a graph based clustering algorithm 
+	where each cluster correspond to an unique device.
+	The algorithm works by creating a graph where the nodes are the
+	probe requests and the links are created between them based on the following
+	rules,
+	A link can go only forward in time.
+	A link can exist between nodes with a 
+		maximum time difference of t (time threshold)
+	A link can go from low to high sequence numbers.
+	A link can exist between nodes with a
+		maximum sequence number difference of s (sequence threshold)
+	A node can have only one incoming link and outgoing link
+		which is the shortest of all possible such links.
+	The nodes are then classified based on the unique connected component they
+	belong to.
+
 	Then we turn our attention to other fields which can provide us information

### 2018-01-30 abstract.tex
final draft for paul and james

--- 2018-01-30 abstract.tex
@@
 	\thispagestyle{firstpage}
-
-	\section{Abstract}
 
@@
 		\begin{center}
-			\includegraphics[width=0.8\textwidth]{outputs/clustering_1.png}
+			\includegraphics
+				[width=0.8\textwidth,trim=4 4 4 4,clip]
+				{outputs/clustering_2.png}
 			\caption{Clustering probe requests based on increasing sequence
 			numbers present in them.}
+			label{fig}
 		\end{center}
@@
 
-	Then we turn our attention to other fields which can provide us information
-	on the device which generated the probe request and hence enabling us to 
-	finger print them uniquely.
-	From a preliminary investigation, we find that the data fields - SSID and length of
-	the tags are very sparse and not useful in identifying the unique device while 
-	OUIs, length of the packet and sequence numbers can be crucial in doing the same.
-	A pair of probes with the same OUI shows the device manufacturer and same length of the packet signifies that 
-	there is a high probability that it was generated by a similar devcie.
+	We then use the fields - OUI, lengths and sequence number,
+	to tackle the noise from devices which anonymising the probe requests.
+	OUI and length are used to split the dataset into
+	groups of probe requests from similar
+	devices and each subset is classified further
+	based on a graph based clustering algorithm 
+	where each cluster corresponds to an unique device.
+	The algorithm works by creating a graph where the probe requests
+	are the nodes and the links are created
+	between them based on the following rules,
 
-	We first split the anonymised probe requests by the OUI and length then run a 
-	graph based partition algorithm on them based on the sequence numbers. This 
-	algortithm links probes with increasing sequence numbers in increasing time within a
-	specific distance. It also enforces the rule that a probe request has only one 
-	incoming and outgoing links. It then returns the membership of the probe requests in
-	distinct connected clusters in the resulting graph thus classifying probe requests
-	with sequentially increasing numbers as ones generated from the same device as shown in Figure 1. Finally
-	all the individual subsets are joined and cleaned for repeating probe requests with
-	unique clusters. This filtering in-turn reduces the difference of the sensor count to the manual count to 
-	-18\%.
+	\begin{enumerate}
+		\item A link can go only forward in time.
+		\item A link can exist between nodes with a 
+			maximum time difference of $\alpha$ (time threshold)
+		\item A link can go from low to high sequence numbers.
+		\item A link can exist between nodes with a
+			maximum sequence number difference of $\beta$ (sequence threshold)
+		\item A node can have only one incoming link and outgoing link
+			which is the shortest of all possible such links.
+	\end{enumerate}
 
-	The important detail to notice is that we have clustered and estimated unique device
-	without the knowledge of the MAC address of the devices hence delivering output 
-	without compromising the users' privacy. The methodology can be applications in 
-	numerous footfall counting projects in retail, urban planning, facilities management etc.
+	The nodes are then classified based on the unique connected component they
+	belong to. This classification is assigned as the unique identifier for the
+	anonymised probe requests.
+	\ref{fig} shows the clustering process for where the black dots show
+	probe requests and red lines connecting them show the clusters
+	corresponding to the same device generated them.
+	We finally combine both normal and anonymised probe requests, 
+	aggregate them based on their unique identifier and remove
+	repeating probe requests which reduces the difference between
+	the sensor counts and the manual counts to -18\%.
+
+	It is important to note that the filtering process is done based on just
+	the information present in the probe requests
+	and their temporal distribution.
+	This makes sure that though the mobile devices are uniquely identified
+	there is no more personal data is generated linking the probe requests
+	to the users of the mobile devices. This method essentially gives us a
+	way to estimate the footfall in real-time without identifying or tracking the
+	mobile devices themselves.
+
+	This Wi-Fi based footfall counting methodology offers a large number of
+	applications and benefits. Since Wi-Fi based sensors are inexpensive and the
+	data model is scalable, it is readily possible to use this methodology for
+	a large network of sensors to gather granular data on pedestrian footfall.
+	Projects such as Smart Street Sensors, can utilise the methodology to
+	overcome the challenges introduced by implementation of MAC address
+	randomisation. Such precise and granular data also enables us in confidently
+	model the pedestrian flow in urban road networks and will be indispensable
+	tool in the smart city framework. This can also be used in understanding and
+	classifying geographical areas based on the spatio temporal distribution of
+	the volume of activity in them.
 

### 2018-01-31 abstract.tex
Fixed figure reference

@@
 			numbers present in them.}
-			label{fig}
+			\label{fig}
 		\end{center}
@@
 	anonymised probe requests.
-	\ref{fig} shows the clustering process for where the black dots show
+	Figure \ref{fig} shows the clustering process for where the black dots show
 	probe requests and red lines connecting them show the clusters

### 2018-01-31 abstract.tex
submitted abstract

@@
 
-	In the past decade Wi-Fi has emerged as the most
-	commonly used technology in providing 
-	internet access to mobile devices such as smartphones, tablets
-	and laptops in public and private spaces.
-	This has resulted in multiple Wi-Fi networks being available at
-	almost every location in an urban environment.
-	Traversing through this overlapping mesh of Wi-Fi networks,
-	modern mobile devices with Wi-Fi antennae regularly
-	broadcast special type of signals known as probe requests to
-	discover available Wi-Fi networks and switch seamlessly
-	between them.
-	This is a hardware level signal which was standardised by IEEE 802.11b/g
-	specification relays information about the source mobile device to
-	any Access Points (AP) available around it.
-	Since this is the first step in establishing
-	a connection between any two devices, it is
-	is universal to any device which uses a Wi-Fi radio to communicate.
-	This makes these probe requests
-	an open, passive, continuous, and wireless source of
-	data available at any urban location which can provide us with clues in
-	understanding the number of people present in the immediate surroundings
-	in real-time with high granularity \citep{freud2015,konto2017}.
-	In this paper, from a set of probe requests collected
-	at a high street location in London,
-	along with manually counted data, we demonstrate
-	that pedestrian footfall can be estimated with considerable accuracy without
-	infringing on the privacy of the mobile users involved.
+	In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
+This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
+Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
+This is a hardware level signal which was standardised by IEEE 802.11b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
+Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
+This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
+Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
+In this paper, using a set of probe requests collected at a high street location in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.
 
@@
 
-	There have been numerous attempts at using Wi-Fi
-	to measure the volume and movement of people in built environment for various
-	applications \citep{zarim2006,sap2015,reki2007}.
-	Though most research observe feasible and favorable results,
-	in recent years, one of the major challenges faced in such attempts
-	has been the MAC address randomisation process.
-	This process aims to protect the users' privacy by anonymising
-	the only globally identifiable portion of the probe requests
-	resulting in a set of probe requests generated by the same device
-	with different random MAC addresses \citep{green2008}.
-	There have been various successful research in breaking this
-	randomisation process to extract real MAC addresses \citep{martin2017}
-	but this usually results in serious risk in infringement of
-	privacy of the users of the mobile devices.
-	There is a clear gap in the research for exploring methodologies
-	which enables us to estimate the number of unique mobile
-	devices from a set of anonymised probe requests
-	without the need to reveal their original MAC addresses.
+	There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
+Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
+This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
+There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
+There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
 
-	A pilot survey was conducted at Oxford street in London
-	on 20 December 2017 from 12:30 to 13:00 where two sets of data
-	were collected on pedestrian footfall through Wi-Fi sensing
-	and manual counting in parallel.
-	The Wi-Fi sensor collected all the probe requests that were 
-	broadcast around the area and recorded
-	the time-stamp at which they were collected,
-	MAC address of the source mobile device
-	(anonymised using a hashing algorithm),
-	organisationally unique identifier (OUI) of the manufacturer of the device,
-	total length of the signal in bits,
-	the strength of the signal reported by the mobile device in dBm,
-	the sequence number of the signal,
-	duration for which the signal was transmitted, 
-	the service set identifier (SSID) of the access point targeted
-	by the probe request and 
-	the length of the extra information (tags) embedded in the packets.
-	The manual count was done by an android application on a mobile phone
-	which records the time-stamp every time the surveyor touches the screen
-	which corresponds to one pedestrian footfall on the sidewalk.
+	A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
+These datasets were collected through Wi-Fi sensing and manual counting in parallel.
+The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
+The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
 
-	Being one of the busiest location in the United Kingdom, the location
-	generated approximately 60,000 probe requests in the 30 minutes
-	interval through the Wi-Fi sensor and 3,722 people were counted manually.
-	When we aggregate the probe requests by their MAC address for every minute,
-	the difference between the sensor counts and the manual counts
-	is observed to be on average 425\%. 
-	This shows that there is a large amount of noise
-	in the data which might include signals from devices outside the area where
-	the manual count was conducted and the anonymised probe requests from the
-	same devices with different MAC addresses.
-	Before we go look at the data
-	in detail, an initial analysis shows that the fields - SSID and tags are
-	very sparse and doesn't provide much information for our cleaning process 
-	and the duration field is closely related to the length of the 
-	probe request and provides no new information.
-	So we remove these fields from further analysis.
+	Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
+When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
+This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
+An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
+In addition, the duration field was closely related to the length of the probe request and provides no new information.
+Therefore, we removed these fields from further analysis.
 
-	We remove the noise from devices outside the area of interest 
-	by removing all the probe requests which report a "low" signal strength.
-	This classification of "high" vs "low" is done through "k-means"
-	classification algorithm.
-	The cutoff point for the collected data is -71dBm.
-	This process of filtering is highly effective and reduces the difference
-	between the sensor counts and manual counts to 30\%. We observe that
-	around 55\% of all probe requests collected are anonymised.
-	We assign the hashed MAC address the unique identifier for the rest of
-	the 45\% and investigate the anonymised probe requests further.
+	We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
+This classification of "high" vs "low" was performed using a k-means classification algorithm.
+The cut-off point for the collected data was -71 dBm.
+This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
+We observed that around 55\% of all probe requests collected were anonymised.
+We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
 
-	We then use the fields - OUI, lengths and sequence number,
-	to tackle the noise from devices which anonymising the probe requests.
-	OUI and length are used to split the dataset into
-	groups of probe requests from similar
-	devices and each subset is classified further
-	based on a graph based clustering algorithm 
-	where each cluster corresponds to an unique device.
-	The algorithm works by creating a graph where the probe requests
-	are the nodes and the links are created
-	between them based on the following rules,
+	We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
+OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
+The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:
 
 	\begin{enumerate}
-		\item A link can go only forward in time.
-		\item A link can exist between nodes with a 
-			maximum time difference of $\alpha$ (time threshold)
-		\item A link can go from low to high sequence numbers.
-		\item A link can exist between nodes with a
-			maximum sequence number difference of $\beta$ (sequence threshold)
-		\item A node can have only one incoming link and outgoing link
-			which is the shortest of all possible such links.
+		\item A link could go only forward in time. 
+		\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
+		\item A link could go from low to high sequence numbers.
+		\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
+		\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
 	\end{enumerate}
 
-	The nodes are then classified based on the unique connected component they
-	belong to. This classification is assigned as the unique identifier for the
-	anonymised probe requests.
-	Figure \ref{fig} shows the clustering process for where the black dots show
-	probe requests and red lines connecting them show the clusters
-	corresponding to the same device generated them.
-	We finally combine both normal and anonymised probe requests, 
-	aggregate them based on their unique identifier and remove
-	repeating probe requests which reduces the difference between
-	the sensor counts and the manual counts to -18\%.
+	The nodes were then classified based on the unique connected component they belonged to.
+This classification was assigned as the unique identifier for the anonymised probe requests.
+Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
+We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
 
-	It is important to note that the filtering process is done based on just
-	the information present in the probe requests
-	and their temporal distribution.
-	This makes sure that though the mobile devices are uniquely identified
-	there is no more personal data is generated linking the probe requests
-	to the users of the mobile devices. This method essentially gives us a
-	way to estimate the footfall in real-time without identifying or tracking the
-	mobile devices themselves.
+	It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
+This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
+This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.
 
-	This Wi-Fi based footfall counting methodology offers a large number of
-	applications and benefits. Since Wi-Fi based sensors are inexpensive and the
-	data model is scalable, it is readily possible to use this methodology for
-	a large network of sensors to gather granular data on pedestrian footfall.
-	Projects such as Smart Street Sensors, can utilise the methodology to
-	overcome the challenges introduced by implementation of MAC address
-	randomisation. Such precise and granular data also enables us in confidently
-	model the pedestrian flow in urban road networks and will be indispensable
-	tool in the smart city framework. This can also be used in understanding and
-	classifying geographical areas based on the spatio temporal distribution of
-	the volume of activity in them.
+	This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
+Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
+Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
+Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
+It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
 

### 2018-05-17 abstract.tex
starting the final paper

@@
 
-\title{Estimating real-time high street footfall from Wi-Fi probe requests}
+\title{Estimating real-time high treat  footfall from Wi-Fi probe requests}
 \author{
@@
 
-	In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
-Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
-This is a hardware level signal which was standardised by IEEE 802.11b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
-Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
-This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
-Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
-In this paper, using a set of probe requests collected at a high street location in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.
+	In the past decade, Wi-Fi has emerged as the most commonly used technolo
+gy in providing internet access to mobile devices such as smartphones, tablets a
+nd laptops in public and private spaces. This has resulted in multiple Wi-Fi net
+works being available at almost every location in urban environments. Traversing
+ through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-
+Fi antennae regularly broadcast a special type of signal known as probe requests
+, in order to discover available Wi-Fi networks and switch seamlessly between th
+em. This is a hardware level signal which was standardised by IEEE 802.11b/g spe
+cification, to relay information about the source mobile device to any Access Po
+int (AP) available around it. Since this is the first step in establishing a con
+nection between any two devices, it is a standard feature of any device which us
+es a Wi-Fi radio to communicate. This makes these probe requests an open, passiv
+e, continuous, and wireless source of data available at any urban location. More
+over, the data can provide us with clues to understand the number of people pres
+ent in the immediate surrounding in real-time and with high granularity \citep{f
+reud2015,konto2017}. In this paper, using a set of probe requests collected at a
+ high street location in London along with manually counted data, we demonstrate
+ that pedestrian footfall can be estimated with considerable accuracy without in
+fringing on the privacy of the mobile users involved.
 
@@
 				{outputs/clustering_2.png}
-			\caption{Clustering probe requests based on increasing sequence
+			\caption{Clustering probe requests based on increasing s
+equence
 			numbers present in them.}
@@
 
-	There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
-Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
-This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
-There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
-There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
+	There have been numerous attempts at using Wi-Fi to measure the volume a
+nd movement of people in the built environment for various applications \citep{z
+arim2006,sap2015,reki2007}. Though most research obtains feasible and favorable 
+results, in recent years, one of the major challenges faced in such attempts has
+ been the MAC address randomisation process. This process aims to protect the us
+ers’ privacy by anonymising the only globally identifiable portion of the prob
+e requests, which results in a set of probe requests generated by the same devic
+e with different random MAC addresses \citep{green2008}. There have been various
+ successful attempts by researchers to breaking this randomisation process in or
+der to extract real MAC addresses, \citep{martin2017} but this usually results i
+n serious risk of infringement of the privacy of the users of the mobile devices
+. There is a clear gap in the research for exploring methodologies which enable 
+us to estimate the number of unique mobile devices from a set of anonymised prob
+e requests, without the need to reveal their original MAC addresses.
 
-	A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
-These datasets were collected through Wi-Fi sensing and manual counting in parallel.
-The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
-The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
+	A pilot survey was conducted on Oxford Street in London in December 2017
+, where two sets of data were collected on pedestrian footfall. These datasets w
+ere collected through Wi-Fi sensing and manual counting in parallel. The Wi-Fi s
+ensor collected all the probe requests which were broadcast around the area, and
+ recorded the following data: the time-stamp at which they were collected, the M
+AC address of the source mobile device (anonymised using a hashing algorithm), t
+he organisationally unique identifier (OUI) of the manufacturer of the device, t
+he total length of the signal in bits, the strength of the signal reported by th
+e mobile device in dBm, the sequence number of the signal, the duration for whic
+h the signal was transmitted, the service set identifier (SSID) of the access po
+int targeted by the probe request, and the length of the extra information (tags
+) embedded in the packets. The manual count was undertaken using an Android appl
+ication on a mobile phone: the researcher touched the phone’s screen every tim
+e an individual pedestrian footfall was counted, and this was recorded as a time
+-stamp.
 
-	Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
-When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
-This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
-An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
-In addition, the duration field was closely related to the length of the probe request and provides no new information.
-Therefore, we removed these fields from further analysis.
+	Being located at one of the busiest retail locations in the United Kingd
+om, the WiFi sensor captured approximately 60,000 probe requests over a 30 minut
+es interval, and 3,722 people were counted manually. When we aggregated the prob
+e requests by their MAC address for every minute, the difference between the sen
+sor counts and the manual counts was observed to be on average 425\%. This sugge
+sted that there was a large amount of noise in the data which might have include
+d signals from devices outside the area where themanual count was conducted, as 
+well as anonymised probe requests from the same devices but with different MAC a
+ddresses. An initial analysis revealed that the fields - SSID and tags - were ve
+ry sparse and did not provide much information for our cleaning process. In addi
+tion, the duration field was closely related to the length of the probe request 
+and provides no new information. Therefore, we removed these fields from further
+ analysis.
 
-	We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
-This classification of "high" vs "low" was performed using a k-means classification algorithm.
-The cut-off point for the collected data was -71 dBm.
-This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
-We observed that around 55\% of all probe requests collected were anonymised.
-We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
+	We eliminated the noise from devices outside the area of interest by rem
+oving all the probe requests which reported a "low" signal strength. This classi
+fication of "high" vs "low" was performed using a k-means classification algorit
+hm. The cut-off point for the collected data was -71 dBm. This process of filter
+ing was highly effective and reduced the difference between the sensor counts an
+d manual counts to 30\%. We observed that around 55\% of all probe requests coll
+ected were anonymised. We assigned the hashed MAC address the unique identifier 
+for the remaining 45\% and investigated the anonymised probe requests further.
 
-	We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
-OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
-The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:
+	We then used the fields - OUI, lengths and sequence number - to tackle t
+he noise from devices which anonymised the probe requests. OUI and length were u
+sed to split the dataset into groups of probe requests from similar devices, and
+ each subset was classified further using a graph based clustering algorithm whe
+re each cluster corresponded to a unique device. The algorithm created a graph w
+here the probe requests represented the nodes, and links are created between the
+m based on the following rules:
 
@@
 		\item A link could go only forward in time. 
-		\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
+		\item A link could exist between nodes with a maximum time diffe
+rence of $\alpha$ (time threshold).
 		\item A link could go from low to high sequence numbers.
-		\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
-		\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
+		\item A link could exist between nodes with a maximum sequence n
+umber difference of $\beta$ (sequence threshold).
+		\item A node could have only one incoming link and one outgoing 
+link, which is the shortest of all such possible links.
 	\end{enumerate}
 
-	The nodes were then classified based on the unique connected component they belonged to.
-This classification was assigned as the unique identifier for the anonymised probe requests.
-Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
-We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
+	The nodes were then classified based on the unique connected component t
+hey belonged to. This classification was assigned as the unique identifier for t
+he anonymised probe requests. Figure \ref{fig} shows the clustering process: the
+ black dots show the probe requests and the red lines connect them into clusters
+ representing those which were generated by the same device. We finally combined
+ both normal and anonymised probe requests, aggregated them based on their uniqu
+e identifier, and removed repeating probe requests which reduced the difference 
+between the sensor counts and the manual counts to -18\%.
 
-	It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
-This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
-This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.
+	It is important to note that the filtering process was done based soley 
+on the information present in the probe requests and their temporal distribution
+. This ensured that although the mobile devices were uniquely identified, there 
+was no further personal data generated by linking the probe requests to the user
+s of the mobile devices. This method essentially gave us a way to estimate the f
+ootfall in real-time without identifying or tracking the mobile devices themselv
+es.
 
-	This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
-Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
-Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
-It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
+	This Wi-Fi based footfall counting methodology offers a large number of 
+applications and benefits for real time spatial analysis. Since Wi-Fi based sens
+ors are inexpensive and the data model is scalable, it is possible to use this m
+ethodology for a large network of sensors to gather granular data on pedestrian 
+footfall. Projects such as SmartStreetSensors \citep{sss2016}, may utilise this 
+methodology to overcome the challenges introduced by the implementation of MAC a
+ddress randomisation. Such precise and granular data also enables us to confiden
+tly model the pedestrian flow in urban road networks, and will be an indispensab
+le tool in the smart city framework. It can also be used to understand and class
+ify geographical areas based on the spatio-temporal distribution of the volume o
+f activity in them.
 

### 2018-06-01 main.tex
added pilot study

\documentclass[11t, a4paper, twocolumn]{article} 
\input{structure.tex}

\title{Estimating real-time highstreet  footfall from Wi-Fi probe requests}
\author{
	\authorstyle{
		Balamurugan Soundararaj\textsuperscript{1}, 
		James Cheshire\textsuperscript{1} and 
		Paul Longley\textsuperscript{1}}
	\newline\newline
	\textsuperscript{1}\institution{
		Department of Geography, 
		University College London, 
		United Kingdom}
}
\date{\today}

\begin{document}

	\maketitle
	\thispagestyle{firstpage}

	\textbf{Abstract : }

	\section{Introduction}\label{intro}

		In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
		This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
		Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
		This is a hardware level signal which was standardised by IEEE 802.
		1b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
		Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
		This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
		Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
		In this paper, using a set of probe requests collected at high street locations in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.

	\section{Previous Work}\label{prev}
		There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
		Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
		This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
		There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
		There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.

	\section{Pilot Study}\label{data}
		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
		The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
		The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.

		Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
		When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
		This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
		An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
		In addition, the duration field was closely related to the length of the probe request and provides no new information.
		Therefore, we removed these fields from further analysis.

		We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
		This classification of "high" vs "low" was performed using a k-means classification algorithm.
		The cut-off point for the collected data was -71 dBm.
		This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
		We observed that around 55\% of all probe requests collected were anonymised.
		We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.

		We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
		OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
		The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:

			\begin{enumerate}
				\item A link could go only forward in time. 
				\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
				\item A link could go from low to high sequence numbers.
				\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
				\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
			\end{enumerate}

		The nodes were then classified based on the unique connected component they belonged to.
		This classification was assigned as the unique identifier for the anonymised probe requests.
		Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
		We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.

		It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
		This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
		This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.

		This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
		Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
		Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
		Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
		It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.

	\printbibliography[title={References}]

\end{document}

### 2018-06-11 main.tex
methodology and pilot study

@@
 
-	\textbf{Abstract : }
+	% \textbf{Abstract : }
 
@@
 
+	\begin{figure*}
+		\begin{center}
+			\includegraphics [width=\textwidth,trim=4 4 4 4,clip] {outputs/clustering_2.png}
+			\caption{Clustering probe requests based on increasing sequence numbers present in them.}
+			\label{clustering_pilot}
+		\end{center}
+	\end{figure*}
+
 	\section{Previous Work}\label{prev}
@@
 
-	\section{Pilot Study}\label{data}
-		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
-		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
+	\section{Methodology}
+		A probe request consists of various information including, MAC address, device capabilities, sequence number, known hotspots etc.
+Out of this we are interested in things which can identify a unique mobile device.
+We collect the WiFi probe requests using the WireShark on hardware such as laptop for the pilot study and raspberry pi based custom sensor.
+j
 		The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
 		The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
-
-		Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
-		When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
-		This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
 			An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
@@
 		Therefore, we removed these fields from further analysis.
-
 		We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
@@
 		The cut-off point for the collected data was -71 dBm.
-		This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
-		We observed that around 55\% of all probe requests collected were anonymised.
-		We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
-
 		We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
@@
 		This classification was assigned as the unique identifier for the anonymised probe requests.
-		Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
+
+	\section{Pilot Study}\label{data}
+		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall with the aim of establishing merit in measuring pedestrian footfall as a function of the number of wifi probe requests collected at a given location.
+		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
+		Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
+
+		When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
+		This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
+		This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
+		We observed that around 55\% of all probe requests collected were anonymised.
+		We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
+
+		Figure \ref{clustering_pilot} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
 		We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
 
+		\begin{figure}
+			\begin{center}
+				\includegraphics [width=0.45\textwidth] {outputs/pilot_counts_comparision.png}
+				\caption{Comparision of counts after filtering with manual counts}
+				\label{comparision_pilot}
+			\end{center}
+		\end{figure}
+
+	\section{Main Study}
+		The k
+
+	\section{Conclusion}
 		It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.

### 2018-06-12 abstract/abstract.tex
reorganised workspace

--- 2018-05-17 abstract.tex
@@
 over, the data can provide us with clues to understand the number of people pres
-ent in the immediate surrounding in real-time and with high granularity \citep{f
-reud2015,konto2017}. In this paper, using a set of probe requests collected at a
+ent in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
+In this paper, using a set of probe requests collected at a
  high street location in London along with manually counted data, we demonstrate
@@
 	There have been numerous attempts at using Wi-Fi to measure the volume a
-nd movement of people in the built environment for various applications \citep{z
-arim2006,sap2015,reki2007}. Though most research obtains feasible and favorable 
+nd movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
+Though most research obtains feasible and favorable 
 results, in recent years, one of the major challenges faced in such attempts has

### 2018-06-12 Temp/graph-partition.r

@@ -1,28 +1,0 @@
-library(tidyverse)
-library(igraph)
-data <- subset_local %>% filter(vendor=="Google")
-data <- subset_local %>% filter(vendor=="Unknown")
-vertices <- data %>%
-	mutate(x = as.integer(time), id = as.numeric(rownames(data))) %>%
-	select(id, x, y=sequence) %>%
-	arrange(x)
-edges <- vertices %>%
-	pull(id) %>%
-	expand.grid(.,.) %>% rename(from=Var1,to=Var2) %>%
-	mutate(from=as.numeric(from), to=as.numeric(to)) %>%
-	filter(from<to) %>%
-	left_join(vertices, by = c("from" = "id")) %>%
-	left_join(vertices, by = c("to" = "id")) %>%
-	mutate(distance = sqrt(((x.y-x.x)**2)+((y.y-y.x)**2))) %>%
-	mutate(tdistance = sqrt((x.y-x.x)**2)) %>%
-	filter(x.y >= x.x, y.y > y.x, distance < 68) %>%
-	group_by(from) %>%
-	summarise(to = to[which(distance == min(distance))][[1]], distance=min(distance)) %>%
-	group_by(to) %>%
-	summarise(from = from[which(distance == min(distance))][[1]])
-
-print(nrow(edges))
-graph <- graph_from_data_frame(edges, vertices = vertices)
-clusters <- clusters(graph)$membership
-plot(graph,vertex.size=2,edge.arrow.size=0,vertex.label=NA,vertex.frame.color="orange")
-plot(graph,vertex.size=2,edge.arrow.size=0,vertex.label=NA,vertex.frame.color=clusters,vertex.color=clusters)

### 2018-06-12 paper/paper.tex
reorganised for efficiency

\documentclass[11t, a4paper, twocolumn]{article} 
\input{structure.tex}
\addbibresource{references.bib}
\title{Estimating real-time highstreet footfall from Wi-Fi probe requests}
\author{
	\authorstyle{
		Balamurugan Soundararaj\textsuperscript{1}, 
		James Cheshire\textsuperscript{1} and 
		Paul Longley\textsuperscript{1}}
	\newline\newline
	\textsuperscript{1} \institution{
		Department of Geography, 
		University College London, 
		United Kingdom } }
\date{\today}
\begin{document}
	\maketitle
	\thispagestyle{firstpage}
	\textbf{Abstract -}
		\input{sections/abstract}
	\section{Introduction}\label{introduction}
		\input{sections/introduction}
	\section{Previous Work}\label{previous_work}
		\input{sections/previous_work}
	\section{Methodology}\label{methodology}
		\input{sections/methodology}
	\section{Pilot Study}\label{pilot_study}
		\input{sections/pilot_study}
	\section{Main Study}\label{main_study}
		\input{sections/main_study}
	\section{Discussion}\label{discussion}
		\input{sections/discussion}
	\section{Conclusion}\label{conclusion}
		\input{sections/discussion}
	\printbibliography[title={References}]
\end{document}

### 2018-06-12 paper/sections/conclusion.tex
reorganised for efficiency

--- 2018-06-01 main.tex
@@
-\documentclass[11t, a4paper, twocolumn]{article} 
-\input{structure.tex}
-
-\title{Estimating real-time highstreet  footfall from Wi-Fi probe requests}
-\author{
-	\authorstyle{
-		Balamurugan Soundararaj\textsuperscript{1}, 
-		James Cheshire\textsuperscript{1} and 
-		Paul Longley\textsuperscript{1}}
-	\newline\newline
-	\textsuperscript{1}\institution{
-		Department of Geography, 
-		University College London, 
-		United Kingdom}
-}
-\date{\today}
-
-\begin{document}
-
-	\maketitle
-	\thispagestyle{firstpage}
-
-	\textbf{Abstract : }
-
-	\section{Introduction}\label{intro}
-
-		In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-		This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
-		Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
-		This is a hardware level signal which was standardised by IEEE 802.
-		1b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
-		Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
-		This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
-		Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
-		In this paper, using a set of probe requests collected at high street locations in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.
-
-	\section{Previous Work}\label{prev}
-		There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
-		Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
-		This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
-		There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
-		There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
-
-	\section{Pilot Study}\label{data}
-		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
-		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
-		The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
-		The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
-
-		Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
-		When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
-		This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
-		An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
-		In addition, the duration field was closely related to the length of the probe request and provides no new information.
-		Therefore, we removed these fields from further analysis.
-
-		We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
-		This classification of "high" vs "low" was performed using a k-means classification algorithm.
-		The cut-off point for the collected data was -71 dBm.
-		This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
-		We observed that around 55\% of all probe requests collected were anonymised.
-		We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
-
-		We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
-		OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
-		The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:
-
-			\begin{enumerate}
-				\item A link could go only forward in time. 
-				\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
-				\item A link could go from low to high sequence numbers.
-				\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
-				\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
-			\end{enumerate}
-
-		The nodes were then classified based on the unique connected component they belonged to.
-		This classification was assigned as the unique identifier for the anonymised probe requests.
-		Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
-		We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
-
 It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
@@
 It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
-
-	\printbibliography[title={References}]
-
-\end{document}

### 2018-06-12 paper/sections/introduction.tex
reorganised for efficiency

In the past decade Wi-Fi has emerged as the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces. This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments. Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests', in order to discover Wi-Fi networks available to them. This helps these devices to connect and switch between the WiFi networks seamlessly.

Probe requests are low level signals standardised by IEEE 802.1b/g specification as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model. This ubiquity and standardisation make them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location. Considering the unprecendented levels of mobile device ownership in recent years, we can in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.

While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users. In this paper, using a set of probe requests and manual counts collected at various high street locations across London, we demonstrate that pedestrian footfall at these locations can be estimated with considerable precision and accuracy while protecting the privacy of the pedestrians.

### 2018-06-12 paper/sections/previous_work.tex
reorganised for efficiency

--- 2018-06-01 main.tex
@@
-\documentclass[11t, a4paper, twocolumn]{article} 
-\input{structure.tex}
-
-\title{Estimating real-time highstreet  footfall from Wi-Fi probe requests}
-\author{
-	\authorstyle{
-		Balamurugan Soundararaj\textsuperscript{1}, 
-		James Cheshire\textsuperscript{1} and 
-		Paul Longley\textsuperscript{1}}
-	\newline\newline
-	\textsuperscript{1}\institution{
-		Department of Geography, 
-		University College London, 
-		United Kingdom}
-}
-\date{\today}
-
-\begin{document}
-
-	\maketitle
-	\thispagestyle{firstpage}
-
-	\textbf{Abstract : }
-
-	\section{Introduction}\label{intro}
-
-		In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-		This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
-		Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
-		This is a hardware level signal which was standardised by IEEE 802.
-		1b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
-		Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
-		This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
-		Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
-		In this paper, using a set of probe requests collected at high street locations in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.
-
-	\section{Previous Work}\label{prev}
 There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
@@
 There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
-
-	\section{Pilot Study}\label{data}
-		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
-		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
-		The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
-		The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
-
-		Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
-		When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
-		This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
-		An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
-		In addition, the duration field was closely related to the length of the probe request and provides no new information.
-		Therefore, we removed these fields from further analysis.
-
-		We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
-		This classification of "high" vs "low" was performed using a k-means classification algorithm.
-		The cut-off point for the collected data was -71 dBm.
-		This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
-		We observed that around 55\% of all probe requests collected were anonymised.
-		We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
-
-		We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
-		OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
-		The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:
-
-			\begin{enumerate}
-				\item A link could go only forward in time. 
-				\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
-				\item A link could go from low to high sequence numbers.
-				\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
-				\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
-			\end{enumerate}
-
-		The nodes were then classified based on the unique connected component they belonged to.
-		This classification was assigned as the unique identifier for the anonymised probe requests.
-		Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
-		We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
-
-		It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
-		This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
-		This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.
-
-		This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
-		Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-		Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
-		Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
-		It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
-
-	\printbibliography[title={References}]
-
-\end{document}

### 2018-06-20 paper/sections/methodology.tex
wednesday start of the day

The overall methodology is to use a WiFi antenna to collect probe requests,
assign a unique identification based on the device that created the request,
aggregate them based on the unique identification for a specific time interval
to create an estimate of number of people at the location.
In this section we look at the characteristics of probe requests in detail,
outline the methodology used to collect these probe requests, 
look at the uncertainities and biases in the process and 
device methods to overcome these issues.

\subsection{Probe Requests}
Probe request is a low level packet sent by a mobile device as a means of
"scanning" for various access points available at a specific location.
This packet consists information about the mobile device
including but not limited to,
Media Access Control address - A two part 12 bit identifier where the first part
identifies the manufacturer of the device and the second part identifies the
device itself. This MAC address is of two types, global - the real identifier
of the device which doesnt change and local - virtual, random addresses used
for temporary situations.
Sequence number of the packet to keep track of the replies.
The access points for which the packet is being sent to.
Capabilities of the device.
We can also infer other things about the packet such as
time at which the packet has been received,
total length of the packet
time it took to transmit the packet
Signal strength of the packet.
All of this can help us identifying the packets and label them when they are transmitted by the same device.

\subsection{Data Collection}
The data collection involved two processes - 
A sensor based collection of probe requests and
a manual count of pedestrians in the area next to the sensor.
We have detailed the sensor design below,

The sensor consists of wireless transponder in a linux/ MacOSX based machine
which is put into a monitor mode and
wireshark software (tshark) is used to sniff, filter and
save the relevant packets.
The filters applied are management frame > probe requests > broadcast. 
This is then parsed, personal information is hashed, 
compressed and sent to a server via web socket protocol.
The server receives, logs and stores the data in a postgresql database,
for further analysis.
The manual count was undertaken using an Android application
on a mobile phone where timestamp of every individual pedestrian footfall
was recorded by touch screen. A schematic is shown in ref figure.

\subsection{Estimating Footfall}
The next step after collecting the probe request is to estimate the footfall or pedestrian activity from them. There are three major challanges in the above data collection methodology.
the first one is deliniating the area of interest.
WiFi and its range depends on lots of factors.
need a way to isolate the noise from the relevant data. depends a lot on microsite so we need a methodology which provides us with a general solution without much tinkering.
Second one is the mac randomisation.
modern phones randomise their MAC addresses to avoid detection and tracking.
This leads to overcounting and explosion of randomised (local) mac addresses.
We need a non MAC dependent method to overcome this.
Finally, inherent mobile phone carrying bias. we need to account for the average mobile ownership. which is steadily increasing and is not 1:1.

\subsubsection{WiFi range}
One of the clues of the distance of the device generating the probe request is
the strength of the signal recieved from it.
The Signal strength varies in inverse square law over distance with a 
propagation constant.
It also depends on lot of micro site, micro temporal factors.
There cannot be a simple rule to fit and filter for all configuration.
Our hypothesis is that in a specific setting and specific source of noise,
there must exist a clear break in the data.
for example, if there is a phone shop next to our sensor where
hundreds of phones regularly send lots of probe requests
we should be able to see a large increase in number of probe requests around
a specific signal strength.
we can identify this sharp change/ break using class interval algorithms such as
k-means, jenks, quantile, etc.

\subsubsection{MAC randomisation}
This is a recent problem. ref in figure. how mac randomisation has caused problem
MAC address has been our unique identifier. Now we need to look for others.
the contenders are length, duration which seem to be unique for device
sets of known wlans and capabilites which can give us unique finger print and 
finally sequence numbers in the packets.
This is a tricky one since it is neither unique nor aggregatable.
we need a method to seperate sequences shown in fig.
We propose a graph based clustering algorithm where
each cluster corresponded to a unique device.
The algorithm creates a graph where the probe requests represented the nodes, 
and links are created between them based on the following rules: 
\begin{enumerate}
	\item A link could go only forward in time. 
	\item A link could exist between nodes with a maximum
		time difference of $\alpha$ (time threshold).
	\item A link could go from low to high sequence numbers.
	\item A link could exist between nodes with a maximum sequence
		number difference of $\beta$ (sequence threshold).
	\item A node could have only one incoming link and
		one outgoing link, which is the
		shortest of all such possible links.
\end{enumerate}
The nodes were then classified based on the unique connected component they belonged to.
This classification was assigned as the unique identifier for the anonymised probe requests
this unique identifier is used instead of MAC to aggregation.

\subsubsection{Mobile phone ownership}
This is both a long term change and result of demographic factors.
Phone ownership is not 1:1. It changes slowly overtime. It also changes
with place to place. It can be more in dense urban centers
and can be low in rural areas. We propose a adjustment factor methodology
we use periodically carried out manual counts to adjust the numbers
to what is reported on ground. The adjustment is as strong as the amount of
ground truth we collected.

### 2018-06-20 paper/sections/pilot_study.tex
wednesday start of the day

--- 2018-06-01 main.tex
@@
-\documentclass[11t, a4paper, twocolumn]{article} 
-\input{structure.tex}
-
-\title{Estimating real-time highstreet  footfall from Wi-Fi probe requests}
-\author{
-	\authorstyle{
-		Balamurugan Soundararaj\textsuperscript{1}, 
-		James Cheshire\textsuperscript{1} and 
-		Paul Longley\textsuperscript{1}}
-	\newline\newline
-	\textsuperscript{1}\institution{
-		Department of Geography, 
-		University College London, 
-		United Kingdom}
-}
-\date{\today}
-
-\begin{document}
-
-	\maketitle
-	\thispagestyle{firstpage}
-
-	\textbf{Abstract : }
-
-	\section{Introduction}\label{intro}
-
-		In the past decade, Wi-Fi has emerged as the most commonly used technology in providing internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-		This has resulted in multiple Wi-Fi networks being available at almost every location in urban environments.
-		Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as probe requests, in order to discover available Wi-Fi networks and switch seamlessly between them.
-		This is a hardware level signal which was standardised by IEEE 802.
-		1b/g specification, to relay information about the source mobile device to any Access Point (AP) available around it.
-		Since this is the first step in establishing a connection between any two devices, it is a standard feature of any device which uses a Wi-Fi radio to communicate.
-		This makes these probe requests an open, passive, continuous, and wireless source of data available at any urban location.
-		Moreover, the data can provide us with clues to understand the number of people present in the immediate surrounding in real-time and with high granularity \citep{freud2015,konto2017}.
-		In this paper, using a set of probe requests collected at high street locations in London along with manually counted data, we demonstrate that pedestrian footfall can be estimated with considerable accuracy without infringing on the privacy of the mobile users involved.
-
-	\section{Previous Work}\label{prev}
-		There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
-		Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
-		This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
-		There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
-		There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
-
-	\section{Pilot Study}\label{data}
-		A pilot survey was conducted on Oxford Street in London in December 2017, where two sets of data were collected on pedestrian footfall.
-		These datasets were collected through Wi-Fi sensing and manual counting in parallel.
-		The Wi-Fi sensor collected all the probe requests which were broadcast around the area, and recorded the following data: the time-stamp at which they were collected, the MAC address of the source mobile device (anonymised using a hashing algorithm), the organisationally unique identifier (OUI) of the manufacturer of the device, the total length of the signal in bits, the strength of the signal reported by the mobile device in dBm, the sequence number of the signal, the duration for which the signal was transmitted, the service set identifier (SSID) of the access point targeted by the probe request, and the length of the extra information (tags) embedded in the packets.
-		The manual count was undertaken using an Android application on a mobile phone: the researcher touched the phone’s screen every time an individual pedestrian footfall was counted, and this was recorded as a time-stamp.
+We conducted a pilot study to get a feel for the data.
+See which fields are relevant and which ones are not.
+to check if the sequencing algorithm holds true.
+to findout which one of the classification algorithm works.
+if these methodology works in real world scenario scalably.
+A pilot survey was conducted on Oxford Street in London in December 2017,
+where two sets of data were collected on pedestrian footfall
+with the aim of establishing merit in measuring pedestrian footfall
+as a function of the number of wifi probe requests
+collected at a given location.
+These datasets were collected through Wi-Fi sensing and 
+manual counting in parallel.
 
 Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
+
 When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
 This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
-		An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
-		In addition, the duration field was closely related to the length of the probe request and provides no new information.
-		Therefore, we removed these fields from further analysis.
-
-		We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
-		This classification of "high" vs "low" was performed using a k-means classification algorithm.
-		The cut-off point for the collected data was -71 dBm.
 This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
@@
 
-		We then used the fields - OUI, lengths and sequence number - to tackle the noise from devices which anonymised the probe requests.
-		OUI and length were used to split the dataset into groups of probe requests from similar devices, and each subset was classified further using a graph based clustering algorithm where each cluster corresponded to a unique device.
-		The algorithm created a graph where the probe requests represented the nodes, and links are created between them based on the following rules:
+An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
+In addition, the duration field was closely related to the length of the probe request and provides no new information.
+Therefore, we removed these fields from further analysis.
+We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
+This classification of "high" vs "low" was performed using a k-means classification algorithm.
+The cut-off point for the collected data was -71 dBm.
 
-			\begin{enumerate}
-				\item A link could go only forward in time. 
-				\item A link could exist between nodes with a maximum time difference of $\alpha$ (time threshold).
-				\item A link could go from low to high sequence numbers.
-				\item A link could exist between nodes with a maximum sequence number difference of $\beta$ (sequence threshold).
-				\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
-			\end{enumerate}
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=\linewidth,trim=4 4 4 4,clip] {images/methodology_clustering.png}
+		\caption{Clustering probe requests based on increasing sequence numbers present in them.}
+		\label{clustering_pilot}
+	\end{center}
+\end{figure}
 
-		The nodes were then classified based on the unique connected component they belonged to.
-		This classification was assigned as the unique identifier for the anonymised probe requests.
-		Figure \ref{fig} shows the clustering process: the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
+Figure \ref{clustering_pilot} shows the clustering process: 
+the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
 We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
 
-		It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
-		This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
-		This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.
-
-		This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
-		Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-		Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
-		Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
-		It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
-
-	\printbibliography[title={References}]
-
-\end{document}
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=\linewidth] {images/pilot_counts_comparision.png}
+		\caption{Comparision of counts after filtering with manual counts}
+		\label{comparision_pilot}
+	\end{center}
+\end{figure}

### 2018-06-20 paper/sections/pilot_study.tex
wed end of the day

@@
 
-An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
+An initial analysis revealed that the fields - SSID and tags -
+were very sparse and did not provide much information for our cleaning process.
 In addition, the duration field was closely related to the length of the probe request and provides no new information.
@@
 Figure \ref{clustering_pilot} shows the clustering process: 
-the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
-We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
+the black dots show the probe requests and the red lines 
+connect them into clusters representing those which were
+generated by the same device.
+We finally combined both normal and anonymised probe requests,
+aggregated them based on their unique identifier, and
+removed repeating probe requests which reduced the difference
+between the sensor counts and the manual counts to -18\%.
 
@@
 \end{figure}
+
+we find that the methodology works as shown by the reduction in the
+mean error
+We find kmeans and quantile best algorithms for filtering
+sequencing works. Even in densely populated areas.
+
+Though this is promising we need to know more to generalise the methodology.
+we do a detailed multi location, longer term study with multiple manual
+verification which is the main study.

### 2018-06-21 paper/sections/main_study.tex
laptop

The aims of the main study are,
test the validity of the signal strength algorithm in different 
micro site conditions.
Test that the sequence number algorithm works in real world
for different locations and different times.
Test if the calibration works over intervals
Finally conclude if we can estimate footfall confidently
with just probe requests.

\subsection{locations}
six locations were selected across central london which
had different types of configurations and specific problems
configuratons are shown in fig.

1) Phone Shop Camden - has phones and bus stops.
2) Restaurant TCR - has seating area on either side.
3) Holborn - High volumne station entry

### 2018-06-21 paper/sections/main_study.tex

@@
 \subsection{locations}
-six locations were selected across central london which
+five locations were selected across central london which
 had different types of configurations and specific problems
-configuratons are shown in fig.
+configuratons are shown in fig. map is shown in figure.
+\begin{enumerate}
+	\item Phone Shop Camden - has phones and bus stops.
+	\item Restaurant TCR - has seating area on either side.
+	\item Holborn Information Kiosk - High volumne station entry
+	\item Restaurant Russell Square - seating on one side and side walk on other
+	\item Shop Charring Cross - sidewalk on one side and phone shop next door
+\end{enumerate}
+installations were carried out over the time period from xxxx to xxxx.
+the data collection happened from xxxx to xxxx. Manual counting was carried out
+with high precision on dates xxxx and aggregated five minutes on xxxx.
+The difference in methods could lead to some inaccuracies in data.
+The overall statistics of data collected. 
+The overally schedule is shown in half page graphic.
 
-1) Phone Shop Camden - has phones and bus stops.
-2) Restaurant TCR - has seating area on either side.
-3) Holborn - High volumne station entry
+\subsection{Signal strength filtering}
+
+\subsection{Device Fingerprinting}
+
+\subsection{Manual Calibration}
+
+\subsection{Discussion}

### 2018-06-24 misc/sample.tex
journal template and more writing

% interacttfvsample.tex
% v1.02 - September 2016

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{,}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

\begin{document}

\articletype{ARTICLE TEMPLATE}

\title{Taylor \& Francis \LaTeX\ template for authors (\textsf{Interact} layout + Reference Style V)}

\author{
\name{A.~N. Author\textsuperscript{a}\thanks{CONTACT A.~N. Author. Email: latex.helpdesk@tandf.co.uk} and John Smith\textsuperscript{b}}
\affil{\textsuperscript{a}Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK; \textsuperscript{b}Institut f\"{u}r Informatik, Albert-Ludwigs-Universit\"{a}t, Freiburg, Germany}
}

\maketitle

\begin{abstract}
This template is for authors who are preparing a manuscript for a Taylor \& Francis journal using the \LaTeX\ document preparation system and the \texttt{interact} class file, which is available via selected journals' home pages on the Taylor \& Francis website.
\end{abstract}

\begin{keywords}
Sections; lists; figures; tables; mathematics; fonts; references; appendices
\end{keywords}

\section{Introduction}

In order to assist authors in the process of preparing a manuscript for a journal, the Taylor \& Francis `\textsf{Interact}' layout style has been implemented as a \LaTeXe\ class file based on the \texttt{article} document class. A \textsc{Bib}\TeX\ bibliography style file and a sample bibliography are also provided in order to assist with the formatting of your references.

Commands that differ from or are provided in addition to standard \LaTeXe\ are described in this document, which is \emph{not} a substitute for a \LaTeXe\ tutorial.

The \texttt{interacttfvsample.tex} file can be used as a template for a manuscript by cutting, pasting, inserting and deleting text as appropriate, using the preamble and the \LaTeX\ environments provided (e.g.\ \verb"\begin{abstract}", \verb"\begin{keywords}").

\subsection{The \textsf{Interact} class file}\label{class}

The \texttt{interact} class file preserves the standard \LaTeXe\ interface such that any document that can be produced using \texttt{article.cls} can also be produced with minimal alteration using the \texttt{interact} class file as described in this document.

If your article is accepted for publication it will be typeset as the journal requires in Minion Pro and/or Myriad Pro. Since most authors will not have these fonts installed, the page make-up is liable to alter slightly with the change of font. Also, the \texttt{interact} class file produces only single-column format, which is preferred for peer review and will be converted to two-column format by the typesetter if necessary during preparation of the proofs. Please therefore do not try to match the typeset format exactly, but use the standard \LaTeX\ fonts instead and ignore details such as slightly long lines of text or figures/tables not appearing in exact synchronization with their citations in the text: these details will be dealt with by the typesetter. Similarly, it is unnecessary to spend time addressing warnings in the log file -- if your .tex file compiles to produce a PDF document that correctly shows how you wish your paper to appear, such warnings will not prevent your source files being imported into the typesetter's software.

\subsection{Submission of manuscripts prepared using \emph{\LaTeX}}

Manuscripts for possible publication should be submitted to the Editors for review as directed in the journal's Instructions for Authors, and in accordance with any technical instructions provided in the journal's ScholarOne Manuscripts or Editorial Manager site. Your \LaTeX\ source file(s), the class file and any graphics files will be required in addition to the final PDF version when final, revised versions of accepted manuscripts are submitted.

Please ensure that any author-defined macros used in your article are gathered together in the preamble of your .tex file, i.e.\ before the \verb"\begin{document}" command. Note that if serious problems are encountered in the coding of a document (missing author-defined macros, for example), the typesetter may resort to rekeying it.

\section{Using the \texttt{interact} class file}

For convenience, simply copy the \texttt{interact.cls} file into the same directory as your manuscript files (you do not need to install it in your \TeX\ distribution).
In order to use the \texttt{interact} document class, replace the command \verb"\documentclass{article}" at the beginning of your document with the command \verb"\documentclass{interact}".

The following document-class options should \emph{not} be used with the \texttt{interact} class file:
\begin{itemize}
  \item \texttt{10pt}, \texttt{11pt}, \texttt{12pt} -- unavailable;
  \item \texttt{oneside}, \texttt{twoside} -- not necessary, \texttt{oneside} is the default;
  \item \texttt{leqno}, \texttt{titlepage} -- should not be used;
  \item \texttt{twocolumn} -- should not be used (see Subsection~\ref{class});
  \item \texttt{onecolumn} -- not necessary as it is the default style.
\end{itemize}
To prepare a manuscript for a journal that is printed in A4 (two column) format, use the \verb"largeformat" document-class option provided by \texttt{interact.cls}; otherwise the class file produces pages sized for B5 (single column) format by default. The \texttt{geometry} package should not be used to make any further adjustments to the page dimensions.

\section{Additional features of the \texttt{interact} class file}

\subsection{Title, authors' names, affiliations, abstracts and article types}

The title should be generated at the beginning of your article using the \verb"\maketitle" command.
In the final version the author name(s) and affiliation(s) must be followed immediately by \verb"\maketitle" as shown below in order for them to be displayed in your PDF document.
To prepare an anonymous version for double-blind peer review, you can put the \verb"\maketitle" between the \verb"\title" and the \verb"\author" in order to hide the author name(s) and affiliation(s) temporarily.
Next you should include the abstract if your article has one, enclosed within an \texttt{abstract} environment.
The \verb"\articletype" command is also provided as an \emph{optional} element which should only be included if your article actually needs it.
For example, the titles for this document begin as follows:
\begin{verbatim}
\articletype{ARTICLE TEMPLATE}

\title{Taylor \& Francis \LaTeX\ template for authors (\textsf{Interact}
layout + Reference Style V)}

\author{
\name{A.~N. Author\textsuperscript{a}\thanks{CONTACT A.~N. Author.
Email: latex.helpdesk@tandf.co.uk} and John Smith\textsuperscript{b}}
\affil{\textsuperscript{a}Taylor \& Francis, 4 Park Square, Milton
Park, Abingdon, UK; \textsuperscript{b}Institut f\"{u}r Informatik,
Albert-Ludwigs-Universit\"{a}t, Freiburg, Germany} }

\maketitle

\begin{abstract}
This template is for authors who are preparing a manuscript for a
Taylor \& Francis journal using the \LaTeX\ document preparation system
and the \texttt{interact} class file, which is available via selected
journals' home pages on the Taylor \& Francis website.
\end{abstract}
\end{verbatim}

An additional abstract in another language (preceded by a translation of the article title) may be included within the \verb"abstract" environment if required.

A graphical abstract may also be included if required. Within the \verb"abstract" environment you can include the code
\begin{verbatim}
\\\resizebox{25pc}{!}{\includegraphics{abstract.eps}}
\end{verbatim}
where the graphical abstract is to appear, where \verb"abstract.eps" is the name of the file containing the graphic (note that \verb"25pc" is the recommended maximum width, expressed in pica, for the graphical abstract in your manuscript).

\subsection{Abbreviations}

A list of abbreviations may be included if required, enclosed within an \texttt{abbreviations} environment, i.e.\ \verb"\begin{abbreviations}"\ldots\verb"\end{abbreviations}", immediately following the \verb"abstract" environment.

\subsection{Keywords}

A list of keywords may be included if required, enclosed within a \texttt{keywords} environment, i.e.\ \verb"\begin{keywords}"\ldots\verb"\end{keywords}". Additional keywords in other languages (preceded by a translation of the word `keywords') may also be included within the \verb"keywords" environment if required.

\subsection{Subject classification codes}

AMS, JEL or PACS classification codes may be included if required. The \texttt{interact} class file provides an \texttt{amscode} environment, i.e.\ \verb"\begin{amscode}"\ldots\verb"\end{amscode}", a \texttt{jelcode} environment, i.e.\ \verb"\begin{jelcode}"\ldots\verb"\end{jelcode}", and a \texttt{pacscode} environment, i.e.\ \verb"\begin{pacscode}"\ldots\verb"\end{pacscode}" to assist with this.

\subsection{Additional footnotes to the title or authors' names}

The \verb"\thanks" command may be used to create additional footnotes to the title or authors' names if required. Footnote symbols for this purpose should be used in the order
$^\ast$~(coded as \verb"$^\ast$"), $\dagger$~(\verb"$\dagger$"), $\ddagger$~(\verb"$\ddagger$"), $\S$~(\verb"$\S$"), $\P$~(\verb"$\P$"), $\|$~(\verb"$\|$"),
$\dagger\dagger$~(\verb"$\dagger\dagger$"), $\ddagger\ddagger$~(\verb"$\ddagger\ddagger$"), $\S\S$~(\verb"$\S\S$"), $\P\P$~(\verb"$\P\P$").

Note that any \verb"footnote"s to the main text will automatically be assigned the superscript symbols 1, 2, 3, etc. by the class file.\footnote{If preferred, the \texttt{endnotes} package may be used to set the notes at the end of your text, before the bibliography. The symbols will be changed to match the style of the journal if necessary by the typesetter.}

\section{Some guidelines for using the standard features of \LaTeX}

\subsection{Sections}

The \textsf{Interact} layout style allows for five levels of section heading, all of which are provided in the \texttt{interact} class file using the standard \LaTeX\ commands \verb"\section", \verb"\subsection", \verb"\subsubsection", \verb"\paragraph" and \verb"\subparagraph". Numbering will be automatically generated for all these headings by default.

\subsection{Lists}

Numbered lists are produced using the \texttt{enumerate} environment, which will number each list item with arabic numerals by default. For example,
\begin{enumerate}
  \item first item
  \item second item
  \item third item
\end{enumerate}
was produced by
\begin{verbatim}
\begin{enumerate}
  \item first item
  \item second item
  \item third item
\end{enumerate}
\end{verbatim}
Alternative numbering styles can be achieved by inserting an optional argument in square brackets to each \verb"item", e.g.\ \verb"\item[(i)] first item"\, to create a list numbered with roman numerals at level one.

Bulleted lists are produced using the \texttt{itemize} environment. For example,
\begin{itemize}
  \item First bulleted item
  \item Second bulleted item
  \item Third bulleted item
\end{itemize}
was produced by
\begin{verbatim}
\begin{itemize}
  \item First bulleted item
  \item Second bulleted item
  \item Third bulleted item
\end{itemize}
\end{verbatim}

\subsection{Figures}

The \texttt{interact} class file will deal with positioning your figures in the same way as standard \LaTeX. It should not normally be necessary to use the optional \texttt{[htb]} location specifiers of the \texttt{figure} environment in your manuscript, although the \texttt{[p]} option -- i.e.\ \verb"\begin{figure}[p]" -- is useful if you are asked to separate figures from the text.

Figure captions should appear below the figure itself, therefore the \verb"\caption" command should appear after the figure.
For example, Figure~\ref{sample-figure} with caption and sub-captions is produced using the following commands:
\begin{verbatim}
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{
\resizebox*{5cm}{!}{\includegraphics{graph1.eps}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{
\resizebox*{5cm}{!}{\includegraphics{graph2.eps}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}
\end{verbatim}
\begin{figure}
\centering
\subfigure[An example of an individual figure sub-caption.]{
\resizebox*{5cm}{!}{\includegraphics{graph1.eps}}}\hspace{5pt}
\subfigure[A slightly shorter sub-caption.]{
\resizebox*{5cm}{!}{\includegraphics{graph2.eps}}}
\caption{Example of a two-part figure with individual sub-captions
 showing that captions are flush left and justified if greater
 than one line of text.} \label{sample-figure}
\end{figure}

To ensure that figures are correctly numbered automatically, the \verb"\label" command should be included just after the \verb"\caption" command, or in its argument.

The \verb"\subfigure" command requires \verb"subfigure.sty", which is called in the preamble of the \texttt{interactapasample.tex} file (in order to allow your choice of an alternative package if preferred) and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Please supply any additional figure macros you use with your article in the preamble of your .tex file, before \verb"\begin{document}".

The source files of any figures will be required when the final, revised version of a manuscript is submitted. Authors should ensure that these are suitable (in terms of lettering size, etc.) for the reductions they envisage.

The \texttt{epstopdf} package can be used to incorporate encapsulated PostScript (.eps) illustrations when using PDF\LaTeX, etc. Please provide the original .eps source files rather than the generated PDF images of those illustrations for production purposes.

\subsection{Tables}

The \texttt{interact} class file will deal with positioning your tables in the same way as standard \LaTeX. It should not normally be necessary to use the optional \texttt{[htb]} location specifiers of the \texttt{table} environment in your manuscript, although the \texttt{[p]} option -- i.e.\ \verb"\begin{table}[p]" -- is useful if you are asked to separate tables from the text.

The \texttt{tabular} environment can be used as illustrated here to produce tables with single horizontal rules at the head, foot and elsewhere as appropriate.
The table caption appears above the body of the table in the \textsf{Interact} layout style, therefore the \verb"\tbl" command should be used before the body of the table.
For example, Table~\ref{sample-table} is produced using the following commands:
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
\begin{verbatim}
\begin{table}
\tbl{Example of a table showing that its caption is as wide as
 the table itself and justified.}
{\begin{tabular}{lcccccc} \toprule
 & \multicolumn{2}{l}{Type} \\ \cmidrule{2-7}
 Class & One & Two & Three & Four & Five & Six \\ \midrule
 Alpha\textsuperscript{a} & A1 & A2 & A3 & A4 & A5 & A6 \\
 Beta & B2 & B2 & B3 & B4 & B5 & B6 \\
 Gamma & C2 & C2 & C3 & C4 & C5 & C6 \\ \bottomrule
\end{tabular}}
\tabnote{\textsuperscript{a}This footnote shows how to include
 footnotes to a table if required.}
\label{sample-table}
\end{table}
\end{verbatim}

To ensure that tables are correctly numbered automatically, the \verb"\label" command should be included just before \verb"\end{table}".

The \verb"\toprule", \verb"\midrule", \verb"\bottomrule" and \verb"\cmidrule" commands require \verb"booktabs.sty", which is called by the \texttt{interact} class file and included in the \textsf{Interact} \LaTeX\ bundle for convenience. Tables produced using the standard commands of the \texttt{tabular} environment are also compatible with the \texttt{interact} class file.

\subsection{Landscape pages}

If a figure or table is too wide to fit the page it will need to be rotated, along with its caption, through 90$^{\circ}$ anticlockwise.
Landscape figures and tables can be produced using the \verb"rotating" package, which is called by the \texttt{interact} class file.
The following commands (for example) can be used to produce such pages.
\begin{verbatim}
\setcounter{figure}{1}
\begin{sidewaysfigure}
\centerline{\epsfbox{figname.eps}}
\caption{Example landscape figure caption.}
\label{landfig}
\end{sidewaysfigure}
\end{verbatim}
\begin{verbatim}
\setcounter{table}{1}
\begin{sidewaystable}
 \tbl{Example landscape table caption.}
  {\begin{tabular}{@{}llllcll}
    .
    .
    .
  \end{tabular}}\label{landtab}
\end{sidewaystable}
\end{verbatim}
Before any such float environment, use the \verb"\setcounter" command as above to fix the numbering of the caption (the value of the counter being the number given to the preceding figure or table). Subsequent captions will then be automatically renumbered accordingly. The \verb"\epsfbox" command requires \verb"epsfig.sty", which is called by the \texttt{interact} class file and is also included in the \textsf{Interact} \LaTeX\ bundle for convenience.

\subsection{Theorem-like structures}

A predefined \verb"proof" environment is provided by the \texttt{amsthm} package (which is called by the \texttt{interact} class file), as follows:
\begin{proof}
More recent algorithms for solving the semidefinite programming relaxation are particularly efficient, because they explore the structure of the MAX-CUT problem.
\end{proof}
\noindent This was produced by simply typing:
\begin{verbatim}
\begin{proof}
More recent algorithms for solving the semidefinite programming
relaxation are particularly efficient, because they explore the
structure of the MAX-CUT problem.
\end{proof}
\end{verbatim}
Other theorem-like environments (theorem, definition, remark, etc.) need to be defined as required, e.g.\ using \verb"\newtheorem{theorem}{Theorem}" in the preamble of your .tex file (see the preamble of \verb"interacttfvsample.tex" for more examples). You can define the numbering scheme for these structures however suits your article best. Please note that the format of the text in these environments may be changed if necessary to match the style of individual journals by the typesetter during preparation of the proofs.

\subsection{Mathematics}

\subsubsection{Displayed mathematics}

The \texttt{interact} class file will set displayed mathematical formulas centred on the page without equation numbers if you use the \texttt{displaymath} environment or the equivalent \verb"\[...\]" construction. For example, the equation
\[
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\]
was typeset using the commands
\begin{verbatim}
\[
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\]
\end{verbatim}

For those of your equations that you wish to be automatically numbered sequentially throughout the text, use the \texttt{equation} environment, e.g.
\begin{equation}
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\end{equation}
was typeset using the commands
\begin{verbatim}
\begin{equation}
 \hat{\theta}_{w_i} = \hat{\theta}(s(t,\mathcal{U}_{w_i}))
\end{equation}
\end{verbatim}

Part numbers for sets of equations may be generated using the \texttt{subequations} environment, e.g.
\begin{subequations} \label{subeqnexample}
\begin{equation}
        \varepsilon \rho w_{tt}(s,t)
        = N[w_{s}(s,t),w_{st}(s,t)]_{s},
        \label{subeqnpart}
\end{equation}
\begin{equation}
        w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0,
\end{equation}
\end{subequations}
which was generated using the commands
\begin{verbatim}
\begin{subequations} \label{subeqnexample}
\begin{equation}
        \varepsilon \rho w_{tt}(s,t)
        = N[w_{s}(s,t),w_{st}(s,t)]_{s},
        \label{subeqnpart}
\end{equation}
\begin{equation}
        w_{tt}(1,t)+N[w_{s}(1,t),w_{st}(1,t)] = 0,
\end{equation}
\end{subequations}
\end{verbatim}
This is made possible by the \texttt{amsmath} package, which is called by the class file. If you put the \verb"\label{}" just after the \verb"\begin{subequations}" line, references will be to the collection of equations, `(\ref{subeqnexample})' in the example above. Or, like the example code above, you can reference each equation individually -- e.g.\ `(\ref{subeqnpart})'.

Displayed mathematics should be given end-of-line punctuation appropriate to the running text sentence of which it forms a part, if required.

\subsubsection{Math fonts}

\paragraph{Superscripts and subscripts}
Superscripts and subscripts will automatically come out in the correct size in a math environment (i.e.\ enclosed by `\verb"$"' delimiters in running text, or within \verb"\[...\]" or the `\texttt{equation}' environment for displayed equations). Sub/superscripts that are physical variables should be italic, whereas those that are labels should be roman (e.g.\ $C_p$, $T_\mathrm{eff}$). If the subscripts or superscripts need to be other than italic, they must be coded individually.

\paragraph{Upright Greek characters and the upright partial derivative sign}
Upright lowercase Greek characters can be obtained by inserting the letter `u' in the control code for the character, e.g.\ \verb"\umu" and \verb"\upi" produce $\umu$ (used, for example, in the symbol for the unit microns -- $\umu\mathrm{m}$) and $\upi$ (the ratio of the circumference of a circle to its diameter). Similarly, the control code for the upright partial derivative $\upartial$ is \verb"\upartial". Bold lowercase as well as uppercase Greek characters can be obtained by \verb"{\bm \gamma}", for example, which gives ${\bm \gamma}$, and \verb"{\bm \Gamma}", which gives ${\bm \Gamma}$.

\section*{Acknowledgement(s)}

An unnumbered section, e.g.\ \verb"\section*{Acknowledgements}", may be used for thanks, etc.\ if required and included \emph{in the non-anonymous version} before any Notes or References.

\section*{Disclosure statement}

An unnumbered section, e.g.\ \verb"\section*{Disclosure statement}", may be used to declare any potential conflict of interest and included \emph{in the non-anonymous version} before any Notes or References, after any Acknowledgements and before any Funding information.

\section*{Funding}

An unnumbered section, e.g.\ \verb"\section*{Funding}", may be used for grant details, etc.\ if required and included \emph{in the non-anonymous version} before any Notes or References.

\section*{Notes on contributor(s)}

An unnumbered section, e.g.\ \verb"\section*{Notes on contributors}", may be included \emph{in the non-anonymous version} if required. A photograph may be added if requested.

\section*{Notes}

An unnumbered `Notes' section may be included before the References (if using the \verb"endnotes" package, use the command \verb"\theendnotes" where the notes are to appear, instead of creating a \verb"\section").

\section{References}

\subsection{References cited in the text}

References should be cited in the text in Taylor \& Francis Reference Style V, separated by commas, e.g.\ `\citep{Got89,Eva94,Azi97}' or `\ldots see \citet[p.~75]{Che51}'. If there are more than two authors, cite the first author's surname followed by \textit{et~al}. For two or more sources by the same author(s) in the same year, use lower-case letters (a, b, c, etc.) with the year to order the entries in the References list and use these letters with the year in the in-text citations, e.g.\ `(Johnson 1994a, 1994b)'. For further details on this reference style, see the Instructions for Authors on the Taylor \& Francis website.

Each bibliographic entry has a key, which is assigned by the author and is used to refer to that entry in the text. In this document, the key \verb"Agu95" in the citation form \verb"\citep{Agu95}" produces `\citep{Agu95}', and the keys \verb"Dow64", \verb"CWW86", \verb"Mer93" in the citation form \verb"\citep{Dow64,CWW86,Mer93}" produce `\citep{Dow64,CWW86,Mer93}'. The appropriate citation style for different situations can be obtained, for example, by \verb"\citet{Ban95}" for `\citet{Ban95}', \verb"\citealt{Gre90}" for `\citealt{Gre90}', or \verb"\citealp{Ano64}" for `\citealp{Ano64}'. Citation of the year alone may be produced by \verb"\citeyear{SCA97}", i.e.\ `\citeyear{SCA97}', or \verb"\citeyearpar{CRE78}", i.e.\ `\citeyearpar{CRE78}', or of the author(s) alone by \verb"\citeauthor{Hol04}", i.e.\ `\citeauthor{Hol04}'. Optional notes may be included at the beginning and/or end of a citation by the use of square brackets, e.g.\ \verb"\citep[see][p.~21]{Pea76}" produces `\citep[see][p.~21]{Pea76}'; \verb"\citep[e.g.][]{Haw94}" produces `\citep[e.g.][]{Haw94}'; \verb"\citet[p.~327]{Sil91}" produces `\citet[p.~327]{Sil91}'. A `plain' \verb"\cite" command will produce the same result as a \verb"\citet", i.e.\ \verb"\cite{Une93}" will produce `\cite{Une93}'.

\subsection{The list of references}

References should be listed at the end of the main text in alphabetical order by authors' surnames, then chronologically (earliest first).
If a reference has no author or editor, order by title. %If a date of publication is impossible to find, use `n.d.' in its place.
A single-author entry precedes a multi-author entry that begins with the same name.
If references have the same author(s), editor(s), etc., arrange by year of publication, with undated works at the end.
%Successive entries by two or more authors when only the first author is the same are alphabetized by co-authors' surnames.
If the reference list contains two or more items by the same author(s) in the same year, add a, b, c, etc.\ and list them alphabetically by title.
If a reference has more than three named authors, list only the first, followed by `\textit{et~al.}'.
The following list shows some sample references prepared in Taylor \& Francis Reference Style V.

\begin{thebibliography}{}

\bibitem[Agutter(1995)]{Agu95}
Agutter, A.J., 1995. \emph{The linguistic significance of current British
  slang}. Thesis (PhD). Edinburgh University.

\bibitem[Anon(1964)]{Ano64}
Anon, 1964. \emph{A history of radio}. London: Beacon Press.

\bibitem[Aziz(1997)]{Azi97}
Aziz, A., 1997. \emph{Method and apparatus for a key management scheme}. United
  States Patent Application 68-665, 27 June.

\bibitem[Bantz(1995)]{Ban95}
Bantz, C.R., 1995. Social dimensions of software development. \emph{In}: J.A.
  Anderson, ed. \emph{Annual review of software management and development}.
  Newbury Park, CA: Sage, 502--510.

\bibitem[Chester(1951)]{Che51}
Chester, D.N., ed., 1951. \emph{Lessons of the British war economy}. Westport,
  CT: Greenwood Press.

\bibitem[Commission for Racial Equality(1978)]{CRE78}
Commission for Racial Equality, 1978. \emph{Multi-racial Britain}. London:
  Commission for Racial Equality.

\bibitem[Cutler \emph{et~al.}(1986)]{CWW86}
Cutler, T., Williams, K., and Williams, J., 1986. \emph{Keynes, Beveridge and
  beyond}. London: Routledge.

\bibitem[Dow(1964)]{Dow64}
Dow, D., 1964. \emph{A history of the world}. 3rd ed. London: Greenfield.

\bibitem[Evans(1994)]{Eva94}
Evans, W.A., 1994. Approaches to intelligent information retrieval.
  \emph{Information Processing and Management}, 7 (2), 174--168.

\bibitem[Gott(1989)]{Got89}
Gott, R., 1989. Crumbs and the capitalists. \emph{The Guardian}, 20~Jan, p.~2.

\bibitem[Greenfield(1990)]{Gre90}
Greenfield, J., 1990. The Sevso Treasure: the legal case. \emph{Apollo},
  132 (3), 14--16.

\bibitem[Hawking(1994)]{Haw94}
Hawking, S.W., 1994. \emph{A brief history of time: an interactive adventure}
  [CD-ROM]. Crunch media.

\bibitem[Holland(2004)]{Hol04}
Holland, M., 2004. \emph{Guide to citing Internet sources} [online]. Poole:
  Bournemouth University. Available from:
  http://www.bournemouth.ac.uk/library/using/guide\_to\_citing\_internet\_sourc.html
  [Accessed 4~November 2004].

\bibitem[Mercer and Smith(1993)]{Mer93}
Mercer, P.A. and Smith, G., 1993. \emph{Private viewdata in the UK}. 2nd ed.
  London: Longman.

\bibitem[Pearce \emph{et~al.}(1976)]{Pea76}
Pearce, I.F., \emph{et~al.}, 1976. \emph{A model of output, employment, wages
  and prices in the UK}. Cambridge: Cambridge University Press.

\bibitem[School Curriculum and Assessment Authority(1997)]{SCA97}
School Curriculum and Assessment Authority, 1997. \emph{Guide to national
  tests}. London: The Stationery Office.

\bibitem[Silver(1991)]{Sil91}
Silver, K., 1991. Electronic mail: The new way to communicate. \emph{In}: D.I.
  Raitt, ed. \emph{9th International Online Information Meeting}, 3--5 December
  1990, London. Oxford: Learned Information, 323--330.

\bibitem[Unesco(1993)]{Une93}
Unesco, 1993. \emph{General information programme}. Paris: Unesco,
  PGI-93/WS/22.

\end{thebibliography}
\bigskip
\noindent This was produced by typing:
\begin{verbatim}
\begin{thebibliography}{}

\bibitem[Agutter(1995)]{Agu95}
Agutter, A.J., 1995. \emph{The linguistic significance of current
 British slang}. Thesis (PhD). Edinburgh University.

\bibitem[Anon(1964)]{Ano64}
Anon, 1964. \emph{A history of radio}. London: Beacon Press.

\bibitem[Aziz(1997)]{Azi97}
Aziz, A., 1997. \emph{Method and apparatus for a key management
 scheme}. United States Patent Application 68-665, 27 June.

\bibitem[Bantz(1995)]{Ban95}
Bantz, C.R., 1995. Social dimensions of software development.
 \emph{In}: J.A. Anderson, ed. \emph{Annual review of software
 management and development}. Newbury Park, CA: Sage, 502--510.

\bibitem[Chester(1951)]{Che51}
Chester, D.N., ed., 1951. \emph{Lessons of the British war economy}.
 Westport, CT: Greenwood Press.

\bibitem[Commission for Racial Equality(1978)]{CRE78}
Commission for Racial Equality, 1978. \emph{Multi-racial Britain}.
 London: Commission for Racial Equality.

\bibitem[Cutler \emph{et~al.}(1986)]{CWW86}
Cutler, T., Williams, K., and Williams, J., 1986. \emph{Keynes,
 Beveridge and beyond}. London: Routledge.

\bibitem[Dow(1964)]{Dow64}
Dow, D., 1964. \emph{A history of the world}. 3rd ed. London:
 Greenfield.

\bibitem[Evans(1994)]{Eva94}
Evans, W.A., 1994. Approaches to intelligent information retrieval.
 \emph{Information Processing and Management}, 7 (2), 174--168.

\bibitem[Gott(1989)]{Got89}
Gott, R., 1989. Crumbs and the capitalists. \emph{The Guardian},
 20~Jan, p.~2.

\bibitem[Greenfield(1990)]{Gre90}
Greenfield, J., 1990. The Sevso Treasure: the legal case.
 \emph{Apollo}, 132 (3), 14--16.

\bibitem[Hawking(1994)]{Haw94}
Hawking, S.W., 1994. \emph{A brief history of time: an interactive
 adventure} [CD-ROM]. Crunch media.

\bibitem[Holland(2004)]{Hol04}
Holland, M., 2004. \emph{Guide to citing Internet sources} [online].
 Poole: Bournemouth University. Available from:
 http://www.bournemouth.ac.uk/library/using/guide\_to\_citing\_internet\_sourc.html
 [Accessed 4~November 2004].

\bibitem[Mercer and Smith(1993)]{Mer93}
Mercer, P.A. and Smith, G., 1993. \emph{Private viewdata in the UK}.
 2nd ed. London: Longman.

\bibitem[Pearce \emph{et~al.}(1976)]{Pea76}
Pearce, I.F., \emph{et~al.}, 1976. \emph{A model of output, employment,
 wages and prices in the UK}. Cambridge: Cambridge University
 Press.

\bibitem[School Curriculum and Assessment Authority(1997)]{SCA97}
School Curriculum and Assessment Authority, 1997. \emph{Guide to
 national tests}. London: The Stationery Office.

\bibitem[Silver(1991)]{Sil91}
Silver, K., 1991. Electronic mail: The new way to communicate.
 \emph{In}: D.I. Raitt, ed. \emph{9th International Online Information
 Meeting}, 3--5 December 1990, London. Oxford: Learned Information,
 323--330.

\bibitem[Unesco(1993)]{Une93}
Unesco, 1993. \emph{General information programme}. Paris: Unesco,
 PGI-93/WS/22.

\end{thebibliography}
\end{verbatim}
\bigskip
\noindent Each entry takes the form:
\begin{verbatim}
\bibitem[authors' names(date of publication)]{key}
 Bibliography entry
\end{verbatim}
where `\texttt{authors' names}' is the list of names to appear where the \verb"bibitem" is cited in the text, and `\texttt{key}' is the tag that is to be used as an argument for the \verb"\cite{}" commands in the text of the article. `\texttt{Bibliography entry}' is the material that is to appear in the list of references, suitably formatted. The commands
\begin{verbatim}
\usepackage{natbib}
\bibpunct[, ]{(}{)}{;}{a}{}{,}
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}
\end{verbatim}
need to be included in the preamble of your .tex file in order to generate the citations and bibliography as described above.

Instead of typing the bibliography by hand, you may prefer to create the list of references using a \textsc{Bib}\TeX\ database. The \texttt{tfv.bst} file needs to be in your working folder or an appropriate directory, and the lines
\begin{verbatim}
\bibliographystyle{tfv}
\bibliography{interacttfvsample}
\end{verbatim}
included where the list of references is to appear, where \texttt{tfv.bst} is the name of the \textsc{Bib}\TeX\ bibliography style file for Taylor \& Francis Reference Style V and \texttt{interacttfvsample.bib} is the bibliographic database included with the \textsf{Interact}-TFV \LaTeX\ bundle (to be replaced with the name of your own .bib file). \LaTeX/\textsc{Bib}\TeX\ will extract from your .bib file only those references that are cited in your .tex file and list them in the References section.

Please include a copy of your .bib file and/or the final generated .bbl file among your source files if your .tex file does not contain a reference list in a \texttt{thebibliography} environment.

\section{Appendices}

Any appendices should be placed after the list of references, beginning with the command \verb"\appendix" followed by the command \verb"\section" for each appendix title, e.g.
\begin{verbatim}
\appendix
\section{This is the title of the first appendix}
\section{This is the title of the second appendix}
\end{verbatim}
produces:\medskip

\noindent\textbf{Appendix A. This is the title of the first appendix}\medskip

\noindent\textbf{Appendix B. This is the title of the second appendix}\medskip

\noindent Subsections, equations, figures, tables, etc.\ within appendices will then be automatically numbered as appropriate. Some theorem-like environments may need to have their counters reset manually (e.g.\ if they are not numbered within sections in the main text). You can achieve this by using \verb"\numberwithin{remark}{section}" (for example) just after the \verb"\appendix" command.

\appendix

\section{Troubleshooting}

Authors may occasionally encounter problems with the preparation of a manuscript using \LaTeX. The appropriate action to take will depend on the nature of the problem:
\begin{enumerate}
\item[(i)] If the problem is with \LaTeX\ itself, rather than with the actual macros, please consult an appropriate \LaTeXe\ manual for initial advice. If the solution cannot be found, or if you suspect that the problem does lie with the macros, then please contact Taylor \& Francis for assistance (\texttt{latex.helpdesk@tandf.co.uk}).
\item[(ii)] Problems with page make-up (e.g.\ occasional overlong lines of text; figures or tables appearing out of order): please do not try to fix these using `hard' page make-up commands -- the typesetter will deal with such problems. (You may, if you wish, draw attention to particular problems when submitting the final version of your manuscript.)
\item[(iii)] If a required font is not available on your system, allow \TeX\ to substitute the font and specify which font is required in a covering letter accompanying your files.
\end{enumerate}

\section{Obtaining the template and class file}

\subsection{Via the Taylor \& Francis website}

This article template and the \texttt{interact} class file may be obtained via the `Instructions for Authors' pages of selected Taylor \& Francis journals.

Please note that the class file calls up the open-source \LaTeX\ packages booktabs.sty, epsfig.sty and rotating.sty, which will, for convenience, unpack with the downloaded template and class file. The template calls for natbib.sty and subfigure.sty, which are also supplied for convenience.

\subsection{Via e-mail}

This article template, the \texttt{interact} class file and the associated open-source \LaTeX\ packages are also available via e-mail. Requests should be addressed to \texttt{latex.helpdesk@tandf.co.uk}, clearly stating for which journal you require the template and class file.

\end{document}

### 2018-06-24 paper/paper.tex
wed end of the day
writing from bus
journal template and more writing

@@
-\documentclass[11t, a4paper, twocolumn]{article} 
-\input{structure.tex}
-\addbibresource{references.bib}
+\documentclass[]{interact}
+\usepackage{epstopdf}
+\usepackage{subfigure}
+\usepackage{natbib}
+\bibpunct[, ]{(}{)}{,}{a}{}{,}
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+\begin{document}
+
 	\title{Estimating real-time highstreet footfall from Wi-Fi probe requests}
 	\author{
-	\authorstyle{
-		Balamurugan Soundararaj\textsuperscript{1}, 
-		James Cheshire\textsuperscript{1} and 
-		Paul Longley\textsuperscript{1}}
-	\newline\newline
-	\textsuperscript{1} \institution{
-		Department of Geography, 
-		University College London, 
-		United Kingdom } }
-\date{\today}
-\begin{document}
+		\name{ Balamurugan Soundararaj\textsuperscript{a}\thanks{CONTACT Balamurugan Soundararaj. Email: s.bala@ucl.ac.uk}, James Cheshire\textsuperscript{a} and Paul Longley\textsuperscript{a} }
+		\affil{\textsuperscript{a}Department of Geography, University College London, United Kingdom.}
+	}
 	\maketitle
-	\thispagestyle{firstpage}
-	\textbf{Abstract -}
-		\input{sections/abstract}
+
+	\begin{abstract}
+		\input{sections/abstract} \end{abstract}
+	\begin{keywords}
+		Highstreet footfall; Wi-Fi Probe requests; Sensors; MAC Randomisation \end{keywords}
 	\section{Introduction}\label{introduction}
@@
 		\input{sections/main_study}
-	\section{Discussion}\label{discussion}
-		\input{sections/discussion}
 	\section{Conclusion}\label{conclusion}
-		\input{sections/discussion}
-	\printbibliography[title={References}]
+		\input{sections/conclusion}
+	\section*{Acknowledgement}
+		\input{sections/acknowledgement}
+	\bibliographystyle{tfv} \bibliography{references}
+
 \end{document}

### 2018-06-24 paper/sections/abstract.tex
journal template and more writing

The accurate measurement human activity with high spatial and temporal granularity is crucial for us to understand the structure and function of built enviroment.
With increasing mobile ownerships, the Wi-Fi probe requests generated by these devices can be an excellent source of data for such measursement.
The major challanges in using Wi-Fi technology for such purpose are precisely delineating the field of measurement and arriving at accurate estimates without compromising the privacy of the users of the mobile devices.
In this paper we demonstrate that, with the application of class intervals and a novel graph based technique, we can overcome the challanges and reliably measure real-time footfall at retail highstreets in London.

### 2018-06-24 paper/sections/introduction.tex
journal template and more writing

@@
 
-Probe requests are low level signals standardised by IEEE 802.1b/g specification as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
+Probe requests are low level signals standardised by IEEE 802.1b/g specification \citep{ieee2013} as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
 This ubiquity and standardisation make them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.

### 2018-06-24 paper/sections/main_study.tex
main analysis
journal template and more writing

@@
 for different locations and different times.
+check if the thresholds are consistent.
 Test if the calibration works over intervals
@@
 
-\subsection{locations}
 five locations were selected across central london which
@@
 \end{enumerate}
-installations were carried out over the time period from xxxx to xxxx.
+Installations were carried out over the time period from xxxx to xxxx.
 the data collection happened from xxxx to xxxx. Manual counting was carried out
@@
 The difference in methods could lead to some inaccuracies in data.
-The overall statistics of data collected. 
-The overally schedule is shown in half page graphic.
+The overall schedule is shown in half page graphic.
 
-\subsection{Signal strength filtering}
+The overall statistics of data collected. How many probe requests.
+Comparision of location in terms of volume, patterns in daily footfall etc.
+The comparision between global and local.
+comparision between different types of vendors.
+Specifics on top 5 manufacturers.
 
-\subsection{Device Fingerprinting}
+We do a daily analysis of distribution of signal strengths.
+The thresholds are shown in the table.
+The average is xxxx and standard deviation is xxx.
+we notice that the variation is lot. There is a definite
+change with the micro site locations.
 
-\subsection{Manual Calibration}
+We see how the signal strength filtering affects the counts.
+compared to manual counts we look at the average mean errors (daily)
+per 5 minutes. The counts go as follows.
+Shown as a red line in the Figure.
 
-\subsection{Discussion}
+We can conclude that even though it has variations, this is
+a good method to reduce the overall level of error.
+
+This is also done for different locations hourly for all the data we had.
+we compare it to the manual counts and see that the average mean error
+has been reduced/increased. The finger print works well for all the locations.
+It also works over a period of time and gives us a comparable and close 
+footfall count to the manual count. The thresholds found in the pilot study
+works as well.
+
+Finally we normalise the sensor counts to match the manual count using
+a fraction/ adjustment factor calculated from the know manual counts.
+we have three sets of counts. We check if the adjustment factor holds the
+same in all three counts across locations. It does with a variations from
+xxx to xxx. The results are shown in the table.
+
+we see that the signal strength filtering works and reduces error by
+xxxxx. there is variation by locations.
+we see that sequence number algorithm works as well. The threshold stays
+constant well and works well across locations. The calibration also works
+and ajustment factor stays consistent short term. This needs more work 
+long term.

### 2018-06-24 paper/sections/methodology.tex
journal template and more writing

@@
-The overall methodology is to use a WiFi antenna to collect probe requests,
-assign a unique identification based on the device that created the request,
-aggregate them based on the unique identification for a specific time interval
-to create an estimate of number of people at the location.
-In this section we look at the characteristics of probe requests in detail,
-outline the methodology used to collect these probe requests, 
-look at the uncertainities and biases in the process and 
-device methods to overcome these issues.
+The primary aim of this research is to enable us to collect a series of probe requests and process them into an usable pedestrian footfall count.
+We do this by using a Wi-Fi receiver to collect probe requests broadcasted by mobile devices, filtering out the background noise and aggregating them based on the device that generated them.
+We begin by looking at the characteristics of probe requests in detail, device a methodology to collect these probe requests in public areas, examine the systemic biases and uncertainties in the data collection method and device data processing methods to overcome these challanges.
+Finally we compare the processed footfall counts to the ground truth recorded by primary surveys.
 
 \subsection{Probe Requests}
-Probe request is a low level packet sent by a mobile device as a means of
-"scanning" for various access points available at a specific location.
-This packet consists information about the mobile device
-including but not limited to,
-Media Access Control address - A two part 12 bit identifier where the first part
-identifies the manufacturer of the device and the second part identifies the
-device itself. This MAC address is of two types, global - the real identifier
-of the device which doesnt change and local - virtual, random addresses used
-for temporary situations.
-Sequence number of the packet to keep track of the replies.
-The access points for which the packet is being sent to.
-Capabilities of the device.
-We can also infer other things about the packet such as
-time at which the packet has been received,
-total length of the packet
-time it took to transmit the packet
-Signal strength of the packet.
-All of this can help us identifying the packets and label them when they are transmitted by the same device.
+Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available access points (AP), quick geo-location by triangulation based known APs, etc.
+These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or model of the devices though there is some variation on the frequency and the information transmitted through them.
+In some cases, such as Android devices, these are broadcast even when the Wi-Fi functionality has been turned off by the user.
+Thus these signals can be used to reliably identify the presence of Wi-Fi enabled mobile devices.
+
+Being a first step of connection initiated by the mobile device, these packets have information regarding the characteristics of the mobile device itself.
+Some of the key information we can infer from these requests are,
+\begin{enumerate}
+	\item Media Access Control (MAC) address which is an unique identifier for the wireless hardware of the mobile device,
+	\item Sequence number of the request for the mobile device to keep track of the responses,
+	\item Timestamp at which the request was received by the AP,
+	\item Total length of the request in number of bits, and 
+	\item The strength of the signal which transmitted the request.
+\end{enumerate}
+The MAC address is the primary unique identifier for the mobile device.
+It has two parts, first part gives information about the manufacturer of the device and the second is uniue to the device.
+The MAC address can be randomised (hence non unique) and is marked as such.
+Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate unique devices.
 
@@
 save the relevant packets.
-The filters applied are management frame > probe requests > broadcast. 
+The filters applied are management frame to probe requests to broadcast. 
 This is then parsed, personal information is hashed, 

### 2018-06-24 analysis/main.r

@@ -1,0 +1,35 @@
+suppressMessages(library('tidyverse'))
+
+if(!exists("locations")) { locations <- read.csv("../data/main_locations.csv") }
+if(!exists("manual")) { manual <- read.csv("../data/main_manual.csv") }
+if(!exists("sensor")) { sensor <- read.csv("../data/main_probes.csv") }
+
+if(!exists("vendors")) {
+	vendors <- (function(url = "https://goo.gl/RzzYwh") {
+		lines <- readLines(url)
+		lines <- lines[!substring(lines, 1, 1) == '#']
+		lines <- lines[!lines == '']
+		lines <- lapply(lines, function(x) {
+							 record <- strsplit(x, '((\t))+')[[1]]
+							 if(is.na(record[4])) { record[4] = '' }
+							 if(is.na(record[3])) { record[3] = '' }
+							 if(length(record) > 4) { record = record[1:4] }
+							 return(record) })
+		lines <- data.frame(matrix(
+								   unlist(lines), 
+								   nrow = length(lines), 
+								   byrow = T), 
+							stringsAsFactors = FALSE)
+		names(lines) <- c("oui", "vendor", "vendor_long", "comments")
+		lines <- lines[nchar(lines$oui) == 8, ]
+		lines$oui <- strsplit(lines$oui, ':') %>% 
+			lapply(function(x) { paste0(x, collapse = '') }) %>%
+			unlist() %>%
+			tolower()
+		return(lines) })() }
+
+if(is.null(sensor$vendor)) {
+	sensor <- merge(sensor, vendors[, 1:2], by = "oui", all.x = TRUE)[, union(names(sensor), names(vendors[, 1:2]))] }
+
+if(exists(sensor$oui_type)) {
+	sensor$oui_type <- ifelse(substr(sensor$oui, 2, 2) %in% c('2', '6', 'e', 'a'), "local", "global") }

### 2018-06-24 paper/sections/abstract.tex
Almost completed methodology

@@
 The major challanges in using Wi-Fi technology for such purpose are precisely delineating the field of measurement and arriving at accurate estimates without compromising the privacy of the users of the mobile devices.
-In this paper we demonstrate that, with the application of class intervals and a novel graph based technique, we can overcome the challanges and reliably measure real-time footfall at retail highstreets in London.
+In this paper we demonstrate that, with the application of class intervals and a novel graph based technique, we can overcome the challanges and reliably measure real-time pedestrian footfall at retail highstreets.

### 2018-06-24 paper/sections/methodology.tex
Almost completed methodology

@@
 
-\subsection{Probe Requests}
 Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available access points (AP), quick geo-location by triangulation based known APs, etc.
@@
 \begin{enumerate}
-	\item Media Access Control (MAC) address which is an unique identifier for the wireless hardware of the mobile device,
-	\item Sequence number of the request for the mobile device to keep track of the responses,
-	\item Timestamp at which the request was received by the AP,
-	\item Total length of the request in number of bits, and 
-	\item The strength of the signal which transmitted the request.
+	\item \textbf{Media Access Control (MAC) address} which is an unique identifier for the wireless hardware of the mobile device,
+	\item \textbf{Sequence number} of the request for the mobile device to keep track of the responses,
+	\item \textbf{Timestamp} at which the request was received by the AP,
+	\item Total \textbf{length} of the request in number of bits, and 
+	\item The \textbf{strength of the signal} which transmitted the request.
 \end{enumerate}
@@
 
-\subsection{Data Collection}
-The data collection involved two processes - 
-A sensor based collection of probe requests and
-a manual count of pedestrians in the area next to the sensor.
-We have detailed the sensor design below,
+Data collection was done with the help of custom sensors built from modifing the Smart street sensor \citep{sss2016} hardware and updating them with custom software.
+The sensor is essentially a Raspberry Pi connected with Wi-Fi and 3G antennae.
+It keeps the Wi-Fi module in `Monitor' mode and uses the open source software - wireshark cite to passively collect all packets sent to `broadcast', marked with type - `management' and subtype - `probe requests'.
+The MAC address in these probe requests is anonymised using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
+A overall schematic of the data collection process is shown in Figure.
+The ground truth on number of pedestrian footfall was recorded using a purpose built Android application cite.
 
-The sensor consists of wireless transponder in a linux/ MacOSX based machine
-which is put into a monitor mode and
-wireshark software (tshark) is used to sniff, filter and
-save the relevant packets.
-The filters applied are management frame to probe requests to broadcast. 
-This is then parsed, personal information is hashed, 
-compressed and sent to a server via web socket protocol.
-The server receives, logs and stores the data in a postgresql database,
-for further analysis.
-The manual count was undertaken using an Android application
-on a mobile phone where timestamp of every individual pedestrian footfall
-was recorded by touch screen. A schematic is shown in ref figure.
+The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following major challanges which arise from our collection methodology.
+\begin{enumerate}
+	\item Background noise - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest.
+This can lead to a signicant background noise at certain locations.
+E.g.
+a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
+	\item MAC randomisation - The mobile devices in the past few years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
+This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor.
+This along with the previous problem can further increase the magnitude of error by several fold.
+	\item Mobile ownership - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall.
+In addition to this, there is a long term overall increase in mobile ownership which may lead to the number of probe requests collected overtime.
+\end{enumerate}
+We propose the following methods to tackle each of these challages.
 
-\subsection{Estimating Footfall}
-The next step after collecting the probe request is to estimate the footfall or pedestrian activity from them. There are three major challanges in the above data collection methodology.
-the first one is deliniating the area of interest.
-WiFi and its range depends on lots of factors.
-need a way to isolate the noise from the relevant data. depends a lot on microsite so we need a methodology which provides us with a general solution without much tinkering.
-Second one is the mac randomisation.
-modern phones randomise their MAC addresses to avoid detection and tracking.
-This leads to overcounting and explosion of randomised (local) mac addresses.
-We need a non MAC dependent method to overcome this.
-Finally, inherent mobile phone carrying bias. we need to account for the average mobile ownership. which is steadily increasing and is not 1:1.
-
-\subsubsection{WiFi range}
+\subsection{Filtering with Signal Strength}
 One of the clues of the distance of the device generating the probe request is
@@
 
-\subsubsection{MAC randomisation}
+\subsection{Clustering with sequence numbers}
 This is a recent problem. ref in figure. how mac randomisation has caused problem
@@
 and links are created between them based on the following rules: 
-\begin{enumerate}
+\begin{itemize}
 	\item A link could go only forward in time. 
@@
 		shortest of all such possible links.
-\end{enumerate}
+\end{itemize}
 The nodes were then classified based on the unique connected component they belonged to.
@@
 
-\subsubsection{Mobile phone ownership}
+\subsection{Calibrating with manual counts}
 This is both a long term change and result of demographic factors.

### 2018-06-24 paper/sections/pilot_study.tex
Almost completed methodology

@@
 	\begin{center}
-		\includegraphics [width=\linewidth,trim=4 4 4 4,clip] {images/methodology_clustering.png}
+		\includegraphics [width=0.85\linewidth,trim=4 4 4 4,clip] {images/methodology_clustering.png}
 		\caption{Clustering probe requests based on increasing sequence numbers present in them.}
@@
 	\begin{center}
-		\includegraphics [width=\linewidth] {images/pilot_counts_comparision.png}
+		\includegraphics [width=0.85\linewidth] {images/pilot_counts_comparision.png}
 		\caption{Comparision of counts after filtering with manual counts}

### 2018-06-24 paper/sections/methodology.tex
Methodology writing complete

@@
 The MAC address is the primary unique identifier for the mobile device.
-It has two parts, first part gives information about the manufacturer of the device and the second is uniue to the device.
+It has two parts, first part is called an Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device.
 The MAC address can be randomised (hence non unique) and is marked as such.
@@
 \begin{enumerate}
-	\item Background noise - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest.
+	\item \textbf{Background noise} - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest.
 This can lead to a signicant background noise at certain locations.
@@
 a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
-	\item MAC randomisation - The mobile devices in the past few years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
+	\item \textbf{MAC randomisation} - The mobile devices in the past few years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
 This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor.
 This along with the previous problem can further increase the magnitude of error by several fold.
-	\item Mobile ownership - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall.
+	\item \textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall.
 In addition to this, there is a long term overall increase in mobile ownership which may lead to the number of probe requests collected overtime.
@@
 \subsection{Filtering with Signal Strength}
-One of the clues of the distance of the device generating the probe request is
-the strength of the signal recieved from it.
-The Signal strength varies in inverse square law over distance with a 
-propagation constant.
-It also depends on lot of micro site, micro temporal factors.
-There cannot be a simple rule to fit and filter for all configuration.
-Our hypothesis is that in a specific setting and specific source of noise,
-there must exist a clear break in the data.
-for example, if there is a phone shop next to our sensor where
-hundreds of phones regularly send lots of probe requests
-we should be able to see a large increase in number of probe requests around
-a specific signal strength.
-we can identify this sharp change/ break using class interval algorithms such as
-k-means, jenks, quantile, etc.
+One of the clues that we can use to estimate the distance between the mobie device and the sensor is the strength of the signal received by the sensor. 
+The obvious approach here is to try and establish a relationship between the signal strength and distance first and use this to filter out the unwanted probe requests.
+This approach has numerous pitfalls and uncertainities since the decay of signal strength with distance is not always constant.
+It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source transponder.
+This severely limits our ability to establishing a simple conversion between reported signal strength and distance.
+There is a need for a method which takes in to account these variables across varous locations.
+
+We hypothesise that in configurations with a specific source of background noise at a constant distance, there must be a distinct break in the number of probe requests reporting signal strength corresponding to that distance.
+For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported singal strength corresponding to the distance between the sensor and the phone shop at any given set of conditions.
+We could identify these breaks in the data using tradition class interval algorithms such as jenks natural breaks, k-means, quantile and hierarchical clustering, etc.
+Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.
+An illustrative example is show in figure.
 
 \subsection{Clustering with sequence numbers}
-This is a recent problem. ref in figure. how mac randomisation has caused problem
-MAC address has been our unique identifier. Now we need to look for others.
-the contenders are length, duration which seem to be unique for device
-sets of known wlans and capabilites which can give us unique finger print and 
-finally sequence numbers in the packets.
-This is a tricky one since it is neither unique nor aggregatable.
-we need a method to seperate sequences shown in fig.
-We propose a graph based clustering algorithm where
-each cluster corresponded to a unique device.
-The algorithm creates a graph where the probe requests represented the nodes, 
-and links are created between them based on the following rules: 
+
+Since our primary unique identifier - MAC adddress, is being anonymised by new devces, we need to find other information present in the probe request for a unique identifier.
+Obvious approach here is to establish a factor of randomisation and adjust the counts for these probe requests based on this factor.
+We found this aproach not feasible, since the proportion of devices which randomise the MAC addresses increases over time.
+There is also a wide variation in the frequency at which the devices randomise the MAC addresses and the method used for the process.
+This leads us to look for a more generalisable approach which is independent of the device model.
+
+From our initial analysis we found that OUI, length of the packet and sequence number of the packet being the most promising information to acheive this.
+First we divide our dataset into sets of probe requests with randomised and non-randomised MAC addresses and keep the MAC address as the unique identifier for the latter set. 
+For randomised ones we futher divide them in to sub categories based on their OUIand length of the packet.
+Since the length tends to stay unique to specific models of devices we are left with the task of identifying the unique mobile devices from within these distinct models.
+
+The proposed algorithm creates a graph where the probe requests represented the nodes, and links are created between them based on the following rules: 
 \begin{itemize}
 	\item A link could go only forward in time. 
-	\item A link could exist between nodes with a maximum
-		time difference of $\alpha$ (time threshold).
 	\item A link could go from low to high sequence numbers.
-	\item A link could exist between nodes with a maximum sequence
-		number difference of $\beta$ (sequence threshold).
-	\item A node could have only one incoming link and
-		one outgoing link, which is the
-		shortest of all such possible links.
+	\item A link could exist between nodes with a maximum time difference of $\alpha$ - time threshold.
+	\item A link could exist between nodes with a maximum sequence number difference of $\beta$ - sequence threshold.
+	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
 \end{itemize}
-The nodes were then classified based on the unique connected component they belonged to.
-This classification was assigned as the unique identifier for the anonymised probe requests
-this unique identifier is used instead of MAC to aggregation.
+The nodes were then classified based on the unique connected component they belong to.
+This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
+Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the maginitude of error is greatly reduced.
 
-\subsection{Calibrating with manual counts}
-This is both a long term change and result of demographic factors.
-Phone ownership is not 1:1. It changes slowly overtime. It also changes
-with place to place. It can be more in dense urban centers
-and can be low in rural areas. We propose a adjustment factor methodology
-we use periodically carried out manual counts to adjust the numbers
-to what is reported on ground. The adjustment is as strong as the amount of
-ground truth we collected.
+\subsection{Calibrating with ground truth}
+Since mobile phone ownership is a external uncertainty to our study and could arise from variety of spatio, temporal and demographic factors, we propose to solve this by using external source of information.
+We hypothesize that an adjustment factor could be arrived at for each location of such survey, comparing the sensor based counts and ground truth and can be used to adjust the data reliably to reflect the ground truth in absolute numbers.
+This calibration can be carried over periodically and the frequecy of which will improve the quality of the estimation.

### 2018-06-24 paper/sections/methodology.tex
methodology figures progress

@@
 The MAC address in these probe requests is anonymised using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
-A overall schematic of the data collection process is shown in Figure.
+A overall schematic of the data collection process is shown in Figure \ref{datacollection_schematic}.
 The ground truth on number of pedestrian footfall was recorded using a purpose built Android application cite.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=\linewidth]{images/datacollection_schematic.jpeg}
+	\caption{Schematic diagram showing the process of collecting probe requests using the sensor} \label{datacollection_schematic}
+\end{figure}
 
@@
 We propose the following methods to tackle each of these challages.
+
+\begin{figure}
+	\centering
+	\subfigure[]{
+	\resizebox*{0.48\linewidth}{!}{\includegraphics{images/pilot_counts_comparision.png}}}
+	\subfigure[]{
+	\resizebox*{0.48\linewidth}{!}{\includegraphics{images/pilot_counts_comparision.png}}}
+	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{sample-figure}
+\end{figure}
 

### 2018-06-24 paper/sections/methodology.tex
methodology complete

@@
 	\centering
-	\subfigure[]{
-	\resizebox*{0.48\linewidth}{!}{\includegraphics{images/pilot_counts_comparision.png}}}
-	\subfigure[]{
-	\resizebox*{0.48\linewidth}{!}{\includegraphics{images/pilot_counts_comparision.png}}}
+	\subfigure[Distribution of signal strengths showing the filtering of background noise]{
+		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_signalstrength.jpeg}}}\hspace{20pt}
+	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
+	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_clustering.jpeg}}}
 	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{sample-figure}
@@
 
-We hypothesise that in configurations with a specific source of background noise at a constant distance, there must be a distinct break in the number of probe requests reporting signal strength corresponding to that distance.
+We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct break in the number of probe requests reporting signal strength corresponding to that distance.
 For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported singal strength corresponding to the distance between the sensor and the phone shop at any given set of conditions.
-We could identify these breaks in the data using tradition class interval algorithms such as jenks natural breaks, k-means, quantile and hierarchical clustering, etc.
+We could identify these breaks in the data using traditional one diamensional classification algorithms such as Jenks natural breaks, k-means, quantile and hierarchical clustering, etc.
 Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.

### 2018-06-24 paper/sections/methodology.tex
methodology reference to figures

@@
 	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_clustering.jpeg}}}
-	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{sample-figure}
+	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
 \end{figure}
@@
 We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct break in the number of probe requests reporting signal strength corresponding to that distance.
-For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported singal strength corresponding to the distance between the sensor and the phone shop at any given set of conditions.
+For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported singal strength corresponding to the distance between the sensor and the phone shop at any given set of conditions as shown in Figure \ref{methodology_schematic}.
 We could identify these breaks in the data using traditional one diamensional classification algorithms such as Jenks natural breaks, k-means, quantile and hierarchical clustering, etc.
 Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.
-An illustrative example is show in figure.
 
@@
 \end{itemize}
-The nodes were then classified based on the unique connected component they belong to.
+The nodes were then classified based on the unique connected component they belong to as shown in Figure \ref{methodology_schematic}.
 This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
@@
 \subsection{Calibrating with ground truth}
-Since mobile phone ownership is a external uncertainty to our study and could arise from variety of spatio, temporal and demographic factors, we propose to solve this by using external source of information.
-We hypothesize that an adjustment factor could be arrived at for each location of such survey, comparing the sensor based counts and ground truth and can be used to adjust the data reliably to reflect the ground truth in absolute numbers.
+Since mobile phone ownership is an external uncertainty to our study and could arise from variety of spatio, temporal and demographic factors, we propose to solve this using external source of information.
+We hypothesize that an adjustment factor could be arrived at for each location of data collection, comparing the sensor based counts and ground truth and it can be used to adjust the data reliably to reflect the ground truth in absolute numbers for the future.
 This calibration can be carried over periodically and the frequecy of which will improve the quality of the estimation.

### 2018-06-24 paper/sections/pilot_study.tex
pilot figures

--- 2018-06-20 paper/sections/pilot_study.tex
@@
-We conducted a pilot study to get a feel for the data.
-See which fields are relevant and which ones are not.
-to check if the sequencing algorithm holds true.
-to findout which one of the classification algorithm works.
-if these methodology works in real world scenario scalably.
-A pilot survey was conducted on Oxford Street in London in December 2017,
-where two sets of data were collected on pedestrian footfall
-with the aim of establishing merit in measuring pedestrian footfall
-as a function of the number of wifi probe requests
-collected at a given location.
-These datasets were collected through Wi-Fi sensing and 
-manual counting in parallel.
+To start we design a small pilot study as a proof of concept with the aim to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
+We also want to find the algorithm which is best suited for classification of signal strengths.
+The data was collected at Oxford Street in London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe requests sets were collected using the Wi-Fi sensor described earlier and pedestrian footfall was manually recorded using the Android app.
+Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests during the period, and 3,722 people were recorded manually.
 
-Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests over a 30 minutes interval, and 3,722 people were counted manually.
+When we just aggregated the probe requests by their MAC address for every minute, the mean error between the sensor counts and the manual counts was observed to be on average 425\%.
+This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where the manual count was conducted and anonymised probe requests with different MAC addresses from a few devices stationed next to the sensor.
+We then classified the probe requests as "high signal strength" and "low signal strength" using `k-means' classification algorithm which resulted in the lowest mean error percentage closely followed by `quantile'.
+The cut-off point or threshold for the collected data was -71 dBm.
+We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low signal strength" below the threshold.
+We found this process of filtering highly effective and reduced the mean minute by minute error to 30\%.
 
-When we aggregated the probe requests by their MAC address for every minute, the difference between the sensor counts and the manual counts was observed to be on average 425\%.
-This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where themanual count was conducted, as well as anonymised probe requests from the same devices but with different MAC addresses.
-This process of filtering was highly effective and reduced the difference between the sensor counts and manual counts to 30\%.
-We observed that around 55\% of all probe requests collected were anonymised.
-We assigned the hashed MAC address the unique identifier for the remaining 45\% and investigated the anonymised probe requests further.
-
-An initial analysis revealed that the fields - SSID and tags - were very sparse and did not provide much information for our cleaning process.
-In addition, the duration field was closely related to the length of the probe request and provides no new information.
-Therefore, we removed these fields from further analysis.
-We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low" signal strength.
-This classification of "high" vs "low" was performed using a k-means classification algorithm.
-The cut-off point for the collected data was -71 dBm.
+We then move on to assign an unique field identifying the mobile device generating the probe requests.
+For the 45\% of the probe requests which were not randoised, we kept the MAC addresses as the unique identifier.
+For the rest of the probe requesrt we applied the clustering alogorithm to assign the unique identifier.
+With the help of the known stationary device (the mobile device used by the surveyor to record pedestrians manually) and through trail and error, We found the time threshold, $\alpha$ to be 15 seconds and threshold for sequence numbers, $\beta$ to be 60.
+Figure \ref{pilot_clustering} shows the clustering of probe requests.
+We finally combine both normal and anonymised probe requests, aggregate them based on their unique identifier which further reduced the mean error to -18\%.
 
@@
 	\begin{center}
-		\includegraphics [width=\linewidth,trim=4 4 4 4,clip] {images/methodology_clustering.png}
-		\caption{Clustering probe requests based on increasing sequence numbers present in them.}
-		\label{clustering_pilot}
+		\includegraphics [width=\linewidth] {images/pilot_clustering.jpeg}
+		\caption{Clustering probe requests based on increasing sequence numbers present in them.
+The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
+		\label{pilot_clustering}
 	\end{center}
@@
 
-Figure \ref{clustering_pilot} shows the clustering process: 
-the black dots show the probe requests and the red lines connect them into clusters representing those which were generated by the same device.
-We finally combined both normal and anonymised probe requests, aggregated them based on their unique identifier, and removed repeating probe requests which reduced the difference between the sensor counts and the manual counts to -18\%.
+The minute by minute comparision of counts from the filtering processes along with the ground truth is shown in Figure \ref{pilot_comparison} From the pilot study, we find that both classification and clustering methods work on complex real world data and results in a final pedestrian counts within a error of 20\%.
+We also find `k-means' and `quantile' are best algorithms for classifying signal strengths.
 
@@
 	\begin{center}
-		\includegraphics [width=\linewidth] {images/pilot_counts_comparision.png}
+		\includegraphics [width=0.85\linewidth] {images/pilot_counts_comparision.png}
 		\caption{Comparision of counts after filtering with manual counts}
-		\label{comparision_pilot}
+		\label{pilot_comparison}
 	\end{center}

### 2018-06-24 analysis/data_plotting.r

@@ -7,6 +7,25 @@
 # ------------------------------------------------------------------------------
 # Plot objects
 # ------------------------------------------------------------------------------
+p_comp <- bind_rows(count_all,count_data_00) %>%
+	filter(time >= as.POSIXct("2017-12-20 12:30:00"),
+		   time <= as.POSIXct("2017-12-20 13:00:00"),
+		   type %in% c("01 Filtered out low signal strength",
+					   "03 Filter out randomised MACs",
+					   "05 Manual counting",
+					   "00 No filtering")) %>%
+	ggplot()+
+	geom_line(aes(time,footfall,col = type),
+			  linejoin="round",
+			  lineend="round",
+			  size = 1.5,
+			  show.legend=FALSE)+
+	theme(legend.position = "bottom") +
+	xlab("") +
+	ylab("") +
+	scale_colour_manual(values = c("#e41a1c","#4daf4a","#377eb8","#984ea3")) +
+	theme(panel.background=element_blank())
+
 
 p_comparison <- bind_rows(count_all,count_data_00) %>%
 	filter(time >= as.POSIXct("2017-12-20 12:30:00"),

### 2018-06-25 paper/sections/methodology.tex
Everything except main study is complete

@@
 a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
-	\item \textbf{MAC randomisation} - The mobile devices in the past few years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
+	\item \textbf{MAC randomisation} - The mobile devices in recent years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
 This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor.

### 2018-06-25 paper/sections/pilot_study.tex
Everything except main study is complete

@@
-To start we design a small pilot study as a proof of concept with the aim to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
+To start we designed a small pilot study with the aim to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
 We also want to find the algorithm which is best suited for classification of signal strengths.
@@
 Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests during the period, and 3,722 people were recorded manually.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=0.95\linewidth] {images/pilot_clustering.jpeg}
+		\caption{Clustering probe requests based on increasing sequence numbers present in them.
+The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
+		\label{pilot_clustering}
+	\end{center}
+\end{figure}
 
@@
 We then move on to assign an unique field identifying the mobile device generating the probe requests.
-For the 45\% of the probe requests which were not randoised, we kept the MAC addresses as the unique identifier.
-For the rest of the probe requesrt we applied the clustering alogorithm to assign the unique identifier.
-With the help of the known stationary device (the mobile device used by the surveyor to record pedestrians manually) and through trail and error, We found the time threshold, $\alpha$ to be 15 seconds and threshold for sequence numbers, $\beta$ to be 60.
+For the 45\% of the probe requests which were not randomised, we kept the MAC addresses as the unique identifier.
+For the rest of the probe requests we applied the clustering algorithm to assign the unique identifier.
+With the help of the known stationary device (the mobile device used by the surveyor to record pedestrians manually) and through trail and error, We found the suitable time threshold, $\alpha$ to be 15 seconds and threshold for sequence numbers, $\beta$ to be 60.
 Figure \ref{pilot_clustering} shows the clustering of probe requests.
@@
 	\begin{center}
-		\includegraphics [width=\linewidth] {images/pilot_clustering.jpeg}
-		\caption{Clustering probe requests based on increasing sequence numbers present in them.
-The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
-		\label{pilot_clustering}
+		\includegraphics [width=0.90\linewidth] {images/pilot_comparison.jpeg}
+		\caption{Comparision of counts after filtering with manual counts}
+		\label{pilot_comparison}
 	\end{center}
@@
 We also find `k-means' and `quantile' are best algorithms for classifying signal strengths.
-
-\begin{figure}
-	\begin{center}
-		\includegraphics [width=0.85\linewidth] {images/pilot_counts_comparision.png}
-		\caption{Comparision of counts after filtering with manual counts}
-		\label{pilot_comparison}
-	\end{center}
-\end{figure}

### 2018-06-25 paper/sections/previous_work.tex
Everything except main study is complete

@@
-There have been numerous attempts at using Wi-Fi to measure the volume and movement of people in the built environment for various applications \citep{zarim2006,sap2015,reki2007}.
-Though most research obtains feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the MAC address randomisation process.
-This process aims to protect the users’ privacy by anonymising the only globally identifiable portion of the probe requests, which results in a set of probe requests generated by the same device with different random MAC addresses \citep{green2008}.
-There have been various successful attempts by researchers to breaking this randomisation process in order to extract real MAC addresses, \citep{martin2017} but this usually results in serious risk of infringement of the privacy of the users of the mobile devices.
+Though WiFi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices by the known locations of APs and the signal strength reported by them from the mobile devices \citep{he2003range, moore2004robust, lamarca2005place}.
+This can overcome the usual shortcoming of GPS, which struggles for precision and accuracy in indoor and densely built environments \citep{zarim2006,kawaguchi2009wifi, xi2010locating}. 
+Utilising this, we can easily and quickly estimate trajectories of the mobile devices just using the WiFi communication the device has with multiple known APs \citep{Sorensenlocation} which can be used similar to the GPS trajectories to understand individual travel patterns (Kim, 2006;\citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
+Such data can also be used in transportation planning and management to estimate travel time \citep{musa2011wiflow} and real time traffic monitoring \citep{abbott2013empirical}.
+
+Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} and globally unique information present in the probe requests one can also successfully track people across these access points \citep{cunche2014linking}, their trajectories \citep{musa2012tracking}, and the interactions between them \citep{cheng2012inferring,barbera2013signals,cunche2014know} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
+Using the semantic information present in these probe requests it even is possible to understand the nature of population at a large scale \citep{di2016mind}.
+
+Though extensive research has been done on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect the users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
+There are various methods which have been devised to overcome this anonymisation process such as decomposition of OUIs where detailed device model information is estimated by analysing an already known dataset of OUIs \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for WiFi \citep{vo2016,bloessl2015scrambler}; and finally, the timing attack where the packet sequence information present in the probe request frame is used \citep{matte2016,cheng2016can}.
+A combination of these methodologies has been proven to produce de-anonymised globally unique device information \citep{vanhoef2016, martin2017}.
+These approaches usually result in serious risk of infringement of the privacy of the users of the mobile devices by revealing their identifiable personal information.
+
 There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
+Such technique has various applications in numerous fields such as uncovering the urban wireless landscape \citep{rose2010mapping}, revealing human activity at large scales \citep{qin2013discovering}, estimating pedestrian numbers in crowds \citep{schauer2014estimating,fukuzaki2015statistical} and even counting people in hyper local scales such as queues \citep{wang2013measuring}
+With enough infrastructure to collect such information we can aim to generate a real-time census of the city \citep{konto2017}.
+It has also been demonstrated by \citep{pinelli2015comparing} through series of experiments on a telecom operator dataset that using such network-driven approach is more advantageous compared to the widely used event-driven approaches.
+
+With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involve.

### 2018-06-25 analysis/data_plotting.r

@@ -15,10 +15,12 @@ p_comp <- bind_rows(count_all,count_data_00) %>%
 					   "05 Manual counting",
 					   "00 No filtering")) %>%
 	ggplot()+
-	geom_line(aes(time,footfall,col = type),
+	geom_line(aes(time,footfall, group = type),
 			  linejoin="round",
 			  lineend="round",
-			  size = 1.5,
+			  size = 0.25,
+			  show.legend=FALSE)+
+	geom_point(aes(time,footfall, shape = type),
 			  show.legend=FALSE)+
 	theme(legend.position = "bottom") +
 	xlab("") +

### 2018-06-26 paper/sections/main_study.tex
main study progress

@@
-The aims of the main study are,
-test the validity of the signal strength algorithm in different 
-micro site conditions.
-Test that the sequence number algorithm works in real world
-for different locations and different times.
-check if the thresholds are consistent.
-Test if the calibration works over intervals
-Finally conclude if we can estimate footfall confidently
-with just probe requests.
+The main study was designed to validate the results of the pilot study over different locations at different times. 
+We choose five different locations across central London, to install the sensors and collect data for a long period of time. 
+We also carry out manual counting on these locations along with this across different time of the day. 
+We then apply the filtering based on signal strength and sequence numbers and compare with manual counts and evaluate the effectiveness of the process with the mean error per minute on these locations.
+Finally we calculate the ajustment factor for the first interval of manual counts and check if that works on the consecutive intervals.
 
-five locations were selected across central london which
-had different types of configurations and specific problems
-configuratons are shown in fig. map is shown in figure.
-\begin{enumerate}
-	\item Phone Shop Camden - has phones and bus stops.
-	\item Restaurant TCR - has seating area on either side.
-	\item Holborn Information Kiosk - High volumne station entry
-	\item Restaurant Russell Square - seating on one side and side walk on other
-	\item Shop Charring Cross - sidewalk on one side and phone shop next door
-\end{enumerate}
-Installations were carried out over the time period from xxxx to xxxx.
-the data collection happened from xxxx to xxxx. Manual counting was carried out
-with high precision on dates xxxx and aggregated five minutes on xxxx.
-The difference in methods could lead to some inaccuracies in data.
-The overall schedule is shown in half page graphic.
+The locations where the data were collected are shown in the table.
+The location are chosen for their variety of configuration and sources of noise.
+Location 1 is the `cleanest' of while location 2 is the one with the most complexity.
+The configuration, installation and data collection schedule is shown in the figure.
 
-The overall statistics of data collected. How many probe requests.
-Comparision of location in terms of volume, patterns in daily footfall etc.
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=\linewidth] {images/main_schedule.jpeg}
+		\caption{Clustering probe requests based on increasing sequence numbers present in them.
+The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
+		\label{pilot_clustering}
+	\end{center}
+\end{figure}
+
+\begin{table}
+	\tbl{Locations where sensors were installed}
+	{\begin{tabular}{clll} 
+		\toprule
+		 ID & Location & Type & Installation notes\\
+		 \midrule
+		 1 & Camden High Street & Phone Shop & Bus stop in front\\
+		 2 & Central St.Giles Piazza & Restaurant & Seating area on both sides\\
+		 3 & Holborn Underground Station & Information Kiosk & Overlooks station entrance\\
+		 4 & Brunswick Center & Fast Food Restaurant & Has seating area on one side\\
+		 5 & The Strand & Regular Shop & Has phone shop next door \\
+		 \bottomrule
+	\end{tabular}}
+	% \tabnote{\textsuperscript{a}This footnote shows how to include footnotes to a table if required.}
+	\label{locations-table}
+\end{table}
+
+\begin{figure}
+	\centering
+	\subfigure[Distribution of signal strengths showing the filtering of background noise]{
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
+	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 25 6,clip]{images/main_signals.jpeg}}}
+	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
+\end{figure}
+
 The comparision between global and local.

### 2018-06-26 paper/sections/pilot_study.tex
main study progress

@@
 \begin{figure}
-	\begin{center}
-		\includegraphics [width=0.95\linewidth] {images/pilot_clustering.jpeg}
-		\caption{Clustering probe requests based on increasing sequence numbers present in them.
-The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
-		\label{pilot_clustering}
-	\end{center}
+	\centering
+	\subfigure[Clustering probe requests based on increasing sequence numbers present in them.]{
+		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_clustering.jpeg}}}\hspace{20pt}
+	\subfigure[Comparision of counts after filtering with manual counts]{
+	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_comparison.jpeg}}}
+	\caption{The results of the clustering process and the comparison with ground truth} \label{methodology_schematic}
 \end{figure}
@@
 Figure \ref{pilot_clustering} shows the clustering of probe requests.
+The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.
 We finally combine both normal and anonymised probe requests, aggregate them based on their unique identifier which further reduced the mean error to -18\%.
-
-\begin{figure}
-	\begin{center}
-		\includegraphics [width=0.90\linewidth] {images/pilot_comparison.jpeg}
-		\caption{Comparision of counts after filtering with manual counts}
-		\label{pilot_comparison}
-	\end{center}
-\end{figure}
 

### 2018-06-26 analysis/main_study.r

@@ -1,0 +1,53 @@
+library(tidyverse)
+library(ggplot2)
+library(RPostgreSQL)
+connection <- dbConnect(dbDriver('PostgreSQL'), dbname = 'bala_ijgis', user = 'bala_read', password = 'readpasswordbala', host = 'cdrc-footfall.geog.ucl.ac.uk')
+manual_schedule <- dbGetQuery(connection,"select new_sensor as sensor, min(date_trunc('minute',timestamp::timestamp)+interval '1 minute') as start, max(date_trunc('minute',timestamp::timestamp)) as end from manual left join locations on locations.id = manual.location group by date(timestamp),new_sensor order by 1;")
+if(!exists("sensor_data")){ 
+	sensor_data <- (function(){
+		data <- dbGetQuery(connection,"select * from sensor limit 0;")
+		for(i in 1:nrow(manual_schedule)) {
+			addition <- dbGetQuery(connection,paste0("select sensor.*, vendor.vendor,case when substring(sensor.oui,2,1) in ('e','a','2','6') then 'local' else 'global' end as type from sensor left join vendor on sensor.oui=vendor.oui where sensor = '",manual_schedule$sensor[i],"' and timestamp between '",manual_schedule$start[i],"' and '",manual_schedule$end[i],"';"))
+			data <- rbind(data,addition) 
+		}
+		return(data)
+	})()
+}
+if(!exists("manual_data")){ 
+	manual_data <- (function(){
+		data <- dbGetQuery(connection,"select * from manual limit 0;")
+		for(i in 1:nrow(manual_schedule)) {
+			addition <- dbGetQuery(connection,paste0("select timestamp, new_sensor as sensor from manual left join locations on locations.id = manual.location where new_sensor = '",manual_schedule$sensor[i],"' and timestamp between '",manual_schedule$start[i],"' and '",manual_schedule$end[i],"';"))
+			data <- rbind(data,addition) 
+		}
+		return(data)
+	})()
+}
+
+sensor_schedule <- dbGetQuery(connection,"select date(timestamp), sensor from sensor where date(timestamp) < '2018-04-01' group by 1,2 order by 1,2;")
+sensor_schedule$sensor <- factor(as.character(sensor_schedule$sensor),levels=c("88","66","55","44","33","22","11")) 
[…]

### 2018-06-26 analysis/old_files/main.r

@@ -1,4 +1,5 @@
 suppressMessages(library('tidyverse'))
+suppressMessages(library('ggplot2'))
 
 if(!exists("locations")) { locations <- read.csv("../data/main_locations.csv") }
 if(!exists("manual")) { manual <- read.csv("../data/main_manual.csv") }

### 2018-06-26 paper/sections/main_study.tex
analysis

@@
 
-\begin{figure}
-	\begin{center}
-		\includegraphics [width=\linewidth] {images/main_schedule.jpeg}
-		\caption{Clustering probe requests based on increasing sequence numbers present in them.
-The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.}
-		\label{pilot_clustering}
-	\end{center}
-\end{figure}
-
 \begin{table}
@@
 		 4 & Brunswick Center & Fast Food Restaurant & Has seating area on one side\\
-		 5 & The Strand & Regular Shop & Has phone shop next door \\
+		 5 & The Strand & Tea Shop & Has phone shop next door \\
 		 \bottomrule
 	\end{tabular}}
-	% \tabnote{\textsuperscript{a}This footnote shows how to include footnotes to a table if required.}
 	\label{locations-table}
@@
 \begin{figure}
+	\begin{center}
+		\includegraphics [width=0.90\linewidth] {images/main_schedule.jpeg}
+		\caption{Days when the sensors were active at the corresponding location. The red square shows that manual data collection was also done.}
+		\label{pilot_clustering}
+	\end{center}
+\end{figure}
+
+Though data was collected for many continous days, for the purposes of comparing with ground truth we just consider the only the data from sensors corresponding to those sessions.
+We have 12 sets of data over 6 different days.
+We have atleast two set manual counts for each location for verification of calibration.
+
+\subsection{signal strength filtering}
+First we see the distribution of the signal strength varies with location and configuration. 
+The density plot for signal strength is shown alogn with configuration in figure.
+We can see that the signal strength distribution shows distinct patterns of high and low when the installation is that there is a clear distant source of noise but this distinction get more and more obscure as we move towards difficult installations.
+For example, location 2 is almost a normally distributed noise as it is too far to pick up any pedestrians but location 5 with a clear view of footpath and a phone shop next door shows clear distinction between the two.
+Intuitively the classfication algorithm should give us better results in the latter. 
+It is important to note that we are dealing with relative singal strengths, this can vary with location and time of the analysis but we should be still be able to differentiate signal from noise. 
+We run the kmeans classification algorithm and filter out the probe requests which are randomised and have signal strengths less than the second break (or the threshold). 
+We then count the number of Unique MAC addresses present in every minute and remove MAC addreses which reappear within 30 minutes of previous appearance. 
+We then compare this with the minute by minute aggreagation for the manual counts and find the average error per minute for the sensor count. 
+The results are shown in the table.We see that the location XXXX has the most error while location XXX has the least error confirming our intuition. 
+We also see that the error follows the complexity of the installation. 
+It is also very promising that this method alone reduces our margin of error by a lot. 
+For some practical purposes which does not require absolute numbers, this should be sufficient. talk about examples of indexes and dashboards and short term trends.
+
+\begin{figure}
 	\centering
-	\subfigure[Distribution of signal strengths showing the filtering of background noise]{
+	\subfigure[Installation configuration of sensors at various locations]{
 		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
-	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
-		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 25 6,clip]{images/main_signals.jpeg}}}
-	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
+	\subfigure[Density distribution of signal strength (lower values show higher signal strength)]{
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/main_signals.jpeg}}}
+	\caption{Distribution of signal strengths across locations} \label{methodology_schematic}
 \end{figure}
@@
 Specifics on top 5 manufacturers.
-
-We do a daily analysis of distribution of signal strengths.
-The thresholds are shown in the table.
-The average is xxxx and standard deviation is xxx.
-we notice that the variation is lot. There is a definite
-change with the micro site locations.
-
-We see how the signal strength filtering affects the counts.
-compared to manual counts we look at the average mean errors (daily)
-per 5 minutes. The counts go as follows.
-Shown as a red line in the Figure.
-
-We can conclude that even though it has variations, this is
-a good method to reduce the overall level of error.
 

### 2018-06-26 analysis/main_study.r

@@ -26,28 +26,116 @@ if(!exists("manual_data")){
 
 sensor_schedule <- dbGetQuery(connection,"select date(timestamp), sensor from sensor where date(timestamp) < '2018-04-01' group by 1,2 order by 1,2;")
 sensor_schedule$sensor <- factor(as.character(sensor_schedule$sensor),levels=c("88","66","55","44","33","22","11")) 
-ggplot(sensor_schedule[ sensor_schedule$sensor %in% c(11,22,33,44,55), ])+geom_tile(aes(date,sensor),color="white",size=5)+theme_void()
+# ggplot(sensor_schedule[ sensor_schedule$sensor %in% c(11,22,33,44,55), ])+geom_tile(aes(date,sensor),color="white",size=5)+theme_void()
+#
+# ggplot(sensor_data)+geom_histogram(aes(-signal),fill="black")+facet_grid(sensor~.)+theme_void()
+#
+# sensor_data %>% 
+# 	filter(-signal>25,type=="local") %>%
+# 	ggplot() +
+# 		geom_density(aes(-signal),fill="black") +
+# 		facet_grid(sensor~.) +
+# 		theme_light() +
+# 		xlab("") +
+# 		ylab("") +
+# 		theme(panel.grid.major=element_blank(),
+# 			  panel.grid.minor=element_blank(),
+# 			  panel.border=element_blank(),
+# 			  strip.text=element_blank(),
+# 			  axis.title.x=element_blank(),
+# 			  axis.text.x=element_blank(),
+# 			  axis.ticks.x=element_blank(),
+# 			  axis.title.y=element_blank(),
+# 			  axis.text.y=element_blank(),
+# 			  axis.ticks.y=element_blank())
+#
+sensor_data_filtered <- (function(){
+	just_global <- sensor_data %>% filter(type=="global")
+	just_local <- sensor_data %>% filter(type=="local")
+	thres_11 <- (just_local %>% filter(signal<(-25),sensor==11) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
+	thres_22 <- (just_local %>% filter(sensor==22) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
+	thres_33 <- (just_local %>% filter(sensor==33) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
[…]

### 2018-06-26 paper/sections/main_study.tex
signal strength complete

@@
 \subsection{signal strength filtering}
-First we see the distribution of the signal strength varies with location and configuration. 
+First we see how the distribution of the signal strength varies with location and configuration. 
 The density plot for signal strength is shown alogn with configuration in figure.
@@
 
+\begin{table}
+	\tbl{Data collected at different locations and the correspondin mean absolute percentage error after each filtering process}
+	{\begin{tabular}{cccc} 
+		\toprule
+			Sensor& MAPE - No Cleaning (\%)&MAPE - Signal Strength (\%)&MAPE - Clustering (\%)\\
+		 \midrule
+			1 & 211 &  56 & \\
+			2 & 934 & 400 & \\
+			3 &  88 &  -3 & \\
+			4 & 501 & 237 & \\
+			5 & 476 & 117 & \\
+		 \bottomrule
+	\end{tabular}}
+	\label{errors-table}
+\end{table}
+
 we see that the signal strength filtering works and reduces error by

### 2018-06-26 analysis/main_study.r

@@ -52,7 +52,7 @@ sensor_schedule$sensor <- factor(as.character(sensor_schedule$sensor),levels=c("
 sensor_data_filtered <- (function(){
 	just_global <- sensor_data %>% filter(type=="global")
 	just_local <- sensor_data %>% filter(type=="local")
-	thres_11 <- (just_local %>% filter(signal<(-25),sensor==11) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
+	thres_11 <- (just_local %>% filter(sensor==11) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
 	thres_22 <- (just_local %>% filter(sensor==22) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
 	thres_33 <- (just_local %>% filter(sensor==33) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
 	thres_44 <- (just_local %>% filter(sensor==44) %>% pull(signal) %>% classIntervals(.,2,"kmeans"))$brks[2]
@@ -79,7 +79,20 @@ is_moving <- function(x) { return( !(((x - (5 * 60)) %in% x) |
 	((x - (12 * 60)) %in% x) |
 	((x - (13 * 60)) %in% x) |
 	((x - (14 * 60)) %in% x) |
-	((x - (15 * 60)) %in% x) ) ) }
+	((x - (16 * 60)) %in% x) |
+	((x - (17 * 60)) %in% x) |
+	((x - (18 * 60)) %in% x) |
+	((x - (19 * 60)) %in% x) |
+	((x - (20 * 60)) %in% x) |
+	((x - (21 * 60)) %in% x) |
+	((x - (22 * 60)) %in% x) |
+	((x - (23 * 60)) %in% x) |
+	((x - (24 * 60)) %in% x) |
+	((x - (26 * 60)) %in% x) |
+	((x - (27 * 60)) %in% x) |
+	((x - (28 * 60)) %in% x) |
+	((x - (29 * 60)) %in% x) |
+	((x - (30 * 60)) %in% x) ) ) }
 
 filter_moving <- function(x) {
 	data_vector <- unlist(x[[1]])
@@ -127,13 +140,16 @@ get_counts <- function(s) {
 	return(comparison)
 }
 
-get_location_parameters <- (function(){
-	counts11 <- get_counts(11)
-	counts22 <- get_counts(22)
-	counts33 <- get_counts(33)
[…]

### 2018-06-27 paper/sections/main_study.tex
starting clustering

@@
 
+\begin{figure}
+	\centering
+	\subfigure[Installation configuration of sensors at various locations]{
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
+	\subfigure[Density distribution of signal strength (lower values show higher signal strength)]{
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/main_signals.jpeg}}}
+	\caption{Distribution of signal strengths across locations} \label{methodology_schematic}
+\end{figure}
+
 \subsection{signal strength filtering}
@@
 We then compare this with the minute by minute aggreagation for the manual counts and find the average error per minute for the sensor count. 
-The results are shown in the table.We see that the location XXXX has the most error while location XXX has the least error confirming our intuition. 
+The results are shown in the table.We see that the location 2 has the most error 400\% confirming our intuition while surprisingly location 3 has the least error -3\%. 
 We also see that the error follows the complexity of the installation. 
-It is also very promising that this method alone reduces our margin of error by a lot. 
-For some practical purposes which does not require absolute numbers, this should be sufficient. talk about examples of indexes and dashboards and short term trends.
+It is also very promising that this method alone reduces our margin of error by a 50 - 100\%
+For some practical purposes which do not require absolute numbers, this should be sufficient. 
+e.g. Indexes of activity and change dashboards and short term trends identification.
 
-\begin{figure}
-	\centering
-	\subfigure[Installation configuration of sensors at various locations]{
-		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
-	\subfigure[Density distribution of signal strength (lower values show higher signal strength)]{
-		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/main_signals.jpeg}}}
-	\caption{Distribution of signal strengths across locations} \label{methodology_schematic}
-\end{figure}
+We see that the success of the signal strength filtering depends on the how we tackle the problems in the installing the devices.
+The more the device in installed in a way that the field of measurement is distinct from its surroundings (however noisy they might be) the more successful this process is.
+Closer sources error causes overcounting as we see in 2, 4 but at the same time highly close field of measurement with no big source of error can cause undercounting such as in 3.
+The complexity of the installation and uniformity of surroundings near the sensors directly coorrelate to the MAPE.
+The most effective way of measuring pedestrian is to have the sensor next to have the sensor next to clear area of pedestrian activity and in case of absence of any major source of noise, we should not do the signal strength filtering at all.
+We can even note the major possible sources of noise around the device and use it to fiugre out the right way of partitioning the data e.g.
+location 2 and 3 should have different pattern of signal strength filtering depending on their configuration.
 
-The comparision between global and local.
-comparision between different types of vendors.
-Specifics on top 5 manufacturers.
-
-This is also done for different locations hourly for all the data we had.
-we compare it to the manual counts and see that the average mean error
-has been reduced/increased. The finger print works well for all the locations.
-It also works over a period of time and gives us a comparable and close 
-footfall count to the manual count. The thresholds found in the pilot study
-works as well.
-
-Finally we normalise the sensor counts to match the manual count using
-a fraction/ adjustment factor calculated from the know manual counts.
-we have three sets of counts. We check if the adjustment factor holds the
-same in all three counts across locations. It does with a variations from
-xxx to xxx. The results are shown in the table.
+\subsection{Clustering based on sequence numbers}
 
@@
 		\toprule
-			Sensor& MAPE - No Cleaning (\%)&MAPE - Signal Strength (\%)&MAPE - Clustering (\%)\\
+			Sensor	& MAPE - No Cleaning & MAPE - Signal Strength	& MAPE - Clustering	\\
+					& 					 & (efficiency)				& (efficiency)		\\
 		 \midrule
-			1 & 211 &  56 & \\
-			2 & 934 & 400 & \\
-			3 &  88 &  -3 & \\
-			4 & 501 & 237 & \\
-			5 & 476 & 117 & \\
+			1 & 211 &  56 ( 73)&	\\
+			2 & 934 & 400 ( 57)& 	\\
+			3 &  88 &  -3 (103)& 	\\
+			4 & 501 & 237 ( 53)& 	\\
+			5 & 476 & 117 ( 75)& 	\\
 		 \bottomrule
@@
 \end{table}
-
-we see that the signal strength filtering works and reduces error by
-xxxxx. there is variation by locations.
-we see that sequence number algorithm works as well. The threshold stays
-constant well and works well across locations. The calibration also works
-and ajustment factor stays consistent short term. This needs more work 
-long term.

### 2018-06-27 analysis/main_study.r

@@ -147,10 +147,12 @@ location_parameters <- (function(){
 		counts[is.na(counts)] <- 1 
 		mape1 <- mean((counts$raw - counts$manual)/counts$manual)*100
 		mape2 <- mean((counts$filtered - counts$manual)/counts$manual)*100
-		data <- rbind(data, data.frame(sensor = i,mape_raw = mape1, mape_sig = mape2))
+		change1 <- (mape1-mape2)/mape1 * 100
+		data <- rbind(data, data.frame(sensor = i,mape_raw = mape1, mape_sig = mape2,change_sig = change1))
 	}
 	return(data)
 })()
+
 
 ggplot(comparison)+geom_line(aes(timestamp,raw,group="raw"),col="red")+geom_line(aes(timestamp,filtered,group="filtered"),col="green")+geom_line(aes(timestamp,manual,group="manual"))
 

### 2018-06-29 paper/sections/main_study.tex
friday morning

@@
 
-\subsection{signal strength filtering}
 First we see how the distribution of the signal strength varies with location and configuration. 
@@
 
-\subsection{Clustering based on sequence numbers}
+In terms of filtering by sequence numbers we see that the the filtering works in all the locations.
+It has the similar effect as the signal strength and brings us furthe closer to the manual count.
+Finally we calculate the adjustment factor from the first 15 minutes and apply that for the rest of the interval.
+We get the counts down to almost 10\% on good locations.
+The locations which still over count are the ones with seating next to them (2 and 4).
+We also see that the trends are preserved with the filtering process.
+
+Finally to summarize, through a simple raspberry pi based wifi sensors, we collected probe requests sent by user's mobile devices, cleaned them based on the reported signal strength, clustered/ fingerprinted them based on sequence numbers, aggregated them to get pedestrian counts.
+We then compared it to the read pedestrian count and developed adjustment factors which work for short term future.
 
@@
 	\tbl{Data collected at different locations and the correspondin mean absolute percentage error after each filtering process}
-	{\begin{tabular}{cccc} 
+	{\begin{tabular}{ccccccc} 
 		\toprule
-			Sensor	& MAPE - No Cleaning & MAPE - Signal Strength	& MAPE - Clustering	\\
-					& 					 & (efficiency)				& (efficiency)		\\
+			Sensor & Signal strength & Adjustment & MAPE & MAPE after & MAPE after & MAPE of\\
+			& threshold & factor & without any & filtering signal & filtering sequence & final adjusted\\
+			& (-dBm) & &  cleaning (\%) & strength (\%) & numbers (\%) & counts (\%)\\
 		 \midrule
-			1 & 211 &  56 ( 73)&	\\
-			2 & 934 & 400 ( 57)& 	\\
-			3 &  88 &  -3 (103)& 	\\
-			4 & 501 & 237 ( 53)& 	\\
-			5 & 476 & 117 ( 75)& 	\\
+			1 & -70 & 1.25 & 259 &  22 & -13 &  9 \\
+			2 & -74 & 0.51 & 928 & 396 & 206 & 55 \\
+			3 & -72 & 1.60 &  87 & -19 & -31 & 10 \\
+			4 & -70 & 0.88 & 498 & 142 &  52 & 33 \\
+			5 & -72 & 0.80 & 473 &  84 &  38 & 11 \\
 		 \bottomrule
@@
 \end{table}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=\linewidth,trim=6 6 6 6,clip] {images/main_comparison.jpeg}
+		\caption{Comparison of all the counts in teh sensors}
+		\label{main_comparison}
+	\end{center}
+\end{figure}

### 2018-06-29 analysis/main_study.r

@@ -1,6 +1,10 @@
 library(tidyverse)
 library(ggplot2)
 library(RPostgreSQL)
+library(igraph)
+library(classInt)
+library(lubridate)
+
 connection <- dbConnect(dbDriver('PostgreSQL'), dbname = 'bala_ijgis', user = 'bala_read', password = 'readpasswordbala', host = 'cdrc-footfall.geog.ucl.ac.uk')
 manual_schedule <- dbGetQuery(connection,"select new_sensor as sensor, min(date_trunc('minute',timestamp::timestamp)+interval '1 minute') as start, max(date_trunc('minute',timestamp::timestamp)) as end from manual left join locations on locations.id = manual.location group by date(timestamp),new_sensor order by 1;")
 if(!exists("sensor_data")){ 
@@ -24,88 +28,183 @@ if(!exists("manual_data")){
 	})()
 }
 
-sensor_schedule <- dbGetQuery(connection,"select date(timestamp), sensor from sensor where date(timestamp) < '2018-04-01' group by 1,2 order by 1,2;")
-sensor_schedule$sensor <- factor(as.character(sensor_schedule$sensor),levels=c("88","66","55","44","33","22","11")) 
-# ggplot(sensor_schedule[ sensor_schedule$sensor %in% c(11,22,33,44,55), ])+geom_tile(aes(date,sensor),color="white",size=5)+theme_void()
-#
-# ggplot(sensor_data)+geom_histogram(aes(-signal),fill="black")+facet_grid(sensor~.)+theme_void()
-#
-# sensor_data %>% 
-# 	filter(-signal>25,type=="local") %>%
-# 	ggplot() +
-# 		geom_density(aes(-signal),fill="black") +
-# 		facet_grid(sensor~.) +
-# 		theme_light() +
-# 		xlab("") +
-# 		ylab("") +
-# 		theme(panel.grid.major=element_blank(),
-# 			  panel.grid.minor=element_blank(),
-# 			  panel.border=element_blank(),
-# 			  strip.text=element_blank(),
-# 			  axis.title.x=element_blank(),
-# 			  axis.text.x=element_blank(),
-# 			  axis.ticks.x=element_blank(),
-# 			  axis.title.y=element_blank(),
-# 			  axis.text.y=element_blank(),
[…]

### 2018-06-29 analysis/data_plotting.r
updated images

@@ -15,19 +15,14 @@ p_comp <- bind_rows(count_all,count_data_00) %>%
 					   "05 Manual counting",
 					   "00 No filtering")) %>%
 	ggplot()+
-	geom_line(aes(time,footfall, group = type),
-			  linejoin="round",
-			  lineend="round",
-			  size = 0.25,
-			  show.legend=FALSE)+
-	geom_point(aes(time,footfall, shape = type),
-			  show.legend=FALSE)+
+	geom_line(aes(time,footfall, group = type, col=type),size = 0.5, show.legend=FALSE)+
+	geom_point(aes(time,footfall, shape = type,col=type),size = 2.5)+
 	theme(legend.position = "bottom") +
 	xlab("") +
 	ylab("") +
-	scale_colour_manual(values = c("#e41a1c","#4daf4a","#377eb8","#984ea3")) +
+	scale_colour_manual(values=c("#333333","#377eb8","#4daf4a","#e41a1c")) +
 	theme(panel.background=element_blank())
-
+p_comp
 
 p_comparison <- bind_rows(count_all,count_data_00) %>%
 	filter(time >= as.POSIXct("2017-12-20 12:30:00"),
@@ -162,3 +157,28 @@ p_local_cluster <- subset_local %>%
 	ylab("") +
 	theme(panel.background=element_blank())
 
+clustering_schematic <- subset_local %>%
+	filter(time >= as.POSIXct("2017-12-20 12:30:00"),
+		   time <= as.POSIXct("2017-12-20 12:40:00"),
+		   sequence < 2000) %>%
+	ggplot() + 
+	geom_point(aes(time,sequence,fill=mac),
+			   size = 4,
+			   shape = 21,
+			   color = "grey",
+			   stroke = 0.5,
+			   show.legend = FALSE) +
+	geom_line(aes(time,
+				  sequence,
+				  group = device_sign),
+			  # arrow = arrow(length=unit(0.1,"cm"), ends="last", type = "closed"),
+			  size = 1,
+			  # color = "#e41a1c",
+			  lineend = "round",
+			  linejoin = "round",
+			  color = "#333333",
+			  show.legend = FALSE) +
+	xlab("") +
+	ylab("") +
+	theme(panel.background=element_blank())
+clustering_schematic

### 2018-06-29 paper/sections/abstract.tex
Paul's comments included

@@
-The accurate measurement human activity with high spatial and temporal granularity is crucial for us to understand the structure and function of built enviroment.
-With increasing mobile ownerships, the Wi-Fi probe requests generated by these devices can be an excellent source of data for such measursement.
-The major challanges in using Wi-Fi technology for such purpose are precisely delineating the field of measurement and arriving at accurate estimates without compromising the privacy of the users of the mobile devices.
+The accurate measurement human activity with high spatial and temporal granularity is crucial for understanding the structure and function of built environment.
+With increasing mobile ownership, the Wi-Fi `probe requests' generated by mobile devices can act as an cheap, scalable and real-time source of data for such establishing such measures.
+One of the major challanges in using Wi-Fi technology for estimating human activity is eliminating biases caused by an uncertain field of measurement and changing technological landscape without compromising the privacy of the users of the mobile devices.
 In this paper we demonstrate that, with the application of class intervals and a novel graph based technique, we can overcome the challanges and reliably measure real-time pedestrian footfall at retail highstreets.

### 2018-06-29 paper/sections/conclusion.tex
journal template and more writing
Everything except main study is complete
friday morning
Paul's comments included

@@
-It is important to note that the filtering process was done based soley on the information present in the probe requests and their temporal distribution.
-This ensured that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
+This paper has made a novel contribution to the measurement of activities in retail centres in real time.
+To this end, it has described the collection and processing of a novel consumer Big Dataset that enables valid measures of levels of footfall activity which has been scaled across a Great Britain wide network of sensors.
+In both conceptual and technical terms, it illustrates the ways in which passively collected consumer data can be ‘hardened’ to render them robust and reliable by using related procedures of internal and external validation.
+Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.
+External validation then entails reconciling adjusted counts with the footfall observed at sample locations.
+The latter procedure makes it possible to generalise from locations at which manual footfall surveys are conducted to all others in the system, and to develop a classification of device locations that are more or less susceptible to noise generation.
+
+It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
+This ensures that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
 This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves.
@@
 Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
+A snapshot showing week's worth of precise footfall in area around charrig cross, London is shown in the \ref{} to demonstrate the potential for such dataset.
 Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
 Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
-It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them.
+It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them which we intend to research in future.

### 2018-06-29 paper/sections/introduction.tex
Paul's comments included

--- 2018-06-12 paper/sections/introduction.tex
@@
-In the past decade Wi-Fi has emerged as the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments.
-Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests', in order to discover Wi-Fi networks available to them.
-This helps these devices to connect and switch between the WiFi networks seamlessly.
+New and developing technologies today provide the infrastructure over which movements and interactions can be measured and monitored in the `sentient city' \citep{amin2017seeing}.
+These include mobile phone networks, which can triangulate user locations relative to networks of masts, use of GPS to locate users of social media services, and WiFi beacon connectivity to access the Internet.
+These technologies offer differing levels of spatial precision, where mobile telephony and WiFi generally being less reliable and offer lower precision than GPS to the end users, while being more advantageous for broader mobility studies \citep{pinelli2015comparing}.
 
-Probe requests are low level signals standardised by IEEE 802.1b/g specification as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
-This ubiquity and standardisation make them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.
-Considering the unprecendented levels of mobile device ownership in recent years, we can in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
+There has been considerable research into the usefulness of these technologies to understand patterns of movement in cities in near real time \citep{candia2008uncovering, gonzalez2008understanding, calabrese2013understanding}.
+Most of this research has focused upon technical specification of accuracy or precision \citep{song2010limits, lane2010survey}, with somewhat less attention devoted to the ways that the characteristics of the technologies and of their human users conspire to create possible bias in the inclusiveness of usage patterns.
+This is perhaps most manifest in the plethora of papers that analyse the distribution and content of publicly available Twitter social media feeds, which likely constitute an extremely biased and self-selecting subset of users that geo-enable their posts, drawn from a user base that itself bears no identified correspondence to any known background population \citep{lansley2016deriving}.
+Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases that may be representative of local populations, may exclude groups such as tourists from distand origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}
+These examples illustrate the issues that underpin the assembly and analysis of consumer data, which can be considered as a distinctive class of Big Data that arise from the interactions between humans and customer-facing organisations such as retailers, domestic energy suppliers, transport providers and suppliers of social media and communications \citep{cdrc2018consumer}.
+Consumer data account for an ever-increasing real share of all of the data that are collected of citizens, but a fundamental characteristic of consumer-led markets is that no single provider has a monopoly in market provision, and therefore issues of market share and segment generate bias in analysis.
+The source and operation of this bias is unknown in the absence of extensive and context sensitive attempts to triangulate consumer data with data of known provenance relating to clearly defined populations \citep{lansley2016deriving}.
+In similar ways to other classes of Big Data, consumer data are best throught of as digital ‘exhaust’, or a by-product created by or harvested from consumer transactions.
 
-While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users.
-In this paper, using a set of probe requests and manual counts collected at various high street locations across London, we demonstrate that pedestrian footfall at these locations can be estimated with considerable precision and accuracy while protecting the privacy of the pedestrians.
+In this paper, we evaluate the value of data collected from a Great Britain wide network of 800 devices, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
+These devices are located in shop windows and record the probes emitted by mobile phones and other WiFi enabled devices.
+These are deemed to be consumer data because devices carried by consumers routinely probe for a consumer service, specifically a WiFi connection.
+Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not Internet connectivity is established.
+Our core motivation is to appraise the usefulnessof WiFi probe records harvested from our network of sensors in order to indicate levels of pedestrian activity.
+More broadly still, in our future research we intend to classify the nationwide network of footfall profiles as part of a programme of research to understand the form and functioning of retail areas at a time of far-reaching structural change for the retail industry.
+
+To this end, it is important to first undertake a thorough conceptual and technical appraisal of our consumer data source.
+In technical terms, screening of information present in the `probes' and classifing them based on their characteristics is essential in order to remove the ones emitted by devices that do not indicate pedestrian activity, such as network enabled printers and other fixed devices.
+Related to this, a method to figerprint Wi-Fi probes is necessary to remove probes from individuals’ devices that in conceptual terms should not be considered part of footfall – as when, for example, an employees is seated in an office within range of the sensor device.
+A calibration of sensor measurement is also essential on two grounds: first, individuals may carry multiple devices, or no device at all; and second, the positioning and orientation of the sensor in the retail unit may lead to systematic over- or under-enumeration.
+These sources of bias in measurement must be accommodated by manual recording of footfall at each location and generalization of these sample survey results to all locations and time periods.
+As we describe in detail below, manual validation of the data needs to be undertaken in parallel with technical profiling of the mix of consumer devices that probe our sensors, since the effectiveness of data cleaning procedures discussed in this paper differ between individual locations and configurations.

### 2018-06-29 paper/sections/previous_work.tex
Paul's comments included

@@
+In the past decade Wi-Fi has emerged as the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
+This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments.
+Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests', in order to discover Wi-Fi networks available to them.
+This helps these devices to connect and switch between the WiFi networks seamlessly.
+
+Probe requests are low level signals standardised by IEEE 802.1b/g specification \citep{ieee2013} as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
+This ubiquity and standardisation make them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.
+Considering the unprecendented levels of mobile device ownership in recent years, we can in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
+
+While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users.
+In this paper, using a set of probe requests and manual counts collected at various high street locations across London, we demonstrate that pedestrian footfall at these locations can be estimated with considerable precision and accuracy while protecting the privacy of the pedestrians.
+
 Though WiFi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices by the known locations of APs and the signal strength reported by them from the mobile devices \citep{he2003range, moore2004robust, lamarca2005place}.
@@
 
-Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} and globally unique information present in the probe requests one can also successfully track people across these access points \citep{cunche2014linking}, their trajectories \citep{musa2012tracking}, and the interactions between them \citep{cheng2012inferring,barbera2013signals,cunche2014know} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
+Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} and using the information present in the probe requests one can also successfully track people across the sensors or access points collecting them \citep{cunche2014linking}, infer their trajectories \citep{musa2012tracking}, and even model interactions between them \citep{cheng2012inferring,barbera2013signals,cunche2014know} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
 Using the semantic information present in these probe requests it even is possible to understand the nature of population at a large scale \citep{di2016mind}.
 
-Though extensive research has been done on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect the users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
+Though extensive research has been carried out on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect the users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
 There are various methods which have been devised to overcome this anonymisation process such as decomposition of OUIs where detailed device model information is estimated by analysing an already known dataset of OUIs \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for WiFi \citep{vo2016,bloessl2015scrambler}; and finally, the timing attack where the packet sequence information present in the probe request frame is used \citep{matte2016,cheng2016can}.
@@
 It has also been demonstrated by \citep{pinelli2015comparing} through series of experiments on a telecom operator dataset that using such network-driven approach is more advantageous compared to the widely used event-driven approaches.
-
 With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involve.

### 2018-06-29 paper/sections/main_study.tex
end of day friday

@@
-The main study was designed to validate the results of the pilot study over different locations at different times. 
-We choose five different locations across central London, to install the sensors and collect data for a long period of time. 
-We also carry out manual counting on these locations along with this across different time of the day. 
-We then apply the filtering based on signal strength and sequence numbers and compare with manual counts and evaluate the effectiveness of the process with the mean error per minute on these locations.
-Finally we calculate the ajustment factor for the first interval of manual counts and check if that works on the consecutive intervals.
-
-The locations where the data were collected are shown in the table.
-The location are chosen for their variety of configuration and sources of noise.
-Location 1 is the `cleanest' of while location 2 is the one with the most complexity.
-The configuration, installation and data collection schedule is shown in the figure.
-
 \begin{table}
@@
 \end{figure}
+
+The main study was designed to validate the results of the pilot study over different locations at different times. 
+We choose five different locations across central London, to install the sensors and collect data for a long period of time. 
+We also carry out manual counting on these locations along with this across different time of the day. 
+We then apply the filtering based on signal strength and sequence numbers and compare with manual counts and evaluate the effectiveness of the process with the mean error per minute on these locations.
+Finally we calculate the ajustment factor for the first interval of manual counts and check if that works on the consecutive intervals.
+
+The locations where the data were collected are shown in the table.
+The location are chosen for their variety of configuration and sources of noise.
+Location 1 is the `cleanest' of while location 2 is the one with the most complexity.
+The configuration, installation and data collection schedule is shown in the figure.
 
@@
 
-Finally to summarize, through a simple raspberry pi based wifi sensors, we collected probe requests sent by user's mobile devices, cleaned them based on the reported signal strength, clustered/ fingerprinted them based on sequence numbers, aggregated them to get pedestrian counts.
-We then compared it to the read pedestrian count and developed adjustment factors which work for short term future.
-
 \begin{table}

### 2018-06-29 paper/sections/pilot_study.tex
end of day friday

@@
 To start we designed a small pilot study with the aim to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
-We also want to find the algorithm which is best suited for classification of signal strengths.
-The data was collected at Oxford Street in London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe requests sets were collected using the Wi-Fi sensor described earlier and pedestrian footfall was manually recorded using the Android app.
-Being located at one of the busiest retail locations in the United Kingdom, the WiFi sensor captured approximately 60,000 probe requests during the period, and 3,722 people were recorded manually.
+We also need to find the algorithm which is best suited for classification of signal strengths to filter out the background noise.
+The data was collected at Oxford Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe requests were collected using the sensor described in Chapter \ref{methodology} and pedestrian footfall was manually recorded using the Android app \citep{bala2018clicker}.
+Being located at one of the busiest retail locations in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe requests during the half hour period, and 3,722 people were recorded manually walking on the sidewalk during that time.
 
@@
 		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_clustering.jpeg}}}\hspace{20pt}
-	\subfigure[Comparision of counts after filtering with manual counts]{
+	\subfigure[Comparision of sensor based counts with the groud truth.]{
 	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_comparison.jpeg}}}
-	\caption{The results of the clustering process and the comparison with ground truth} \label{methodology_schematic}
+	\caption{The results of the pilot study demonstrating the validity of the deviced methodology.} \label{pilot_results}
 \end{figure}
 
-When we just aggregated the probe requests by their MAC address for every minute, the mean error between the sensor counts and the manual counts was observed to be on average 425\%.
-This suggested that there was a large amount of noise in the data which might have included signals from devices outside the area where the manual count was conducted and anonymised probe requests with different MAC addresses from a few devices stationed next to the sensor.
-We then classified the probe requests as "high signal strength" and "low signal strength" using `k-means' classification algorithm which resulted in the lowest mean error percentage closely followed by `quantile'.
-The cut-off point or threshold for the collected data was -71 dBm.
-We eliminated the noise from devices outside the area of interest by removing all the probe requests which reported a "low signal strength" below the threshold.
-We found this process of filtering highly effective and reduced the mean minute by minute error to 30\%.
+As a first step we just aggregated the probe requests by their MAC address for every minute to generate a minute by minute count of the number of people around the sensor assuming each MAC address corresponds to a mobile device and hence a pedestrian.
+We then compare this preliminary `footfall' count to the actual number of pedestrian recorded manually to check for the robustness.
+We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there is no `zero records' issue.
+We find that the MAPE in the raw counts compared to the ground truth is around 425\%.
+This suggests the presence of large amount of noise in the data which might be generated due to sources of uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need for filtering the data.
 
-We then move on to assign an unique field identifying the mobile device generating the probe requests.
-For the 45\% of the probe requests which were not randomised, we kept the MAC addresses as the unique identifier.
-For the rest of the probe requests we applied the clustering algorithm to assign the unique identifier.
-With the help of the known stationary device (the mobile device used by the surveyor to record pedestrians manually) and through trail and error, We found the suitable time threshold, $\alpha$ to be 15 seconds and threshold for sequence numbers, $\beta$ to be 60.
-Figure \ref{pilot_clustering} shows the clustering of probe requests.
-The dots are individual probe requests and the red lines connect probe requests within the same cluster which are generated by the same mobile device.
-We finally combine both normal and anonymised probe requests, aggregate them based on their unique identifier which further reduced the mean error to -18\%.
+We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, heirarchical clusters, fisher and jeanks.
+We found that while heirarchical clustering and jenks were resource intensive, kmeans gave quickest results with the lowest MAPE closely followed by quantile.
+The cut-off point or threshold for the collected data with which we could classify them as high and low was -71 dBm.
+We then removed all the probe requests which reported `low signal strength' and repeated the same aggregation process as before to produce footfall count. 
+This process resulted in a footfall count with a net MAPE of 30\%.
+Though the results are encouraging we are still not completely confident that our filtering process is indeed removing noise or has any correlation the configuration of sensor or position of the mobile devices.
+These need to be addressed with a larger survey with multiple location of varying orientations.
 
-The minute by minute comparision of counts from the filtering processes along with the ground truth is shown in Figure \ref{pilot_comparison} From the pilot study, we find that both classification and clustering methods work on complex real world data and results in a final pedestrian counts within a error of 20\%.
-We also find `k-means' and `quantile' are best algorithms for classifying signal strengths.
+The next challenge was to identify probe requests which are generated by the same device irrespective of MAC randomisation process.
+We use the algorithm defined in Chapter \ref{methodology} and assign a unique identifier or signature to each probe request independent of the MAC address. 
+Since there is neither prior research nor documentation on a universal behaviour of phones in randomising their MAC addresses, we use the stationary device - the one used for manual counting as a reference and find out the suitable time threshold, $\alpha$ and threshold for sequence numbers, $\beta$ to be 16s and 60 respectively via trial and error.
+This process is done on top the filtering done based on signal strength and only for the probe requests with randomised MAC addresses.
+Figure \ref{pilot_results} shows the results of this clustering process on a small set of randomised probe requests.
+The probe request with different randomised MAC address is shown by the colored points and the line joining them shows the ones belonging to the same cluster hence expected to be generated by the same device.
+We finally aggregate the probe requests as before but with the device signature rather than just MAC addresses this results in a footfall count with a MAPE of -18\%. 
+A comparision of minute by minute counts resulting from different filtering processes along with the ground truth is shown in Figure \ref{pilot_results} showing the promising effectiveness of the methods.
+
+To conclude, from the pilot study we found that both classification and clustering methods we devices work on complex real world data and results in a final pedestrian counts within a MAPE of 20\%.
+We also found `k-means' and `quantile' are best algorithms for classifying signal strengths and the threshold for time and sequence numbers for the clustering algorithm is around 16 and 60 respectively.

### 2018-06-29 paper/sections/previous_work.tex
end of day friday

@@
 This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments.
-Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests', in order to discover Wi-Fi networks available to them.
+Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests' in order to discover Wi-Fi networks available to them.
 This helps these devices to connect and switch between the WiFi networks seamlessly.

### 2018-06-30 paper/paper.tex
Almost completed methodology
Everything except main study is complete
Paul's comments included
end of day friday
start proofreading

@@
 \newtheorem{notation}{Notation}
+\usepackage{color,soul}
 
@@
 
-	\title{Estimating real-time highstreet footfall from Wi-Fi probe requests}
+	\title{Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests}
 	\author{
@@
 	\maketitle
-
 	\begin{abstract}
@@
 	\begin{keywords}
-		Highstreet footfall; Wi-Fi Probe requests; Sensors; MAC Randomisation \end{keywords}
+		Pedestrian footfall; Urban sensing; Wi-Fi probe requests; MAC Randomisation \end{keywords}
 	\section{Introduction}\label{introduction}
 		\input{sections/introduction}
-	\section{Previous Work}\label{previous_work}
+	\section{Background}\label{previous_work}
 		\input{sections/previous_work}

### 2018-06-30 paper/sections/abstract.tex
start proofreading

--- 2018-06-24 paper/sections/abstract.tex
@@
-The accurate measurement human activity with high spatial and temporal granularity is crucial for us to understand the structure and function of built enviroment.
-With increasing mobile ownerships, the Wi-Fi probe requests generated by these devices can be an excellent source of data for such measursement.
-The major challanges in using Wi-Fi technology for such purpose are precisely delineating the field of measurement and arriving at accurate estimates without compromising the privacy of the users of the mobile devices.
-In this paper we demonstrate that, with the application of class intervals and a novel graph based technique, we can overcome the challanges and reliably measure real-time footfall at retail highstreets in London.
+The accurate measurement of human activity with high spatial and temporal granularity is crucial for understanding the structure and function of the built environment.
+With increasing mobile ownership, the Wi-Fi `probe requests' generated by mobile devices can act as a cheap, scalable and real-time source of data for establishing such measures.
+The two major challanges we face in using these probe requests for estimating human activity are, filtering the noise generated by the uncertain field of measurement and clustering anonymised probe requests generated by the same devices together without compromising the privacy of the users.
+In this paper we demonstrate that we can overcome these challanges by using class intervals and a novel graph based technique for filtering and clustering the probe requests which in turn, enables us to reliably measure real-time pedestrian footfall at retail highstreets.

### 2018-06-30 paper/sections/conclusion.tex
end of day friday
start proofreading

@@
 In both conceptual and technical terms, it illustrates the ways in which passively collected consumer data can be ‘hardened’ to render them robust and reliable by using related procedures of internal and external validation.
-Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.
-External validation then entails reconciling adjusted counts with the footfall observed at sample locations.
-The latter procedure makes it possible to generalise from locations at which manual footfall surveys are conducted to all others in the system, and to develop a classification of device locations that are more or less susceptible to noise generation.
 
-It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
+Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
 This ensures that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
 This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves. 
+External validation then entails reconciling adjusted counts with the footfall observed at sample locations.
+This procedure makes it possible to generalise from locations at which manual footfall surveys are conducted to all others in the system, and to develop a classification of device locations that are more or less susceptible to noise generation.
 
@@
 Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-A snapshot showing week's worth of precise footfall in area around charrig cross, London is shown in the \ref{} to demonstrate the potential for such dataset.
+A snapshot showing week's worth of precise footfall in area around Charring cross, London is shown in Figure \ref{} to demonstrate the potential for such dataset.
 Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.

### 2018-06-30 paper/sections/introduction.tex
start proofreading

@@
 Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases that may be representative of local populations, may exclude groups such as tourists from distand origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}
+
 These examples illustrate the issues that underpin the assembly and analysis of consumer data, which can be considered as a distinctive class of Big Data that arise from the interactions between humans and customer-facing organisations such as retailers, domestic energy suppliers, transport providers and suppliers of social media and communications \citep{cdrc2018consumer}.
@@
 
-In this paper, we evaluate the value of data collected from a Great Britain wide network of 800 devices, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
+In this paper, through a set of experiments, we evaluate the value of data collected from a Great Britain wide network of 800 devices, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
 These devices are located in shop windows and record the probes emitted by mobile phones and other WiFi enabled devices.
@@
 These sources of bias in measurement must be accommodated by manual recording of footfall at each location and generalization of these sample survey results to all locations and time periods.
-As we describe in detail below, manual validation of the data needs to be undertaken in parallel with technical profiling of the mix of consumer devices that probe our sensors, since the effectiveness of data cleaning procedures discussed in this paper differ between individual locations and configurations.
+As we describe in detail below, manual validation of the data needs to be undertaken in parallel with technical profiling of the mix of consumer mobile devices that probe our sensors, since the effectiveness of data cleaning procedures discussed in this paper differ between individual locations and configurations.

### 2018-06-30 paper/sections/pilot_study.tex
start proofreading

@@
-To start we designed a small pilot study with the aim to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
-We also need to find the algorithm which is best suited for classification of signal strengths to filter out the background noise.
+To start we designed a small pilot study to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
+We also aim to find the algorithm which is best suited for classification of signal strengths to filter out the background noise.
 The data was collected at Oxford Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe requests were collected using the sensor described in Chapter \ref{methodology} and pedestrian footfall was manually recorded using the Android app \citep{bala2018clicker}.

### 2018-06-30 paper/sections/previous_work.tex
start proofreading

@@
-In the past decade Wi-Fi has emerged as the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
+In the past decade Wi-Fi has emerged as one of the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
 This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments. 
-Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as 'Probe Requests' in order to discover Wi-Fi networks available to them.
-This helps these devices to connect and switch between the WiFi networks seamlessly.
+Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as `Probe Requests' in order to discover Wi-Fi networks available to them.
+This helps these devices to connect and switch between the Wi-Fi networks seamlessly.
 
@@
 Considering the unprecendented levels of mobile device ownership in recent years, we can in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
-
 While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users.
@@
 
-Though WiFi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices by the known locations of APs and the signal strength reported by them from the mobile devices \citep{he2003range, moore2004robust, lamarca2005place}.
+Though Wi-Fi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices by the known locations of APs and the signal strength reported by them from the mobile devices \citep{he2003range, moore2004robust, lamarca2005place}.
 This can overcome the usual shortcoming of GPS, which struggles for precision and accuracy in indoor and densely built environments \citep{zarim2006,kawaguchi2009wifi, xi2010locating}. 
-Utilising this, we can easily and quickly estimate trajectories of the mobile devices just using the WiFi communication the device has with multiple known APs \citep{Sorensenlocation} which can be used similar to the GPS trajectories to understand individual travel patterns (Kim, 2006;\citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
+Utilising this, we can easily and quickly estimate trajectories of the mobile devices just using the Wi-Fi communication the device has with multiple known APs \citep{Sorensenlocation} which can be used similar to the GPS trajectories to understand individual travel patterns (Kim, 2006;\citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
 Such data can also be used in transportation planning and management to estimate travel time \citep{musa2011wiflow} and real time traffic monitoring \citep{abbott2013empirical}.
@@
 Though extensive research has been carried out on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect the users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
-There are various methods which have been devised to overcome this anonymisation process such as decomposition of OUIs where detailed device model information is estimated by analysing an already known dataset of OUIs \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for WiFi \citep{vo2016,bloessl2015scrambler}; and finally, the timing attack where the packet sequence information present in the probe request frame is used \citep{matte2016,cheng2016can}.
+There are various methods which have been devised to overcome this anonymisation process such as decomposition of OUIs where detailed device model information is estimated by analysing an already known dataset of OUIs \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and finally, the timing attack where the packet sequence information present in the probe request frame is used \citep{matte2016,cheng2016can}.
 A combination of these methodologies has been proven to produce de-anonymised globally unique device information \citep{vanhoef2016, martin2017}.
@@
 It has also been demonstrated by \citep{pinelli2015comparing} through series of experiments on a telecom operator dataset that using such network-driven approach is more advantageous compared to the widely used event-driven approaches.
-With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involve.
+With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involved.

### 2018-07-01 paper/sections/conclusion.tex
Final edits

@@
 This paper has made a novel contribution to the measurement of activities in retail centres in real time.
-To this end, it has described the collection and processing of a novel consumer Big Dataset that enables valid measures of levels of footfall activity which has been scaled across a Great Britain wide network of sensors.
+To this end, it has described the collection and processing of a novel consumer Big Dataset that enables valid measures of levels of footfall activity which has been scaled across a Great Britain wide network of sensors \citep{sss2016}.
 In both conceptual and technical terms, it illustrates the ways in which passively collected consumer data can be ‘hardened’ to render them robust and reliable by using related procedures of internal and external validation.
 
-Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
+Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.It is important to note that the filtering process work based solely on the information present in the probe requests and their temporal distribution.
 This ensures that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
@@
 Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-A snapshot showing week's worth of precise footfall in area around Charring cross, London is shown in Figure \ref{} to demonstrate the potential for such dataset.
+A snapshot showing week's worth of precise footfall in area around Charring cross, London is shown in Figure \ref{main_study_counts} to demonstrate the potential for such dataset.
 Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.

### 2018-07-01 paper/sections/introduction.tex
Final edits

@@
-New and developing technologies today provide the infrastructure over which movements and interactions can be measured and monitored in the `sentient city' \citep{amin2017seeing}.
-These include mobile phone networks, which can triangulate user locations relative to networks of masts, use of GPS to locate users of social media services, and WiFi beacon connectivity to access the Internet.
-These technologies offer differing levels of spatial precision, where mobile telephony and WiFi generally being less reliable and offer lower precision than GPS to the end users, while being more advantageous for broader mobility studies \citep{pinelli2015comparing}.
+New and developing technologies today provide the infrastructure over which movements and interactions of people can be measured and monitored in the `sentient city' \citep{amin2017seeing}.
+These include mobile phone networks, which can triangulate user locations relative to networks of masts, use of GPS to locate users of social media services, and Wi-Fi beacon connectivity to access the Internet.
+These technologies offer differing levels of spatial precision, where mobile telephony and Wi-Fi generally being less reliable and offer lower precision than GPS to the end users, while being more advantageous for broader mobility studies \citep{pinelli2015comparing}.
 
@@
 This is perhaps most manifest in the plethora of papers that analyse the distribution and content of publicly available Twitter social media feeds, which likely constitute an extremely biased and self-selecting subset of users that geo-enable their posts, drawn from a user base that itself bears no identified correspondence to any known background population \citep{lansley2016deriving}.
-Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases that may be representative of local populations, may exclude groups such as tourists from distand origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}
+Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases representative of local populations, may exclude groups such as tourists from distant origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}
 
@@
 The source and operation of this bias is unknown in the absence of extensive and context sensitive attempts to triangulate consumer data with data of known provenance relating to clearly defined populations \citep{lansley2016deriving}.
-In similar ways to other classes of Big Data, consumer data are best throught of as digital ‘exhaust’, or a by-product created by or harvested from consumer transactions.
+In similar ways to other classes of Big Data, consumer data are best thought of as digital ‘exhaust’, or a by-product created by or harvested from consumer transactions.
 
-In this paper, through a set of experiments, we evaluate the value of data collected from a Great Britain wide network of 800 devices, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
-These devices are located in shop windows and record the probes emitted by mobile phones and other WiFi enabled devices.
-These are deemed to be consumer data because devices carried by consumers routinely probe for a consumer service, specifically a WiFi connection.
-Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not Internet connectivity is established.
-Our core motivation is to appraise the usefulnessof WiFi probe records harvested from our network of sensors in order to indicate levels of pedestrian activity.
+In this paper, through a set of experiments, we evaluate the value of data collected from a Great Britain wide network of 800 devices \citep{sss2016}, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
+These devices are located in shop windows and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
+These are deemed to be consumer data because devices carried by consumers routinely probe for a consumer service, specifically a Wi-Fi connection.
+Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not internet connectivity is established.
+Our core motivation is to appraise the usefulness of Wi-Fi probe requests harvested from our network of sensors in order to indicate levels of pedestrian activity.
 More broadly still, in our future research we intend to classify the nationwide network of footfall profiles as part of a programme of research to understand the form and functioning of retail areas at a time of far-reaching structural change for the retail industry.
@@
 To this end, it is important to first undertake a thorough conceptual and technical appraisal of our consumer data source.
-In technical terms, screening of information present in the `probes' and classifing them based on their characteristics is essential in order to remove the ones emitted by devices that do not indicate pedestrian activity, such as network enabled printers and other fixed devices.
-Related to this, a method to figerprint Wi-Fi probes is necessary to remove probes from individuals’ devices that in conceptual terms should not be considered part of footfall – as when, for example, an employees is seated in an office within range of the sensor device.
+In technical terms, screening the information present in the `probes requests' and classifying them based on their characteristics is essential in order to remove the ones emitted by devices that do not indicate pedestrian activity, such as network enabled printers and other fixed devices.
+Related to this, a method to fingerprint Wi-Fi probes is necessary to remove probes from individuals’ devices that in conceptual terms should not be considered part of footfall – as when, for example, an employees is seated in an office within range of the sensor device.
 A calibration of sensor measurement is also essential on two grounds: first, individuals may carry multiple devices, or no device at all; and second, the positioning and orientation of the sensor in the retail unit may lead to systematic over- or under-enumeration.

### 2018-07-01 paper/sections/main_study.tex
Final edits

@@
 		\includegraphics [width=0.90\linewidth] {images/main_schedule.jpeg}
-		\caption{Days when the sensors were active at the corresponding location. The red square shows that manual data collection was also done.}
-		\label{pilot_clustering}
+		\caption{Data collection schedule showing the days when sensors were active at their corresponding locations. The red squares show that manual counting of pedestrians was also done on that day.}
+		\label{main_schedule}
 	\end{center}
@@
 The main study was designed to validate the results of the pilot study over different locations at different times.
-We choose five different locations across central London, to install the sensors and collect data for a long period of time. 
-We also carry out manual counting on these locations along with this across different time of the day. 
-We then apply the filtering based on signal strength and sequence numbers and compare with manual counts and evaluate the effectiveness of the process with the mean error per minute on these locations.
-Finally we calculate the ajustment factor for the first interval of manual counts and check if that works on the consecutive intervals.
+We start by choosing five different locations across central London, install the sensors and collect data for extended periods of time.
+We also carry out manual counting at these locations across different times of the day. 
+We then apply the methodologies discussed earlier to arrive at estimated pedestrian footfall and compare them with corresponding manual counts.
+We finally evaluate the effectiveness of the processes with the Mean Absolute Percentage Error (MAPE) at these locations and report our findings.
 
-The locations where the data were collected are shown in the table.
-The location are chosen for their variety of configuration and sources of noise.
-Location 1 is the `cleanest' of while location 2 is the one with the most complexity.
-The configuration, installation and data collection schedule is shown in the figure.
-
-Though data was collected for many continous days, for the purposes of comparing with ground truth we just consider the only the data from sensors corresponding to those sessions.
-We have 12 sets of data over 6 different days.
-We have atleast two set manual counts for each location for verification of calibration.
+The locations at which the data were collected are shown in Table \ref{locations-table}.
+The locations are chosen for their diverse site conditions and unique sources of noise around the potential location of the sensors.
+The position of the sensor at these locations with respect to the context is shown the Figure \ref{main_signal_strength}.
+We can see that Location 5 is the `cleanest' with one clear stationary source of noise (phone shop) of while location 2 is the one with the most complexity with seating area all around the sensor.
+The sensors were operational through out February and March while manual counts were conducted in these location in half hour sessions on at least two different days.
+For the purposes of comparing with ground truth we just considered the data from sensors which corresponding to the 12 sets of available manual counts.
+The schedule of data collection is shown in Figure \ref{main_schedule}
 
@@
 	\centering
-	\subfigure[Installation configuration of sensors at various locations]{
+	\subfigure[Installation configuration of sensors at the survey locations (not to scale)]{
 		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
-	\subfigure[Density distribution of signal strength (lower values show higher signal strength)]{
+	\subfigure[Density distribution of signal strength reported in collected probe requests (lower values show higher signal strength)]{
 		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/main_signals.jpeg}}}
-	\caption{Distribution of signal strengths across locations} \label{methodology_schematic}
+	\caption{Distribution of signal strengths across locations} \label{main_signal_strength}
 \end{figure}
 
-First we see how the distribution of the signal strength varies with location and configuration. 
-The density plot for signal strength is shown alogn with configuration in figure.
-We can see that the signal strength distribution shows distinct patterns of high and low when the installation is that there is a clear distant source of noise but this distinction get more and more obscure as we move towards difficult installations.
-For example, location 2 is almost a normally distributed noise as it is too far to pick up any pedestrians but location 5 with a clear view of footpath and a phone shop next door shows clear distinction between the two.
-Intuitively the classfication algorithm should give us better results in the latter. 
-It is important to note that we are dealing with relative singal strengths, this can vary with location and time of the analysis but we should be still be able to differentiate signal from noise. 
-We run the kmeans classification algorithm and filter out the probe requests which are randomised and have signal strengths less than the second break (or the threshold). 
-We then count the number of Unique MAC addresses present in every minute and remove MAC addreses which reappear within 30 minutes of previous appearance. 
-We then compare this with the minute by minute aggreagation for the manual counts and find the average error per minute for the sensor count. 
-The results are shown in the table.We see that the location 2 has the most error 400\% confirming our intuition while surprisingly location 3 has the least error -3\%. 
-We also see that the error follows the complexity of the installation. 
-It is also very promising that this method alone reduces our margin of error by a 50 - 100\%
-For some practical purposes which do not require absolute numbers, this should be sufficient. 
-e.g. Indexes of activity and change dashboards and short term trends identification.
+We start by looking at the distribution of the signal strength reported by the probe requests across the locations.
+From the density plot shown in Figure \ref{main_signal_strength} we can see that there is clear relation between the distribution of the signal strength and the distance and complexity of the source of noise.
+We can see that while location 5 shows clean difference between low and high signal strengths, location 2 is almost normally distributed.
+Intuitively we expect that location 2 and 4 must be harder to classify than location 1 3 and 5. 
+We run the k-means classification algorithm and filter out the probe requests which are randomised and have signal strengths less than the second break (threshold). 
+It is important to note that we are dealing with relative thresholds of signal strengths which can vary with location and time of the analysis.
+We then aggregate then probe requests for every minute by counting the number of Unique MAC addresses present in every minute.
+We also remove devices that dwell around the sensor by removing the MAC addresses which reappear from within the past hour.
+The results of the first stage of filtering process along with the thresholds are shown in Table \ref{errors-table}.
+Confirming our intuition, we see that the location 2 has the most MAPE followed by location 4 while rest of them have highly reduced MAPE.
+It is significant that this method alone reduces our margin of error by a 50 - 100\% from the raw counts without any cleaning.
+This makes the signal strength filtering quick and ideal method for practical applications which doesn't require absolute numbers such as creating large aggregated indexes to show long term trends. 
+We also see that the success of the signal strength filtering can be improved significantly by installing sensors such that the pedestrians and noise are at different distances from the sensor and are the field of measurement is distinct from the surroundings however noisy it might be.
 
-We see that the success of the signal strength filtering depends on the how we tackle the problems in the installing the devices.
-The more the device in installed in a way that the field of measurement is distinct from its surroundings (however noisy they might be) the more successful this process is.
-Closer sources error causes overcounting as we see in 2, 4 but at the same time highly close field of measurement with no big source of error can cause undercounting such as in 3.
-The complexity of the installation and uniformity of surroundings near the sensors directly coorrelate to the MAPE.
-The most effective way of measuring pedestrian is to have the sensor next to have the sensor next to clear area of pedestrian activity and in case of absence of any major source of noise, we should not do the signal strength filtering at all.
-We can even note the major possible sources of noise around the device and use it to fiugre out the right way of partitioning the data e.g.
-location 2 and 3 should have different pattern of signal strength filtering depending on their configuration.
-
-In terms of filtering by sequence numbers we see that the the filtering works in all the locations.
-It has the similar effect as the signal strength and brings us furthe closer to the manual count.
+In terms of filtering by sequence numbers we see that the filtering works in all the locations.
+It has the similar effect as the signal strength and brings us further closer to the manual count.
 Finally we calculate the adjustment factor from the first 15 minutes and apply that for the rest of the interval.
@@
 \begin{table}
-	\tbl{Data collected at different locations and the correspondin mean absolute percentage error after each filtering process}
+	\tbl{Data collected at different locations and the corresponding mean absolute percentage error after each filtering process}
 	{\begin{tabular}{ccccccc} 
@@
 		\includegraphics [width=\linewidth,trim=6 6 6 6,clip] {images/main_comparison.jpeg}
-		\caption{Comparison of all the counts in teh sensors}
+		\caption{Comparison of all the counts in the sensors}
 		\label{main_comparison}
@@
 \end{figure}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics [width=\linewidth] {images/main_schedule.jpeg}
+		\caption{A week of pedestrian footfall at Camden Highstreet collected by the methodology. The counts are aggregated for 5 minute intervals.}
+		\label{main_study_counts}
+	\end{center}
+\end{figure}

### 2018-07-01 paper/sections/previous_work.tex
Final edits

@@
 Probe requests are low level signals standardised by IEEE 802.1b/g specification \citep{ieee2013} as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
-This ubiquity and standardisation make them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.
-Considering the unprecendented levels of mobile device ownership in recent years, we can in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
+This ubiquity and standardisation makes them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.
+Considering the unprecendented levels of mobile device ownership in recent years, we can, in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
 While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users.
@@
 
-Though Wi-Fi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices by the known locations of APs and the signal strength reported by them from the mobile devices \citep{he2003range, moore2004robust, lamarca2005place}.
+Though Wi-Fi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices from the locations of known access points(AP) and the signal strength reported by them \citep{he2003range, moore2004robust, lamarca2005place}.
 This can overcome the usual shortcoming of GPS, which struggles for precision and accuracy in indoor and densely built environments \citep{zarim2006,kawaguchi2009wifi, xi2010locating}. 
-Utilising this, we can easily and quickly estimate trajectories of the mobile devices just using the Wi-Fi communication the device has with multiple known APs \citep{Sorensenlocation} which can be used similar to the GPS trajectories to understand individual travel patterns (Kim, 2006;\citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
+Utilising this, we can easily and quickly estimate trajectories of the mobile devices \citep{Sorensenlocation,musa2012tracking} which can be used similar to the GPS trajectories to understand individual travel patterns \citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
 Such data can also be used in transportation planning and management to estimate travel time \citep{musa2011wiflow} and real time traffic monitoring \citep{abbott2013empirical}.
-
-Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} and using the information present in the probe requests one can also successfully track people across the sensors or access points collecting them \citep{cunche2014linking}, infer their trajectories \citep{musa2012tracking}, and even model interactions between them \citep{cheng2012inferring,barbera2013signals,cunche2014know} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
+Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} along with information present in the probe requests one even model interactions between the users \citep{cheng2012inferring,barbera2013signals,cunche2014know,cunche2014linking} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
 Using the semantic information present in these probe requests it even is possible to understand the nature of population at a large scale \citep{di2016mind}.
 
-Though extensive research has been carried out on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect the users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
-There are various methods which have been devised to overcome this anonymisation process such as decomposition of OUIs where detailed device model information is estimated by analysing an already known dataset of OUIs \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and finally, the timing attack where the packet sequence information present in the probe request frame is used \citep{matte2016,cheng2016can}.
+Though extensive research has been carried out on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect their users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
+Various methods have been devised to overcome this anonymisation process such as estimating the device model information from a known dataset of manufacturers and device behaviours \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and timing attack where the packet sequence information along with information elements present in the probe request frame is used \citep{matte2016,cheng2016can}.
 A combination of these methodologies has been proven to produce de-anonymised globally unique device information \citep{vanhoef2016, martin2017}.
@@
 
-There is a clear gap in the research for exploring methodologies which enable us to estimate the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original MAC addresses.
-Such technique has various applications in numerous fields such as uncovering the urban wireless landscape \citep{rose2010mapping}, revealing human activity at large scales \citep{qin2013discovering}, estimating pedestrian numbers in crowds \citep{schauer2014estimating,fukuzaki2015statistical} and even counting people in hyper local scales such as queues \citep{wang2013measuring}
-With enough infrastructure to collect such information we can aim to generate a real-time census of the city \citep{konto2017}.
-It has also been demonstrated by \citep{pinelli2015comparing} through series of experiments on a telecom operator dataset that using such network-driven approach is more advantageous compared to the widely used event-driven approaches.
+There is a clear gap in research for exploring methodologies for estimating the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original device information.
+Such technique has various applications such as uncovering the urban wireless landscape \citep{rose2010mapping}, revealing human activity at large scales \citep{qin2013discovering}, estimating pedestrian numbers in crowds \citep{schauer2014estimating,fukuzaki2015statistical} and even counting people in hyper local scales such as queues \citep{wang2013measuring}
+With enough infrastructure to collect such information we can even aim to generate a real-time census of the city \citep{konto2017}.
 With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involved.

### 2018-07-01 paper/sections/main_study.tex
final commit

@@
 
-In terms of filtering by sequence numbers we see that the filtering works in all the locations.
-It has the similar effect as the signal strength and brings us further closer to the manual count.
-Finally we calculate the adjustment factor from the first 15 minutes and apply that for the rest of the interval.
-We get the counts down to almost 10\% on good locations.
-The locations which still over count are the ones with seating next to them (2 and 4).
-We also see that the trends are preserved with the filtering process.
+We then run the sequence numbers based clustering process on the rest of the probe requests to reduce the MAPE by almost 50-100\% on all the sensors except for location 3.
+Location 3 is an outlier among all the other sensors since it is the only one with large amount of pedestrians very close to the sensor.
+This may be causing the over filtering caused by the previous process.
+We finally run the calibration process where we calculate the adjustment factors shown in Table \ref{errors_table}.
+We the use them to adjust the counts to achieve a MAPE ranging from 10 - 50\%.
+We can see that the sensors with people moving right next to them tend to under-count with our methodology while sensors with seating next to them tend to over-count significantly.
+However, using the filtering process, we can reduce the error to almost 10\% closer to that of the ground truth.
 
 \begin{table}
-	\tbl{Data collected at different locations and the corresponding mean absolute percentage error after each filtering process}
+	\tbl{Results of footfall estimation at each location as Mean Absolute Percentage Error (MAPE) after each step of the filtering process}
 	{\begin{tabular}{ccccccc} 
@@
 		\includegraphics [width=\linewidth,trim=6 6 6 6,clip] {images/main_comparison.jpeg}
-		\caption{Comparison of all the counts in the sensors}
+		\caption{Comparison of the filtering process with the ground truth in all the locations.}
 		\label{main_comparison}
@@
 	\begin{center}
-		\includegraphics [width=\linewidth] {images/main_schedule.jpeg}
-		\caption{A week of pedestrian footfall at Camden Highstreet collected by the methodology. The counts are aggregated for 5 minute intervals.}
+		\includegraphics [width=\linewidth] {images/main_footfall.jpeg}
+		\caption{A week of pedestrian footfall at the Strand, London collected by the methodology. The counts are aggregated for 5 minute intervals.}
 		\label{main_study_counts}

### 2018-07-01 analysis/main_study.r

@@ -302,3 +302,83 @@ counts_plot <- counts %>%
 		scale_color_manual(values=c("#e41a1c","#377eb8","#ff7f00","#4daf4a","#333333"))
 
 dbDisconnect(connection)
+
+single_sensor <- dbGetQuery(connection,"select sensor.*, vendor.vendor,case when substring(sensor.oui,2,1) in ('e','a','2','6') then 'local' else 'global' end as type from sensor left join vendor on sensor.oui=vendor.oui where sensor = 55 and date(timestamp) > '2018-02-18' and date(timestamp) < '2018-02-26';")
+
+single_sensor_c <- single_sensor %>%
+	mutate(date = as.Date(timestamp),
+		   hour = format(timestamp,"%H"),
+		   five = format(floor_date(timestamp,"5 minutes"),"%M")) %>%
+	select(-sensor,-length,-oui) %>%
+	mutate(interval = paste(date,hour,five,sep="-"))
+
+filter_signal <- function(d) {
+	if(nrow(d)>0) {
+		sig <- d$sig[d$type=="local"]
+		thres <- classIntervals(sig,2,"kmeans")$brks[2]
+		return(d[!(d$type=="local" & d$signal<thres),])
+	}
+}
+
+filter_sequence <- function(data, ds=200, dt=60) {
+	if(nrow(data)>0) {
+	data1 <- data[data$type=="global",]
+	data1$signature <- paste(data1$date,data1$mac,sep="-")
+	data <- data[data$type=="local",]
+	vertices <- data %>%
+		mutate(x = as.integer(timestamp), id = as.numeric(rownames(data))) %>%
+		select(id, x, y=sequence) %>%
+		arrange(x)
+	edges <- vertices %>%
+		pull(id) %>%
+		expand.grid(.,.) %>% rename(from=Var1,to=Var2) %>%
+		mutate(from=as.numeric(from), to=as.numeric(to)) %>%
+		filter(from<to) %>%
+		left_join(vertices, by = c("from" = "id")) %>%
+		left_join(vertices, by = c("to" = "id")) %>%
+		mutate(sdistance = sqrt((y.y-y.x)**2)) %>%
+		mutate(tdistance = sqrt((x.y-x.x)**2)) %>%
+		mutate(distance = sqrt((tdistance**2)+(sdistance**2))) %>%
+		filter(x.y >= x.x, y.y > y.x, tdistance < dt, sdistance < ds) %>%
+		group_by(from) %>%
+		summarise(to = to[which(distance == min(distance))][[1]],
[…]

### 2018-07-01 final_submission/sections/main_study.tex
submitted

--- 2018-07-01 paper/sections/main_study.tex
@@
 We also remove devices that dwell around the sensor by removing the MAC addresses which reappear from within the past hour.
-The results of the first stage of filtering process along with the thresholds are shown in Table \ref{errors-table}.
+The results of the first stage of filtering process along with the thresholds are shown in Table \ref{errors_table}.
 Confirming our intuition, we see that the location 2 has the most MAPE followed by location 4 while rest of them have highly reduced MAPE.
@@
 
-We then run the sequence numbers based clustering process on the rest of the probe requests to reduce the MAPE by almost 50-100\% on all the sensors except for location 3.
+We then run the sequence numbers based clustering process on the rest of the probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except for location 3.
 Location 3 is an outlier among all the other sensors since it is the only one with large amount of pedestrians very close to the sensor.
@@
 	\end{tabular}}
-	\label{errors-table}
+	\label{errors_table}
 \end{table}

### 2018-07-01 final_submission/sections/pilot_study.tex
submitted

--- 2018-06-30 paper/sections/pilot_study.tex
@@
-To start we designed a small pilot study to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail highstreet.
+To start we designed a small pilot study to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail high street.
 We also aim to find the algorithm which is best suited for classification of signal strengths to filter out the background noise.
@@
 		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_clustering.jpeg}}}\hspace{20pt}
-	\subfigure[Comparision of sensor based counts with the groud truth.]{
+	\subfigure[Comparison of sensor based counts with the ground truth.]{
 	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_comparison.jpeg}}}
-	\caption{The results of the pilot study demonstrating the validity of the deviced methodology.} \label{pilot_results}
+	\caption{The results of the pilot study demonstrating the validity of the methodology.} \label{pilot_results}
 \end{figure}
@@
 We then compare this preliminary `footfall' count to the actual number of pedestrian recorded manually to check for the robustness.
-We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there is no `zero records' issue.
+We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there are no intervals without any footfall.
 We find that the MAPE in the raw counts compared to the ground truth is around 425\%.
@@
 
-We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, heirarchical clusters, fisher and jeanks.
-We found that while heirarchical clustering and jenks were resource intensive, kmeans gave quickest results with the lowest MAPE closely followed by quantile.
+We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, hierarchical clusters, fisher and jenks.
+We found that while hierarchical clustering and jenks were resource intensive, k-means gave quickest results with the lowest MAPE closely followed by quantile.
 The cut-off point or threshold for the collected data with which we could classify them as high and low was -71 dBm.
@@
 We finally aggregate the probe requests as before but with the device signature rather than just MAC addresses this results in a footfall count with a MAPE of -18\%. 
-A comparision of minute by minute counts resulting from different filtering processes along with the ground truth is shown in Figure \ref{pilot_results} showing the promising effectiveness of the methods.
+A comparison of minute by minute counts resulting from different filtering processes along with the ground truth is shown in Figure \ref{pilot_results} showing the promising effectiveness of the methods.
 

### 2018-11-03 review/response.md
started drafting responses

First review comments
=====================

Title : Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests
Manuscript id : IJGIS-2018-0340
Comments dated : 28-Oct-2018

1 Editor
--------

1.1 Overall Comments
---------------------

1. While this is an important and timing topic dealing with data uncertainty
and privacy in the mobile age, the reviewers have raised concerns on
	- Clarity of the methodology
	- Lack of discussion of wifi-based human activity studies in the big data age
	- Proper use of terminology
	- Performance of the algorithm etc.

2. In lieu of these comments, they have recommended to make substantial revisions
to your manuscript

3. Also, since this is a "smart cities" related special issue, I ask you to
	- Add more discussions on smart cities in the introduction and 
	  conclusion sections and
	- Link your research more tightly to this theme.

>>> Thank you for the very helpful comments of the reviewers of our paper/
>>> As requested, we have responded to all comments without adding
>>> significant length to the paper. We have also extended our discussion of
>>> how effective measurement of footfall is relevant to understanding the
>>> functioning of the Smart City.
>>> 
>>> If any of what follows and what has been uploaded is unclear, please do
>>> not hesitate to contact me.
>>> 
>>> Yours sincerely
>>> Balamurugan Soundararaj
>>> (on behalf also of Paul Longley and James Cheshire)

1.2 General Instructions
------------------------

1. Please try to avoid making the paper longer (keep in mind that the average
length of papers in this journal is 18-20 published pages). It is important that
you ensure that the spelling and grammar is of a high standard.

2. Please also highlight the changes to your manuscript within the document by
using the track changes mode in MS Word or by using bold or colored text.

3. To revise your manuscript, please select the link below,
https://mc.manuscriptcentral.com/ijgis?URL_MASK=a32caf3bbd49412cac8d1cfa9d7ebaf8
which will take you your decision letter, and where you need to
resubmit your manuscript.

4. If you have used the on-line LaTeX template, upload all those files and a pdf.

5. Please check your English spelling and grammar.

6. There are two hard limits that you need to bear in mind, 
	- Image size - Please be aware that the total number of pixels 
	  (height X width) in each image must be less than 40,000,000 (40 mp).
	- Total file size - please make sure that the combined, uncompressed,
	  file sizes do not exceed 20Mb. The internal LZW compression, which 
	  is an option in TIF, is acceptable however.
	- Please do not save B/W or Greyscale images as RGB.

7. Equations and tables need to be in an editable format.

8. You need to check that the paper title in ScholarOne is the same as the
title on the paper and that the Title page contains all authors’ names and
affiliations as you would want them to appear.

9. When submitting your revised manuscript, you will be able to respond to
the comments made by the reviewers in the space provided.  You can use this
space to document any changes you make to the original manuscript.  In order
to expedite the processing of the revised manuscript, please be as specific as
possible in your response to the reviewers.

10. Please delete any redundant files before completing the submission.

2 Reviewer 1
------------

2.1 Overall Summary
-------------------

This paper is about the estimation of pedestrian footfall using Wi-Fi
signal emitted by mobile devices.
It deals with the problem of estimating the footfall size based on a noisy
and biased data. More specifically it addresses the following issues that may
lead to overcounting:
 - MAC randomization that may cause a device to generate several MAC addresses;
 - Signals received from outside the interest zone; unequal adoption
   of Wi-Fi smartphone in the population.
To address this issue, a number of techniques are introduced:
 - clustering received signal strength to remove device that are outside
   the monitored area;
 - clustering probes based on sequence number and timing;
 - finally an adjustment factor is applied to correct the estimated value.
   The adjustment factor is calibrated for each location using
   a manual counting.
The author conducted a pilot experiment to identify the clustering 
algorithms and parameters, and then they conducted a second range of 
experiments at different location over an extended period of time. During 
each of those experiments, ground-truth count was collected manually.
The results show that the presented approach can significantly reduce the
counting error compared to a naive solution consisting in counting the 
distinct number of MAC observed.
Nevertheless, there are cases (2) where the error is still large (50%),
even if in other cases (1) the error is only of 9%.

>>> This is a correct surmisal of the remit of the paper. We have also added#
>>> material to the introduction and conclusion of the paper to better link
>>> this remit to the Smart Cities discourse.

2.2 General Comments
--------------------

1. The idea of clustering frames using sequence number has been introduced 
in [1]. This work needs to be cited when introducing the proposed clustering
algorithm.
[1] Mathy Vanhoef, Célestin Matte, Mathieu Cunche, Leonardo Cardoso, Frank
Piessens. Why MAC Address Randomization is not Enough: An Analysis of Wi-Fi
Network Discovery Mechanisms. ACM AsiaCCS, May 2016, Xi’an, China. 2016,

>>> Thank you. We have included a citiation to the source, as well as one 
>>> other recent addition to this fast-developing literature.

2. The author considered that the hashing of MAC address is a good way to 
protect users privacy, as some MAC address collected by the device may not be 
random. However hashing of MAC address cannot be considered as a proper 
anonymization technique (see [2] and 
http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/ ); 
at best it can be considered as an obfuscation. 
[2] Levent Demir, Mathieu Cunche, Cédric Lauradoux. Analysing the privacy 
policies of Wi-Fi trackers. Workshop on Physical Analytics, Jun 2014, Bretton 
Woods, United States. ACM, 2014

>>> Thank you for pointing this out. In the revised version of the paper
>>> we have made reference to this discussion, while also making clear the 
>>> anonymisation is ensured by the system architecture that processess and
>>> curates the data. Spedicifically the weekly rotation of the random salt
>>> with which the hashing process is carried out.

3. There is a number of commercial solutions doing counting based on Wi-Fi 
signal. Although the detail of their solution is not public, it would be 
worthwhile to acknowledge their existence.

>>> Thank you - this has been done.

4. In section 3.2, it is explained that data is divided into two sets 
corresponding on random and non-random MAC. Which method is used to perform 
this classification ?

>>> The identification of a MAC address as randomised or non-randomised is done
>>> looking at the second character of the vendor part of the MAC address.
>>> The addresses with E, A, 2, 6 at these locations are identified to be
>>> randomised MAC addresses. This has been included in the revised paper.

5. How was the sensor positioned compared to the area monitored used for 
ground truth ? A more detailed description is required regarding the positioning
of the sensor and the size and topology of the surveyed area.

>>> This is a good point, and is very important for our validation work. We
>>> have included a more detailed and explicit discussion of the way that field
>>> observation was used to calibrate and cross validate the Wi-Fi results.

2.3 Clarifications & Corrections
--------------------------------

1. The topic addressed in this work is interesting and timely. The proposed
approach is sound and shows promising results. The fact that the authors 
conducted field experiments to evaluate their solution is a positive point. 
Yet before publication, I believe that there are a number of points that needs 
to be clarified or corrected. Regarding the technical aspects I have a
number of minor remarks:

>>> Thank you very much for these very positive comments. The paper reports on
>>> extensive methodological research and field implementation and is, we 
>>> believe, cutting edge in its finding. It also has important implications 
>>> for our understanding of the functioning of smart cities, and these
>>> implications are discussed in greater detail in the revised version.

2. "WiFi beacon technologies to access the Internet". In Wi-Fi, beacons are
management frames used by Access Point for service discovery. They do not
provide Internet access. "Wi-Fi Access Point to access ..."

>>> Thank you. We have corrected and clarified the point.

3. "Wi-Fi antennae regularly broadcast a special type of signal" -> the antenna
is just the part ensuring that the signal is correctly emitted, but it is not 
the source. In this context it is better to use a more general concept and refer
to the "Wi-Fi network interface" (or Wi-Fi interface) as the the element
broadcasting a signal.

>>> Thank you. Point clarified in the revised version.

4. The IEEE 802.11 b/g is just a subpart of the 802.11 specification. Here you 
can refer to the 802.11 specifications as a whole : [“802.11-2012 - IEEE 
Standard for Information technology–Telecommunications and information exchange
between systems Local and metropolitan area networks–Specific requirements 
Part 11: Wireless LAN Medium Access Control (MAC) and Physical Layer (PHY) 
Specifications,” IEEE Std 802.11-2012 (Revision of IEEE Std 802.11-2007),
pp. 1–2793, Mar. 2012.

>>> Thank you. Noted and added to the revised version.

5. "the first step in establishing a Wi-Fi based connection " is also known as
service discovery

>>> Thank you. Added to the revised version

6. "Wi-Fi functionality has been turned off by the user", regarding this point
you may be interested in :

7. "Media Access Control (MAC) address which is an unique identifier for the 
wireless hardware of the mobile device"  The transmitter field of the 802.11
frame can indeed contain a unique MAC address  but can also contain a random 
one as noted by the author. Here it would be better to avoid stating that the 
content of this field is always a unique MAC.

>>> Thank you. Again, this has been noted in the revised version of the paper.

8. "The strength of the signal which transmitted the request.". In fact the 
Received Signal Strength Indicator, a value providing an indication on the 
strength of the received signal as seen by the receiver, not the strength at
the emitter side.

>>> Thank you. This distinction has been noted in the revised version of the
>>> paper.

9. Could you provide more information on the application used for counting 
(Clicker)

>>> This simple application was developed specifically for the project for
>>> Android platform. More details on the application has been included in 
>>> the revised manuscript.

10. "Though the recycling of sequence number r after 4000 leads to multiple
classifications reported on single device, the magnitude of error is greatly
reduced." This sentence is not clear. And the cycle length is not 4000 but 
4096 = 2^12.

>>> Thank you. This point has been boted and accommodated into the revised
>>> version.

11. Regarding the issue of cycling sequence number, one could use a distance 
with a modulo : di,j = di - dj mod 4096.

>>> Thank you very much for the suggestion. From out tests on using this 
>>> method for distance calculation results in a change of --% of the result.
>>> Which in our opinion doesnt influence the results in any significant way
>>> considering the context.

2.4 Typos and other remarks
--------------------------------

1. Challanges

2. "These are deemed to be consumer data because devices carried by consumers 
routinely probe for a consumer service, specifically a Wi-Fi connection.
Monitoring the probes from such devices provides an indication" This sentence 
is not clear.

3. "Though Wi-Fi F is a ‘location-less’ technology, ", please clarify what 
you mean by that.

4. "privacy infringement" may not be the best word as infrigement relate to a 
law or a regulation. Privacy breach or leak may be a better choice.

5. Wireshark and not WireShark.

6. "We hypothesize that an adjustment factor could be arrived at for each 
location of data collection, ""This calibration can be carried over periodically
and the frequency to improve the quality of the estimation." something wrong in 
both sentences.

>>> Thak you for picking up these minor lapses. All have been corrected or 
>>> clarified in the revised version of the paper.

3 Reviewer 2
------------

3.1 Overall Summary
-------------------

1. This paper addressed the data quality issue of Wi-Fi-based probe requests in 
estimating pedestrian counts. The authors proposed several methods to improve
the accuracy of estimating the number of unique mobile devices from a set of 
anonymized probe requests without revealing their original device information.

2. This is a crucial topic in mobile data uncertainty, and the manuscript has 
the potential of becoming a useful reference in the field. However, it was not 
organized effectively and missing many details in the methodology.  Before
addressing these problems, it is difficult to assess the reliability and 
scientific merit of the case study.

3.2 General Comments
--------------------

1. First, the literature review section adequately covered Wi-Fi-based human 
activity studies, but it is in lack of an overview of modeling human activity 
in the big data era. I would suggest the authors add an overarching paragraph 
(or a sub-section) in Section 2 discussing the importance and challenges of 
modeling human activities and urban dynamics based on various types of big 
geodata, such as mobile phone records, location-based social media, Bluetooth 
data, etc.

2. The conclusion did not address most of the important results from the case study and should be expanded.

3.3 Corrections and Clarifications
----------------------------------

Second, the methodology should be better clarified. There are many details 
missing or in need of a better justification. For example,

1. P4 l50, without providing any explanation, the authors hypothesized that 
sequence number and length of the packet are sufficient to estimate the number 
of unique devices. This argument was not elaborated until 2.5 pages later 
in section 3.2.

2. P6 l27, please provide more information regarding how you defined the 
threshold for low/high signal strength when eliminating the background noise. 
This was also unclear in the case study on P8, where the authors repeatedly 
mixed up classification and clustering. Neither k-means nor hierarchical 
clustering is considered a “classification algorithm” as claimed in p8 l26. 
Very few details were provided regarding the parameter setting of 
these analyses.

3. P6 l30-31, in the phone shop example, the authors mentioned that it is 
possible to identify background noise based on a sharp rise of the request 
number; however, this will also eliminate the regular pedestrian flow in that 
shop and cause additional inaccuracy. Please clarify.

4. P7 l21, how often does the recycling of sequence number occur, and how will 
this impact the reliability of your analysis? Please provide more details.

5. P7 Section 3.3, please be more specific about the calibration process and 
the external source of information to be used here.

6. P7 l45, the analysis in this research is based on one single sensor for each 
location; however, in reality, it is very common to estimate travel flows based 
on multiple sensors (e.g., nearby stores in a shopping mall). How would your 
method address duplication in multiple sensors?

7. P10 l48, how was the adjustment factor determined?

4 Reviewer 3
------------

4.1 Overall Summary
-------------------

1. This paper presents Wi-Fi based footfall counting methodology, which is timely and interesting topic. The method can be performed to estimate human activity such as pedestrian footfall from Wi-Fi probe requests.

4.2 General Comments
--------------------

Some areas where I would like to see more detail:

1. It would have been interesting to understand the temporal characteristics for
a week pedestrian footfall. How to consider this situation in your study? Why do
you choose time range from 12:30 to 13:00 hrs ?

2. Will you give more detail information about the data and the parameters for 
various clustering algorithms? Including the number of records, the number of 
clusters.

4.3 Corrections and Clarifications
----------------------------------

More specific comments can be found in the following.

1. Minor formatting issues and typos needs to be fixed.

2. Figure 2, no value and units in the x-axis and y-axis.

3. Table 1, and Figure 4 are in wrong location.

4. Related work: How about WiFi-based real-time data analytics?

5. Methodology: Will you draw a diagram of the Methodology?

4. Performance Analysis: How about the performance for collecting data using 
PostgreSQL? Did you compare Apache Flink and Spark Streaming for real-time data 
collecting?

7. How about the performance for different clustering methods in your study？

8. The reference list needs to be written in a standard format. Some reference papers are lacking specific pages .Please use a standard way of reference list.

### 2018-11-05 review/response.md
responses

@@
 
->>> Thank you for the very helpful comments of the reviewers of our paper/
->>> As requested, we have responded to all comments without adding
->>> significant length to the paper. We have also extended our discussion of
->>> how effective measurement of footfall is relevant to understanding the
->>> functioning of the Smart City.
->>> 
->>> If any of what follows and what has been uploaded is unclear, please do
->>> not hesitate to contact me.
->>> 
->>> Yours sincerely
->>> Balamurugan Soundararaj
->>> (on behalf also of Paul Longley and James Cheshire)
+> Thank you for the very helpful comments of the reviewers of our paper/
+> As requested, we have responded to all comments without adding
+> significant length to the paper. We have also extended our discussion of
+> how effective measurement of footfall is relevant to understanding the
+> functioning of the Smart City.
+> 
+> If any of what follows and what has been uploaded is unclear, please do
+> not hesitate to contact me.
+> 
+> Yours sincerely
+> Balamurugan Soundararaj
+> (on behalf also of Paul Longley and James Cheshire)
 
@@
 
->>> This is a correct surmisal of the remit of the paper. We have also added#
->>> material to the introduction and conclusion of the paper to better link
->>> this remit to the Smart Cities discourse.
+> This is a correct surmisal of the remit of the paper. We have also added#
+> material to the introduction and conclusion of the paper to better link
+> this remit to the Smart Cities discourse.
 
@@
 
->>> Thank you. We have included a citiation to the source, as well as one 
->>> other recent addition to this fast-developing literature.
+> Thank you. We have included a citiation to the source, as well as one 
+> other recent addition to this fast-developing literature.
 
@@
 
->>> Thank you for pointing this out. In the revised version of the paper
->>> we have made reference to this discussion, while also making clear the 
->>> anonymisation is ensured by the system architecture that processess and
->>> curates the data. Spedicifically the weekly rotation of the random salt
->>> with which the hashing process is carried out.
+> Thank you for pointing this out. In the revised version of the paper
+> we have made reference to this discussion, while also making clear the 
+> anonymisation is ensured by the system architecture that processess and
+> curates the data. Spedicifically the weekly rotation of the random salt
+> with which the hashing process is carried out.
 
@@
 
->>> Thank you - this has been done.
+> Thank you - this has been done.
 
@@
 
->>> The identification of a MAC address as randomised or non-randomised is done
->>> looking at the second character of the vendor part of the MAC address.
->>> The addresses with E, A, 2, 6 at these locations are identified to be
->>> randomised MAC addresses. This has been included in the revised paper.
+> The identification of a MAC address as randomised or non-randomised is done
+> looking at the second character of the vendor part of the MAC address.
+> The addresses with E, A, 2, 6 at these locations are identified to be
+> randomised MAC addresses. This has been included in the revised paper.
 
@@
 
->>> This is a good point, and is very important for our validation work. We
->>> have included a more detailed and explicit discussion of the way that field
->>> observation was used to calibrate and cross validate the Wi-Fi results.
+> This is a good point, and is very important for our validation work. We
+> have included a more detailed and explicit discussion of the way that field
+> observation was used to calibrate and cross validate the Wi-Fi results.
 
@@
 
->>> Thank you very much for these very positive comments. The paper reports on
->>> extensive methodological research and field implementation and is, we 
->>> believe, cutting edge in its finding. It also has important implications 
->>> for our understanding of the functioning of smart cities, and these
->>> implications are discussed in greater detail in the revised version.
+> Thank you very much for these very positive comments. The paper reports on
+> extensive methodological research and field implementation and is, we 
+> believe, cutting edge in its finding. It also has important implications 
+> for our understanding of the functioning of smart cities, and these
+> implications are discussed in greater detail in the revised version.
 
@@
 
->>> Thank you. We have corrected and clarified the point.
+> Thank you. We have corrected and clarified the point.
 
@@
 
->>> Thank you. Point clarified in the revised version.
+> Thank you. Point clarified in the revised version.
 
@@
 
->>> Thank you. Noted and added to the revised version.
+> Thank you. Noted and added to the revised version.
 
@@
 
->>> Thank you. Added to the revised version
+> Thank you. Added to the revised version
 
@@
 
->>> Thank you. Again, this has been noted in the revised version of the paper.
+> Thank you. Again, this has been noted in the revised version of the paper.
 
@@
 
->>> Thank you. This distinction has been noted in the revised version of the
->>> paper.
+> Thank you. This distinction has been noted in the revised version of the
+> paper.
 
@@
 
->>> This simple application was developed specifically for the project for
->>> Android platform. More details on the application has been included in 
->>> the revised manuscript.
+> This simple application was developed specifically for the project for
+> Android platform. More details on the application has been included in 
+> the revised manuscript.
 
@@
 
->>> Thank you. This point has been boted and accommodated into the revised
->>> version.
+> Thank you. This point has been boted and accommodated into the revised
+> version.
 
@@
 
->>> Thank you very much for the suggestion. From out tests on using this 
->>> method for distance calculation results in a change of --% of the result.
->>> Which in our opinion doesnt influence the results in any significant way
->>> considering the context.
+> Thank you very much for the suggestion. From out tests on using this 
+> method for distance calculation results in a change of --% of the result.
+> Which in our opinion doesnt influence the results in any significant way
+> considering the context.
 
@@
 
->>> Thak you for picking up these minor lapses. All have been corrected or 
->>> clarified in the revised version of the paper.
+> Thank you for picking up these minor lapses. All have been corrected or 
+> clarified in the revised version of the paper.
 
@@
 
+> This is correct, although the intended emphasis upon disclosure is a feature
+> rather than a principal motivation. We have also added further discussion of
+> how these principally technical issues relate to smart city data
+> infrastructure issues.
+
 2. This is a crucial topic in mobile data uncertainty, and the manuscript has 
@@
 scientific merit of the case study.
+
+>
 

### 2018-11-05 review/response.md
responses

@@
 
->
+> Thank you very much for this endorsement of the motivation for our paper.
+> We hope that following responses and the amendments to the paper will allay
+> your concerns about the methodology and application.
 
@@
 
-2. The conclusion did not address most of the important results from the case study and should be expanded.
+2. The conclusion did not address most of the important results from the case 
+study and should be expanded.
+
+> This is a welcome comment, and we have added material in the
+> introduction and conclusion as to the value of new Big Data sources for
+> developing activity-based representations of behaviour and circumstances
+> in the smart city, This chimes with much of our related research, but we
+> write cognisant of the article length restrictions of IJGIS.
 
@@
 
+> This material has now been amended and resequenced.
+
 2. P6 l27, please provide more information regarding how you defined the 
@@
 
+> Thank you. The reference to clustering has been amended. The
+> specifics of the thresholds used have been included, and the techniques
+> justified in more detail.
+
 3. P6 l30-31, in the phone shop example, the authors mentioned that it is 
@@
 
+> Thank you. Clarified in the revised version. It is know with preliminary 
+> investigation that the received signal strength indicator decreases with
+> increasing distance between the mobile device and the base station. This
+> change is also found to be exponential as opposed to linear. With this context
+> When we install a sensor to have a clear view of the pedestrian footpath
+> in front of it, we make sure that all the devices beyond a certain distance
+> will report similarly low signal strength. This threhold is idenfied using
+> the kmeans algorithm to find the natural break in the data. We have found that
+> this successfully isolates a farther stationary source of large number of
+> probe requests such as a phone shop or a restaurant next door, Bustops etc.
+> This doesn't work when the source of error is within the range of the
+> footfall. As shown in our examples 2,4. Here our methodology fails to provide
+> sufficient accuracy.
+
 4. P7 l21, how often does the recycling of sequence number occur, and how will 
@@
 
+> The recycling of the sequence number occurs every 4096 counts. From our
+> experiments we observed that a device staying next to device actively searching
+> recycles the sequence number 1-2 times in a 30 mins interval. On average
+> this is expected to be much less than this. when the device doesn't probe
+> as frequently. We have included more details regarding the cycling of sequence
+> numbers and have mentioned method to overcomet this as mentioned by other
+> reviewer. The change in implementing the method is estimated to be --% which
+> we found to be not significant in our context.
+
 5. P7 Section 3.3, please be more specific about the calibration process and 
 the external source of information to be used here.
+
+> This section has been expanded to include more details. The calibration process
+> is adjusting the future counts based on a sample manual counts done at the
+> location. The external source of information used is the manual counts which is
+> collected outside the sensor
 

### 2018-11-05 review/response.md
responses

@@
 
+> This is a valid concern. Which we tried to clarify in the revised manuscript.
+> Our motivation is not to attempt to track the movements of individuals,
+> although we do attempt to accommodate measurement variations attributable to 
+> micro-site location of the sensors. In related work we have carried out 
+> controlled comparison of sensor measures obtained from adjacent store
+> locations, and examined correspondence with retail unit function. This enables
+> identification of the differences in dwell time between, for example,
+> fast food outlets and clothes shops. This full analysis requires
+> addition of store till data and the analysis is detailed. For this
+> reason we include only a short reference to this extension of our
+> analysis here.
+
 7. P10 l48, how was the adjustment factor determined?
+
+> The adjustment factor is calculated as the ration between the manual counts to
+> the counts calculated from sensor for the sample period. This 
+> is used to adjust future counts at the same locations.
 

### 2018-11-05 review/response.md
responses

@@
 > reason we include only a short reference to this extension of our
-> analysis here.
+> analysis here. We also aim this method to count footfall accurately at 
+> locations rather than the flow of footfall between those locations. But as 
+> demonstarted in
+> Lee J, Nemati S, Silva I, Edwards BA, Butler JP, Malhotra A. Transfer entropy 
+> estimation and directional coupling change detection in biomedical time series.
+> Biomed Eng Online. 2012;11:19. Published 2012 Apr 13.
+> doi:10.1186/1475-925X-11-19
+> and
+> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free 
+> measure of effective connectivity for the neurosciences.
+> J Comput Neurosci. 2010;30(1):45-67.
+> we can infer the relation or flow between the locations using other
+> probabilistic approaches.
 

### 2018-11-05 review/response.md
responses

@@
 
-2. The conclusion did not address most of the important results from the case 
-study and should be expanded.
-
 > This is a welcome comment, and we have added material in the
@@
 > write cognisant of the article length restrictions of IJGIS.
+
+2. The conclusion did not address most of the important results from the case 
+study and should be expanded.
+
+> Yes, we agree, and some wider conclusions about the implications of
+> our research for better understanding the functioning of the smart city
+> have now been added.
 
@@
 
+> Yes, thank you.
+
 4.2 General Comments
@@
 
+> Thank you. Our related work examines daily, weekly and seasonal
+> variation, and wil be the subject of future papers. There was no reason for
+> selecting the time in the pilot study. the location was chosen for it it known
+> to have high number of randomised probe requests and has one of the highest
+> footfall in London. For the main study the counting was done such that each
+> location has atleast 2/3 counts in the same time interval.
+
 2. Will you give more detail information about the data and the parameters for 
@@
 clusters.
+
+> Thank you. This has been corrected in the revised version of the
+> paper.
 
@@
 
+> We have resequenced some of the material in response to Referee 2
+> and, cognisant of over-all length requirements, think that this (rather
+> than a diagram) addresses your concerns.
+
 4. Performance Analysis: How about the performance for collecting data using 

### 2018-11-05 review/response.md
responses

@@
 various clustering algorithms? Including the number of records, the number of 
-clusters.
+clusters. How about the performance for different clustering methods 
+in your study？
 
-> Thank you. This has been corrected in the revised version of the
-> paper.
+> Thank you. This part has been expanded and clarified in the revised manuscript. The
+> performance of the algorithms were tested on the basis of the time taken for
+> processing and the relative improvement in results. We found the k-means gives
+> us the best data for the least processing time. The graph based clustering
+> was done using parameters which were arrived at using trial and error
+> controlling for a known device.
 
@@
 1. Minor formatting issues and typos needs to be fixed.
+Figure 2, no value and units in the x-axis and y-axis.
+Table 1, and Figure 4 are in wrong location.
 
-2. Figure 2, no value and units in the x-axis and y-axis.
+> Thank you. This has been corrected in the revised version of the
 
-3. Table 1, and Figure 4 are in wrong location.
+2. Related work: How about WiFi-based real-time data analytics?
 
-4. Related work: How about WiFi-based real-time data analytics?
+> We have included reference to such work in our extended discussion of
+> the relevance of the work to the smart cities discourse.
 
-5. Methodology: Will you draw a diagram of the Methodology?
+3. Methodology: Will you draw a diagram of the Methodology?
 
@@
 
-7. How about the performance for different clustering methods in your study？
+> This is an interesting point in the context of big data and tools for tackling
+> big data. Though the data set at hand is large enough for desktop computing to
+> be unfeasible, it is neither big nor fast enough for a large scale big data
+> systems. A toolkit with opensource, free text processing tools and relational
+> database management system was enough to handle the compexity, size and speed
+> of the data. The toolkit employed here involves PostgreSQL, Bash, Unix tools
+> for text processing (sed, awk, sort, join), NodeJs and R counting.
+> This was a proven toolkit used in the larger project (smart street sensor)
+> providing near real-time delivery of data.
 
-8. The reference list needs to be written in a standard format. Some reference papers are lacking specific pages .Please use a standard way of reference list.
+5. The reference list needs to be written in a standard format. Some reference 
+papers are lacking specific pages .Please use a standard way of reference list.
+
+> Thank you. This has been corrected in the revised version of the paper.

### 2018-11-14 paper/sections/abstract.tex
wednesday modifications

@@
 With increasing mobile ownership, the Wi-Fi `probe requests' generated by mobile devices can act as a cheap, scalable and real-time source of data for establishing such measures.
-The two major challanges we face in using these probe requests for estimating human activity are, filtering the noise generated by the uncertain field of measurement and clustering anonymised probe requests generated by the same devices together without compromising the privacy of the users.
-In this paper we demonstrate that we can overcome these challanges by using class intervals and a novel graph based technique for filtering and clustering the probe requests which in turn, enables us to reliably measure real-time pedestrian footfall at retail highstreets.
+The two major challenges we face in using these probe requests for estimating human activity are, filtering the noise generated by the uncertain field of measurement and clustering anonymised probe requests generated by the same devices together without compromising the privacy of the users.
+In this paper we demonstrate that we can overcome these challenges by using class intervals and a novel graph based technique for filtering and clustering the probe requests which in turn, enables us to reliably measure real-time pedestrian footfall at retail highstreets.

### 2018-11-14 paper/sections/introduction.tex
wednesday modifications

@@
 These devices are located in shop windows and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
-These are deemed to be consumer data because devices carried by consumers routinely probe for a consumer service, specifically a Wi-Fi connection.
+The data collected from these devices are deemed to be consumer data because devices carried by consumers routinely probe for Wi-Fi connection which is a consumer service.
 Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not internet connectivity is established.

### 2018-11-14 paper/sections/main_study.tex
wednesday modifications

@@
+The main study was designed to validate the results of the pilot study over different locations at different times.
+We start by choosing five different locations across central London, install the sensors and collect data for extended periods of time.
+We also carry out manual counting at these locations across different times of the day. 
+We then apply the methodologies discussed earlier to arrive at estimated pedestrian footfall and compare them with corresponding manual counts.
+We finally evaluate the effectiveness of the processes with the Mean Absolute Percentage Error (MAPE) at these locations and report our findings.
+
 \begin{table}
@@
 \end{figure}
-
-The main study was designed to validate the results of the pilot study over different locations at different times.
-We start by choosing five different locations across central London, install the sensors and collect data for extended periods of time.
-We also carry out manual counting at these locations across different times of the day. 
-We then apply the methodologies discussed earlier to arrive at estimated pedestrian footfall and compare them with corresponding manual counts.
-We finally evaluate the effectiveness of the processes with the Mean Absolute Percentage Error (MAPE) at these locations and report our findings.
 

### 2018-11-14 paper/sections/methodology.tex
Paul's comments included
wednesday modifications

@@
-The primary aim of this research is to enable us to collect a series of probe requests and process them into an usable pedestrian footfall count. 
-We do this by using a Wi-Fi receiver to collect probe requests broadcasted by mobile devices, filtering out the background noise and aggregating them based on the device that generated them.
-We begin by looking at the characteristics of probe requests in detail, device a methodology to collect these probe requests in public areas, examine the systemic biases and uncertainties in the data collection method and device data processing methods to overcome these challanges.
+The primary aim of this research is to enable us to collect a series of probe requests and process them into a usable pedestrian footfall count. 
+We do this by using a Wi-Fi receiver to collect probe requests broadcast by mobile devices, filtering out the background noise and aggregating them based on the device that generated them.
+In this section, we begin by looking at the characteristics of probe requests in detail, device a methodology to collect these probe requests in public areas, examine the systemic biases and uncertainties in the data collection method and device data processing methods to overcome these challenges.
 Finally we compare the processed footfall counts to the ground truth recorded by primary surveys.
 
-Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available access points (AP), quick geo-location by triangulation based known APs, etc.
+Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available APs, quick geolocation by triangulation based known APs, etc.
 These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or model of the devices though there is some variation on the frequency and the information transmitted through them.
@@
 Thus these signals can be used to reliably identify the presence of Wi-Fi enabled mobile devices.
-
 Being a first step of connection initiated by the mobile device, these packets have information regarding the characteristics of the mobile device itself.
@@
 	\item \textbf{Sequence number} of the request for the mobile device to keep track of the responses,
-	\item \textbf{Timestamp} at which the request was received by the AP,
+	\item \textbf{Time stamp} at which the request was received by the AP,
 	\item Total \textbf{length} of the request in number of bits, and 
@@
 The MAC address is the primary unique identifier for the mobile device.
-It has two parts, first part is called an Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device.
-The MAC address can be randomised (hence non unique) and is marked as such.
-Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate unique devices.
+It has two parts, first part is the Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device.
+In modern devices, to protect the users privacy, the MAC address can also be randomised (hence non unique) and is marked as such.
+Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate the number unique devices.
 
-Data collection was done with the help of custom sensors built from modifing the Smart street sensor \citep{sss2016} hardware and updating them with custom software.
-The sensor is essentially a Raspberry Pi connected with Wi-Fi and 3G antennae.
-It keeps the Wi-Fi module in `Monitor' mode and uses the open source software - wireshark cite to passively collect all packets sent to `broadcast', marked with type - `management' and subtype - `probe requests'.
-The MAC address in these probe requests is anonymised using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
-A overall schematic of the data collection process is shown in Figure \ref{datacollection_schematic}.
-The ground truth on number of pedestrian footfall was recorded using a purpose built Android application cite.
+Data collection was done with the help of custom sensors built from modifying the hardware used in Smart street sensors \citep{sss2016} and updating them with custom software.
+The sensor is essentially a Raspberry-Pi device with Wi-Fi and 3G modules.
+It keeps the Wi-Fi module in `Monitor' mode and uses the open source software - Wireshark \citep{wireshark2} to passively collect all packets sent to `broadcast', marked with type as management' and subtype as `probe requests'.
+The MAC address in these probe requests is anonymised at the device level using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
+A overall schematic of the data collection and storage process is shown in Figure \ref{datacollection_schematic}.
+The ground truth on number of pedestrian footfall was recorded using a custom mobile application - Clicker \citep{bala2018clicker}
 
@@
 	\includegraphics[width=\linewidth]{images/datacollection_schematic.jpeg}
-	\caption{Schematic diagram showing the process of collecting probe requests using the sensor} \label{datacollection_schematic}
+	\caption{Schematic diagram showing the process of collecting and storing probe requests using the sensor} \label{datacollection_schematic}
 \end{figure}
 
-The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following major challanges which arise from our collection methodology.
+The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following potential uncertainties which arise from our collection methodology.
 \begin{enumerate}
 	\item \textbf{Background noise} - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest.
-This can lead to a signicant background noise at certain locations.
+This can lead to a significant background noise at certain locations.
 E.g.
@@
 \end{enumerate}
-We propose the following methods to tackle each of these challages.
+We propose the following internal and external validation methods to tackle each of these uncertainties.
 
@@
 	\centering
-	\subfigure[Distribution of signal strengths showing the filtering of background noise]{
+	\subfigure[Distribution of signal strengths (dBm) showing the filtering of background noise]{
 		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_signalstrength.jpeg}}}\hspace{20pt}
@@
 \subsection{Filtering with Signal Strength}
-One of the clues that we can use to estimate the distance between the mobie device and the sensor is the strength of the signal received by the sensor. 
+One of the clues that we can use to estimate the distance between the mobile device and the sensor is the strength of the signal received by the sensor. 
 The obvious approach here is to try and establish a relationship between the signal strength and distance first and use this to filter out the unwanted probe requests.
-This approach has numerous pitfalls and uncertainities since the decay of signal strength with distance is not always constant.
-It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source transponder.
-This severely limits our ability to establishing a simple conversion between reported signal strength and distance.
-There is a need for a method which takes in to account these variables across varous locations.
+This approach was found not to be feasible since the decay of signal strength with distance is not always constant.
+It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source.
+This severely limits our ability in establishing a simple conversion between reported signal strength and distance.
+There is a need for a method which takes in to account these variables across various locations.
 
-We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct break in the number of probe requests reporting signal strength corresponding to that distance.
-For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported singal strength corresponding to the distance between the sensor and the phone shop at any given set of conditions as shown in Figure \ref{methodology_schematic}.
-We could identify these breaks in the data using traditional one diamensional classification algorithms such as Jenks natural breaks, k-means, quantile and hierarchical clustering, etc.
+We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct pattern in the number of probe requests reporting signal strength corresponding to that distance.
+For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported signal strength corresponding to the distance between the sensor and the phone shop irrespective of the local conditions as shown in Figure \ref{methodology_schematic}.
+We could identify these breaks in the data using traditional one dimensional classification algorithms such as `jenks natural breaks', `k-means', `quantile' and `hierarchical clustering', etc.
 Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.
@@
 
-Since our primary unique identifier - MAC adddress, is being anonymised by new devces, we need to find other information present in the probe request for a unique identifier.
+Since our primary unique identifier - MAC address, is being anonymised by new devices, we need to find other information present in the probe request for use as a unique identifier.
 Obvious approach here is to establish a factor of randomisation and adjust the counts for these probe requests based on this factor.
-We found this aproach not feasible, since the proportion of devices which randomise the MAC addresses increases over time.
+We found this approach to be not feasible since the proportion of devices which randomise the MAC addresses increases over time.
 There is also a wide variation in the frequency at which the devices randomise the MAC addresses and the method used for the process.
-This leads us to look for a more generalisable approach which is independent of the device model.
+This lead us to look for a more generalisable approach which is independent of the device model.
 
-From our initial analysis we found that OUI, length of the packet and sequence number of the packet being the most promising information to acheive this.
+From our initial analysis we found that OUI, length of the packet and sequence number of the packet being the most promising information to achieve this.
 First we divide our dataset into sets of probe requests with randomised and non-randomised MAC addresses and keep the MAC address as the unique identifier for the latter set. 
-For randomised ones we futher divide them in to sub categories based on their OUIand length of the packet.
-Since the length tends to stay unique to specific models of devices we are left with the task of identifying the unique mobile devices from within these distinct models.
+For randomised ones we further divide them in to sub categories based on their OUI and length of the packet.
+Since the length tends to stay unique to specific models of devices we are left with the task of identifying the unique mobile devices from within these sets.
 
@@
 	\item A link could exist between nodes with a maximum sequence number difference of $\beta$ - sequence threshold.
-	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links.
+	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links in terms of both time and sequence number.
 \end{itemize}
@@
 This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
-Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the maginitude of error is greatly reduced.
+Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the magnitude of error is greatly reduced.
 
-\subsection{Calibrating with ground truth}
-Since mobile phone ownership is an external uncertainty to our study and could arise from variety of spatio, temporal and demographic factors, we propose to solve this using external source of information.
+\subsection{Calibrating with Ground Truth}
+Since mobile device ownership is an external uncertainty to our study and could arise from variety of spatio - temporal and demographic factors, we propose to solve this using external source of information.
 We hypothesize that an adjustment factor could be arrived at for each location of data collection, comparing the sensor based counts and ground truth and it can be used to adjust the data reliably to reflect the ground truth in absolute numbers for the future.
-This calibration can be carried over periodically and the frequecy of which will improve the quality of the estimation.
+This calibration can be carried over periodically and the frequency to improve the quality of the estimation.

### 2018-11-14 paper/sections/pilot_study.tex
wednesday modifications

@@
 Being located at one of the busiest retail locations in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe requests during the half hour period, and 3,722 people were recorded manually walking on the sidewalk during that time.
+
+As a first step we just aggregated the probe requests by their MAC address for every minute to generate a minute by minute count of the number of people around the sensor assuming each MAC address corresponds to a mobile device and hence a pedestrian.
+We then compare this preliminary `footfall' count to the actual number of pedestrian recorded manually to check for the robustness.
+We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there are no intervals without any footfall.
+We find that the MAPE in the raw counts compared to the ground truth is around 425\%.
+This suggests the presence of large amount of noise in the data which might be generated due to sources of uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need for filtering the data.
+
+We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, hierarchical clusters, fisher and jenks.
+We found that while hierarchical clustering and jenks were resource intensive, k-means gave quickest results with the lowest MAPE closely followed by quantile.
+The cut-off point or threshold for the collected data with which we could classify them as high and low was -71 dBm.
+We then removed all the probe requests which reported `low signal strength' and repeated the same aggregation process as before to produce footfall count. 
+This process resulted in a footfall count with a net MAPE of 30\%.
+Though the results are encouraging we are still not completely confident that our filtering process is indeed removing noise or has any correlation the configuration of sensor or position of the mobile devices.
+These need to be addressed with a larger survey with multiple location of varying orientations.
 
@@
 		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_clustering.jpeg}}}\hspace{20pt}
-	\subfigure[Comparision of sensor based counts with the groud truth.]{
+	\subfigure[Comparison of sensor based counts with the ground truth.]{
 	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_comparison.jpeg}}}
-	\caption{The results of the pilot study demonstrating the validity of the deviced methodology.} \label{pilot_results}
+	\caption{The results of the pilot study demonstrating the validity of the methodology.} \label{pilot_results}
 \end{figure}
-
-As a first step we just aggregated the probe requests by their MAC address for every minute to generate a minute by minute count of the number of people around the sensor assuming each MAC address corresponds to a mobile device and hence a pedestrian.
-We then compare this preliminary `footfall' count to the actual number of pedestrian recorded manually to check for the robustness.
-We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there is no `zero records' issue.
-We find that the MAPE in the raw counts compared to the ground truth is around 425\%.
-This suggests the presence of large amount of noise in the data which might be generated due to sources of uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need for filtering the data.
-
-We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, heirarchical clusters, fisher and jeanks.
-We found that while heirarchical clustering and jenks were resource intensive, kmeans gave quickest results with the lowest MAPE closely followed by quantile.
-The cut-off point or threshold for the collected data with which we could classify them as high and low was -71 dBm.
-We then removed all the probe requests which reported `low signal strength' and repeated the same aggregation process as before to produce footfall count. 
-This process resulted in a footfall count with a net MAPE of 30\%.
-Though the results are encouraging we are still not completely confident that our filtering process is indeed removing noise or has any correlation the configuration of sensor or position of the mobile devices.
-These need to be addressed with a larger survey with multiple location of varying orientations.
 

### 2018-11-14 paper/sections/previous_work.tex
wednesday modifications

@@
 
-Though Wi-Fi is a `location-less' technology, there are reliable methods to triangulate the location of Wi-Fi enabled mobile devices from the locations of known access points(AP) and the signal strength reported by them \citep{he2003range, moore2004robust, lamarca2005place}.
+Though, unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly inferred from Wi-Fi, there are reliable methods to triangulate the location of mobile devices from the locations of known access points(AP) and the signal strength reported by them \citep{he2003range, moore2004robust, lamarca2005place}.
 This can overcome the usual shortcoming of GPS, which struggles for precision and accuracy in indoor and densely built environments \citep{zarim2006,kawaguchi2009wifi, xi2010locating}. 
@@
 A combination of these methodologies has been proven to produce de-anonymised globally unique device information \citep{vanhoef2016, martin2017}.
-These approaches usually result in serious risk of infringement of the privacy of the users of the mobile devices by revealing their identifiable personal information.
+These approaches usually result in serious risk of breach of privacy of the users of the mobile devices by revealing their identifiable personal information.
 
@@
 With enough infrastructure to collect such information we can even aim to generate a real-time census of the city \citep{konto2017}.
-With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without infringing the privacy of the users involved.
+With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without risking the breach of privacy of the users involved.

### 2018-11-14 review/response.md
wednesday modifications
Wed 14 Nov 13:44:49 GMT 2018

@@
 
-> Thank you. Our related work examines daily, weekly and seasonal
-> variation, and wil be the subject of future papers. There was no reason for
+> Thank you. Our related work (cited below) examines daily, weekly and seasonal
+> variation, and wil be the subject of future papers.
+
+> Lugomer, K., & Longley, P. (2018). Towards a Comprehensive Temporal 
+> Classification of Footfall Patterns in the Cities of Great Britain (Short
+> Paper). In LIPIcs-Leibniz International Proceedings in Informatics (Vol. 114).
+> Schloss Dagstuhl-Leibniz-Zentrum fuer Informatik.
+
+> There was no reason for
 > selecting the time in the pilot study. the location was chosen for it it known
@@
 
-> Thank you. This has been corrected in the revised version of the
+> Thank you. This has been corrected in the revised version of the manuscript.
+> Figure 2 is just a schematic showing the idea of cut off and clustering hence
+> there are now values on the axes. We have included the units in the description
 

### 2018-11-14 review/response.md
Wed 14 Nov 16:20:32 GMT 2018

@@
 > demonstarted in
+
 > Lee J, Nemati S, Silva I, Edwards BA, Butler JP, Malhotra A. Transfer entropy 
@@
 > doi:10.1186/1475-925X-11-19
+
 > and
+
 > Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free 
@@
 > J Comput Neurosci. 2010;30(1):45-67.
+
 > we can infer the relation or flow between the locations using other
@@
 
-2. Will you give more detail information about the data and the parameters for 
+2. Will you give more detailed information about the data and the parameters for 
 various clustering algorithms? Including the number of records, the number of 

### 2018-11-18 analysis/singal_strength.r
signal strength comparison

@@ -1,0 +1,69 @@
+rm(list = ls())
+
+library(tidyverse)
+library(classInt)
+library(ggplot2)
+library(igraph)
+library(lubridate)
+
+signal_comparision <- function(method,records) {
+	print(method)
+	t1 <- Sys.time()
+	sensor <- "../data/final_sensor_data.csv" %>% 
+		read.csv(stringsAsFactors = FALSE) %>% as_tibble() %>%
+		select(time, signal, mac, oui, vendor, type, sequence) %>%
+		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS")) %>%
+		arrange(time) %>% head(records)
+	llogg <- capture.output({threshold <- classIntervals(sensor$signal, 2, method)$brks[2]})
+	print(threshold)
+	add_signature <- function(data, ds, dt) {
+		vertices <- data %>%
+			mutate(x = as.integer(time), id = as.numeric(rownames(data))) %>%
+			select(id, x, y=sequence) %>%
+			arrange(x)
+		edges <- vertices %>%
+			pull(id) %>% expand.grid(.,.) %>% rename(from=Var1,to=Var2) %>%
+			mutate(from=as.numeric(from), to=as.numeric(to)) %>%
+			filter(from<to) %>%
+			left_join(vertices, by = c("from" = "id")) %>%
+			left_join(vertices, by = c("to" = "id")) %>%
+			mutate(sdis = sqrt((y.y-y.x)**2)) %>%
+			mutate(tdis = sqrt((x.y-x.x)**2)) %>%
+			filter(x.y >= x.x, y.y > y.x, tdis < dt, sdis < ds) %>%
+			group_by(from) %>% summarise(to = to[which(tdis == min(tdis))][[1]], tdis = min(tdis)) %>% 
+			group_by(to) %>% summarise(from = from[which(tdis == min(tdis))][[1]])
+		graph <- graph_from_data_frame(edges, vertices = vertices)
+		data$device_sign <- paste(data$vendor, clusters(graph)$membership, sep = "-")
+		return(data) }
+	sensor <- sensor  %>% filter(signal > threshold) %>% 
+		inner_join(read.csv("../data/final_vendors_data.csv", stringsAsFactors=FALSE), by = "vendor") %>%
+		filter(is_mobile == TRUE) %>% mutate(is_mobile = NULL)
+	subset_global <- sensor %>% filter(type == "global") %>% mutate(device_sign = mac)
[…]

### 2018-11-18 analysis/singal_strength.r
minor fixes

@@ -6,16 +6,15 @@ library(ggplot2)
 library(igraph)
 library(lubridate)
 
-signal_comparision <- function(method,records) {
+signal_comparison <- function(method) {
 	print(method)
 	t1 <- Sys.time()
 	sensor <- "../data/final_sensor_data.csv" %>% 
 		read.csv(stringsAsFactors = FALSE) %>% as_tibble() %>%
 		select(time, signal, mac, oui, vendor, type, sequence) %>%
-		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS")) %>%
-		arrange(time) %>% head(records)
+		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS"))
 	llogg <- capture.output({threshold <- classIntervals(sensor$signal, 2, method)$brks[2]})
-	print(threshold)
+	print(threshold); time <- as.numeric(Sys.time()-t1); print(time)
 	add_signature <- function(data, ds, dt) {
 		vertices <- data %>%
 			mutate(x = as.integer(time), id = as.numeric(rownames(data))) %>%
@@ -56,14 +55,12 @@ signal_comparision <- function(method,records) {
 		mutate(diff = (footfall.y - footfall.x)/ footfall.x) %>%
 		pull(diff) %>% mean
 	mape <- (mape * 100) %>% round %>% as.character %>% paste("%")
-	time <- as.numeric(Sys.time()-t1)
-	print(mape); print(time); cat("\n\n");
+	print(mape);
 	return(list(manual_counts = manual_counts,
 				sensor_counts = sensor_counts,
 				threshold = threshold,
 				mape = mape,
 				time = time))
 }
-methods <- c( "sd", "equal", "pretty", "quantile", "kmeans", "hclust", "bclust", "fisher")
-for(i in methods) { 
-	data <- signal_comparision(i,30000) }
+# methods <- c( "sd", "equal", "pretty", "quantile", "kmeans", "hclust", "bclust", "fisher","jenks")
+#for(i in methods) { data <- signal_comparision(i) }

### 2018-11-18 paper/sections/pilot_study.tex
Thu 15 Nov 15:14:58 GMT 2018
fixes

@@
 We use the algorithm defined in Chapter \ref{methodology} and assign a unique identifier or signature to each probe request independent of the MAC address. 
-Since there is neither prior research nor documentation on a universal behaviour of phones in randomising their MAC addresses, we use the stationary device - the one used for manual counting as a reference and find out the suitable time threshold, $\alpha$ and threshold for sequence numbers, $\beta$ to be 16s and 60 respectively via trial and error.
+Since there is neither prior research nor documentation on a universal behaviour of phones in randomising their MAC addresses, we use the stationary device - the one used for manual counting as a reference and find out the suitable time threshold, $\alpha$ and threshold for sequence numbers, $\beta$ to be 16 seconds and 60 respectively via trial and error.
 This process is done on top the filtering done based on signal strength and only for the probe requests with randomised MAC addresses.

### 2018-11-18 analysis/signal_strength.r

@@ -6,13 +6,14 @@ library(ggplot2)
 library(igraph)
 library(lubridate)
 
-signal_comparison <- function(method) {
+signal_comparison <- function(method,records) {
 	print(method)
 	t1 <- Sys.time()
 	sensor <- "../data/final_sensor_data.csv" %>% 
 		read.csv(stringsAsFactors = FALSE) %>% as_tibble() %>%
 		select(time, signal, mac, oui, vendor, type, sequence) %>%
-		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS"))
+		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS")) %>%
+		arrange(time) %>% head(records)
 	llogg <- capture.output({threshold <- classIntervals(sensor$signal, 2, method)$brks[2]})
 	print(threshold); time <- as.numeric(Sys.time()-t1); print(time)
 	add_signature <- function(data, ds, dt) {

### 2018-11-20 review/response.md
fixes
Tue 20 Nov 13:38:53 GMT 2018

@@
 
-> This is a correct surmisal of the remit of the paper. We have also added#
+> This is a correct surmisal of the remit of the paper. We have also added
 > material to the introduction and conclusion of the paper to better link
@@
 
-> The adjustment factor is calculated as the ration between the manual counts to
+> The adjustment factor is calculated as the ratio between the manual counts to
 > the counts calculated from sensor for the sample period. This 
-> is used to adjust future counts at the same locations.
+> is used to adjust future counts at the same locations. We have included
+> the clarification on the revised manuscript.
 
@@
 
-1.
-This paper presents Wi-Fi based footfall counting methodology, which is timely and interesting topic.
-The method can be performed to estimate human activity such as pedestrian footfall from Wi-Fi probe requests.
+1. This paper presents Wi-Fi based footfall counting methodology, which is 
+timely and interesting topic. The method can be performed to estimate human
+activity such as pedestrian footfall from Wi-Fi probe requests.
 
-> Yes, thank you.
+> Yes, Thank you for your encouraging comment.
 
@@
 
-> Thank you. This part has been expanded and clarified in the revised manuscript. The
-> performance of the algorithms were tested on the basis of the time taken for
-> processing and the relative improvement in results. We found the k-means gives
-> us the best data for the least processing time. The graph based clustering
-> was done using parameters which were arrived at using trial and error
-> controlling for a known device.
+> Thank you. This part has been expanded and clarified in the revised manuscript. 
+> The performance of the algorithms used for one dimensional classification were 
+> evaluated on the basis of the time taken for processing and the resulting 
+> mean absolute percentage error. We found that the k-means gives
+> us the best data for the quickest processing time. The graph based clustering
+> was done using two parameters - sequence threshold of 60 and time threshold of
+> 16 seconds. These values were arrived arrived at using trial and error by using
+> a know device as the control mechanism. This has been illustrated further in 
+> the revised manuscript.
 
@@
 > This is an interesting point in the context of big data and tools for tackling
-> big data. Though the data set at hand is large enough for desktop computing to
-> be unfeasible, it is neither big nor fast enough for a large scale big data
-> systems. A toolkit with opensource, free text processing tools and relational
-> database management system was enough to handle the compexity, size and speed
-> of the data. The toolkit employed here involves PostgreSQL, Bash, Unix tools
-> for text processing (sed, awk, sort, join), NodeJs and R counting.
-> This was a proven toolkit used in the larger project (smart street sensor)
-> providing near real-time delivery of data.
+> big data. Though the data set here is large enough to make desktop computing
+> unfeasible, it is neither big nor fast enough for a large scale big data
+> system. An opensource and free toolkit comprising of text processing tools 
+> and relational database management system was enough 
+> to handle the compexity, size and speed of the data.
+> The toolkit employed here involves PostgreSQL, Bash, Unix tools
+> for text processing (sed, awk, sort, join), Node.js and R.
+> The same toolkit proved to be useful in the larger project 
+> (smart street sensor) providing near real-time delivery of data.
 

### 2018-11-20 paper/sections/conclusion.tex
Tue 20 Nov 13:38:53 GMT 2018

--- 2018-06-30 paper/sections/conclusion.tex
@@
-This paper has made a novel contribution to the measurement of activities in retail centres in real time.
-To this end, it has described the collection and processing of a novel consumer Big Dataset that enables valid measures of levels of footfall activity which has been scaled across a Great Britain wide network of sensors.
-In both conceptual and technical terms, it illustrates the ways in which passively collected consumer data can be ‘hardened’ to render them robust and reliable by using related procedures of internal and external validation.
+Sentient technologies make possible measurement of the human activities that are
+the life blood of the smart city. Yet the data that they harvest are frequently
+relevant only to sub-groups within society that avail themselves of particular
+goods and services – such as social media applications, transport modes or
+retail offers. In each of these cases, it is necessary to remember that the
+resulting data are a by-products of consumer transactions, and will as a
+consequence, only pertain to users of the relevant good or service. If the smart
+city is to be socially inclusive, it therefore follows that sentient data must
+represent entire populations, whether by design or by triangulation with
+external, population wide, sources. This usually poses a non-trivial task, since
+the ebbs and flows of smart device-enabled citizens rarely pertain to any
+clearly defined population in either administrative or functional terms Massam
+1973.
 
-Internal validation addresses the issues of screening out device probes that do not indicate footfall, and the further screening of device probes to ‘fingerprint’ the effects of MAC randomization.It is important to note that the filtering process work based soley on the information present in the probe requests and their temporal distribution.
-This ensures that although the mobile devices were uniquely identified, there was no further personal data generated by linking the probe requests to the users of the mobile devices.
-This method essentially gave us a way to estimate the footfall in real-time without identifying or tracking the mobile devices themselves. 
-External validation then entails reconciling adjusted counts with the footfall observed at sample locations.
-This procedure makes it possible to generalise from locations at which manual footfall surveys are conducted to all others in the system, and to develop a classification of device locations that are more or less susceptible to noise generation.
+Our objective here has been to collect, rather than re-use, data on smart city
+functioning, by recording Wi-Fi probes and ultimately reconciling them with
+manual counts in order to infer ambient populations. The internal validation
+methodology set out in the technical sections of this paper, allied to external
+validation from pedestrian counts, renders the method inclusive and robust when
+recording activity levels in retail centres in real time. We have described the
+collection and processing of a novel consumer Big Dataset that enables valid
+measures of levels of footfall activity which has been scaled across a Great
+Britain wide network of sensors CDRC 2016.  In both conceptual and technical
+terms, it illustrates the ways in which passively collected consumer data can be
+‘hardened’ to render them robust and reliable by using related procedures of
+internal and external validation.
 
-This Wi-Fi based footfall counting methodology offers a large number of applications and benefits for real time spatial analysis.
-Since Wi-Fi based sensors are inexpensive and the data model is scalable, it is possible to use this methodology for a large network of sensors to gather granular data on pedestrian footfall.
-A snapshot showing week's worth of precise footfall in area around Charring cross, London is shown in Figure \ref{} to demonstrate the potential for such dataset.
-Projects such as SmartStreetSensors \citep{sss2016}, may utilise this methodology to overcome the challenges introduced by the implementation of MAC address randomisation.
-Such precise and granular data also enables us to confidently model the pedestrian flow in urban road networks, and will be an indispensable tool in the smart city framework.
-It can also be used to understand and classify geographical areas based on the spatio-temporal distribution of the volume of activity in them which we intend to research in future.
+Internal validation addresses the issues of screening out device probes that do
+not indicate footfall, and the further screening of device probes to
+‘fingerprint’ the effects of MAC randomization. It is important to note that the
+filtering process work based solely on the information present in the probe
+requests and their temporal distribution. This ensures that although the mobile
+devices were uniquely identified, there was no further personal data generated
+by linking the probe requests to the users of the mobile devices. This method
+essentially gave us a way to estimate the footfall in real-time without
+identifying or tracking the mobile devices themselves. External validation then
+entails reconciling adjusted counts with the footfall observed at sample
+locations. This procedure makes it possible to generalise from locations at
+which manual footfall surveys are conducted to all others in the system, and to
+develop a classification of device locations that are more or less susceptible
+to noise generation.
+
+This Wi-Fi based footfall counting methodology offers a large number of
+applications and benefits for real time spatial analysis. Since Wi-Fi based
+sensors are inexpensive and the data model is scalable, it is possible to use
+this methodology for a large network of sensors to gather granular data on
+pedestrian footfall. A snapshot showing week’s worth of precise footfall in area
+around Charring cross, London is shown in Figure 7 to demonstrate the potential
+for such dataset. Projects such as SmartStreetSensors CDRC 2016, may utilise
+this methodology to overcome the challenges introduced by the implementation of
+MAC address randomisation.
+
+The vicissitudes of MAC randomisation, and the provisions of privacy legislation
+such as EU General Data Protection Regulations mitigate against tracking
+individuals across the smart city using this approach.  This can be modelled
+using agent-based methods Heppenstall et al 2012, however. In our own research
+we have also begun to link store time-lagged till receipts to footfall, and have
+used such data to better understand the dwell times that characterise such
+different retail uses as stores with window displays and fast food restaurants.
+Such analysis not only provides a more nuanced picture of movement through
+retail areas, but also enables valorisation of micro sites within retail
+centres. In the UK, for example, this is of immediate practical importance in
+evaluating business rates on properties, and has still wider implications for
+the setting of retail unit rental values. There are obvious extensions to
+understanding the ebbs and flows of activities in the 24 hour smart city.
+
+More broadly still, extensions to this strand of smart city research are likely
+to seek to differentiate the quality of different elements within footfall
+according to mission e.g. travel to adjacent workplace zones, leisure, etc and
+personal characteristics such as spending power. In this respect, future
+research may not only simulate linkage of harmonised footfall counts between
+sensor locations, but also link these in turn to disaggregate origin-destination
+matrices for bikeshare and other public transport modes.  Our own investigations
+will consider these and other challenges to understanding the functioning of the
+sentient city.

### 2018-11-20 paper/sections/introduction.tex
Tue 20 Nov 13:38:53 GMT 2018

@@
-New and developing technologies today provide the infrastructure over which movements and interactions of people can be measured and monitored in the `sentient city' \citep{amin2017seeing}.
-These include mobile phone networks, which can triangulate user locations relative to networks of masts, use of GPS to locate users of social media services, and Wi-Fi beacon connectivity to access the Internet.
-These technologies offer differing levels of spatial precision, where mobile telephony and Wi-Fi generally being less reliable and offer lower precision than GPS to the end users, while being more advantageous for broader mobility studies \citep{pinelli2015comparing}.
+New and developing ‘smart’ technologies today provide the infrastructure over
+which movements and interactions of people can be measured and monitored in the
+‘sentient city’ (Amin and Thrift 2017). This is making it possible to
+reinvigorate conceptualise the city as the locus of human activities
+supplementing night time geographies of residence (Martin et al 2015) with
+geographies of shopping behaviour (Lloyd and Cheshire 2018*), workzone
+geographies (Singleton and Longley 2018) and studies of movement trajectories
+(Campbell et al 2008). This is rendering activity-based conceptions of human
+behaviour central to analysis of hardship and opportunity in and around the
+smart city (Venerandi et al 2015).
 
-There has been considerable research into the usefulness of these technologies to understand patterns of movement in cities in near real time \citep{candia2008uncovering, gonzalez2008understanding, calabrese2013understanding}.
-Most of this research has focused upon technical specification of accuracy or precision \citep{song2010limits, lane2010survey}, with somewhat less attention devoted to the ways that the characteristics of the technologies and of their human users conspire to create possible bias in the inclusiveness of usage patterns.
-This is perhaps most manifest in the plethora of papers that analyse the distribution and content of publicly available Twitter social media feeds, which likely constitute an extremely biased and self-selecting subset of users that geo-enable their posts, drawn from a user base that itself bears no identified correspondence to any known background population \citep{lansley2016deriving}.
-Even analysis of mobile phone data, usually derived from industry players that have significant market share and user bases representative of local populations, may exclude groups such as tourists from distant origins or subscribers to third party services that share distinctive characteristics \citep{di2016mind}
+Sentient technologies include mobile phone networks, which can triangulate user
+locations relative to networks of masts, use of GPS to locate users of social
+media services, and Wi-Fi beacon connectivity to access the Internet. These
+technologies offer differing levels of spatial precision, where mobile telephony
+and Wi-Fi generally being less reliable and offer lower precision than GPS to
+the end users, while being more advantageous for broader mobility studies
+(Pinelli et al.  2015). There has been considerable research into the usefulness
+of these technologies to understand patterns of movement in cities in near real
+time (Candia et al. 2008, Gonzalez et al. 2008, Calabrese et al. 2013).  Most of
+this research has focused upon technical specification of accuracy or precision
+(Song et al. 2010, Lane et al. 2010), with somewhat less attention devoted to
+the ways that the characteristics of the technologies and of their human users
+conspire to create possible bias in representing usage patterns across the
+entire smart city.
 
-These examples illustrate the issues that underpin the assembly and analysis of consumer data, which can be considered as a distinctive class of Big Data that arise from the interactions between humans and customer-facing organisations such as retailers, domestic energy suppliers, transport providers and suppliers of social media and communications \citep{cdrc2018consumer}.
-Consumer data account for an ever-increasing real share of all of the data that are collected of citizens, but a fundamental characteristic of consumer-led markets is that no single provider has a monopoly in market provision, and therefore issues of market share and segment generate bias in analysis.
-The source and operation of this bias is unknown in the absence of extensive and context sensitive attempts to triangulate consumer data with data of known provenance relating to clearly defined populations \citep{lansley2016deriving}.
-In similar ways to other classes of Big Data, consumer data are best thought of as digital ‘exhaust’, or a by-product created by or harvested from consumer transactions.
+Even analysis of mobile phone data, usually derived from industry players that
+have significant market share and user bases representative of local
+populations, may exclude groups such as tourists from distant origins or
+subscribers to third party services that share distinctive characteristics
+\citep{di2016mind}
 
-In this paper, through a set of experiments, we evaluate the value of data collected from a Great Britain wide network of 800 devices \citep{sss2016}, installed in order to characterize the footfall patterns of a scientifically balanced sample of retail centres.
-These devices are located in shop windows and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
-The data collected from these devices are deemed to be consumer data because devices carried by consumers routinely probe for Wi-Fi connection which is a consumer service.
-Monitoring the probes from such devices provides an indication of the presence of their users, regardless of whether or not internet connectivity is established.
-Our core motivation is to appraise the usefulness of Wi-Fi probe requests harvested from our network of sensors in order to indicate levels of pedestrian activity.
-More broadly still, in our future research we intend to classify the nationwide network of footfall profiles as part of a programme of research to understand the form and functioning of retail areas at a time of far-reaching structural change for the retail industry.
+These examples illustrate the issues that underpin the assembly and analysis of
+consumer data, which can be considered as a distinctive class of Big Data that
+arise from the interactions between humans and customer-facing organisations
+such as retailers, domestic energy suppliers, transport providers and suppliers
+of social media and communications \citep{cdrc2018consumer}.  Consumer data
+account for an ever-increasing real share of all of the data that are collected
+of citizens, but a fundamental characteristic of consumer-led markets is that no
+single provider has a monopoly in market provision, and therefore issues of
+market share and segment generate bias in analysis.  The source and operation of
+this bias is unknown in the absence of extensive and context sensitive attempts
+to triangulate consumer data with data of known provenance relating to clearly
+defined populations \citep{lansley2016deriving}.  In similar ways to other
+classes of Big Data, consumer data are best thought of as digital ‘exhaust’, or
+a by-product created by or harvested from consumer transactions.
 
-To this end, it is important to first undertake a thorough conceptual and technical appraisal of our consumer data source.
-In technical terms, screening the information present in the `probes requests' and classifying them based on their characteristics is essential in order to remove the ones emitted by devices that do not indicate pedestrian activity, such as network enabled printers and other fixed devices.
-Related to this, a method to fingerprint Wi-Fi probes is necessary to remove probes from individuals’ devices that in conceptual terms should not be considered part of footfall – as when, for example, an employees is seated in an office within range of the sensor device.
-A calibration of sensor measurement is also essential on two grounds: first, individuals may carry multiple devices, or no device at all; and second, the positioning and orientation of the sensor in the retail unit may lead to systematic over- or under-enumeration.
-These sources of bias in measurement must be accommodated by manual recording of footfall at each location and generalization of these sample survey results to all locations and time periods.
-As we describe in detail below, manual validation of the data needs to be undertaken in parallel with technical profiling of the mix of consumer mobile devices that probe our sensors, since the effectiveness of data cleaning procedures discussed in this paper differ between individual locations and configurations.
+In this paper, through a set of experiments, we evaluate the value of data
+collected from a Great Britain wide network of 800 devices \citep{sss2016}
+installed in order to characterize the footfall patterns of a scientifically
+balanced sample of retail centres.  These devices are located in shop windows
+and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
+The data collected from these devices are deemed to be consumer data because
+devices carried by consumers routinely probe for Wi-Fi connection which is a
+consumer service.  Monitoring the probes from such devices provides an
+indication of the presence of their users, regardless of whether or not internet
+connectivity is established.  Our core motivation is to appraise the usefulness
+of Wi-Fi probe requests harvested from our network of sensors in order to
+indicate levels of pedestrian activity.  More broadly still, in our future
+research we intend to classify the nationwide network of footfall profiles as
+part of a programme of research to understand the form and functioning of retail
+areas at a time of far-reaching structural change for the retail industry.
+
+To this end, it is important to first undertake a thorough conceptual and
+technical appraisal of our consumer data source.  In technical terms, screening
+the information present in the `probes requests' and classifying them based on
+their characteristics is essential in order to remove the ones emitted by
+devices that do not indicate pedestrian activity, such as network enabled
+printers and other fixed devices.  Related to this, a method to fingerprint
+Wi-Fi probes is necessary to remove probes from individuals’ devices that in
+conceptual terms should not be considered part of footfall – as when, for
+example, an employees is seated in an office within range of the sensor device.
+A calibration of sensor measurement is also essential on two grounds: first
+individuals may carry multiple devices, or no device at all; and second, the
+positioning and orientation of the sensor in the retail unit may lead to
+systematic over- or under-enumeration.  These sources of bias in measurement
+must be accommodated by manual recording of footfall at each location and
+generalization of these sample survey results to all locations and time periods.
+As we describe in detail below, manual validation of the data needs to be
+undertaken in parallel with technical profiling of the mix of consumer mobile
+devices that probe our sensors, since the effectiveness of data cleaning
+procedures discussed in this paper differ between individual locations and
+configurations.

### 2018-11-20 paper/sections/main_study.tex
Tue 20 Nov 13:38:53 GMT 2018

@@
-The main study was designed to validate the results of the pilot study over different locations at different times.
-We start by choosing five different locations across central London, install the sensors and collect data for extended periods of time.
-We also carry out manual counting at these locations across different times of the day. 
-We then apply the methodologies discussed earlier to arrive at estimated pedestrian footfall and compare them with corresponding manual counts.
-We finally evaluate the effectiveness of the processes with the Mean Absolute Percentage Error (MAPE) at these locations and report our findings.
+The methodology set out above is next implemented in five different Central
+London locations at different times. Sensors are installed and data collected
+for extended periods of time. We also carry out manual counting at these
+locations across different times of the day. We then apply the methodologies
+discussed earlier to arrive at estimated pedestrian footfall and compare them
+with corresponding manual counts.  We finally evaluate the effectiveness of the
+processes with the Mean Absolute Percentage Error (MAPE) at these locations and
+report our findings.
 
@@
 
-The locations at which the data were collected are shown in Table \ref{locations-table}.
-The locations are chosen for their diverse site conditions and unique sources of noise around the potential location of the sensors.
-The position of the sensor at these locations with respect to the context is shown the Figure \ref{main_signal_strength}.
-We can see that Location 5 is the `cleanest' with one clear stationary source of noise (phone shop) of while location 2 is the one with the most complexity with seating area all around the sensor.
-The sensors were operational through out February and March while manual counts were conducted in these location in half hour sessions on at least two different days.
-For the purposes of comparing with ground truth we just considered the data from sensors which corresponding to the 12 sets of available manual counts.
-The schedule of data collection is shown in Figure \ref{main_schedule}
+The locations at which the data were collected are shown in Table
+\ref{locations-table}. The locations are chosen for their diverse site
+conditions and unique sources of noise around the potential location of the
+sensors. The position of the sensor at these locations with respect to the
+context is shown the Figure \ref{main_signal_strength}. We can see that Location
+5 is the `cleanest' with one clear stationary source of noise (phone shop) of
+while location 2 is the one with the most complexity with seating area all
+around the sensor.  The sensors were operational through out February and March
+while manual counts were conducted in these location in half hour sessions on at
+least two different days. For the purposes of comparing with ground truth we
+just considered the data from sensors which corresponding to the 12 sets of
+available manual counts. The schedule of data collection is shown in Figure
+\ref{main_schedule}
 
@@
 
-We start by looking at the distribution of the signal strength reported by the probe requests across the locations.
-From the density plot shown in Figure \ref{main_signal_strength} we can see that there is clear relation between the distribution of the signal strength and the distance and complexity of the source of noise.
-We can see that while location 5 shows clean difference between low and high signal strengths, location 2 is almost normally distributed.
-Intuitively we expect that location 2 and 4 must be harder to classify than location 1 3 and 5. 
-We run the k-means classification algorithm and filter out the probe requests which are randomised and have signal strengths less than the second break (threshold). 
-It is important to note that we are dealing with relative thresholds of signal strengths which can vary with location and time of the analysis.
-We then aggregate then probe requests for every minute by counting the number of Unique MAC addresses present in every minute.
-We also remove devices that dwell around the sensor by removing the MAC addresses which reappear from within the past hour.
-The results of the first stage of filtering process along with the thresholds are shown in Table \ref{errors_table}.
-Confirming our intuition, we see that the location 2 has the most MAPE followed by location 4 while rest of them have highly reduced MAPE.
-It is significant that this method alone reduces our margin of error by a 50 - 100\% from the raw counts without any cleaning.
-This makes the signal strength filtering quick and ideal method for practical applications which doesn't require absolute numbers such as creating large aggregated indexes to show long term trends. 
-We also see that the success of the signal strength filtering can be improved significantly by installing sensors such that the pedestrians and noise are at different distances from the sensor and are the field of measurement is distinct from the surroundings however noisy it might be.
+We start by looking at the distribution of the signal strength reported by the
+probe requests across the locations. From the density plot shown in Figure
+\ref{main_signal_strength} we can see that there is clear relation between the
+distribution of the signal strength and the distance and complexity of the
+source of noise. We can see that while location 5 shows clean difference between
+low and high signal strengths, location 2 is almost normally distributed.
+Intuitively we expect that location 2 and 4 must be harder to classify than
+location 1 3 and 5.  We run the k-means clustering algorithm and filter out the
+probe requests which are randomised and have signal strengths less than the
+second break (threshold).  It is important to note that we are dealing with
+relative thresholds of signal strengths which can vary with location and time of
+the analysis.  We then aggregate then probe requests for every minute by
+counting the number of Unique MAC addresses present in every minute. We also
+remove devices that dwell around the sensor by removing the MAC addresses which
+reappear from within the past hour.  The results of the first stage of filtering
+process along with the thresholds are shown in Table \ref{errors_table}.
+Confirming our intuition, we see that the location 2 has the most MAPE followed
+by location 4 while rest of them have highly reduced MAPE.  It is significant
+that this method alone reduces our margin of error by a 50 - 100\% from the raw
+counts without any cleaning.  This makes the signal strength filtering quick and
+ideal method for practical applications which doesn't require absolute numbers
+such as creating large aggregated indexes to show long term trends.  We also see
+that the success of the signal strength filtering can be improved significantly
+by installing sensors such that the pedestrians and noise are at different
+distances from the sensor and are the field of measurement is distinct from the
+surroundings however noisy it might be.
 
-We then run the sequence numbers based clustering process on the rest of the probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except for location 3.
-Location 3 is an outlier among all the other sensors since it is the only one with large amount of pedestrians very close to the sensor.
-This may be causing the over filtering caused by the previous process.
-We finally run the calibration process where we calculate the adjustment factors shown in Table \ref{errors_table}.
-We the use them to adjust the counts to achieve a MAPE ranging from 10 - 50\%.
-We can see that the sensors with people moving right next to them tend to under-count with our methodology while sensors with seating next to them tend to over-count significantly.
-However, using the filtering process, we can reduce the error to almost 10\% closer to that of the ground truth.
+We then run the sequence numbers based clustering process on the rest of the
+probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except
+for location 3.  Location 3 is an outlier among all the other sensors since it
+is the only one with large amount of pedestrians very close to the sensor.  This
+may be causing the over filtering caused by the previous process.  We finally
+run the calibration process where we calculate the adjustment factors from the
+ratio between the manual counts to the counts calculated for the sample period
+as shown in Table \ref{errors_table}.  We the use them to adjust the counts to
+achieve a MAPE ranging from 10 - 50\%. We can see that the sensors with people
+moving right next to them tend to under-count with our methodology while sensors
+with seating next to them tend to over-count significantly. However, using the
+filtering process, we can reduce the error to almost 10\% closer to that of the
+ground truth.
 

### 2018-11-20 paper/sections/methodology.tex
Tue 20 Nov 13:38:53 GMT 2018

@@
-The primary aim of this research is to enable us to collect a series of probe requests and process them into a usable pedestrian footfall count. 
-We do this by using a Wi-Fi receiver to collect probe requests broadcast by mobile devices, filtering out the background noise and aggregating them based on the device that generated them.
-In this section, we begin by looking at the characteristics of probe requests in detail, device a methodology to collect these probe requests in public areas, examine the systemic biases and uncertainties in the data collection method and device data processing methods to overcome these challenges.
-Finally we compare the processed footfall counts to the ground truth recorded by primary surveys.
+The primary aim of this research is to enable us to collect a series of probe
+requests and process them into a usable pedestrian footfall count. We do this by
+using a Wi-Fi receiver to collect probe requests broadcast by mobile devices,
+filtering out the background noise and aggregating them based on the device that
+generated them. In this section, we begin by looking at the characteristics of
+probe requests in detail, device a methodology to collect these probe requests
+in public areas, examine the systemic biases and uncertainties in the data
+collection method and device data processing methods to overcome these
+challenges. Finally we compare the processed footfall counts to the ground truth
+recorded by primary surveys.
 
-Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available APs, quick geolocation by triangulation based known APs, etc.
-These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or model of the devices though there is some variation on the frequency and the information transmitted through them.
-In some cases, such as Android devices, these are broadcast even when the Wi-Fi functionality has been turned off by the user.
-Thus these signals can be used to reliably identify the presence of Wi-Fi enabled mobile devices.
-Being a first step of connection initiated by the mobile device, these packets have information regarding the characteristics of the mobile device itself.
-Some of the key information we can infer from these requests are,
+Probe requests are a special type of management packets broadcast by Wi-Fi
+enabled devices as part of the various functions such as scanning for available
+APs, quick geolocation by triangulation based known APs, etc. These are
+broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
+model of the devices though there is some variation on the frequency and the
+information transmitted through them. In some cases, such as Android devices,
+these are broadcast even when the Wi-Fi functionality has been turned off by the
+user. Thus these signals can be used to reliably identify the presence of Wi-Fi
+enabled mobile devices. Being a first step of connection initiated by the mobile
+device, these packets have information regarding the characteristics of the
+mobile device itself. Some of the key information we can infer from these
+requests are,
+
 \begin{enumerate} 
-	\item \textbf{Media Access Control (MAC) address} which is an unique identifier for the wireless hardware of the mobile device,
-	\item \textbf{Sequence number} of the request for the mobile device to keep track of the responses,
+\item \textbf{Media Access Control (MAC) address} which is an unique identifier
+	for the wireless hardware of the mobile device, 
+\item \textbf{Sequence number} of the request for the mobile device to keep
+	track of the responses, 
 \item \textbf{Time stamp} at which the request was received by the AP, 
@@
 \end{enumerate}
-The MAC address is the primary unique identifier for the mobile device.
-It has two parts, first part is the Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device.
-In modern devices, to protect the users privacy, the MAC address can also be randomised (hence non unique) and is marked as such.
-Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate the number unique devices.
 
-Data collection was done with the help of custom sensors built from modifying the hardware used in Smart street sensors \citep{sss2016} and updating them with custom software.
-The sensor is essentially a Raspberry-Pi device with Wi-Fi and 3G modules.
-It keeps the Wi-Fi module in `Monitor' mode and uses the open source software - Wireshark \citep{wireshark2} to passively collect all packets sent to `broadcast', marked with type as management' and subtype as `probe requests'.
-The MAC address in these probe requests is anonymised at the device level using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
-A overall schematic of the data collection and storage process is shown in Figure \ref{datacollection_schematic}.
-The ground truth on number of pedestrian footfall was recorded using a custom mobile application - Clicker \citep{bala2018clicker}
+The MAC address is the primary unique identifier for the mobile device. It has
+two parts, first part is the Organisation Unique Identifier (OUI) which gives
+information about the manufacturer of the device and the second is unique to the
+device. In modern devices, to protect the users privacy, the MAC address can
+also be randomised (hence non unique) and is marked as such. Though sequence
+number and length of the packet are not strictly unique, we hypothesize that we
+can use them to estimate the number unique devices as demonstrated by
+\citep{vanhoef2016}.
+
+Data collection was done with the help of custom sensors built from modifying
+the hardware used in Smart street sensors \citep{sss2016} and updating them with
+custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
+3G modules. It keeps the Wi-Fi module in `Monitor' mode and uses the open source
+software - Wireshark \citep{wireshark2} to passively collect all packets sent to
+`broadcast', marked with type as management' and subtype as `probe requests'.
+The MAC address in these probe requests is anonymised at the device level using
+a cryptographic hashing algorithm and transmitted through 3G connection to a
+central database via web-sockets protocol, where it is stored in a PostgreSQL
+database for further analysis. A overall schematic of the data collection and
+storage process is shown in Figure \ref{datacollection_schematic}. The ground
+truth on number of pedestrian footfall was recorded using a custom mobile
+application - Clicker \citep{bala2018clicker}
+
+\begin{figure} 
+\centering \includegraphics[width=\linewidth]
+	{images/datacollection_schematic.jpeg}
+\caption 
+	{Schematic diagram showing the process of collecting and storing probe
+	requests using the sensor}
+\label{datacollection_schematic} 
+\end{figure}
+
+The next step after collecting data was to estimate the footfall or pedestrian
+activity from them. We identified the following potential uncertainties which
+arise from our collection methodology.
+
+\begin{enumerate} 
+\item 
+\textbf{Background noise} - since the extent to which Wi-Fi signals travel
+		differs subject to various factors such as interference and humidity, it
+		is close to impossible to restrict our data collection to a finite area
+		of interest. This can lead to a significant background noise at certain
+		locations. E.g. a phone shop or a bus stop located next to the study
+		area can increase the number of probe requests received by the sensor.
+		It is important to note that this method may not work effectively on
+		study locations with complex configurations such as the source of noise
+		and the area of study being located at the same distance from the
+		sensor. This aspect is explored in detail in the broader case study in
+		the following sections.
+\item 
+\textbf{MAC randomisation} - The mobile devices in recent years have been using
+		randomised 'local' MAC addresses for probe requests to protect the users
+		from being tracked. This makes it impossible to tell if the probe
+		requests are being sent by the same mobile device which is being
+		stationed next to the sensor. This along with the previous problem can
+		further increase the magnitude of error by several fold.
+\item
+\textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely
+		across geography and demography, we cannot assume that every mobile
+		device translates to one pedestrian footfall. In addition to this, there
+		is a long term overall increase in mobile ownership which may lead to
+		the number of probe requests collected overtime.
+\end{enumerate}
+
+We propose the following internal and external validation methods to tackle each
+of these uncertainties.
 
@@
 \centering 
-	\includegraphics[width=\linewidth]{images/datacollection_schematic.jpeg}
-	\caption{Schematic diagram showing the process of collecting and storing probe requests using the sensor} \label{datacollection_schematic}
-\end{figure}
-
-The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following potential uncertainties which arise from our collection methodology.
-\begin{enumerate}
-	\item \textbf{Background noise} - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest.
-This can lead to a significant background noise at certain locations.
-E.g.
-a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
-	\item \textbf{MAC randomisation} - The mobile devices in recent years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked.
-This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor.
-This along with the previous problem can further increase the magnitude of error by several fold.
-	\item \textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall.
-In addition to this, there is a long term overall increase in mobile ownership which may lead to the number of probe requests collected overtime.
-\end{enumerate}
-We propose the following internal and external validation methods to tackle each of these uncertainties.
-
-\begin{figure}
-	\centering
-	\subfigure[Distribution of signal strengths (dBm) showing the filtering of background noise]{
-		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_signalstrength.jpeg}}}\hspace{20pt}
-	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
-	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_clustering.jpeg}}}
-	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
+\subfigure [Distribution of signal strengths (dBm) showing the filtering of
+		background noise] {\resizebox*{0.45\linewidth}{!}
+		{\includegraphics{images/methodology_signalstrength.jpeg}}}
+\hspace{20pt}
+\subfigure [Clustering probe requests as nodes in a graph using increasing
+		sequence numbers] {\resizebox*{0.45\linewidth}{!}
+		{\includegraphics{images/methodology_clustering.jpeg}}}
+\caption {Schematic diagrams explaining the methods for filtering by signal
+	strength and clustering using sequence numbers}
+\label{methodology_schematic}
 \end{figure}
@@
 \subsection{Filtering with Signal Strength}
-One of the clues that we can use to estimate the distance between the mobile device and the sensor is the strength of the signal received by the sensor. 
-The obvious approach here is to try and establish a relationship between the signal strength and distance first and use this to filter out the unwanted probe requests.
-This approach was found not to be feasible since the decay of signal strength with distance is not always constant.
-It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source.
-This severely limits our ability in establishing a simple conversion between reported signal strength and distance.
-There is a need for a method which takes in to account these variables across various locations.
 
-We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct pattern in the number of probe requests reporting signal strength corresponding to that distance.
-For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported signal strength corresponding to the distance between the sensor and the phone shop irrespective of the local conditions as shown in Figure \ref{methodology_schematic}.
-We could identify these breaks in the data using traditional one dimensional classification algorithms such as `jenks natural breaks', `k-means', `quantile' and `hierarchical clustering', etc.
-Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.
+One of the clues that we can use to estimate the distance between the mobile
+device and the sensor is the strength of the signal received by the sensor. The
+obvious approach here is to try and establish a relationship between the signal
+strength and distance first and use this to filter out the unwanted probe
+requests. This approach was found not to be feasible since the decay of signal
+strength with distance is not always constant. It varies with atmospheric
+conditions, presence of obstructions between the source and target, the nature
+of these obstructions and the strength (power level) of the source. This
+severely limits our ability in establishing a simple conversion between reported
+signal strength and distance. There is a need for a method which takes in to
+account these variables across various locations.
+
+We hypothesise that in configurations where a specific source of background
+noise is at a constant distance, there must be a distinct pattern in the number
+of probe requests reporting signal strength corresponding to that distance. For
+example, if there is a phone shop next to our sensor where hundreds of phones
+regularly send probe requests there should be a sharp rise of number of probe
+requests with reported signal strength corresponding to the distance between the
+sensor and the phone shop irrespective of the local conditions as shown in
+Figure \ref{methodology_schematic}. We could identify these breaks in the data
+using traditional one dimensional clustering algorithms such as `jenks natural
+breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
+only looking for the break in the data and not for absolute values, the
+methodology should apply for all the variations due micro site conditions thus
+reducing the overall noise in the collected data.
 
@@
 
-Since our primary unique identifier - MAC address, is being anonymised by new devices, we need to find other information present in the probe request for use as a unique identifier.
-Obvious approach here is to establish a factor of randomisation and adjust the counts for these probe requests based on this factor.
-We found this approach to be not feasible since the proportion of devices which randomise the MAC addresses increases over time.
-There is also a wide variation in the frequency at which the devices randomise the MAC addresses and the method used for the process.
-This lead us to look for a more generalisable approach which is independent of the device model.
+Since our primary unique identifier - MAC address, is being anonymised by new
+devices, we need to find other information present in the probe request for use
+as a unique identifier. Obvious approach here is to establish a factor of
+randomisation and adjust the counts for these probe requests based on this
+factor. We found this approach to be not feasible since the proportion of
+devices which randomise the MAC addresses increases over time. There is also a
+wide variation in the frequency at which the devices randomise the MAC addresses
+and the method used for the process. This lead us to look for a more
+generalisable approach which is independent of the device model.
 
-From our initial analysis we found that OUI, length of the packet and sequence number of the packet being the most promising information to achieve this.
-First we divide our dataset into sets of probe requests with randomised and non-randomised MAC addresses and keep the MAC address as the unique identifier for the latter set. 
-For randomised ones we further divide them in to sub categories based on their OUI and length of the packet.
-Since the length tends to stay unique to specific models of devices we are left with the task of identifying the unique mobile devices from within these sets.
+From our initial analysis we found that OUI, length of the packet and sequence
+number of the packet being the most promising information to achieve this.
+First we divide our dataset into sets of probe requests with randomised and
+non-randomised MAC addresses and keep the MAC address as the unique identifier
+for the latter set. For randomised ones we further divide them in to sub
+categories based on their OUI and length of the packet. Since the length tends
+to stay unique to specific models of devices we are left with the task of
+identifying the unique mobile devices from within these sets.
 
-The proposed algorithm creates a graph where the probe requests represented the nodes, and links are created between them based on the following rules: 
+The proposed algorithm creates a graph where the probe requests represented the
+nodes, and links are created between them based on the following rules:
+
 \begin{itemize} 
@@
 \item A link could go from low to high sequence numbers. 
-	\item A link could exist between nodes with a maximum time difference of $\alpha$ - time threshold.
-	\item A link could exist between nodes with a maximum sequence number difference of $\beta$ - sequence threshold.
-	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links in terms of both time and sequence number.
+\item A link could exist between nodes with a maximum time difference of
+	$\alpha$ - time threshold. 
+\item A link could exist between nodes with a maximum sequence number
+	difference of $\beta$ - sequence threshold.
+\item A node could have only one incoming link and one outgoing link, which is
+	the shortest of all such possible links in terms of both time and sequence
+	number.
 \end{itemize}
-The nodes were then classified based on the unique connected component they belong to as shown in Figure \ref{methodology_schematic}.
-This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
-Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the magnitude of error is greatly reduced.
+
+The nodes were then assigned a unique id based on the unique connected component
+they belong to as shown in Figure \ref{methodology_schematic}. This unique
+identifier is used in the place of MAC address for aggregation for the
+anonymised probe requests. Though the recycling of sequence number after 4096
+leads to multiple unique ids being reported from a single device, the magnitude
+of error is greatly reduced.
 
 \subsection{Calibrating with Ground Truth}
-Since mobile device ownership is an external uncertainty to our study and could arise from variety of spatio - temporal and demographic factors, we propose to solve this using external source of information.
-We hypothesize that an adjustment factor could be arrived at for each location of data collection, comparing the sensor based counts and ground truth and it can be used to adjust the data reliably to reflect the ground truth in absolute numbers for the future.
-This calibration can be carried over periodically and the frequency to improve the quality of the estimation.
+
+Since mobile device ownership is an external uncertainty to our study and could
+arise from variety of spatio - temporal and demographic factors, we propose to
+solve this using external source of information. We hypothesize that an
+adjustment factor could be arrived at for each location of data collection by
+comparing the sensor based counts and ground truth and it can be used to adjust
+the data reliably to reflect the ground truth in absolute numbers for the
+future. This calibration can be carried out periodically and frequently to
+improve the quality of the estimation over longer periods of time.

### 2018-11-20 paper/sections/pilot_study.tex
Tue 20 Nov 13:38:53 GMT 2018

@@
-To start we designed a small pilot study to validate the classification and clustering methodology against the scale and complexity of data collected on a open public area such as a retail high street.
-We also aim to find the algorithm which is best suited for classification of signal strengths to filter out the background noise.
-The data was collected at Oxford Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe requests were collected using the sensor described in Chapter \ref{methodology} and pedestrian footfall was manually recorded using the Android app \citep{bala2018clicker}.
-Being located at one of the busiest retail locations in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe requests during the half hour period, and 3,722 people were recorded manually walking on the sidewalk during that time.
+\begin{table}
+\tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
+{\begin{tabular}{lcc} 
+\toprule
+	 Algorithm					& Time (s) 	& MAPE (\%) \\
+\midrule
+	 Quantile					& 0.002 	&  27\% \\
+	 K-Means			 		& 0.007 	& -23\% \\
+	 Hierarchical Clustering	& 172.520 	& -9\% \\
+	 Bagged Clustering 			& 0.135 	& -30\% \\
+	 Fisher 					& 3.034 	& -30\% \\
+	 Jenks Natural Break 		& 556.279 	& -30\% \\
+ \bottomrule
+\end{tabular}}
+\label{classification-table}
+\end{table}
 
-As a first step we just aggregated the probe requests by their MAC address for every minute to generate a minute by minute count of the number of people around the sensor assuming each MAC address corresponds to a mobile device and hence a pedestrian.
-We then compare this preliminary `footfall' count to the actual number of pedestrian recorded manually to check for the robustness.
-We use Mean Absolute Percentage Error (MAPE) measure of robustness of the count since it provides a simple and quick measurement while the street conditions ensure that there are no intervals without any footfall.
-We find that the MAPE in the raw counts compared to the ground truth is around 425\%.
-This suggests the presence of large amount of noise in the data which might be generated due to sources of uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need for filtering the data.
+To start we designed a small pilot study to validate the filtering and
+clustering methodology against the scale and complexity of data collected on an
+open public area such as a retail high street.  We also aim to find the
+algorithm which is best suited for classification of signal strengths as 'low'
+and 'high' to filter out the background noise.  The data was collected at Oxford
+Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe
+requests were collected using the sensor described in Chapter \ref{methodology}
+and pedestrian footfall was manually recorded using the Android app
+\citep{bala2018clicker}.  Being located at one of the busiest retail locations
+in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe
+requests during the half hour period, and 3,722 people were recorded manually
+walking on the sidewalk during that time.
 
-We then classified the probe requests as ones with "high signal strength" and "low signal strength" using various classification algorithm such as k-means, quantile, hierarchical clusters, fisher and jenks.
-We found that while hierarchical clustering and jenks were resource intensive, k-means gave quickest results with the lowest MAPE closely followed by quantile.
-The cut-off point or threshold for the collected data with which we could classify them as high and low was -71 dBm.
-We then removed all the probe requests which reported `low signal strength' and repeated the same aggregation process as before to produce footfall count. 
-This process resulted in a footfall count with a net MAPE of 30\%.
-Though the results are encouraging we are still not completely confident that our filtering process is indeed removing noise or has any correlation the configuration of sensor or position of the mobile devices.
-These need to be addressed with a larger survey with multiple location of varying orientations.
+As a first step we just aggregated the probe requests by their MAC address for
+every minute to generate a minute by minute count of the number of people around
+the sensor assuming each MAC address corresponds to a mobile device and hence a
+pedestrian.  We then compare this preliminary `footfall' count to the actual
+number of pedestrians recorded manually to check for the robustness. We use Mean
+Absolute Percentage Error (MAPE) measure of robustness of the count since it
+provides a simple and quick measurement while the street conditions ensure that
+there are no intervals without any footfall.  We find that the MAPE in the raw
+counts compared to the ground truth is around 425\%.  This suggests the presence
+of large amount of noise in the data which might be generated due to sources of
+uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need
+for filtering the data.
+
+We then classified the probe requests as ones with "high signal strength" and
+"low signal strength" using various one dimensional clustering algorithms such
+as k-means, quantile, hierarchical clustering, bagged clustering, fisher and
+jenks natural breaks.  The results are shown in Table
+\ref{classification-table}.  We found that while hierarchical clustering and
+jenks gave us fairly low errors, they were too resource intensive for practical
+use with a larger dataset.  We also found that k-means gave quickest results
+with the lowest MAPE closely followed by quantile algorithm.  The cut-off point
+or threshold for the collected data with which we could classify them as high
+and low was -71 dBm.  We then removed all the probe requests which reported `low
+signal strength' and repeated the same aggregation process as before to produce
+footfall count.  This process resulted in a footfall count with a net MAPE of
+30\%.  Though the results are encouraging we are still not completely confident
+that our filtering process is indeed removing noise or has any correlation the
+configuration of sensor or position of the mobile devices.  These need to be
+addressed with a larger survey with multiple location of varying orientations.
+
+\begin{figure}
+\begin{center}
+\vspace{10pt}
+\includegraphics [width=\linewidth,trim=10 30 10 75,clip]
+	{images/pilot_clustering_params.png}
+\caption{The clustering process was repeated with both increasing sequence
+	number threshold ($\alpha$) and time threshold ($\beta$), until we arrived
+	at the lowest parameters where the know device (black line) is clustered as
+	a single device.}
+\label{pilot_clustering_params}
+\end{center}
+\end{figure}
 
@@
 \centering
-	\subfigure[Clustering probe requests based on increasing sequence numbers present in them.]{
-		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_clustering.jpeg}}}\hspace{20pt}
-	\subfigure[Comparison of sensor based counts with the ground truth.]{
-	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/pilot_comparison.jpeg}}}
-	\caption{The results of the pilot study demonstrating the validity of the methodology.} \label{pilot_results}
+\subfigure[Clustering probe requests based on increasing sequence numbers 
+	present in them.]
+	{\resizebox*{0.45\linewidth}{!}
+	{\includegraphics{images/pilot_clustering.jpeg}}}
+\hspace{20pt}
+\subfigure[Comparison of sensor based counts with the ground truth.]
+	{\resizebox*{0.45\linewidth}{!}
+	{\includegraphics{images/pilot_comparison.jpeg}}}
+\caption{The results of the pilot study demonstrating the validity of the
+	methodology.} 
+\label{pilot_results}
 \end{figure}
 
-The next challenge was to identify probe requests which are generated by the same device irrespective of MAC randomisation process.
-We use the algorithm defined in Chapter \ref{methodology} and assign a unique identifier or signature to each probe request independent of the MAC address. 
-Since there is neither prior research nor documentation on a universal behaviour of phones in randomising their MAC addresses, we use the stationary device - the one used for manual counting as a reference and find out the suitable time threshold, $\alpha$ and threshold for sequence numbers, $\beta$ to be 16 seconds and 60 respectively via trial and error.
-This process is done on top the filtering done based on signal strength and only for the probe requests with randomised MAC addresses.
-Figure \ref{pilot_results} shows the results of this clustering process on a small set of randomised probe requests.
-The probe request with different randomised MAC address is shown by the colored points and the line joining them shows the ones belonging to the same cluster hence expected to be generated by the same device.
-We finally aggregate the probe requests as before but with the device signature rather than just MAC addresses this results in a footfall count with a MAPE of -18\%. 
-A comparison of minute by minute counts resulting from different filtering processes along with the ground truth is shown in Figure \ref{pilot_results} showing the promising effectiveness of the methods.
+The next challenge was to identify probe requests which are generated by the
+same device irrespective of MAC randomisation process.  We use the algorithm
+defined in Chapter \ref{methodology} and assign a unique identifier or signature
+to each probe request independent of the MAC address.  Since there is neither
+prior research nor documentation on a universal behaviour of phones in
+randomising their MAC addresses, we use the stationary device - the one used for
+manual counting as a reference and find out the suitable time threshold,
+$\alpha$ and threshold for sequence numbers, $\beta$ to be 16 seconds and 60
+respectively via trial and error. This process is shown in Figure
+\ref{pilot_clustering_params} This process is done on top the filtering done
+based on signal strength and only for the probe requests with randomised MAC
+addresses.  Figure \ref{pilot_results} shows the results of this clustering
+process on a small set of randomised probe requests.  The probe request with
+different randomised MAC address is shown by the colored points and the line
+joining them shows the ones belonging to the same cluster hence expected to be
+generated by the same device.  We finally aggregate the probe requests as before
+but with the device signature rather than just MAC addresses this results in a
+footfall count with a MAPE of -18\%.  A comparison of minute by minute counts
+resulting from different filtering processes along with the ground truth is
+shown in Figure \ref{pilot_results} showing the promising effectiveness of the
+methods.
 
-To conclude, from the pilot study we found that both classification and clustering methods we devices work on complex real world data and results in a final pedestrian counts within a MAPE of 20\%.
-We also found `k-means' and `quantile' are best algorithms for classifying signal strengths and the threshold for time and sequence numbers for the clustering algorithm is around 16 and 60 respectively.
+To conclude, from the pilot study we found that both filtering and clustering
+methods we devices work on complex real world data and results in a final
+pedestrian counts within a MAPE of 20\%. We also found `k-means' and `quantile'
+are best algorithms for classifying signal strengths and the threshold for time
+and sequence numbers for the clustering algorithm is around 16 and 60
+respectively.

### 2018-11-20 paper/sections/previous_work.tex
Tue 20 Nov 13:38:53 GMT 2018

@@
-In the past decade Wi-Fi has emerged as one of the most commonly used technology in providing high speed internet access to mobile devices such as smartphones, tablets and laptops in public and private spaces.
-This has resulted in multiple Wi-Fi networks being available at almost every location in dense urban environments. 
-Traversing through this overlapping mesh of Wi-Fi networks, modern mobile devices with Wi-Fi antennae regularly broadcast a special type of signal known as `Probe Requests' in order to discover Wi-Fi networks available to them.
-This helps these devices to connect and switch between the Wi-Fi networks seamlessly.
+In the past decade Wi-Fi has emerged as one of the most commonly used technology
+in providing high speed internet access to mobile devices such as smartphones,
+tablets and laptops in public and private spaces. This has resulted in multiple
+Wi-Fi networks being available at almost every location in dense urban
+environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
+mobile devices with Wi-Fi antennae regularly broadcast a special type of signal
+known as `Probe Requests' in order to discover Wi-Fi networks available to them.
+This helps these devices to connect and switch between the Wi-Fi networks
+seamlessly.
 
-Probe requests are low level signals standardised by IEEE 802.1b/g specification \citep{ieee2013} as the first step in establishing a Wi-Fi based connection between two devices and is implemented in any Wi-Fi capable device irrespective of the manufacturer or the model.
-This ubiquity and standardisation makes them an excellent source of open, passive, continuous, and wireless data generated by Wi-Fi capable devices present at any given time and location.
-Considering the unprecendented levels of mobile device ownership in recent years, we can, in turn use this data to understand the population distribution in highly dynamic urban environments with high spatial and temporal granularity \citep{freud2015,konto2017}.
-While a Wi-Fi based method to collect data offers us various advantages such as, easy scalability and efficiency in terms of cost and time, It also introduces few systematic biases, uncertainities in the collected data along with the serious risk of infringing on the privacy of the mobile users.
-In this paper, using a set of probe requests and manual counts collected at various high street locations across London, we demonstrate that pedestrian footfall at these locations can be estimated with considerable precision and accuracy while protecting the privacy of the pedestrians.
+Probe requests are low level signals standardised by IEEE 802.1b/g specification
+\citep{ieee2013} as the first step in establishing a Wi-Fi based connection
+between two devices and is implemented in any Wi-Fi capable device irrespective
+of the manufacturer or the model. This ubiquity and standardisation makes them
+an excellent source of open, passive, continuous, and wireless data generated by
+Wi-Fi capable devices present at any given time and location. Considering the
+unprecendented levels of mobile device ownership in recent years, we can, in
+turn use this data to understand the population distribution in highly dynamic
+urban environments with high spatial and temporal granularity
+\citep{freud2015,konto2017}. While a Wi-Fi based method to collect data offers
+us various advantages such as, easy scalability and efficiency in terms of cost
+and time, It also introduces few systematic biases, uncertainities in the
+collected data along with the serious risk of infringing on the privacy of the
+mobile users. In this paper, using a set of probe requests and manual counts
+collected at various high street locations across London, we demonstrate that
+pedestrian footfall at these locations can be estimated with considerable
+precision and accuracy while protecting the privacy of the pedestrians.
 
-Though, unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly inferred from Wi-Fi, there are reliable methods to triangulate the location of mobile devices from the locations of known access points(AP) and the signal strength reported by them \citep{he2003range, moore2004robust, lamarca2005place}.
-This can overcome the usual shortcoming of GPS, which struggles for precision and accuracy in indoor and densely built environments \citep{zarim2006,kawaguchi2009wifi, xi2010locating}. 
-Utilising this, we can easily and quickly estimate trajectories of the mobile devices \citep{Sorensenlocation,musa2012tracking} which can be used similar to the GPS trajectories to understand individual travel patterns \citep{reki2007,Sap2015}, crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular \citep{lu2010vehicle} and pedestrian movement \citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}.
-Such data can also be used in transportation planning and management to estimate travel time \citep{musa2011wiflow} and real time traffic monitoring \citep{abbott2013empirical}.
-Using techniques demonstrated by \cite{franklin2006passive} and \cite{pang2007802} along with information present in the probe requests one even model interactions between the users \citep{cheng2012inferring,barbera2013signals,cunche2014know,cunche2014linking} such as predicting which of them are most likely to meet again \citep{cunche2012know}.
-Using the semantic information present in these probe requests it even is possible to understand the nature of population at a large scale \citep{di2016mind}.
+Though, unlike GPS, the location of the Wi-Fi enabled mobile device cannot be
+directly inferred from Wi-Fi, there are reliable methods to triangulate the
+location of mobile devices from the locations of known access points(AP) and the
+signal strength reported by them
+\citep{he2003range,moore2004robust,lamarca2005place}. This can overcome the
+usual shortcoming of GPS, which struggles for precision and accuracy in indoor
+and densely built environments
+\citep{zarim2006,kawaguchi2009wifi,xi2010locating}. Utilising this, we can
+easily and quickly estimate trajectories of the mobile devices
+\citep{Sorensenlocation,musa2012tracking} which can be used similar to the GPS
+trajectories to understand individual travel patterns \citep{reki2007,Sap2015},
+crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular
+\citep{lu2010vehicle} and pedestrian movement
+\citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}. Such data can
+also be used in transportation planning and management to estimate travel time
+\citep{musa2011wiflow} and real time traffic monitoring
+\citep{abbott2013empirical}. Using techniques demonstrated by
+\cite{franklin2006passive} and \cite{pang2007802} along with information present
+in the probe requests one even model interactions between the users
+\citep{cheng2012inferring,barbera2013signals,cunche2014know,cunche2014linking}
+such as predicting which of them are most likely to meet again
+\citep{cunche2012know}. Using the semantic information present in these probe
+requests it even is possible to understand the nature of population at a large
+scale \citep{di2016mind}.
 
-Though extensive research has been carried out on this subject with feasible and favorable results, in recent years, one of the major challenges faced in such attempts has been the increasing attempt by mobile phone manufacturers to protect their users’ privacy by anonymising the globally identifiable portion of the probe requests, \citep{green2008}.
-Various methods have been devised to overcome this anonymisation process such as estimating the device model information from a known dataset of manufacturers and device behaviours \citep{martin2016decomposition}; Scrambler attack using a small part of the physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and timing attack where the packet sequence information along with information elements present in the probe request frame is used \citep{matte2016,cheng2016can}.
-A combination of these methodologies has been proven to produce de-anonymised globally unique device information \citep{vanhoef2016, martin2017}.
-These approaches usually result in serious risk of breach of privacy of the users of the mobile devices by revealing their identifiable personal information.
+Though extensive research has been carried out on this subject with feasible and
+favorable results, in recent years, one of the major challenges faced in such
+attempts has been the increasing attempt by mobile phone manufacturers to
+protect their users’ privacy by anonymising the globally identifiable portion of
+the probe requests, \citep{green2008}. Various methods have been devised to
+overcome this anonymisation process such as estimating the device model
+information from a known dataset of manufacturers and device behaviours
+\citep{martin2016decomposition}; Scrambler attack using a small part of the
+physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and
+timing attack where the packet sequence information along with information
+elements present in the probe request frame is used
+\citep{matte2016,cheng2016can}. A combination of these methodologies has been
+proven to produce de-anonymised globally unique device information
+\citep{vanhoef2016,martin2017}. These approaches usually result in serious risk
+of breach of privacy of the users of the mobile devices by revealing their
+identifiable personal information.
 
-There is a clear gap in research for exploring methodologies for estimating the number of unique mobile devices from a set of anonymised probe requests, without the need to reveal their original device information.
-Such technique has various applications such as uncovering the urban wireless landscape \citep{rose2010mapping}, revealing human activity at large scales \citep{qin2013discovering}, estimating pedestrian numbers in crowds \citep{schauer2014estimating,fukuzaki2015statistical} and even counting people in hyper local scales such as queues \citep{wang2013measuring}
-With enough infrastructure to collect such information we can even aim to generate a real-time census of the city \citep{konto2017}.
-With this background we set out to device and implement a methodology to reliably estimate human activity such as pedestrian footfall from Wi-Fi probe requests without risking the breach of privacy of the users involved.
+There is a clear gap in research for exploring methodologies for estimating the
+number of unique mobile devices from a set of anonymised probe requests, without
+the need to reveal their original device information. Such technique has
+various applications such as uncovering the urban wireless landscape
+\citep{rose2010mapping}, revealing human activity at large scales
+\citep{qin2013discovering}, estimating pedestrian numbers in crowds
+\citep{schauer2014estimating,fukuzaki2015statistical} and even counting people
+in hyper local scales such as queues \citep{wang2013measuring} With enough
+infrastructure to collect such information we can even aim to generate a
+real-time census of the city \citep{konto2017}. With this background we set out
+to device and implement a methodology to reliably estimate human activity such
+as pedestrian footfall from Wi-Fi probe requests without risking the breach of
+privacy of the users involved.

### 2018-11-20 analysis/parameters.r

@@ -1,0 +1,54 @@
+library(tidyverse); library(classInt); library(ggplot2); library(igraph); library(lubridate)
+
+get_clustered <- function(thres_s,thres_t) {
+	sensor <- "../data/final_sensor_data.csv" %>% 
+		read.csv(stringsAsFactors = FALSE) %>% as_tibble() %>%
+		select(time, signal, mac, oui, vendor, type, sequence) %>%
+		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS"))
+	threshold <- classIntervals(sensor$signal, 2, "kmeans")$brks[2]
+	add_signature <- function(data, ds, dt) {
+		vertices <- data %>%
+			mutate(x = as.integer(time), id = as.numeric(rownames(data))) %>%
+			select(id, x, y=sequence) %>%
+			arrange(x)
+		edges <- vertices %>%
+			pull(id) %>% expand.grid(.,.) %>% rename(from=Var1,to=Var2) %>%
+			mutate(from=as.numeric(from), to=as.numeric(to)) %>%
+			filter(from<to) %>%
+			left_join(vertices, by = c("from" = "id")) %>%
+			left_join(vertices, by = c("to" = "id")) %>%
+			mutate(sdis = sqrt((y.y-y.x)**2)) %>%
+			mutate(tdis = sqrt((x.y-x.x)**2)) %>%
+			filter(x.y >= x.x, y.y > y.x, tdis < dt, sdis < ds) %>%
+			group_by(from) %>% summarise(to = to[which(tdis == min(tdis))][[1]], tdis = min(tdis)) %>% 
+			group_by(to) %>% summarise(from = from[which(tdis == min(tdis))][[1]])
+		graph <- graph_from_data_frame(edges, vertices = vertices)
+		data$device_sign <- paste(data$vendor, clusters(graph)$membership, sep = "-")
+		return(data) }
+	sensor <- sensor  %>% filter(signal > threshold) %>% 
+		inner_join(read.csv("../data/final_vendors_data.csv", stringsAsFactors=FALSE), by = "vendor") %>%
+		filter(is_mobile == TRUE) %>% mutate(is_mobile = NULL) %>%
+		filter(type == "local") %>% split(f = {.$vendor}) %>%
+		lapply(add_signature, ds = thres_s, dt = thres_t) %>% bind_rows()
+	return(sensor) }
+
+get_df <- function(range_x,range_y) { 
[…]

### 2018-11-20 analysis/signal_strength.r

@@ -8,14 +8,13 @@ library(lubridate)
 
 signal_comparison <- function(method,records) {
 	print(method)
-	t1 <- Sys.time()
 	sensor <- "../data/final_sensor_data.csv" %>% 
 		read.csv(stringsAsFactors = FALSE) %>% as_tibble() %>%
 		select(time, signal, mac, oui, vendor, type, sequence) %>%
 		mutate(time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS")) %>%
 		arrange(time) %>% head(records)
-	llogg <- capture.output({threshold <- classIntervals(sensor$signal, 2, method)$brks[2]})
-	print(threshold); time <- as.numeric(Sys.time()-t1); print(time)
+	print(system.time(llogg <- capture.output({threshold <- classIntervals(sensor$signal, 2, method)$brks[2]})))
+	print(threshold)
 	add_signature <- function(data, ds, dt) {
 		vertices <- data %>%
 			mutate(x = as.integer(time), id = as.numeric(rownames(data))) %>%

### 2018-11-21 review/response.md
Wed 21 Nov 17:08:08 GMT 2018

@@
 Title : Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests
+
 Manuscript id : IJGIS-2018-0340
+
 Comments dated : 28-Oct-2018
@@
 	- Clarity of the methodology
-	- Lack of discussion of wifi-based human activity studies in the big data age
+	- Lack of discussion of wifi-based human activity studies in the big data
+	age
 	- Proper use of terminology
@@
 
-2. In lieu of these comments, they have recommended to make substantial revisions
-to your manuscript
+2. In lieu of these comments, they have recommended to make substantial
+revisions to your manuscript
 
 3. Also, since this is a "smart cities" related special issue, I ask you to
-	- Add more discussions on smart cities in the introduction and 
-	  conclusion sections and
+	- Add more discussions on smart cities in the introduction and conclusion
+	sections and
 	- Link your research more tightly to this theme.
 
-> Thank you for the very helpful comments of the reviewers of our paper/
-> As requested, we have responded to all comments without adding
-> significant length to the paper. We have also extended our discussion of
-> how effective measurement of footfall is relevant to understanding the
-> functioning of the Smart City.
+> Thank you for the very helpful comments of the reviewers of our paper/ As
+> requested, we have responded to all comments without adding significant
+> length to the paper. We have also extended our discussion of how effective
+> measurement of footfall is relevant to understanding the functioning of the
+> Smart City.
 > 
-> If any of what follows and what has been uploaded is unclear, please do
-> not hesitate to contact me.
+> If any of what follows and what has been uploaded is unclear, please do not
+> hesitate to contact me.
 > 
-> Yours sincerely
-> Balamurugan Soundararaj
+> Yours sincerely,
+> Balamurugan Soundararaj,
 > (on behalf also of Paul Longley and James Cheshire)
@@
 1. Please try to avoid making the paper longer (keep in mind that the average
-length of papers in this journal is 18-20 published pages). It is important that
-you ensure that the spelling and grammar is of a high standard.
+   length of papers in this journal is 18-20 published pages). It is important
+that you ensure that the spelling and grammar is of a high standard.
 
@@
    https://mc.manuscriptcentral.com/ijgis?URL_MASK=a32caf3bbd49412cac8d1cfa9d7ebaf8
-which will take you your decision letter, and where you need to
-resubmit your manuscript.
+   which will take you your decision letter, and where you need to resubmit
+   your manuscript.
 
-4. If you have used the on-line LaTeX template, upload all those files and a pdf.
+4. If you have used the on-line LaTeX template, upload all those files and a
+   pdf.
 
@@
 6. There are two hard limits that you need to bear in mind, 
-	- Image size - Please be aware that the total number of pixels 
-	  (height X width) in each image must be less than 40,000,000 (40 mp).
-	- Total file size - please make sure that the combined, uncompressed,
-	  file sizes do not exceed 20Mb. The internal LZW compression, which 
-	  is an option in TIF, is acceptable however.
+	- Image size - Please be aware that the total number of pixels (height X
+			width) in each image must be less than 40,000,000 (40 mp).
+	- Total file size - please make sure that the combined, uncompressed, file
+	sizes do not exceed 20Mb. The internal LZW compression, which is an option
+	in TIF, is acceptable however.
 	- Please do not save B/W or Greyscale images as RGB.
@@
 
-9. When submitting your revised manuscript, you will be able to respond to
-the comments made by the reviewers in the space provided.  You can use this
-space to document any changes you make to the original manuscript.  In order
-to expedite the processing of the revised manuscript, please be as specific as
+9. When submitting your revised manuscript, you will be able to respond to the
+comments made by the reviewers in the space provided.  You can use this space
+to document any changes you make to the original manuscript.  In order to
+expedite the processing of the revised manuscript, please be as specific as
 possible in your response to the reviewers.
@@
 10. Please delete any redundant files before completing the submission.
+
+> Thank you for the detailed instructions. All of these have been checked in
+> revised manuscript for the final submission.
 
@@
 
-This paper is about the estimation of pedestrian footfall using Wi-Fi
-signal emitted by mobile devices.
-It deals with the problem of estimating the footfall size based on a noisy
-and biased data. More specifically it addresses the following issues that may
-lead to overcounting:
+This paper is about the estimation of pedestrian footfall using Wi-Fi signal
+emitted by mobile devices.  It deals with the problem of estimating the
+footfall size based on a noisy and biased data. More specifically it addresses
+the following issues that may lead to over counting:
+
 - MAC randomization that may cause a device to generate several MAC addresses;
- - Signals received from outside the interest zone; unequal adoption
-   of Wi-Fi smartphone in the population.
-To address this issue, a number of techniques are introduced:
- - clustering received signal strength to remove device that are outside
-   the monitored area;
+ - Signals received from outside the interest zone; unequal adoption of Wi-Fi
+ smartphone in the population.  To address this issue, a number of techniques
+ are introduced:
+ - clustering received signal strength to remove device that are outside the
+ monitored area;
  - clustering probes based on sequence number and timing;
  - finally an adjustment factor is applied to correct the estimated value.
-   The adjustment factor is calibrated for each location using
-   a manual counting.
-The author conducted a pilot experiment to identify the clustering 
-algorithms and parameters, and then they conducted a second range of 
-experiments at different location over an extended period of time. During 
-each of those experiments, ground-truth count was collected manually.
-The results show that the presented approach can significantly reduce the
-counting error compared to a naive solution consisting in counting the 
-distinct number of MAC observed.
-Nevertheless, there are cases (2) where the error is still large (50%),
-even if in other cases (1) the error is only of 9%.
+
+The adjustment factor is calibrated for each location using a manual counting.
+The author conducted a pilot experiment to identify the clustering algorithms
+and parameters, and then they conducted a second range of experiments at
+different location over an extended period of time. During each of those
+experiments, ground-truth count was collected manually.  The results show that
+the presented approach can significantly reduce the counting error compared to
+a naive solution consisting in counting the distinct number of MAC observed.
+Nevertheless, there are cases (2) where the error is still large (50%), even if
+in other cases (1) the error is only of 9%.
 
 > This is a correct surmisal of the remit of the paper. We have also added
-> material to the introduction and conclusion of the paper to better link
-> this remit to the Smart Cities discourse.
+> material to the introduction and conclusion of the paper to better link this
+> remit to the Smart Cities discourse.
 
@@
 algorithm.
+
 [1] Mathy Vanhoef, Célestin Matte, Mathieu Cunche, Leonardo Cardoso, Frank
@@
 > anonymisation is ensured by the system architecture that processess and
-> curates the data. Spedicifically the weekly rotation of the random salt
-> with which the hashing process is carried out.
+> curates the data. Specifically we rotated the random salt used to hash the
+> MAC addresses regulary to prevent any kind of tracking of the users across a
+> longer period of time with which the hashing process is carried out.
 
@@
 Yet before publication, I believe that there are a number of points that needs
-to be clarified or corrected. Regarding the technical aspects I have a
-number of minor remarks:
+to be clarified or corrected. Regarding the technical aspects I have a number
+of minor remarks:
 
@@
 > extensive methodological research and field implementation and is, we
-> believe, cutting edge in its finding. It also has important implications 
-> for our understanding of the functioning of smart cities, and these
-> implications are discussed in greater detail in the revised version.
+> believe, cutting edge in its finding. It also has important implications for
+> our understanding of the functioning of smart cities, and these implications
+> are discussed in greater detail in the revised version.
 
@@
 is just the part ensuring that the signal is correctly emitted, but it is not
-the source. In this context it is better to use a more general concept and refer
-to the "Wi-Fi network interface" (or Wi-Fi interface) as the the element
+the source. In this context it is better to use a more general concept and
+refer to the "Wi-Fi network interface" (or Wi-Fi interface) as the the element
 broadcasting a signal.
@@
 Standard for Information technology–Telecommunications and information exchange
-between systems Local and metropolitan area networks–Specific requirements 
-Part 11: Wireless LAN Medium Access Control (MAC) and Physical Layer (PHY) 
-Specifications,” IEEE Std 802.11-2012 (Revision of IEEE Std 802.11-2007),
-pp. 1–2793, Mar. 2012.
+between systems Local and metropolitan area networks–Specific requirements Part
+11: Wireless LAN Medium Access Control (MAC) and Physical Layer (PHY)
+Specifications,” IEEE Std 802.11-2012 (Revision of IEEE Std 802.11-2007), pp.
+1–2793, Mar. 2012.
 
@@
 > This simple application was developed specifically for the project for
-> Android platform. More details on the application has been included in 
-> the revised manuscript.
+> Android platform. More details on the application has been included in the
+> revised manuscript.
 
@@
 classifications reported on single device, the magnitude of error is greatly
-reduced." This sentence is not clear. And the cycle length is not 4000 but 
-4096 = 2^12.
+reduced." This sentence is not clear. And the cycle length is not 4000 but 4096
+= 2^12.
 
@@
 
-> Thank you very much for the suggestion. From out tests on using this 
-> method for distance calculation results in a change of --% of the result.
-> Which in our opinion doesnt influence the results in any significant way
-> considering the context.
+> Thank you very much for the suggestion. From tests carried out on sample of
+> 1910 probe requests from google phones in the pilot study we found that
+> around 0.5% reset their mac addresses.  Which in our opinion doesn't
+> influence the results in any significant way considering the context.
 

### 2018-11-21 paper/sections/introduction.tex
Wed 21 Nov 17:08:08 GMT 2018

@@
 locations relative to networks of masts, use of GPS to locate users of social
-media services, and Wi-Fi beacon connectivity to access the Internet. These
-technologies offer differing levels of spatial precision, where mobile telephony
-and Wi-Fi generally being less reliable and offer lower precision than GPS to
-the end users, while being more advantageous for broader mobility studies
+media services, and Wi-Fi access points connectivity to access the Internet.
+These technologies offer differing levels of spatial precision, where mobile
+telephony and Wi-Fi generally being less reliable and offer lower precision than
+GPS to the end users, while being more advantageous for broader mobility studies
 (Pinelli et al.  2015). There has been considerable research into the usefulness

### 2018-11-21 paper/sections/methodology.tex
Wed 21 Nov 17:08:08 GMT 2018

@@
 information transmitted through them. In some cases, such as Android devices,
-these are broadcast even when the Wi-Fi functionality has been turned off by the
-user. Thus these signals can be used to reliably identify the presence of Wi-Fi
-enabled mobile devices. Being a first step of connection initiated by the mobile
-device, these packets have information regarding the characteristics of the
-mobile device itself. Some of the key information we can infer from these
-requests are,
+these are broadcast even when the Wi-Fi functionality has been turned off by
+the user so that device can immediately connet to networks when the
+functionality is switched back on. Since some devices even use the probe
+requests as a less accurate form of localisation they keep sending probe
+requests as well. Thus these signals can be used to reliably identify the
+presence of Wi-Fi enabled mobile devices. Being a first step of connection
+initiated by the mobile device, these packets have information regarding the
+characteristics of the mobile device itself. Some of the key information we can
+infer from these requests are,
 
 \begin{enumerate} 
-\item \textbf{Media Access Control (MAC) address} which is an unique identifier
+\item \textbf{Media Access Control (MAC) address} which is an name identifying
 	for the wireless hardware of the mobile device, 
@@
 \item Total \textbf{length} of the request in number of bits, and 
-\item The \textbf{strength of the signal} which transmitted the request.
+\item The \textbf{strength of the signal} received by the mobile device.
 \end{enumerate}
@@
 The MAC address is the primary unique identifier for the mobile device. It has
-two parts, first part is the Organisation Unique Identifier (OUI) which gives
-information about the manufacturer of the device and the second is unique to the
-device. In modern devices, to protect the users privacy, the MAC address can
-also be randomised (hence non unique) and is marked as such. Though sequence
-number and length of the packet are not strictly unique, we hypothesize that we
-can use them to estimate the number unique devices as demonstrated by
-\citep{vanhoef2016}.
+two parts, first part is the Organisationally Unique Identifier (OUI) which
+gives information about the manufacturer of the device and the second is unique
+to the device. In modern devices, to protect the users privacy, the MAC address
+can also be randomised (hence non unique) and is marked as such. Though
+sequence number of the packet is strictly unique to a mobile device, we
+hypothesize that we can use them to estimate the number unique devices as
+demonstrated by \citep{vanhoef2016}, where optional information present in the
+probe requests - Information Elements (IE) along with the probe requests have
+been used to fingerprint the devices. This approach has become increasinly
+difficult as mobile phone manufacturers have severely limited the IEs present
+in the probe requests thus leading us to explore methods which uses only the
+sequence numbers. This also affects the established commercial solutions using
+Wi-Fi probe requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.
+There have been another solution proposed by \citep{hong2018crowdprobe} where
+the authors tried to solve the similar problem using a hidden markov models
+based trajectory inference alogorithm but the scope is limited to enclosed,
+exit controlled public spaces such as shopping malls, railway stations, etc.
 
@@
 `broadcast', marked with type as management' and subtype as `probe requests'.
-The MAC address in these probe requests is anonymised at the device level using
+The MAC address in these probe requests is obfuscated at the device level using
 a cryptographic hashing algorithm and transmitted through 3G connection to a
 central database via web-sockets protocol, where it is stored in a PostgreSQL
-database for further analysis. A overall schematic of the data collection and
-storage process is shown in Figure \ref{datacollection_schematic}. The ground
-truth on number of pedestrian footfall was recorded using a custom mobile
-application - Clicker \citep{bala2018clicker}
+database for further analysis. The random salt used in the hashing algorithm was
+rotated regularly to further mitigate the risk of de-anonymisation of the hash.
+Though hashing cannot completely ensure anonymisation as discussed in
+\citep{demir2014analysing} it can sufficiently obfuscate the data which along
+with a secure process of data handling gives us reasonable security. A overall
+schematic of the data collection and storage process is shown in Figure
+\ref{datacollection_schematic}. The ground truth on number of pedestrian
+footfall was recorded using a custom Android application - Clicker
+\citep{bala2018clicker}. This app logs accurate timestamps for each time the
+surveryor records a pedestrian crossing the designated cordon line at the
+location. In addition to counting the pedestrians manually this also makes the
+mobile device used constantly probe for networks which also gives us a know
+device to calibrate our methodology against.
 
@@
 
-We hypothesise that in configurations where a specific source of background
+We assume that in configurations where a specific source of background
 noise is at a constant distance, there must be a distinct pattern in the number
@@
 
-From our initial analysis we found that OUI, length of the packet and sequence
-number of the packet being the most promising information to achieve this.
+From our initial look at the data we found that OUI and the sequence
+number of the packet is the most promising information to achieve this.
 First we divide our dataset into sets of probe requests with randomised and
-non-randomised MAC addresses and keep the MAC address as the unique identifier
-for the latter set. For randomised ones we further divide them in to sub
-categories based on their OUI and length of the packet. Since the length tends
-to stay unique to specific models of devices we are left with the task of
-identifying the unique mobile devices from within these sets.
+non-randomised MAC addresses by looking at the second character of the vendor
+part of the MAC address and if it is E, A, 2 or 6, then these addresses are
+identified to be randomised. We keep the MAC address as the unique
+identifier for the non-randomised requests and further divide the randomised ones in to
+sub categories based on their OUI. We then  
+identify unique mobile devices from within these sets and assign a unique identifier to each device.
 
@@
 arise from variety of spatio - temporal and demographic factors, we propose to
-solve this using external source of information. We hypothesize that an
-adjustment factor could be arrived at for each location of data collection by
-comparing the sensor based counts and ground truth and it can be used to adjust
-the data reliably to reflect the ground truth in absolute numbers for the
-future. This calibration can be carried out periodically and frequently to
-improve the quality of the estimation over longer periods of time.
+solve this using a sample count done manually at each location.  We can then
+calculate an  adjustment factor or an offset for each location by comparing the
+sensor based counts and ground truth and in turn it can be used to adjust the
+data reliably to reflect the ground truth in absolute numbers for the future.
+This calibration can be carried out periodically at these locations to improve
+the quality of the estimation.

### 2018-11-21 paper/sections/pilot_study.tex
Wed 21 Nov 17:08:08 GMT 2018

@@
 	\midrule
-	 Quantile					& 0.002 	&  27\% \\
-	 K-Means			 		& 0.007 	& -23\% \\
-	 Hierarchical Clustering	& 172.520 	& -9\% \\
-	 Bagged Clustering 			& 0.135 	& -30\% \\
-	 Fisher 					& 3.034 	& -30\% \\
-	 Jenks Natural Break 		& 556.279 	& -30\% \\
+	 Quantile					& 0.002 	&  27 \% \\
+	 K-Means			 		& 0.007 	& -23 \% \\
+	 Hierarchical Clustering	& 172.520 	&  -9 \% \\
+	 Bagged Clustering 			& 0.135 	& -30 \% \\
+	 Fisher 					& 3.034 	& -30 \% \\
+	 Jenks Natural Break 		& 556.279 	& -30 \% \\
 	 \bottomrule
@@
 requests during the half hour period, and 3,722 people were recorded manually
-walking on the sidewalk during that time.
+walking on the sidewalk during that time. The count was done at the location in
+front of the store where the surveyor carried the sensor on a backpack for the
+automated count and used a phone for doing the manual count. The sensor is kept
+as close to the store window as possible and the manual count is done as a
+cordon count in fron of the store.
 

### 2018-11-21 paper/sections/previous_work.tex
Wed 21 Nov 17:08:08 GMT 2018

@@
 environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
-mobile devices with Wi-Fi antennae regularly broadcast a special type of signal
-known as `Probe Requests' in order to discover Wi-Fi networks available to them.
-This helps these devices to connect and switch between the Wi-Fi networks
-seamlessly.
+mobile devices with Wi-Fi network interface regularly broadcast a special type
+of signal known as `Probe Requests' in order to discover Wi-Fi networks
+available to them.  This helps these devices to connect and switch between the
+Wi-Fi networks seamlessly.
 
-Probe requests are low level signals standardised by IEEE 802.1b/g specification
-\citep{ieee2013} as the first step in establishing a Wi-Fi based connection
-between two devices and is implemented in any Wi-Fi capable device irrespective
-of the manufacturer or the model. This ubiquity and standardisation makes them
-an excellent source of open, passive, continuous, and wireless data generated by
-Wi-Fi capable devices present at any given time and location. Considering the
-unprecendented levels of mobile device ownership in recent years, we can, in
-turn use this data to understand the population distribution in highly dynamic
-urban environments with high spatial and temporal granularity
-\citep{freud2015,konto2017}. While a Wi-Fi based method to collect data offers
-us various advantages such as, easy scalability and efficiency in terms of cost
-and time, It also introduces few systematic biases, uncertainities in the
-collected data along with the serious risk of infringing on the privacy of the
-mobile users. In this paper, using a set of probe requests and manual counts
-collected at various high street locations across London, we demonstrate that
-pedestrian footfall at these locations can be estimated with considerable
-precision and accuracy while protecting the privacy of the pedestrians.
+Probe requests are low level signals standardised by IEEE 802.11 specification
+\citep{ieee2016} for service discovery and is implemented in any Wi-Fi capable
+device irrespective of the manufacturer or the model. This ubiquity and
+standardisation makes them an excellent source of open, passive, continuous,
+and wireless data generated by Wi-Fi capable devices present at any given time
+and location. Considering the unprecendented levels of mobile device ownership
+in recent years, we can, in turn use this data to understand the population
+distribution in highly dynamic urban environments with high spatial and
+temporal granularity \citep{freud2015,konto2017}. While a Wi-Fi based method to
+collect data offers us various advantages such as, easy scalability and
+efficiency in terms of cost and time, It also introduces few systematic biases,
+uncertainities in the collected data along with the serious risk of infringing
+on the privacy of the mobile users. In this paper, using a set of probe
+requests and manual counts collected at various high street locations across
+London, we demonstrate that pedestrian footfall at these locations can be
+estimated with considerable precision and accuracy while protecting the privacy
+of the pedestrians.
 

### 2018-11-21 analysis/rotating_sequence.r

@@ -1,0 +1,16 @@
+rm(list=ls())
+options(width = as.integer(Sys.getenv("COLUMNS")))
+
+library(tidyverse)
+
+data <- read.csv("../data/final_sensor_data.csv") %>% mutate(id=1:60040,time = as.POSIXct(time, format = "%b %d, %Y %H:%M:%OS")) %>% filter(type == "local") %>% filter(vendor == "Google") %>% select(id, time,seq = sequence)
+
+dist_fun <- function(a, b) { return( ifelse (a > b, 4096 - a + b, 0) ) }
+
+mat <- data %>% pull(id) %>% expand.grid(.,.) %>% 
+	rename(x=Var1,y=Var2) %>% mutate(x=as.numeric(x), y=as.numeric(y)) %>%
+	filter(x<y) %>%
+	left_join(data, by = c("x" = "id")) %>%
+	left_join(data, by = c("y" = "id")) %>%
+	mutate(tdis = as.numeric(time.y - time.x), sdis = dist_fun(seq.x,seq.y) ) %>%
+	filter(tdis < 17, sdis < 61, sdis != 0 )

### 2018-11-21 review/response.md
Wed 21 Nov 17:10:16 GMT 2018

@@
 > measurement of footfall is relevant to understanding the functioning of the
-> Smart City.
-> 
-> If any of what follows and what has been uploaded is unclear, please do not
-> hesitate to contact me.
+> Smart City. If any of what follows and what has been uploaded is unclear,
+> please do not hesitate to contact me.
 > 
 > Yours sincerely,
+>
 > Balamurugan Soundararaj,
+>
 > (on behalf also of Paul Longley and James Cheshire)

### 2018-11-21 review/response.md
Wed 21 Nov 17:11:26 GMT 2018
Pre final

@@
 > Yours sincerely,  
->
 > Balamurugan Soundararaj,  
->
 > (on behalf also of Paul Longley and James Cheshire)
@@
 
-5. How was the sensor positioned compared to the area monitored used for 
-ground truth ? A more detailed description is required regarding the positioning
-of the sensor and the size and topology of the surveyed area.
+5. How was the sensor positioned compared to the area monitored used for ground
+truth ? A more detailed description is required regarding the positioning of the
+sensor and the size and topology of the surveyed area.
 
@@
 > Thank you. This part has been expanded and clarified in the revised manuscript. 
-> The performance of the algorithms used for one dimensional classification were 
+> The performance of the algorithms used for one dimensional clustering were 
 > evaluated on the basis of the time taken for processing and the resulting 

### 2018-11-21 paper/sections/introduction.tex
Pre final

@@
 technical appraisal of our consumer data source.  In technical terms, screening
-the information present in the `probes requests' and classifying them based on
+the information present in the `probes requests' and clustering them based on
 their characteristics is essential in order to remove the ones emitted by

### 2018-11-21 paper/sections/methodology.tex
Pre final

@@
 anonymised probe requests. Though the recycling of sequence number after 4096
-leads to multiple unique ids being reported from a single device, the magnitude
-of error is greatly reduced.
+leads to multiple unique ids being reported from a single device from a sample
+experiment done on randomised probe requests sent by "Google" devices we found
+that the sequence number is reset for around 0.05\% of them. We hence assume
+this to be inconsequential.
 

### 2018-11-21 paper/sections/pilot_study.tex
Pre final

@@
 pedestrian counts within a MAPE of 20\%. We also found `k-means' and `quantile'
-are best algorithms for classifying signal strengths and the threshold for time
+are best algorithms for clustering signal strengths and the threshold for time
 and sequence numbers for the clustering algorithm is around 16 and 60

### 2018-11-21 analysis/main_study.r

@@ -5,7 +5,7 @@ library(igraph)
 library(classInt)
 library(lubridate)
 
-connection <- dbConnect(dbDriver('PostgreSQL'), dbname = 'bala_ijgis', user = 'bala_read', password = 'readpasswordbala', host = 'cdrc-footfall.geog.ucl.ac.uk')
+connection <- dbConnect(dbDriver('PostgreSQL'), dbname = 'bala_ijgis', user = 'bala_read', password = 'balareadpassword', host = 'cdrc-footfall.geog.ucl.ac.uk')
 manual_schedule <- dbGetQuery(connection,"select new_sensor as sensor, min(date_trunc('minute',timestamp::timestamp)+interval '1 minute') as start, max(date_trunc('minute',timestamp::timestamp)) as end from manual left join locations on locations.id = manual.location group by date(timestamp),new_sensor order by 1;")
 if(!exists("sensor_data")){ 
 	sensor_data <- (function(){

### 2018-11-25 review/response.md
changes

@@
 
-> Thank you for the detailed instructions. All of these have been checked in
-> revised manuscript for the final submission.
+> Thank you for the detailed instructions. All of these have been taken in
+> consideration and checked for creating the revised manuscript for the final
+> submission.
 
@@
 
-1. The idea of clustering frames using sequence number has been introduced 
-in [1]. This work needs to be cited when introducing the proposed clustering
+1. The idea of clustering frames using sequence number has been introduced in
+   [1]. This work needs to be cited when introducing the proposed clustering
    algorithm.
@@
 
-> Thank you. We have included a citiation to the source, as well as one 
-> other recent addition to this fast-developing literature.
+> Thank you. We have included a citiation to the source, as well as one other
+> recent addition to this fast-developing literature.
 
@@
    anonymization technique (see [2] and
-http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/ ); 
-at best it can be considered as an obfuscation. 
-[2] Levent Demir, Mathieu Cunche, Cédric Lauradoux. Analysing the privacy 
-policies of Wi-Fi trackers. Workshop on Physical Analytics, Jun 2014, Bretton 
-Woods, United States. ACM, 2014
+   http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/
+   ); at best it can be considered as an obfuscation.  [2] Levent Demir, Mathieu
+   Cunche, Cédric Lauradoux. Analysing the privacy policies of Wi-Fi trackers.
+   Workshop on Physical Analytics, Jun 2014, Bretton Woods, United States. ACM,
+   2014
 

### 2018-11-25 review/response.md
Sun 25 Nov 22:14:38 GMT 2018

@@
 
-This paper is about the estimation of pedestrian footfall using Wi-Fi signal
+1. This paper is about the estimation of pedestrian footfall using Wi-Fi signal
    emitted by mobile devices.  It deals with the problem of estimating the
-footfall size based on a noisy and biased data. More specifically it addresses
-the following issues that may lead to over counting:
+   footfall size based on a noisy and biased data. More specifically it
+   addresses the following issues that may lead to over counting:
 
@@
 
-The adjustment factor is calibrated for each location using a manual counting.
-The author conducted a pilot experiment to identify the clustering algorithms
-and parameters, and then they conducted a second range of experiments at
-different location over an extended period of time. During each of those
-experiments, ground-truth count was collected manually.  The results show that
-the presented approach can significantly reduce the counting error compared to
-a naive solution consisting in counting the distinct number of MAC observed.
-Nevertheless, there are cases (2) where the error is still large (50%), even if
-in other cases (1) the error is only of 9%.
+2. The adjustment factor is calibrated for each location using a manual
+   counting.  The author conducted a pilot experiment to identify the clustering
+   algorithms and parameters, and then they conducted a second range of
+   experiments at different location over an extended period of time. During
+   each of those experiments, ground-truth count was collected manually.  The
+   results show that the presented approach can significantly reduce the
+   counting error compared to a naive solution consisting in counting the
+   distinct number of MAC observed.  Nevertheless, there are cases (2) where the
+   error is still large (50%), even if in other cases (1) the error is only of
+   9%.
 
@@
 
-> Thank you for pointing this out. In the revised version of the paper
-> we have made reference to this discussion, while also making clear the 
-> anonymisation is ensured by the system architecture that processess and
-> curates the data. Specifically we rotated the random salt used to hash the
-> MAC addresses regulary to prevent any kind of tracking of the users across a
-> longer period of time with which the hashing process is carried out.
+> Thank you for pointing this out. In the revised version of the paper we have
+> made reference to this discussion, while also making clear the anonymisation
+> is ensured by the system architecture that processess and curates the data.
+> Specifically we rotated the random salt used to hash the MAC addresses
+> regulary to prevent any kind of tracking of the users across a longer period
+> of time with which the hashing process is carried out.
 
@@
 > The identification of a MAC address as randomised or non-randomised is done
-> looking at the second character of the vendor part of the MAC address.
-> The addresses with E, A, 2, 6 at these locations are identified to be
-> randomised MAC addresses. This has been included in the revised paper.
+> looking at the second character of the vendor part of the MAC address.  The
+> addresses with E, A, 2, 6 at these locations are identified to be randomised
+> MAC addresses. This has been included in the revised paper.
 
 5. How was the sensor positioned compared to the area monitored used for ground
-truth ? A more detailed description is required regarding the positioning of the
-sensor and the size and topology of the surveyed area.
+   truth ? A more detailed description is required regarding the positioning of
+   the sensor and the size and topology of the surveyed area.
 
-> This is a good point, and is very important for our validation work. We
-> have included a more detailed and explicit discussion of the way that field
+> This is a good point, and is very important for our validation work. We have
+> included a more detailed and explicit discussion of the way that field
 > observation was used to calibrate and cross validate the Wi-Fi results.
@@
    conducted field experiments to evaluate their solution is a positive point.
-Yet before publication, I believe that there are a number of points that needs
-to be clarified or corrected. Regarding the technical aspects I have a number
-of minor remarks:
+   Yet before publication, I believe that there are a number of points that
+   needs to be clarified or corrected. Regarding the technical aspects I have a
+   number of minor remarks:
 
 > Thank you very much for these very positive comments. The paper reports on
-> extensive methodological research and field implementation and is, we
-> believe, cutting edge in its finding. It also has important implications for
-> our understanding of the functioning of smart cities, and these implications
-> are discussed in greater detail in the revised version.
+> extensive methodological research and field implementation and is, we believe,
+> cutting edge in its finding. It also has important implications for our
+> understanding of the functioning of smart cities, and these implications are
+> discussed in greater detail in the revised version.
 
@@
    the source. In this context it is better to use a more general concept and
-refer to the "Wi-Fi network interface" (or Wi-Fi interface) as the the element
-broadcasting a signal.
+   refer to the "Wi-Fi network interface" (or Wi-Fi interface) as the the
+   element broadcasting a signal.
 
@@
    can refer to the 802.11 specifications as a whole : [“802.11-2012 - IEEE
-Standard for Information technology–Telecommunications and information exchange
-between systems Local and metropolitan area networks–Specific requirements Part
-11: Wireless LAN Medium Access Control (MAC) and Physical Layer (PHY)
-Specifications,” IEEE Std 802.11-2012 (Revision of IEEE Std 802.11-2007), pp.
-1–2793, Mar. 2012.
+   Standard for Information technology–Telecommunications and information
+   exchange between systems Local and metropolitan area networks–Specific
+   requirements Part 11: Wireless LAN Medium Access Control (MAC) and Physical
+   Layer (PHY) Specifications,” IEEE Std 802.11-2012 (Revision of IEEE Std
+   802.11-2007), pp.  1–2793, Mar. 2012.
 
@@
 
-> This simple application was developed specifically for the project for
-> Android platform. More details on the application has been included in the
-> revised manuscript.
+> This simple application was developed specifically for the project for Android
+> platform. More details on the application has been included in the revised
+> manuscript.
 
@@
 	classifications reported on single device, the magnitude of error is greatly
-reduced." This sentence is not clear. And the cycle length is not 4000 but 4096
-= 2^12.
+	reduced." This sentence is not clear. And the cycle length is not 4000 but
+	4096 = 2^12.
 

### 2018-11-25 review/response.md
Sun 25 Nov 22:25:41 GMT 2018

@@
 
-8. You need to check that the paper title in ScholarOne is the same as the
-title on the paper and that the Title page contains all authors’ names and
+8. You need to check that the paper title in ScholarOne is the same as the title
+   on the paper and that the Title page contains all authors’ names and
    affiliations as you would want them to appear.
@@
 > Thank you very much for the suggestion. From tests carried out on sample of
-> 1910 probe requests from google phones in the pilot study we found that
-> around 0.5% reset their mac addresses.  Which in our opinion doesn't
-> influence the results in any significant way considering the context.
+> 1910 probe requests from google phones in the pilot study we found that around
+> 0.5% reset their mac addresses. Which in our opinion doesn't influence the
+> results in any significant way considering the context.
 
@@
 
-3. "Though Wi-Fi F is a ‘location-less’ technology, ", please clarify what 
-you mean by that.
+3. "Though Wi-Fi F is a ‘location-less’ technology, ", please clarify what you
+   mean by that.
 
@@
 6. "We hypothesize that an adjustment factor could be arrived at for each
-location of data collection, ""This calibration can be carried over periodically
-and the frequency to improve the quality of the estimation." something wrong in 
-both sentences.
+   location of data collection, ""This calibration can be carried over
+   periodically and the frequency to improve the quality of the estimation."
+   something wrong in both sentences.
 
@@
    the accuracy of estimating the number of unique mobile devices from a set of
-anonymized probe requests without revealing their original device information.
+   anonymized probe requests without revealing their original device
+   information.
 
@@
 2. This is a crucial topic in mobile data uncertainty, and the manuscript has
-the potential of becoming a useful reference in the field. However, it was not 
-organized effectively and missing many details in the methodology.  Before
+   the potential of becoming a useful reference in the field. However, it was
+   not organized effectively and missing many details in the methodology. Before
    addressing these problems, it is difficult to assess the reliability and
@@
 
-> Thank you very much for this endorsement of the motivation for our paper.
-> We hope that following responses and the amendments to the paper will allay
-> your concerns about the methodology and application.
+> Thank you very much for this endorsement of the motivation for our paper.  We
+> hope that following responses and the amendments to the paper will allay your
+> concerns about the methodology and application.
 
@@
 
-> This is a welcome comment, and we have added material in the
-> introduction and conclusion as to the value of new Big Data sources for
-> developing activity-based representations of behaviour and circumstances
-> in the smart city, This chimes with much of our related research, but we
-> write cognisant of the article length restrictions of IJGIS.
+> This is a welcome comment, and we have added material in the introduction and
+> conclusion as to the value of new Big Data sources for developing
+> activity-based representations of behaviour and circumstances in the smart
+> city, This chimes with much of our related research, but we write cognisant of
+> the article length restrictions of IJGIS.
 
@@
 
-> Yes, we agree, and some wider conclusions about the implications of
-> our research for better understanding the functioning of the smart city
-> have now been added.
+> Yes, we agree, and some wider conclusions about the implications of our
+> research for better understanding the functioning of the smart city have now
+> been added.
 
@@
 1. P4 l50, without providing any explanation, the authors hypothesized that
-sequence number and length of the packet are sufficient to estimate the number 
-of unique devices. This argument was not elaborated until 2.5 pages later 
-in section 3.2.
+   sequence number and length of the packet are sufficient to estimate the
+   number of unique devices. This argument was not elaborated until 2.5 pages
+   later in section 3.2.
 
@@
    clustering is considered a “classification algorithm” as claimed in p8 l26.
-Very few details were provided regarding the parameter setting of 
-these analyses.
+   Very few details were provided regarding the parameter setting of these
+   analyses.
 
-> Thank you. The reference to clustering has been amended. The
-> specifics of the thresholds used have been included, and the techniques
-> justified in more detail.
+> Thank you. The reference to clustering has been amended. The specifics of the
+> thresholds used have been included, and the techniques justified in more
+> detail.
 
@@
 > change is also found to be exponential as opposed to linear. With this context
-> When we install a sensor to have a clear view of the pedestrian footpath
-> in front of it, we make sure that all the devices beyond a certain distance
-> will report similarly low signal strength. This threhold is idenfied using
-> the kmeans algorithm to find the natural break in the data. We have found that
+> When we install a sensor to have a clear view of the pedestrian footpath in
+> front of it, we make sure that all the devices beyond a certain distance will
+> report similarly low signal strength. This threhold is idenfied using the
+> kmeans algorithm to find the natural break in the data. We have found that
 > this successfully isolates a farther stationary source of large number of
@@
 > The recycling of the sequence number occurs every 4096 counts. From our
-> experiments we observed that a device staying next to device actively searching
-> recycles the sequence number 1-2 times in a 30 mins interval. On average
-> this is expected to be much less than this. when the device doesn't probe
-> as frequently. We have included more details regarding the cycling of sequence
-> numbers and have mentioned method to overcomet this as mentioned by other
-> reviewer. The change in implementing the method is estimated to be --% which
-> we found to be not significant in our context.
+> experiments we observed that a device staying next to device actively
+> searching recycles the sequence number 1-2 times in a 30 mins interval. On
+> average this is expected to be much less than this. when the device doesn't
+> probe as frequently. We have included more details regarding the cycling of
+> sequence numbers and have mentioned method to overcomet this as mentioned by
+> other reviewer. The change in implementing the method is estimated to be --%
+> which we found to be not significant in our context.
 
-5. P7 Section 3.3, please be more specific about the calibration process and 
-the external source of information to be used here.
+5. P7 Section 3.3, please be more specific about the calibration process and the
+   external source of information to be used here.
 
-> This section has been expanded to include more details. The calibration process
-> is adjusting the future counts based on a sample manual counts done at the
-> location. The external source of information used is the manual counts which is
-> collected outside the sensor
+> This section has been expanded to include more details. The calibration
+> process is adjusting the future counts based on a sample manual counts done at
+> the location. The external source of information used is the manual counts
+> which is collected outside the sensor
 
 6. P7 l45, the analysis in this research is based on one single sensor for each
-location; however, in reality, it is very common to estimate travel flows based 
-on multiple sensors (e.g., nearby stores in a shopping mall). How would your 
-method address duplication in multiple sensors?
+   location; however, in reality, it is very common to estimate travel flows
+   based on multiple sensors (e.g., nearby stores in a shopping mall). How would
+   your method address duplication in multiple sensors?
 
@@
 > locations, and examined correspondence with retail unit function. This enables
-> identification of the differences in dwell time between, for example,
-> fast food outlets and clothes shops. This full analysis requires
-> addition of store till data and the analysis is detailed. For this
-> reason we include only a short reference to this extension of our
-> analysis here. We also aim this method to count footfall accurately at 
-> locations rather than the flow of footfall between those locations. But as 
-> demonstarted in
+> identification of the differences in dwell time between, for example, fast
+> food outlets and clothes shops. This full analysis requires addition of store
+> till data and the analysis is detailed. For this reason we include only a
+> short reference to this extension of our analysis here. We also aim this
+> method to count footfall accurately at locations rather than the flow of
+> footfall between those locations. But as demonstarted in
 
 > Lee J, Nemati S, Silva I, Edwards BA, Butler JP, Malhotra A. Transfer entropy
-> estimation and directional coupling change detection in biomedical time series.
-> Biomed Eng Online. 2012;11:19. Published 2012 Apr 13.
+> estimation and directional coupling change detection in biomedical time
+> series.  Biomed Eng Online. 2012;11:19. Published 2012 Apr 13.
 > doi:10.1186/1475-925X-11-19
@@
 
-> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free 
-> measure of effective connectivity for the neurosciences.
-> J Comput Neurosci. 2010;30(1):45-67.
+> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free measure
+> of effective connectivity for the neurosciences.  J Comput Neurosci.
+> 2010;30(1):45-67.
 
@@
 > The adjustment factor is calculated as the ratio between the manual counts to
-> the counts calculated from sensor for the sample period. This 
-> is used to adjust future counts at the same locations. We have included
-> the clarification on the revised manuscript.
+> the counts calculated from sensor for the sample period. This is used to
+> adjust future counts at the same locations. We have included the clarification
+> on the revised manuscript.
 
@@
 1. It would have been interesting to understand the temporal characteristics for
-a week pedestrian footfall. How to consider this situation in your study? Why do
-you choose time range from 12:30 to 13:00 hrs ?
+   a week pedestrian footfall. How to consider this situation in your study? Why
+   do you choose time range from 12:30 to 13:00 hrs ?
 
@@
 
-> There was no reason for
-> selecting the time in the pilot study. the location was chosen for it it known
-> to have high number of randomised probe requests and has one of the highest
-> footfall in London. For the main study the counting was done such that each
-> location has atleast 2/3 counts in the same time interval.
+> There was no reason for selecting the time in the pilot study. the location
+> was chosen for it it known to have high number of randomised probe requests
+> and has one of the highest footfall in London. For the main study the counting
+> was done such that each location has atleast 2/3 counts in the same time
+> interval.
 
@@
    various clustering algorithms? Including the number of records, the number of
-clusters. How about the performance for different clustering methods 
-in your study？
+   clusters. How about the performance for different clustering methods in your
+   study？
 
-> Thank you. This part has been expanded and clarified in the revised manuscript. 
-> The performance of the algorithms used for one dimensional clustering were 
-> evaluated on the basis of the time taken for processing and the resulting 
-> mean absolute percentage error. We found that the k-means gives
+> Thank you. This part has been expanded and clarified in the revised
+> manuscript.  The performance of the algorithms used for one dimensional
+> clustering were evaluated on the basis of the time taken for processing and
+> the resulting mean absolute percentage error. We found that the k-means gives
 > us the best data for the quickest processing time. The graph based clustering
 > was done using two parameters - sequence threshold of 60 and time threshold of
-> 16 seconds. These values were arrived arrived at using trial and error by using
-> a know device as the control mechanism. This has been illustrated further in 
-> the revised manuscript.
+> 16 seconds. These values were arrived arrived at using trial and error by
+> using a know device as the control mechanism. This has been illustrated
+> further in the revised manuscript.
 
@@
 
-1. Minor formatting issues and typos needs to be fixed.
-Figure 2, no value and units in the x-axis and y-axis.
-Table 1, and Figure 4 are in wrong location.
+1. Minor formatting issues and typos needs to be fixed.  Figure 2, no value and
+   units in the x-axis and y-axis.  Table 1, and Figure 4 are in wrong location.
 
@@
 > Figure 2 is just a schematic showing the idea of cut off and clustering hence
-> there are now values on the axes. We have included the units in the description
+> there are now values on the axes. We have included the units in the
+> description
 
@@
 
-> We have included reference to such work in our extended discussion of
-> the relevance of the work to the smart cities discourse.
+> We have included reference to such work in our extended discussion of the
+> relevance of the work to the smart cities discourse.
 
@@
 
-> We have resequenced some of the material in response to Referee 2
-> and, cognisant of over-all length requirements, think that this (rather
-> than a diagram) addresses your concerns.
+> We have resequenced some of the material in response to Referee 2 and,
+> cognisant of over-all length requirements, think that this (rather than a
+> diagram) addresses your concerns.
 
 4. Performance Analysis: How about the performance for collecting data using
-PostgreSQL? Did you compare Apache Flink and Spark Streaming for real-time data 
-collecting?
+   PostgreSQL? Did you compare Apache Flink and Spark Streaming for real-time
+   data collecting?
 
@@
 > unfeasible, it is neither big nor fast enough for a large scale big data
-> system. An opensource and free toolkit comprising of text processing tools 
-> and relational database management system was enough 
-> to handle the compexity, size and speed of the data.
-> The toolkit employed here involves PostgreSQL, Bash, Unix tools
-> for text processing (sed, awk, sort, join), Node.js and R.
-> The same toolkit proved to be useful in the larger project 
-> (smart street sensor) providing near real-time delivery of data.
+> system. An opensource and free toolkit comprising of text processing tools and
+> relational database management system was enough to handle the compexity, size
+> and speed of the data.  The toolkit employed here involves PostgreSQL, Bash,
+> Unix tools for text processing (sed, awk, sort, join), Node.js and R.  The
+> same toolkit proved to be useful in the larger project (smart street sensor)
+> providing near real-time delivery of data.
 
 5. The reference list needs to be written in a standard format. Some reference
-papers are lacking specific pages .Please use a standard way of reference list.
+   papers are lacking specific pages .Please use a standard way of reference
+   list.
 

### 2018-11-26 review/response.md
Sun 25 Nov 22:26:39 GMT 2018
Sun 25 Nov 22:29:59 GMT 2018
ready to send for proof read

@@
-First review comments
-=====================
+## First review comments
 
@@
 
-1 Editor
---------
+## 1 Editor
 
-1.1 Overall Comments
----------------------
+### 1.1 Overall Comments
 
@@
 > 
-> Yours sincerely,  
-> Balamurugan Soundararaj,  
-> (on behalf also of Paul Longley and James Cheshire)
+> Yours sincerely,  Balamurugan Soundararaj,  (on behalf also of Paul Longley
+> and James Cheshire)
 
-1.2 General Instructions
-------------------------
+### 1.2 General Instructions
 
@@
 
-2. Please also highlight the changes to your manuscript within the document by
-   using the track changes mode in MS Word or by using bold or colored text.
+2. Please also highlight the changes to your manuscript within the document
+by using the track changes mode in MS Word or by using bold or coloured text.
 
@@
 https://mc.manuscriptcentral.com/ijgis?URL_MASK=a32caf3bbd49412cac8d1cfa9d7ebaf8
-   which will take you your decision letter, and where you need to resubmit
-   your manuscript.
+which will take you your decision letter, and where you need to resubmit your
+manuscript.
 
@@
 	  width) in each image must be less than 40,000,000 (40 mp).
-	- Total file size - please make sure that the combined, uncompressed, file
-	sizes do not exceed 20Mb. The internal LZW compression, which is an option
-	in TIF, is acceptable however.
+	- Total file size - please make sure that the combined, uncompressed,
+	  file sizes do not exceed 20Mb. The internal LZW compression, which is
+an option in TIF, is acceptable however.
 	- Please do not save B/W or Greyscale images as RGB.
@@
 
-8. You need to check that the paper title in ScholarOne is the same as the title
-   on the paper and that the Title page contains all authors’ names and
+8. You need to check that the paper title in ScholarOne is the same as the
+title on the paper and that the Title page contains all authors’ names and
 affiliations as you would want them to appear.
 
-9. When submitting your revised manuscript, you will be able to respond to the
-   comments made by the reviewers in the space provided. You can use this space
-   to document any changes you make to the original manuscript. In order to
-   expedite the processing of the revised manuscript, please be as specific as
-   possible in your response to the reviewers.
+9. When submitting your revised manuscript, you will be able to respond to
+the comments made by the reviewers in the space provided. You can use this
+space to document any changes you make to the original manuscript. In order
+to expedite the processing of the revised manuscript, please be as specific
+as possible in your response to the reviewers.
 
@@
 
-2 Reviewer 1
-------------
+## 2 Reviewer-1
 
-2.1 Overall Summary
--------------------
+### 2.1 Overall Summary
 
-1. This paper is about the estimation of pedestrian footfall using Wi-Fi signal
-   emitted by mobile devices. It deals with the problem of estimating the
+1. This paper is about the estimation of pedestrian footfall using Wi-Fi
+signal emitted by mobile devices. It deals with the problem of estimating the
 footfall size based on a noisy and biased data. More specifically it
@@
 
- - MAC randomization that may cause a device to generate several MAC addresses;
+	- MAC randomization that may cause a device to generate several MAC
+	   addresses;
 	- Signals received from outside the interest zone; unequal adoption of Wi-Fi
- smartphone in the population. To address this issue, a number of techniques
- are introduced:
+	  smartphone in the population. To address this issue, a number of
+	  techniques are introduced:
 	- clustering received signal strength to remove device that are outside the
@@
 > This is a correct surmisal of the remit of the paper. We have also added
-> material to the introduction and conclusion of the paper to better link this
-> remit to the Smart Cities discourse.
+> material to the introduction and conclusion of the paper to better link
+> this remit to the Smart Cities discourse.
 
-2.2 General Comments
---------------------
+### 2.2 General Comments
 
@@
 anonymization technique (see [2] and
-   http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/
-   ); at best it can be considered as an obfuscation. [2] Levent Demir, Mathieu
+http://webpolicy.org/2014/03/19/questionable-crypto-in-retail-analytics/); at
+best it can be considered as an obfuscation. [2] Levent Demir, Mathieu
 Cunche, Cédric Lauradoux. Analysing the privacy policies of Wi-Fi trackers.
@@
 
-> Thank you for pointing this out. In the revised version of the paper we have
-> made reference to this discussion, while also making clear the anonymisation
-> is ensured by the system architecture that processess and curates the data.
-> Specifically we rotated the random salt used to hash the MAC addresses
-> regulary to prevent any kind of tracking of the users across a longer period
-> of time with which the hashing process is carried out.
+> Thank you for pointing this out. In the revised version of the paper we
+> have made reference to this discussion, while also making clear the
+> anonymisation is ensured by the system architecture that processess and
+> curates the data.  Specifically we rotated the random salt used to hash the
+> MAC addresses regulary to prevent any kind of tracking of the users across
+> a longer period of time with which the hashing process is carried out.
 
@@
 > looking at the second character of the vendor part of the MAC address. The
-> addresses with E, A, 2, 6 at these locations are identified to be randomised
-> MAC addresses. This has been included in the revised paper.
+> addresses with E, A, 2, 6 at these locations are identified to be
+> randomised MAC addresses. This has been included in the revised paper.
 
-5. How was the sensor positioned compared to the area monitored used for ground
-   truth ? A more detailed description is required regarding the positioning of
-   the sensor and the size and topology of the surveyed area.
+5. How was the sensor positioned compared to the area monitored used for
+ground truth ? A more detailed description is required regarding the
+positioning of the sensor and the size and topology of the surveyed area.
 
-> This is a good point, and is very important for our validation work. We have
-> included a more detailed and explicit discussion of the way that field
+> This is a good point, and is very important for our validation work. We
+> have included a more detailed and explicit discussion of the way that field
 > observation was used to calibrate and cross validate the Wi-Fi results.
 
-2.3 Clarifications & Corrections
---------------------------------
+### 2.3 Clarifications & Corrections
 
@@
 > Thank you very much for these very positive comments. The paper reports on
-> extensive methodological research and field implementation and is, we believe,
-> cutting edge in its finding. It also has important implications for our
-> understanding of the functioning of smart cities, and these implications are
-> discussed in greater detail in the revised version.
+> extensive methodological research and field implementation and is, we
+> believe, cutting edge in its finding. It also has important implications
+> for our understanding of the functioning of smart cities, and these
+> implications are discussed in greater detail in the revised version.
 
-2. "WiFi beacon technologies to access the Internet". In Wi-Fi, beacons are
+2. "Wi-Fi beacon technologies to access the Internet". In Wi-Fi, beacons are
 management frames used by Access Point for service discovery. They do not
@@
 
-3. "Wi-Fi antennae regularly broadcast a special type of signal" -> the antenna
-   is just the part ensuring that the signal is correctly emitted, but it is not
-   the source. In this context it is better to use a more general concept and
-   refer to the "Wi-Fi network interface" (or Wi-Fi interface) as the the
-   element broadcasting a signal.
+3. "Wi-Fi antennae regularly broadcast a special type of signal" -> the
+antenna is just the part ensuring that the signal is correctly emitted, but
+it is not the source. In this context it is better to use a more general
+concept and refer to the "Wi-Fi network interface" (or Wi-Fi interface) as
+the the element broadcasting a signal.
 
@@
 
-4. The IEEE 802.11 b/g is just a subpart of the 802.11 specification. Here you
-   can refer to the 802.11 specifications as a whole : [“802.11-2012 - IEEE
+4. The IEEE 802.11 b/g is just a sub part of the 802.11 specification. Here
+you can refer to the 802.11 specifications as a whole : [“802.11-2012 - IEEE
 Standard for Information technology–Telecommunications and information
@@
 
-5. "the first step in establishing a Wi-Fi based connection " is also known as
-   service discovery
+5. "the first step in establishing a Wi-Fi based connection " is also known
+as service discovery
 
@@
 
-6. "Wi-Fi functionality has been turned off by the user", regarding this point
-   you may be interested in :
+6. "Wi-Fi functionality has been turned off by the user", regarding this
+point you may be interested in :
 
@@
 
-> This simple application was developed specifically for the project for Android
-> platform. More details on the application has been included in the revised
-> manuscript.
+> This simple application was developed specifically for the project for
+> Android platform. More details on the application has been included in the
+> revised manuscript.
 
@@
 
-> Thank you very much for the suggestion. From tests carried out on sample of
-> 1910 probe requests from google phones in the pilot study we found that around
-> 0.5% reset their mac addresses. Which in our opinion doesn't influence the
-> results in any significant way considering the context.
+> Thank you very much for the suggestion. We carried out some tests on a
+> sample data of 1910 probe requests corresponding to just google phones with
+> randomised MAC addresses captured in the pilot study and found that only
+> around 0.5% of the probe requests encountered the recycling of sequence
+> number. Which, in our opinion, doesn't influence the results in any
+> significant way considering the context. We have also included this in our
+> revised manuscript for the benefit of future research.
 
-2.4 Typos and other remarks
---------------------------------
+### 2.4 Typos and other remarks
 
@@
 
-3. "Though Wi-Fi F is a ‘location-less’ technology, ", please clarify what you
-   mean by that.
+3. "Though Wi-Fi F is a ‘location-less’ technology, ", please clarify what
+you mean by that.
 
@@
 
-3 Reviewer 2
-------------
+## 3 Reviewer-2
 
-3.1 Overall Summary
--------------------
+### 3.1 Overall Summary
 
-1. This paper addressed the data quality issue of Wi-Fi-based probe requests in
-   estimating pedestrian counts. The authors proposed several methods to improve
-   the accuracy of estimating the number of unique mobile devices from a set of
-   anonymized probe requests without revealing their original device
+1. This paper addressed the data quality issue of Wi-Fi-based probe requests
+in estimating pedestrian counts. The authors proposed several methods to
+improve the accuracy of estimating the number of unique mobile devices from a
+set of anonymized probe requests without revealing their original device
 information.
 
-> This is correct, although the intended emphasis upon disclosure is a feature
-> rather than a principal motivation. We have also added further discussion of
-> how these principally technical issues relate to smart city data
-> infrastructure issues.
+> This is correct, although the intended emphasis upon disclosure is a
+> feature rather than a principal motivation. We have also added further
+> discussion of how these principally technical issues relate to smart city
+> data infrastructure issues.
 
@@
 
-> Thank you very much for this endorsement of the motivation for our paper.  We
-> hope that following responses and the amendments to the paper will allay your
-> concerns about the methodology and application.
+> Thank you very much for this endorsement of the motivation for our paper.
+> We hope that following responses and the amendments to the paper will allay
+> your concerns about the methodology and application.
 
-3.2 General Comments
---------------------
+### 3.2 General Comments
 
@@
 
-> This is a welcome comment, and we have added material in the introduction and
-> conclusion as to the value of new Big Data sources for developing
+> This is a welcome comment, and we have added material in the introduction
+> and conclusion as to the value of new Big Data sources for developing
 > activity-based representations of behaviour and circumstances in the smart
-> city, This chimes with much of our related research, but we write cognisant of
-> the article length restrictions of IJGIS.
+> city, This chimes with much of our related research, but we write cognisant
+> of the article length restrictions of IJGIS.
 
@@
 > Yes, we agree, and some wider conclusions about the implications of our
-> research for better understanding the functioning of the smart city have now
-> been added.
+> research for better understanding the functioning of the smart city have
+> now been added.
 
-3.3 Corrections and Clarifications
-----------------------------------
+### 3.3 Corrections and Clarifications
 
@@
 
-> Thank you. The reference to clustering has been amended. The specifics of the
-> thresholds used have been included, and the techniques justified in more
-> detail.
+> Thank you. The reference to clustering has been amended. The specifics of
+> the thresholds used have been included, and the techniques justified in
+> more detail.
 
@@
 > increasing distance between the mobile device and the base station. This
-> change is also found to be exponential as opposed to linear. With this context
-> When we install a sensor to have a clear view of the pedestrian footpath in
-> front of it, we make sure that all the devices beyond a certain distance will
-> report similarly low signal strength. This threhold is idenfied using the
-> kmeans algorithm to find the natural break in the data. We have found that
-> this successfully isolates a farther stationary source of large number of
-> probe requests such as a phone shop or a restaurant next door, Bustops etc.
-> This doesn't work when the source of error is within the range of the
-> footfall. As shown in our examples 2,4. Here our methodology fails to provide
-> sufficient accuracy.
+> change is also found to be exponential as opposed to linear. With this
+> context When we install a sensor to have a clear view of the pedestrian
+> footpath in front of it, we make sure that all the devices beyond a certain
+> distance will report similarly low signal strength. This threhold is
+> idenfied using the kmeans algorithm to find the natural break in the data.
+> We have found that this successfully isolates a farther stationary source
+> of large number of probe requests such as a phone shop or a restaurant next
+> door, Bustops etc.  This doesn't work when the source of error is within
+> the range of the footfall. As shown in our examples 2,4. Here our
+> methodology fails to provide sufficient accuracy.
 
-4. P7 l21, how often does the recycling of sequence number occur, and how will
-   this impact the reliability of your analysis? Please provide more details.
+4. P7 l21, how often does the recycling of sequence number occur, and how
+will this impact the reliability of your analysis? Please provide more
+details.
 
@@
 > probe as frequently. We have included more details regarding the cycling of
-> sequence numbers and have mentioned method to overcomet this as mentioned by
-> other reviewer. The change in implementing the method is estimated to be --%
-> which we found to be not significant in our context.
+> sequence numbers and have mentioned method to overcomet this as mentioned
+> by other reviewer. The change in implementing the method is estimated to be
+> --% which we found to be not significant in our context.
 
-5. P7 Section 3.3, please be more specific about the calibration process and the
-   external source of information to be used here.
+5. P7 Section 3.3, please be more specific about the calibration process and
+the external source of information to be used here.
 
-> This section has been expanded to include more details. The calibration
-> process is adjusting the future counts based on a sample manual counts done at
-> the location. The external source of information used is the manual counts
-> which is collected outside the sensor
+> Thanks for pointing this out. Following your suggestion, this section has
+> been expanded and clarified to include more details. The calibration is
+> done by adjusting the future counts based on a ratio between sample manual
+> counts and corresponding sensor counts at the location. The external source
+> of information used is the manual counts which is collected outside the
+> sensor
 
-6. P7 l45, the analysis in this research is based on one single sensor for each
-   location; however, in reality, it is very common to estimate travel flows
-   based on multiple sensors (e.g., nearby stores in a shopping mall). How would
-   your method address duplication in multiple sensors?
+6. P7 l45, the analysis in this research is based on one single sensor for
+each location; however, in reality, it is very common to estimate travel
+flows based on multiple sensors (e.g., nearby stores in a shopping mall). How
+would your method address duplication in multiple sensors?
 
-> This is a valid concern. Which we tried to clarify in the revised manuscript.
-> Our motivation is not to attempt to track the movements of individuals,
-> although we do attempt to accommodate measurement variations attributable to
-> micro-site location of the sensors. In related work we have carried out
-> controlled comparison of sensor measures obtained from adjacent store
-> locations, and examined correspondence with retail unit function. This enables
-> identification of the differences in dwell time between, for example, fast
-> food outlets and clothes shops. This full analysis requires addition of store
-> till data and the analysis is detailed. For this reason we include only a
-> short reference to this extension of our analysis here. We also aim this
-> method to count footfall accurately at locations rather than the flow of
-> footfall between those locations. But as demonstarted in
-
-> Lee J, Nemati S, Silva I, Edwards BA, Butler JP, Malhotra A. Transfer entropy
-> estimation and directional coupling change detection in biomedical time
-> series.  Biomed Eng Online. 2012;11:19. Published 2012 Apr 13.
+> This is a valid concern. Which we tried to clarify in the revised
+> manuscript.  Our motivation is not to attempt to track the movements of
+> individuals, although we do attempt to accommodate measurement variations
+> attributable to micro-site location of the sensors. In related work, we
+> have carried out controlled comparison of sensor measures obtained from
+> adjacent store locations, and examined correspondence with retail unit
+> function. This enables identification of the differences in dwell time
+> between, for example, fast food outlets and clothes shops. This full
+> analysis requires addition of store till data and the analysis is detailed.
+> For this reason we include only a short reference to this extension of our
+> analysis here. We also aim this method to count footfall accurately at
+> locations rather than the flow of footfall between those locations. But as
+> demonstrated in we can infer the relation or flow between the locations
+> using other probabilistic approaches.
+>
+> Lee J, Nemati S, Silva I, Edwards BA, Butler JP, Malhotra A. Transfer
+> entropy estimation and directional coupling change detection in biomedical
+> time series. Biomed Eng Online. 2012;11:19. Published 2012 Apr 13.
 > doi:10.1186/1475-925X-11-19
-
-> and
-
-> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free measure
-> of effective connectivity for the neurosciences.  J Comput Neurosci.
+>
+> Vicente R, Wibral M, Lindner M, Pipa G. Transfer entropy--a model-free
+> measure of effective connectivity for the neurociences. J Comput Neurosci.
 > 2010;30(1):45-67.
-
-> we can infer the relation or flow between the locations using other
-> probabilistic approaches.
 
@@
 
-> The adjustment factor is calculated as the ratio between the manual counts to
-> the counts calculated from sensor for the sample period. This is used to
-> adjust future counts at the same locations. We have included the clarification
-> on the revised manuscript.
+> The adjustment factor is calculated as the ratio between the manual counts
+> to the counts calculated from sensor for the sample period. This is used to
+> adjust future counts at the same locations. We have included the
+> clarification on the revised manuscript.
 
-4 Reviewer 3
-------------
+## 4 Reviewer-3
 
-4.1 Overall Summary
--------------------
+### 4.1 Overall Summary
 
@@
 
-4.2 General Comments
---------------------
+### 4.2 General Comments
 
@@
 
-1. It would have been interesting to understand the temporal characteristics for
-   a week pedestrian footfall. How to consider this situation in your study? Why
-   do you choose time range from 12:30 to 13:00 hrs ?
+1. It would have been interesting to understand the temporal characteristics
+for a week pedestrian footfall. How to consider this situation in your study?
+Why do you choose time range from 12:30 to 13:00 hrs ?
 
-> Thank you. Our related work (cited below) examines daily, weekly and seasonal
-> variation, and wil be the subject of future papers.
-
+> Thank you. Our related work (cited below) examines daily, weekly and
+> seasonal variation, and wil be the subject of future papers.
+>
 > Lugomer, K., & Longley, P. (2018). Towards a Comprehensive Temporal
 > Classification of Footfall Patterns in the Cities of Great Britain (Short
-> Paper). In LIPIcs-Leibniz International Proceedings in Informatics (Vol. 114).
-> Schloss Dagstuhl-Leibniz-Zentrum fuer Informatik.
-
+> Paper). In LIPIcs-Leibniz International Proceedings in Informatics (Vol.
+> 114).  Schloss Dagstuhl-Leibniz-Zentrum fuer Informatik.
+>
 > There was no reason for selecting the time in the pilot study. the location
 > was chosen for it it known to have high number of randomised probe requests
-> and has one of the highest footfall in London. For the main study the counting
-> was done such that each location has atleast 2/3 counts in the same time
-> interval.
+> and has one of the highest footfall in London. For the main study the
+> counting was done such that each location has atleast 2/3 counts in the
+> same time interval.
 
-2. Will you give more detailed information about the data and the parameters for
-   various clustering algorithms? Including the number of records, the number of
-   clusters. How about the performance for different clustering methods in your
-   study？
+2. Will you give more detailed information about the data and the parameters
+for various clustering algorithms? Including the number of records, the
+number of clusters. How about the performance for different clustering
+methods in your study？
 
@@
 > clustering were evaluated on the basis of the time taken for processing and
-> the resulting mean absolute percentage error. We found that the k-means gives
-> us the best data for the quickest processing time. The graph based clustering
-> was done using two parameters - sequence threshold of 60 and time threshold of
-> 16 seconds. These values were arrived arrived at using trial and error by
-> using a know device as the control mechanism. This has been illustrated
-> further in the revised manuscript.
+> the resulting mean absolute percentage error. We found that the k-means
+> gives us the best data for the quickest processing time. The graph based
+> clustering was done using two parameters - sequence threshold of 60 and
+> time threshold of 16 seconds. These values were arrived arrived at using
+> trial and error by using a know device as the control mechanism. This has
+> been illustrated further in the revised manuscript.
 
-4.3 Corrections and Clarifications
-----------------------------------
+### 4.3 Corrections and Clarifications
 
@@
 
-1. Minor formatting issues and typos needs to be fixed.  Figure 2, no value and
-   units in the x-axis and y-axis.  Table 1, and Figure 4 are in wrong location.
+1. Minor formatting issues and typos needs to be fixed. Figure 2, no value
+and units in the x-axis and y-axis. Table 1, and Figure 4 are in wrong
+location.
 
-> Thank you. This has been corrected in the revised version of the manuscript.
-> Figure 2 is just a schematic showing the idea of cut off and clustering hence
-> there are now values on the axes. We have included the units in the
-> description
+> Thank you. This has been corrected in the revised version of the
+> manuscript.  Figure 2 is just a schematic showing the idea of cut off and
+> clustering hence there are now values on the axes. We have included the
+> units in the description
 
@@
 
-> This is an interesting point in the context of big data and tools for tackling
-> big data. Though the data set here is large enough to make desktop computing
-> unfeasible, it is neither big nor fast enough for a large scale big data
-> system. An opensource and free toolkit comprising of text processing tools and
-> relational database management system was enough to handle the compexity, size
-> and speed of the data.  The toolkit employed here involves PostgreSQL, Bash,
-> Unix tools for text processing (sed, awk, sort, join), Node.js and R.  The
-> same toolkit proved to be useful in the larger project (smart street sensor)
-> providing near real-time delivery of data.
+> This is an interesting point in the context of big data and tools for
+> tackling big data. Though the data set here is large enough to make desktop
+> computing unfeasible, it is neither big nor fast enough for a large scale
+> big data system. An opensource and free toolkit comprising of text
+> processing tools and relational database management system was enough to
+> handle the compexity, size and speed of the data. The toolkit employed here
+> involves PostgreSQL, Bash, Unix tools for text processing (sed, awk, sort,
+> join), Node.js and R. The same toolkit proved to be useful in the larger
+> project (smart street sensor) providing near real-time delivery of data.
 
-5. The reference list needs to be written in a standard format. Some reference
-   papers are lacking specific pages .Please use a standard way of reference
-   list.
+5. The reference list needs to be written in a standard format. Some
+reference papers are lacking specific pages .Please use a standard way of
+reference list.
 

### 2018-11-26 paper/sections/conclusion.tex
ready to send for proof read

@@
 measures of levels of footfall activity which has been scaled across a Great
-Britain wide network of sensors CDRC 2016.  In both conceptual and technical
+Britain wide network of sensors \citep{cdrc2018consumer}.  In both conceptual and technical
 terms, it illustrates the ways in which passively collected consumer data can be
@@
 around Charring cross, London is shown in Figure 7 to demonstrate the potential
-for such dataset. Projects such as SmartStreetSensors CDRC 2016, may utilise
+for such dataset. Projects such as SmartStreetSensors \citep{cdrc2018consumer}, may utilise
 this methodology to overcome the challenges introduced by the implementation of
@@
 individuals across the smart city using this approach.  This can be modelled
-using agent-based methods Heppenstall et al 2012, however. In our own research
+using agent-based methods \citep{heppenstall2011agent}, however. In our own research
 we have also begun to link store time-lagged till receipts to footfall, and have

### 2018-11-26 paper/sections/introduction.tex
ready to send for proof read

@@
 which movements and interactions of people can be measured and monitored in the
-‘sentient city’ (Amin and Thrift 2017). This is making it possible to
+‘sentient city’ \citep{amin2017seeing}. This is making it possible to
 reinvigorate conceptualise the city as the locus of human activities
-supplementing night time geographies of residence (Martin et al 2015) with
-geographies of shopping behaviour (Lloyd and Cheshire 2018*), workzone
-geographies (Singleton and Longley 2018) and studies of movement trajectories
-(Campbell et al 2008). This is rendering activity-based conceptions of human
-behaviour central to analysis of hardship and opportunity in and around the
-smart city (Venerandi et al 2015).
+supplementing night time geographies of residence \citep{martin2015developing}
+with geographies of shopping behaviour \citep{lloyd2018detecting}, workzone
+geographies \citep{singleton2018london} and studies of movement trajectories
+\citep{campbell2008transforming}. This is rendering activity-based conceptions
+of human behaviour central to analysis of hardship and opportunity in and
+around the smart city \citep{venerandi2015measuring}.
 
@@
 These technologies offer differing levels of spatial precision, where mobile
-telephony and Wi-Fi generally being less reliable and offer lower precision than
-GPS to the end users, while being more advantageous for broader mobility studies
-(Pinelli et al.  2015). There has been considerable research into the usefulness
-of these technologies to understand patterns of movement in cities in near real
-time (Candia et al. 2008, Gonzalez et al. 2008, Calabrese et al. 2013).  Most of
-this research has focused upon technical specification of accuracy or precision
-(Song et al. 2010, Lane et al. 2010), with somewhat less attention devoted to
+telephony and Wi-Fi generally being less reliable and offer lower precision
+than GPS to the end users, while being more advantageous for broader mobility
+studies \citep{pinelli2015comparing}. There has been considerable research into
+the usefulness of these technologies to understand patterns of movement in
+cities in near real time \citep{candia2008uncovering,
+gonzalez2008understanding, calabrese2013understanding}.  Most of this research
+has focused upon technical specification of accuracy or precision
+\citep{song2010limits,lane2010survey}, with somewhat less attention devoted to
 the ways that the characteristics of the technologies and of their human users
@@
 account for an ever-increasing real share of all of the data that are collected
-of citizens, but a fundamental characteristic of consumer-led markets is that no
-single provider has a monopoly in market provision, and therefore issues of
-market share and segment generate bias in analysis.  The source and operation of
-this bias is unknown in the absence of extensive and context sensitive attempts
-to triangulate consumer data with data of known provenance relating to clearly
-defined populations \citep{lansley2016deriving}.  In similar ways to other
-classes of Big Data, consumer data are best thought of as digital ‘exhaust’, or
-a by-product created by or harvested from consumer transactions.
+of citizens, but a fundamental characteristic of consumer-led markets is that
+no single provider has a monopoly in market provision, and therefore issues of
+market share and segment generate bias in analysis.  The source and operation
+of this bias is unknown in the absence of extensive and context sensitive
+attempts to triangulate consumer data with data of known provenance relating to
+clearly defined populations \citep{lansley2016deriving}.  In similar ways to
+other classes of Big Data, consumer data are best thought of as digital
+‘exhaust’, or a by-product created by or harvested from consumer transactions.
 
@@
 consumer service.  Monitoring the probes from such devices provides an
-indication of the presence of their users, regardless of whether or not internet
-connectivity is established.  Our core motivation is to appraise the usefulness
-of Wi-Fi probe requests harvested from our network of sensors in order to
-indicate levels of pedestrian activity.  More broadly still, in our future
-research we intend to classify the nationwide network of footfall profiles as
-part of a programme of research to understand the form and functioning of retail
-areas at a time of far-reaching structural change for the retail industry.
+indication of the presence of their users, regardless of whether or not
+internet connectivity is established.  Our core motivation is to appraise the
+usefulness of Wi-Fi probe requests harvested from our network of sensors in
+order to indicate levels of pedestrian activity.  More broadly still, in our
+future research we intend to classify the nationwide network of footfall
+profiles as part of a programme of research to understand the form and
+functioning of retail areas at a time of far-reaching structural change for the
+retail industry.
 
@@
 must be accommodated by manual recording of footfall at each location and
-generalization of these sample survey results to all locations and time periods.
-As we describe in detail below, manual validation of the data needs to be
-undertaken in parallel with technical profiling of the mix of consumer mobile
-devices that probe our sensors, since the effectiveness of data cleaning
+generalization of these sample survey results to all locations and time
+periods.  As we describe in detail below, manual validation of the data needs
+to be undertaken in parallel with technical profiling of the mix of consumer
+mobile devices that probe our sensors, since the effectiveness of data cleaning
 procedures discussed in this paper differ between individual locations and

### 2018-12-03 paper/sections/introduction.tex
intro and conclusions corrected

@@
 ‘sentient city’ \citep{amin2017seeing}. This is making it possible to
-reinvigorate conceptualise the city as the locus of human activities
+reinvigorate conceptualisations of city as the locus of human activities
 supplementing night time geographies of residence \citep{martin2015developing}
@@
 \citep{campbell2008transforming}. This is rendering activity-based conceptions
-of human behaviour central to analysis of hardship and opportunity in and
-around the smart city \citep{venerandi2015measuring}.
+of human behaviour central to the analysis of hardship and opportunity in, and
+around, the smart city \citep{venerandi2015measuring}.
 
 Sentient technologies include mobile phone networks, which can triangulate user
-locations relative to networks of masts, use of GPS to locate users of social
-media services, and Wi-Fi access points connectivity to access the Internet.
-These technologies offer differing levels of spatial precision, where mobile
-telephony and Wi-Fi generally being less reliable and offer lower precision
-than GPS to the end users, while being more advantageous for broader mobility
+locations relative to networks of masts, the use of GPS to locate users of
+social media services, and Wi-Fi access points providing internet access.  These
+technologies offer differing levels of spatial precision, where mobile telephony
+and Wi-Fi are generally less reliable and offer lower precision than GPS to the
+end users, whilst simultaneously being more advantageous for broader mobility
 studies \citep{pinelli2015comparing}. There has been considerable research into
-the usefulness of these technologies to understand patterns of movement in
-cities in near real time \citep{candia2008uncovering,
-gonzalez2008understanding, calabrese2013understanding}.  Most of this research
-has focused upon technical specification of accuracy or precision
-\citep{song2010limits,lane2010survey}, with somewhat less attention devoted to
-the ways that the characteristics of the technologies and of their human users
-conspire to create possible bias in representing usage patterns across the
-entire smart city.
-
-Even analysis of mobile phone data, usually derived from industry players that
-have significant market share and user bases representative of local
-populations, may exclude groups such as tourists from distant origins or
-subscribers to third party services that share distinctive characteristics
-\citep{di2016mind}
+the utility of these technologies to understand patterns of movement in cities
+in near real time \citep{candia2008uncovering, gonzalez2008understanding,
+calabrese2013understanding}.  Most of this research has focused upon technical
+specification of accuracy or precision \citep{song2010limits,lane2010survey},
+with somewhat less attention devoted to the ways that the characteristics of the
+technologies and of their human users, conspire to create possible bias in
+representing usage patterns across the entire smart city. Analysis of mobile
+phone data, usually derived from industry players that have significant market
+share and user bases representative of local populations, may also exclude
+groups such as tourists from distant origins or subscribers to third party
+services that share distinctive characteristics \citep{di2016mind}.
 
 These examples illustrate the issues that underpin the assembly and analysis of
-consumer data, which can be considered as a distinctive class of Big Data that
-arise from the interactions between humans and customer-facing organisations
-such as retailers, domestic energy suppliers, transport providers and suppliers
-of social media and communications \citep{cdrc2018consumer}.  Consumer data
-account for an ever-increasing real share of all of the data that are collected
-of citizens, but a fundamental characteristic of consumer-led markets is that
-no single provider has a monopoly in market provision, and therefore issues of
-market share and segment generate bias in analysis.  The source and operation
-of this bias is unknown in the absence of extensive and context sensitive
-attempts to triangulate consumer data with data of known provenance relating to
-clearly defined populations \citep{lansley2016deriving}.  In similar ways to
-other classes of Big Data, consumer data are best thought of as digital
-‘exhaust’, or a by-product created by or harvested from consumer transactions.
+consumer data. Consumer data can be considered as a distinctive class of Big
+Data which arise from the interactions between humans and customer-facing
+organisations such as retailers, domestic energy suppliers, transport providers
+and suppliers of social media and communications \citep{cdrc2018consumer}.
+Consumer data account for an ever-increasing real share of all of the data that
+are collected about citizens, but a fundamental characteristic of consumer-led
+markets is that no single provider has a monopoly in market provision; therefore
+issues of market share and segment generate bias in the analysis.  The source
+and operation of this bias is unknown in the absence of extensive and context
+sensitive attempts to triangulate consumer data with data of known provenance
+relating to clearly defined populations \citep{lansley2016deriving}.  In similar
+ways to other classes of Big Data, consumer data are best thought of as digital
+‘exhaust’, or a by-product created by, or harvested from, consumer transactions.
 
 In this paper, through a set of experiments, we evaluate the value of data
-collected from a Great Britain wide network of 800 devices \citep{sss2016}
-installed in order to characterize the footfall patterns of a scientifically
-balanced sample of retail centres.  These devices are located in shop windows
+collected a network of 800 devices \citep{sss2016} installed across Great
+Britain in order to characterize the footfall patterns of a scientifically
+balanced sample of retail centres.  These devices are located in shop windows,
 and record the probes emitted by mobile phones and other Wi-Fi enabled devices.
@@
 consumer service.  Monitoring the probes from such devices provides an
-indication of the presence of their users, regardless of whether or not
-internet connectivity is established.  Our core motivation is to appraise the
-usefulness of Wi-Fi probe requests harvested from our network of sensors in
-order to indicate levels of pedestrian activity.  More broadly still, in our
-future research we intend to classify the nationwide network of footfall
-profiles as part of a programme of research to understand the form and
-functioning of retail areas at a time of far-reaching structural change for the
-retail industry.
+indication of the presence of their users, regardless of whether or not internet
+connectivity is established.  Our core motivation is to appraise the usefulness
+of Wi-Fi probe requests harvested from our network of sensors as a method of
+indicating levels of pedestrian activity.  More broadly still, in our future
+research we intend to classify the nationwide network of footfall profiles as
+part of a programme of research to understand the form and function of retail
+areas at a time of far-reaching structural change for the retail industry.
 
-To this end, it is important to first undertake a thorough conceptual and
+To this end, it was important to first undertake a thorough conceptual and
 technical appraisal of our consumer data source.  In technical terms, screening
-the information present in the `probes requests' and clustering them based on
-their characteristics is essential in order to remove the ones emitted by
-devices that do not indicate pedestrian activity, such as network enabled
-printers and other fixed devices.  Related to this, a method to fingerprint
-Wi-Fi probes is necessary to remove probes from individuals’ devices that in
-conceptual terms should not be considered part of footfall – as when, for
-example, an employees is seated in an office within range of the sensor device.
-A calibration of sensor measurement is also essential on two grounds: first
-individuals may carry multiple devices, or no device at all; and second, the
-positioning and orientation of the sensor in the retail unit may lead to
-systematic over- or under-enumeration.  These sources of bias in measurement
-must be accommodated by manual recording of footfall at each location and
-generalization of these sample survey results to all locations and time
-periods.  As we describe in detail below, manual validation of the data needs
-to be undertaken in parallel with technical profiling of the mix of consumer
-mobile devices that probe our sensors, since the effectiveness of data cleaning
-procedures discussed in this paper differ between individual locations and
-configurations.
+the information present in the `probe requests' and clustering them based on
+their characteristics was essential in order to remove those emitted by devices
+which do not indicate pedestrian activity, such as network enabled printers and
+other fixed devices.  Related to this, a method to fingerprint Wi-Fi probes was
+necessary in order to remove probes from individuals’ devices that in conceptual
+terms should not be considered part of footfall; for instance when , an employee
+is seated in an office within range of the sensor device.  A calibration of
+sensor measurement was also essential on two grounds: first individuals may
+carry multiple devices, or no device at all; and second, the positioning and
+orientation of the sensor in the retail unit may lead to systematic over- or
+under-enumeration.  These sources of bias in measurement must be accommodated by
+manual recording of footfall at each location and the generalization of these
+sample survey results to all locations and time periods.  As we describe in
+detail below, manual validation of the data needed to be undertaken in parallel
+with the technical profiling of the mix of consumer mobile devices that probed
+our sensors, since the effectiveness of data cleaning procedures discussed in
+this paper differ between individual locations and configurations.

### 2018-12-03 paper/sections/main_study.tex
intro and conclusions corrected

@@
-The methodology set out above is next implemented in five different Central
-London locations at different times. Sensors are installed and data collected
-for extended periods of time. We also carry out manual counting at these
-locations across different times of the day. We then apply the methodologies
-discussed earlier to arrive at estimated pedestrian footfall and compare them
-with corresponding manual counts.  We finally evaluate the effectiveness of the
-processes with the Mean Absolute Percentage Error (MAPE) at these locations and
-report our findings.
+The methodology set out above was implemented in five different Central
+London locations at different times. Sensors were installed and data collected
+for extended periods of time. We also carried out manual counting at these
+locations across different times of the day. We then applied the methodologies
+discussed earlier to arrive at estimated pedestrian footfall and compared them
+with the corresponding manual counts.  We finally evaluated the effectiveness of the
+processes with the Mean Absolute Percentage Error (MAPE) at the locations and
+report our findings below.
 

### 2018-12-03 paper/sections/main_study.tex
end of day commits

@@
-The methodology set out above was implemented in five different Central
-London locations at different times. Sensors were installed and data collected
-for extended periods of time. We also carried out manual counting at these
-locations across different times of the day. We then applied the methodologies
-discussed earlier to arrive at estimated pedestrian footfall and compared them
-with the corresponding manual counts.  We finally evaluated the effectiveness of the
+The methodology set out above was implemented in five different Central London
+locations at different times. Sensors were installed and data collected for
+extended periods of time. We also carried out manual counting at these locations
+across different times of the day. We then applied the methodologies discussed
+earlier to arrive at estimated pedestrian footfall and compared them with the
+corresponding manual counts.  We finally evaluated the effectiveness of the
 processes with the Mean Absolute Percentage Error (MAPE) at the locations and
@@
 The locations at which the data were collected are shown in Table
-\ref{locations-table}. The locations are chosen for their diverse site
+\ref{locations-table}. The locations were chosen for their diverse site
 conditions and unique sources of noise around the potential location of the
@@
 context is shown the Figure \ref{main_signal_strength}. We can see that Location
-5 is the `cleanest' with one clear stationary source of noise (phone shop) of
-while location 2 is the one with the most complexity with seating area all
-around the sensor.  The sensors were operational through out February and March
-while manual counts were conducted in these location in half hour sessions on at
-least two different days. For the purposes of comparing with ground truth we
-just considered the data from sensors which corresponding to the 12 sets of
+5 is the `cleanest' with one clear stationary source of noise (phone shop)
+while location 2 is the most complex due to the proximity of seating areas
+to the sensor.  The sensors were operational through out February and March,
+while manual counts were conducted in these locations in half hour sessions on at
+least two different days. For the purposes of comparing with ground truth, we
+considered the data from sensors which correspond to the 12 sets of
 available manual counts. The schedule of data collection is shown in Figure
-\ref{main_schedule}
+\ref{main_schedule}.
 

### 2018-12-06 paper/sections/main_study.tex
main study changes

@@
 
-We start by looking at the distribution of the signal strength reported by the
+We begin by looking at the distribution of the signal strength reported by the
 probe requests across the locations. From the density plot shown in Figure
-\ref{main_signal_strength} we can see that there is clear relation between the
-distribution of the signal strength and the distance and complexity of the
-source of noise. We can see that while location 5 shows clean difference between
-low and high signal strengths, location 2 is almost normally distributed.
-Intuitively we expect that location 2 and 4 must be harder to classify than
-location 1 3 and 5.  We run the k-means clustering algorithm and filter out the
-probe requests which are randomised and have signal strengths less than the
-second break (threshold).  It is important to note that we are dealing with
-relative thresholds of signal strengths which can vary with location and time of
-the analysis.  We then aggregate then probe requests for every minute by
-counting the number of Unique MAC addresses present in every minute. We also
-remove devices that dwell around the sensor by removing the MAC addresses which
-reappear from within the past hour.  The results of the first stage of filtering
-process along with the thresholds are shown in Table \ref{errors_table}.
-Confirming our intuition, we see that the location 2 has the most MAPE followed
-by location 4 while rest of them have highly reduced MAPE.  It is significant
-that this method alone reduces our margin of error by a 50 - 100\% from the raw
-counts without any cleaning.  This makes the signal strength filtering quick and
-ideal method for practical applications which doesn't require absolute numbers
-such as creating large aggregated indexes to show long term trends.  We also see
-that the success of the signal strength filtering can be improved significantly
-by installing sensors such that the pedestrians and noise are at different
+\ref{main_signal_strength}, we can observe that there is clear relation between
+the distribution of the signal strength and the distance and complexity of the
+source of noise. We can see that while location 5 shows clean difference
+between low and high signal strengths, location 2 is almost normally
+distributed.  Intuitively we expected that location 2 and 4 would be harder to
+classify than locations 1, 3 and 5.  We ran the k-means clustering algorithm
+and filtered out the probe requests which were randomised and had signal
+strengths less than the second break (threshold).  It is important to note that
+we were dealing with relative thresholds of signal strengths which can vary
+with location and time of the analysis.  We then aggregated the probe requests
+by counting the number of Unique MAC addresses present in every minute. We also
+removed devices that dwelled around the sensor by removing the MAC
+addresses which reappeared within the previous hour.
+
+The results of the first stage of the filtering process along with the
+thresholds are shown in Table \ref{errors_table}.  Confirming our intuition, we
+see that the location 2 has the most MAPE followed by location 4 while rest of
+them have highly reduced MAPE.  It is significant that this method alone
+reduces our margin of error by a 50 - 100\% from the raw counts without any
+cleaning.  This makes the signal strength filtering quick and ideal method for
+practical applications which doesn't require absolute numbers such as creating
+large aggregated indexes to show long term trends.  We also see that the
+success of the signal strength filtering can be improved significantly by
+installing sensors such that the pedestrians and noise are at different
 distances from the sensor and are the field of measurement is distinct from the

### 2018-12-07 paper/sections/introduction.tex
more corrections

@@
 social media services, and Wi-Fi access points providing internet access. These
-technologies offer differing levels of spatial precision, where mobile telephony
-and Wi-Fi are generally less reliable and offer lower precision than GPS to the
-end users, whilst simultaneously being more advantageous for broader mobility
-studies \citep{pinelli2015comparing}. There has been considerable research into
-the utility of these technologies to understand patterns of movement in cities
-in near real time \citep{candia2008uncovering, gonzalez2008understanding,
-calabrese2013understanding}.  Most of this research has focused upon technical
-specification of accuracy or precision \citep{song2010limits,lane2010survey},
-with somewhat less attention devoted to the ways that the characteristics of the
-technologies and of their human users, conspire to create possible bias in
-representing usage patterns across the entire smart city. Analysis of mobile
-phone data, usually derived from industry players that have significant market
-share and user bases representative of local populations, may also exclude
-groups such as tourists from distant origins or subscribers to third party
-services that share distinctive characteristics \citep{di2016mind}.
+technologies offer differing levels of spatial precision, where mobile
+telephony and Wi-Fi are generally less reliable and offer lower precision than
+GPS to the end users, whilst simultaneously being more advantageous for broader
+mobility studies \citep{pinelli2015comparing}. There has been considerable
+research into the utility of these technologies to understand patterns of
+movement in cities in near real time \citep{candia2008uncovering,
+gonzalez2008understanding, calabrese2013understanding}. Most of this research
+has focused upon technical specification of accuracy or precision
+\citep{song2010limits,lane2010survey}, with somewhat less attention devoted to
+the ways that the characteristics of the technologies and of their human users,
+conspire to create possible bias in representing usage patterns across the
+entire smart city. Analysis of mobile phone data, usually derived from industry
+players that have significant market share and user bases representative of
+local populations, may also exclude groups such as tourists from distant
+origins or subscribers to third party services that share distinctive
+characteristics \citep{di2016mind}.
 
@@
 are collected about citizens, but a fundamental characteristic of consumer-led
-markets is that no single provider has a monopoly in market provision; therefore
-issues of market share and segment generate bias in the analysis.  The source
-and operation of this bias is unknown in the absence of extensive and context
-sensitive attempts to triangulate consumer data with data of known provenance
-relating to clearly defined populations \citep{lansley2016deriving}.  In similar
-ways to other classes of Big Data, consumer data are best thought of as digital
-‘exhaust’, or a by-product created by, or harvested from, consumer transactions.
+markets is that no single provider has a monopoly in market provision;
+therefore issues of market share and segment generate bias in the analysis. The
+source and operation of this bias is unknown in the absence of extensive and
+context sensitive attempts to triangulate consumer data with data of known
+provenance relating to clearly defined populations \citep{lansley2016deriving}.
+In similar ways to other classes of Big Data, consumer data are best thought of
+as digital ‘exhaust’, or a by-product created by, or harvested from, consumer
+transactions.
 
@@
 consumer service. Monitoring the probes from such devices provides an
-indication of the presence of their users, regardless of whether or not internet
-connectivity is established.  Our core motivation is to appraise the usefulness
-of Wi-Fi probe requests harvested from our network of sensors as a method of
-indicating levels of pedestrian activity.  More broadly still, in our future
-research we intend to classify the nationwide network of footfall profiles as
-part of a programme of research to understand the form and function of retail
-areas at a time of far-reaching structural change for the retail industry.
+indication of the presence of their users, regardless of whether or not
+internet connectivity is established. Our core motivation is to appraise the
+usefulness of Wi-Fi probe requests harvested from our network of sensors as a
+method of indicating levels of pedestrian activity. More broadly still, in our
+future research we intend to classify the nationwide network of footfall
+profiles as part of a programme of research to understand the form and function
+of retail areas at a time of far-reaching structural change for the retail
+industry.
 
@@
 other fixed devices. Related to this, a method to fingerprint Wi-Fi probes was
-necessary in order to remove probes from individuals’ devices that in conceptual
-terms should not be considered part of footfall; for instance when , an employee
-is seated in an office within range of the sensor device.  A calibration of
-sensor measurement was also essential on two grounds: first individuals may
-carry multiple devices, or no device at all; and second, the positioning and
-orientation of the sensor in the retail unit may lead to systematic over- or
-under-enumeration.  These sources of bias in measurement must be accommodated by
-manual recording of footfall at each location and the generalization of these
-sample survey results to all locations and time periods.  As we describe in
-detail below, manual validation of the data needed to be undertaken in parallel
-with the technical profiling of the mix of consumer mobile devices that probed
-our sensors, since the effectiveness of data cleaning procedures discussed in
-this paper differ between individual locations and configurations.
+necessary in order to remove probes from individuals’ devices that in
+conceptual terms should not be considered part of footfall; for instance when,
+an employee is seated in an office within range of the sensor device. A
+calibration of sensor measurement was also essential on two grounds: first
+individuals may carry multiple devices, or no device at all; and second, the
+positioning and orientation of the sensor in the retail unit may lead to
+systematic over- or under-enumeration. These sources of bias in measurement
+must be accommodated by manual recording of footfall at each location and the
+generalization of these sample survey results to all locations and time
+periods. As we describe in detail below, manual validation of the data needed
+to be undertaken in parallel with the technical profiling of the mix of
+consumer mobile devices that probed our sensors, since the effectiveness of
+data cleaning procedures discussed in this paper differ between individual
+locations and configurations.

### 2018-12-07 paper/sections/pilot_study.tex
more corrections

@@
-\begin{table}
+begin{table}
 \tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
@@
 
-To start we designed a small pilot study to validate the filtering and
-clustering methodology against the scale and complexity of data collected on an
-open public area such as a retail high street.  We also aim to find the
-algorithm which is best suited for classification of signal strengths as 'low'
-and 'high' to filter out the background noise.  The data was collected at Oxford
-Street, London on 20 December 2017 from 12:30 to 13:00 hrs, where Wi-Fi probe
-requests were collected using the sensor described in Chapter \ref{methodology}
-and pedestrian footfall was manually recorded using the Android app
-\citep{bala2018clicker}.  Being located at one of the busiest retail locations
-in the United Kingdom, the Wi-Fi sensor captured approximately 60,000 probe
-requests during the half hour period, and 3,722 people were recorded manually
-walking on the sidewalk during that time. The count was done at the location in
-front of the store where the surveyor carried the sensor on a backpack for the
-automated count and used a phone for doing the manual count. The sensor is kept
-as close to the store window as possible and the manual count is done as a
-cordon count in fron of the store.
+To start, we designed a small pilot study to validate the filtering and
+clustering methodology against the scale and complexity of data collected in an
+open public area such as a retail high street. We also aimed to find the
+algorithm which was best suited for the classification of signal strengths as
+'low' and 'high' in order to filter out the background noise. The data was
+collected at Oxford Street, London on 20 December 2017 from 12:30 to 13:00 hrs,
+Wi-Fi probe requests were collected using the sensor described in Section
+\ref{methodology} and pedestrian footfall was manually recorded using the
+Android app - Clicker \citep{bala2018clicker}. Being located at one of the
+busiest retail locations in the United Kingdom, the Wi-Fi sensor captured
+approximately 60,000 probe requests during the half hour period; 3,722 people
+were manually recorded walking on the pavement during that time. The surveyor
+positioned himself at the front of a store while carrying the sensor in a
+backpack and counted people walking by the store on the pavement (3m wide
+approximately) using a mobile phone. The sensor was kept as close to the store
+window as possible, and the manual count was done as a cordon count in front of
+the store.
 
-As a first step we just aggregated the probe requests by their MAC address for
-every minute to generate a minute by minute count of the number of people around
-the sensor assuming each MAC address corresponds to a mobile device and hence a
-pedestrian.  We then compare this preliminary `footfall' count to the actual
-number of pedestrians recorded manually to check for the robustness. We use Mean
-Absolute Percentage Error (MAPE) measure of robustness of the count since it
-provides a simple and quick measurement while the street conditions ensure that
-there are no intervals without any footfall.  We find that the MAPE in the raw
-counts compared to the ground truth is around 425\%.  This suggests the presence
-of large amount of noise in the data which might be generated due to sources of
-uncertainties discussed in Chapter \ref{methodology} thus demonstrating the need
-for filtering the data.
+As a first step we aggregated the probe requests by their MAC addresses for
+every minute to generate a minute by minute count of the number of people near
+the sensor. We assumed that each MAC address corresponded to a mobile device and
+hence a pedestrian. We then compared this preliminary `footfall' count to the
+actual number of pedestrians recorded manually to check for it's robustness. We
+used Mean Absolute Percentage Error (MAPE) as a measure of robustness of the
+count, since it provided a simple and quick measurement and the street
+conditions ensured that there are no intervals without any footfall. We found
+that the MAPE in the raw counts compared to the ground truth was around 425\%.
+This suggests the presence of large amount of noise in the data which may have
+been generated by the sources of uncertainties discussed in Section
+\ref{methodology} thus demonstrating the need for filtering the data.
 
-We then classified the probe requests as ones with "high signal strength" and
-"low signal strength" using various one dimensional clustering algorithms such
-as k-means, quantile, hierarchical clustering, bagged clustering, fisher and
-jenks natural breaks.  The results are shown in Table
-\ref{classification-table}.  We found that while hierarchical clustering and
-jenks gave us fairly low errors, they were too resource intensive for practical
-use with a larger dataset.  We also found that k-means gave quickest results
-with the lowest MAPE closely followed by quantile algorithm.  The cut-off point
-or threshold for the collected data with which we could classify them as high
-and low was -71 dBm.  We then removed all the probe requests which reported `low
-signal strength' and repeated the same aggregation process as before to produce
-footfall count.  This process resulted in a footfall count with a net MAPE of
-30\%.  Though the results are encouraging we are still not completely confident
-that our filtering process is indeed removing noise or has any correlation the
-configuration of sensor or position of the mobile devices.  These need to be
-addressed with a larger survey with multiple location of varying orientations.
+We then classified the probe requests as `high signal strength' and `low signal
+strength' using various one dimensional clustering algorithms such as k-means,
+quantile, hierarchical clustering, bagged clustering, fisher and jenks natural
+breaks. The results are shown in Table \ref{classification-table}. We found that
+while hierarchical clustering and jenks gave us fairly low errors, they were too
+resource intensive for practical use with a larger dataset. We also found that
+k-means gave the quickest results with the lowest MAPE, closely followed by
+quantile algorithm. The cut-off point or threshold for the collected data with
+which we could classify as high and low was -71 dBm. We then removed all the
+probe requests which reported `low signal strength' and repeated the same
+aggregation process as before to produce footfall count. This process resulted
+in a footfall count with a net MAPE of 30\%. Although the results are
+encouraging we are still not completely confident that our filtering process is
+removing noise or has any correlation the configuration of sensor or position of
+the mobile devices. These concerns need to be addressed with a larger survey
+with multiple locations of varying orientations.
 
@@
         \caption{The clustering process was repeated with both increasing sequence
-	number threshold ($\alpha$) and time threshold ($\beta$), until we arrived
-	at the lowest parameters where the know device (black line) is clustered as
-	a single device.}
+            number threshold ($\alpha$) and time threshold ($\beta$), until we
+            arrived at the lowest parameters where the know device (black line) is
+            clustered as a single device.}
         \label{pilot_clustering_params}
@@
 
-The next challenge was to identify probe requests which are generated by the
-same device irrespective of MAC randomisation process.  We use the algorithm
-defined in Chapter \ref{methodology} and assign a unique identifier or signature
-to each probe request independent of the MAC address.  Since there is neither
-prior research nor documentation on a universal behaviour of phones in
-randomising their MAC addresses, we use the stationary device - the one used for
-manual counting as a reference and find out the suitable time threshold,
-$\alpha$ and threshold for sequence numbers, $\beta$ to be 16 seconds and 60
-respectively via trial and error. This process is shown in Figure
-\ref{pilot_clustering_params} This process is done on top the filtering done
-based on signal strength and only for the probe requests with randomised MAC
-addresses.  Figure \ref{pilot_results} shows the results of this clustering
-process on a small set of randomised probe requests.  The probe request with
-different randomised MAC address is shown by the colored points and the line
-joining them shows the ones belonging to the same cluster hence expected to be
-generated by the same device.  We finally aggregate the probe requests as before
-but with the device signature rather than just MAC addresses this results in a
-footfall count with a MAPE of -18\%.  A comparison of minute by minute counts
-resulting from different filtering processes along with the ground truth is
-shown in Figure \ref{pilot_results} showing the promising effectiveness of the
-methods.
+The next challenge was to identify the probe requests which are generated by the
+same device irrespective of the MAC randomisation process. We used the algorithm
+defined in Section \ref{methodology} and assigned a unique identifier or
+signature to each probe request, independent of their the MAC addresses. Since
+we didn't know the nature or frequency of the MAC address randomisation process,
+we used the surveyor's mobile device as a reference. As the surveyor's device
+was being actively used to count pedestrians and it's Wi-Fi module was kept
+active without establishing connection to any network, the device was known to
+be continuously probing for new networks. We also knew that the OUI of the
+device was 'Google' and the device was regularly randomising it's MAC address,
+thus providing us an excellent reference with which we could optimise the
+parameters for our clustering algorithm. Using this reference device we observed
+that the threshold for time $\alpha$ and the threshold for sequence numbers,
+$\beta$ are 16 seconds and 60 respectively via trial and error. This process is
+shown in Figure \ref{pilot_clustering_params}. This was undertaken on top the
+filtering done based on signal strength, and only for the probe requests with
+randomised MAC addresses. Figure \ref{pilot_results} shows the results of this
+clustering process on a small set of randomised probe requests. The probe
+requests with different randomised MAC address are shown by the coloured points
+and the lines joining them show that those probe requests were most likely be
+generated by the same device. We finally aggregated the probe requests as we did
+before but with the device signature rather than MAC addresses. This results in
+a footfall count with a MAPE of -18\% compared to the manual count. A comparison
+of minute by minute counts resulting from different filtering processes along
+with the ground truth is shown in Figure \ref{pilot_results} illustrating the
+promising effectiveness of the methods.
 
-To conclude, from the pilot study we found that both filtering and clustering
-methods we devices work on complex real world data and results in a final
-pedestrian counts within a MAPE of 20\%. We also found `k-means' and `quantile'
-are best algorithms for clustering signal strengths and the threshold for time
-and sequence numbers for the clustering algorithm is around 16 and 60
-respectively.
+To conclude, from the pilot study we found that both the filtering and the
+clustering methods we devised worked on complex real world data and resulted in
+final pedestrian counts within a MAPE of 20\%. We also found that `k-means' and
+`quantile' are best algorithms for clustering signal strengths. Finally, we
+observed that the best thresholds for time and sequence numbers in the
+clustering algorithm is around 16 and 60 respectively.

### 2018-12-07 paper/sections/previous_work.tex
more corrections

@@
-In the past decade Wi-Fi has emerged as one of the most commonly used technology
-in providing high speed internet access to mobile devices such as smartphones,
-tablets and laptops in public and private spaces. This has resulted in multiple
-Wi-Fi networks being available at almost every location in dense urban
-environments. Traversing through this overlapping mesh of Wi-Fi networks, modern
-mobile devices with Wi-Fi network interface regularly broadcast a special type
-of signal known as `Probe Requests' in order to discover Wi-Fi networks
-available to them.  This helps these devices to connect and switch between the
-Wi-Fi networks seamlessly.
+In the past decade, Wi-Fi has emerged as one of the most commonly used
+technologies in providing high speed internet access to mobile devices such as
+smartphones, tablets and laptops in public and private spaces. This has
+resulted in multiple Wi-Fi networks being available at almost every location in
+dense urban environments. Traversing through this overlapping mesh of Wi-Fi
+networks, modern mobile devices with Wi-Fi network interfaces regularly
+broadcast a special type of signal known as `Probe Requests' in order to
+discover the Wi-Fi networks available to them. This helps these devices to
+connect and switch between the Wi-Fi networks seamlessly.
 
 Probe requests are low level signals standardised by IEEE 802.11 specification
-\citep{ieee2016} for service discovery and is implemented in any Wi-Fi capable
-device irrespective of the manufacturer or the model. This ubiquity and
+\citep{ieee2016} for service discovery, and are implemented in any Wi-Fi
+capable device irrespective of the manufacturer or the model. This ubiquity and
 standardisation makes them an excellent source of open, passive, continuous,
@@
 distribution in highly dynamic urban environments with high spatial and
-temporal granularity \citep{freud2015,konto2017}. While a Wi-Fi based method to
+temporal granularity \citep{freud2015, konto2017}. While a Wi-Fi based method to
 collect data offers us various advantages such as, easy scalability and
-efficiency in terms of cost and time, It also introduces few systematic biases,
-uncertainities in the collected data along with the serious risk of infringing
-on the privacy of the mobile users. In this paper, using a set of probe
-requests and manual counts collected at various high street locations across
-London, we demonstrate that pedestrian footfall at these locations can be
-estimated with considerable precision and accuracy while protecting the privacy
-of the pedestrians.
+efficiency in terms of cost and time, it also introduces few systematic biases
+and uncertainities in the collected data along with the serious risk of
+infringing on the privacy of the mobile users. In this paper, using a set of
+probe requests and manual counts collected at various high street locations
+across London, we demonstrate that pedestrian footfall at these locations can
+be estimated with considerable precision and accuracy while protecting the
+privacy of the pedestrians.
 
-Though, unlike GPS, the location of the Wi-Fi enabled mobile device cannot be
-directly inferred from Wi-Fi, there are reliable methods to triangulate the
-location of mobile devices from the locations of known access points(AP) and the
-signal strength reported by them
-\citep{he2003range,moore2004robust,lamarca2005place}. This can overcome the
-usual shortcoming of GPS, which struggles for precision and accuracy in indoor
-and densely built environments
-\citep{zarim2006,kawaguchi2009wifi,xi2010locating}. Utilising this, we can
+Unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly
+inferred from Wi-Fi, however there are reliable methods to triangulate the
+location of mobile devices from the locations of known access points (AP) and
+the signal strength reported by them \citep{he2003range, moore2004robust,
+lamarca2005place}. This can overcome the usual shortcoming of GPS, which
+struggles for precision and accuracy in indoor and densely built environments
+\citep{zarim2006, kawaguchi2009wifi, xi2010locating}. Utilising this, we can
 easily and quickly estimate trajectories of the mobile devices
-\citep{Sorensenlocation,musa2012tracking} which can be used similar to the GPS
-trajectories to understand individual travel patterns \citep{reki2007,Sap2015},
-crowd behaviour \citep{abedi2013bluetooth,mowafi2013tracking}, vehicular
-\citep{lu2010vehicle} and pedestrian movement
-\citep{xu2013pedestrian,fukuzaki2014pedestrian,wang2016gait}. Such data can
+\citep{Sorensenlocation, musa2012tracking} which can be used similary to the
+GPS trajectories to understand individual travel patterns \citep{reki2007,
+Sap2015}, crowd behaviour \citep{abedi2013bluetooth, mowafi2013tracking},
+vehicular \citep{lu2010vehicle} and pedestrian movement
+\citep{xu2013pedestrian, fukuzaki2014pedestrian, wang2016gait}. Such data can
 also be used in transportation planning and management to estimate travel time
@@
 \citep{abbott2013empirical}. Using techniques demonstrated by
-\cite{franklin2006passive} and \cite{pang2007802} along with information present
-in the probe requests one even model interactions between the users
-\citep{cheng2012inferring,barbera2013signals,cunche2014know,cunche2014linking}
-such as predicting which of them are most likely to meet again
-\citep{cunche2012know}. Using the semantic information present in these probe
-requests it even is possible to understand the nature of population at a large
-scale \citep{di2016mind}.
+\cite{franklin2006passive} and \cite{pang2007802}, along with information
+present in the probe requests, one can even model interactions between the
+users \citep{cheng2012inferring, barbera2013signals, cunche2014know,
+cunche2014linking} such as predicting which of them are most likely to meet
+again \citep{cunche2012know}. Using the semantic information present in these
+probe requests it even is possible to understand the nature of population at a
+large scale \citep{di2016mind}.
 
-Though extensive research has been carried out on this subject with feasible and
-favorable results, in recent years, one of the major challenges faced in such
-attempts has been the increasing attempt by mobile phone manufacturers to
-protect their users’ privacy by anonymising the globally identifiable portion of
-the probe requests, \citep{green2008}. Various methods have been devised to
+Although extensive research has been carried out on this subject with feasible
+and favorable results, in recent years, one of the major challenges faced in
+such attempts has been the increasing attempt by mobile phone manufacturers to
+protect their users’ privacy by anonymising the globally identifiable portion
+of the probe requests \citep{green2008}. Various methods have been devised to
 overcome this anonymisation process such as estimating the device model
@@
 \citep{martin2016decomposition}; Scrambler attack using a small part of the
-physical layer specification for Wi-Fi \citep{vo2016,bloessl2015scrambler}; and
-timing attack where the packet sequence information along with information
-elements present in the probe request frame is used
-\citep{matte2016,cheng2016can}. A combination of these methodologies has been
-proven to produce de-anonymised globally unique device information
-\citep{vanhoef2016,martin2017}. These approaches usually result in serious risk
-of breach of privacy of the users of the mobile devices by revealing their
+physical layer specification for Wi-Fi \citep{vo2016, bloessl2015scrambler};
+and timing attack where the packet sequence information along with information
+elements present in the probe request frame is used \citep{matte2016,
+cheng2016can}. A combination of these methodologies has been proven to produce
+de-anonymised globally unique device information \citep{vanhoef2016,
+martin2017}. These approaches usually result in serious risk of breach of
+privacy of the users of the mobile devices by revealing their
 identifiable personal information.
 
-There is a clear gap in research for exploring methodologies for estimating the
-number of unique mobile devices from a set of anonymised probe requests, without
-the need to reveal their original device information. Such technique has
-various applications such as uncovering the urban wireless landscape
+There is a clear gap in the research for exploring methodologies for estimating
+the number of unique mobile devices from a set of anonymised probe requests,
+without the need to reveal their original device information. Such a technique
+has various applications such as uncovering the urban wireless landscape
 \citep{rose2010mapping}, revealing human activity at large scales
 \citep{qin2013discovering}, estimating pedestrian numbers in crowds
-\citep{schauer2014estimating,fukuzaki2015statistical} and even counting people
-in hyper local scales such as queues \citep{wang2013measuring} With enough
-infrastructure to collect such information we can even aim to generate a
-real-time census of the city \citep{konto2017}. With this background we set out
-to device and implement a methodology to reliably estimate human activity such
-as pedestrian footfall from Wi-Fi probe requests without risking the breach of
-privacy of the users involved.
+\citep{schauer2014estimating, fukuzaki2015statistical}, and even counting
+people in hyper local scales such as queues \citep{wang2013measuring}. With
+enough infrastructure to collect such information we can even aim to generate a
+real-time census of the city \citep{konto2017}. With this background, we set
+out to devise and implement a methodology to reliably estimate human activity
+such as pedestrian footfall from Wi-Fi probe requests without risking a breach
+of privacy of the users involved.

### 2018-12-07 paper/sections/methodology.tex
methodology corrections

@@
-The primary aim of this research is to enable us to collect a series of probe
-requests and process them into a usable pedestrian footfall count. We do this by
-using a Wi-Fi receiver to collect probe requests broadcast by mobile devices,
-filtering out the background noise and aggregating them based on the device that
-generated them. In this section, we begin by looking at the characteristics of
-probe requests in detail, device a methodology to collect these probe requests
-in public areas, examine the systemic biases and uncertainties in the data
-collection method and device data processing methods to overcome these
-challenges. Finally we compare the processed footfall counts to the ground truth
-recorded by primary surveys.
+The primary aim of this research was to enable us to collect a series of probe
+requests and process them into a usable pedestrian footfall count. We did this
+by using a Wi-Fi receiver to collect probe requests broadcast by mobile
+devices, filtering out the background noise, and aggregating them based on the
+device that generated them. In this section, we examine the
+characteristics of probe requests in detail, devise a methodology to collect
+these probe requests in public areas, examine the systemic biases and
+uncertainties in the data collection method, and devise data processing methods
+to overcome these challenges. Finally, we compare the processed footfall counts
+to the ground truth recorded by primary surveys.
 
-Probe requests are a special type of management packets broadcast by Wi-Fi
-enabled devices as part of the various functions such as scanning for available
-APs, quick geolocation by triangulation based known APs, etc. These are
-broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
-model of the devices though there is some variation on the frequency and the
-information transmitted through them. In some cases, such as Android devices,
-these are broadcast even when the Wi-Fi functionality has been turned off by
-the user so that device can immediately connet to networks when the
-functionality is switched back on. Since some devices even use the probe
-requests as a less accurate form of localisation they keep sending probe
-requests as well. Thus these signals can be used to reliably identify the
-presence of Wi-Fi enabled mobile devices. Being a first step of connection
-initiated by the mobile device, these packets have information regarding the
-characteristics of the mobile device itself. Some of the key information we can
-infer from these requests are,
+Probe requests are a special type of management packet broadcast by Wi-Fi
+enabled devices as part of their various functions such as scanning for
+available APs and quick geolocation by triangulation based known APs, etc,.
+These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer,
+type or model of the devices, although there is some variation in the frequency
+and the content of the information transmitted through them. In some cases, such
+as Android devices, these are broadcast even when the Wi-Fi functionality has
+been turned off by the user so that the device can immediately connect to
+networks when the functionality is switched back on. Since some devices even use
+the probe requests as a less accurate form of localisation, they continuously
+send probe requests when Wi-Fi has been switched off. Thus, these signals can be
+used to reliably identify the presence of Wi-Fi enabled mobile devices. Being a
+first step of connection initiated by the mobile device, these packets have
+information regarding the characteristics of the mobile device itself. Some of
+the key information we can infer from these requests are,
 
 \begin{enumerate} 
-\item \textbf{Media Access Control (MAC) address} which is an name identifying
-	for the wireless hardware of the mobile device, 
+\item \textbf{Media Access Control (MAC) address} which is a name identifying
+	the wireless hardware of the mobile device, 
 \item \textbf{Sequence number} of the request for the mobile device to keep
@@
 
-The MAC address is the primary unique identifier for the mobile device. It has
-two parts, first part is the Organisationally Unique Identifier (OUI) which
-gives information about the manufacturer of the device and the second is unique
-to the device. In modern devices, to protect the users privacy, the MAC address
-can also be randomised (hence non unique) and is marked as such. Though
-sequence number of the packet is strictly unique to a mobile device, we
-hypothesize that we can use them to estimate the number unique devices as
-demonstrated by \citep{vanhoef2016}, where optional information present in the
-probe requests - Information Elements (IE) along with the probe requests have
-been used to fingerprint the devices. This approach has become increasinly
-difficult as mobile phone manufacturers have severely limited the IEs present
-in the probe requests thus leading us to explore methods which uses only the
-sequence numbers. This also affects the established commercial solutions using
-Wi-Fi probe requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.
-There have been another solution proposed by \citep{hong2018crowdprobe} where
-the authors tried to solve the similar problem using a hidden markov models
-based trajectory inference alogorithm but the scope is limited to enclosed,
+The MAC address is the primary identifier for the mobile device and has two
+parts. The first part is the Organisationally Unique Identifier (OUI) which
+provides information about the manufacturer of the device and the second part is
+the identifier for the device. In modern devices, to protect users' privacy, the
+second part of the MAC address can also be randomised and hence may not be
+unique to devices. When the MAC address is randomised, it is marked as such by
+setting a specific bit in the probe request packet as 1. Although sequence
+number of the packet is strictly unique to a mobile device, we hypothesize that
+we can use them to estimate the number of unique devices as demonstrated by
+\citep{vanhoef2016}; where optional information present in the probe requests -
+Information Elements (IE) along with the sequence numbers, have been used to
+fingerprint the devices. This approach has become increasingly difficult as
+mobile phone manufacturers have severely limited the IEs present in the probe
+requests thus leading us to explore methods which use only the sequence numbers.
+This also affects the established commercial solutions using Wi-Fi probe
+requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.  There has
+been another solution proposed by \citep{hong2018crowdprobe} where the authors
+tried to solve the similar problem using a hidden markov models based trajectory
+inference algorithm but the scope of this research was limited to enclosed,
 exit controlled public spaces such as shopping malls, railway stations, etc.
@@
 Data collection was done with the help of custom sensors built from modifying
-the hardware used in Smart street sensors \citep{sss2016} and updating them with
+the hardware used in Smart Street Sensors \citep{sss2016} and updating them with
 custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
-3G modules. It keeps the Wi-Fi module in `Monitor' mode and uses the open source
+3G modules. It keeps the Wi-Fi module in `Monitor mode' and uses the open source
 software - Wireshark \citep{wireshark2} to passively collect all packets sent to
-`broadcast', marked with type as management' and subtype as `probe requests'.
+`broadcast', marked with type as `management', and subtype `probe requests'.
 The MAC address in these probe requests is obfuscated at the device level using
@@
 Though hashing cannot completely ensure anonymisation as discussed in
-\citep{demir2014analysing} it can sufficiently obfuscate the data which along
-with a secure process of data handling gives us reasonable security. A overall
+\citep{demir2014analysing}, it can sufficiently obfuscate the data; which along
+with a secure process of data handling, gives us reasonable security. An overall
 schematic of the data collection and storage process is shown in Figure
-\ref{datacollection_schematic}. The ground truth on number of pedestrian
+\ref{datacollection_schematic}. The ground truth on the number of pedestrian
 footfall was recorded using a custom Android application - Clicker
-\citep{bala2018clicker}. This app logs accurate timestamps for each time the
+\citep{bala2018clicker}. This app logs accurate timestamps each time the
 surveryor records a pedestrian crossing the designated cordon line at the
-location. In addition to counting the pedestrians manually this also makes the
-mobile device used constantly probe for networks which also gives us a know
-device to calibrate our methodology against.
+location. In addition to counting the pedestrians manually, this procedure
+results in the device broadcasting probe requests regularly, which in turn,
+gives us a `known device' to calibrate our methodology against.
 
@@
 
-The next step after collecting data was to estimate the footfall or pedestrian
-activity from them. We identified the following potential uncertainties which
-arise from our collection methodology.
+After collecting data, we began estimating the footfall or pedestrian
+activity from them by identifying the following potential uncertainties arising
+from our data collection method.
 

### 2018-12-07 paper/sections/methodology.tex
All corrections completed

@@
 requests and process them into a usable pedestrian footfall count. We did this
-by using a Wi-Fi receiver to collect probe requests broadcast by mobile
-devices, filtering out the background noise, and aggregating them based on the
-device that generated them. In this section, we examine the
-characteristics of probe requests in detail, devise a methodology to collect
-these probe requests in public areas, examine the systemic biases and
-uncertainties in the data collection method, and devise data processing methods
-to overcome these challenges. Finally, we compare the processed footfall counts
-to the ground truth recorded by primary surveys.
+by using a Wi-Fi receiver to collect probe requests broadcast by mobile devices,
+filtering out the background noise, and aggregating them based on the device
+that generated them. In this section, we examine the characteristics of probe
+requests in detail, devise a methodology to collect these probe requests in
+public areas, examine the systemic biases and uncertainties in the data
+collection method, and devise data processing methods to overcome these
+challenges. Finally, we compare the processed footfall counts to the ground
+truth recorded by primary surveys.
 
@@
 enabled devices as part of their various functions such as scanning for
-available APs and quick geolocation by triangulation based known APs, etc,.
+available APs and quick geolocation by triangulation based known APs, etc.
 These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer,
@@
 tried to solve the similar problem using a hidden markov models based trajectory
-inference algorithm but the scope of this research was limited to enclosed,
-exit controlled public spaces such as shopping malls, railway stations, etc.
+inference algorithm but the scope of this research was limited to enclosed, exit
+controlled public spaces such as shopping malls, railway stations, etc.
 
@@
 
-After collecting data, we began estimating the footfall or pedestrian
-activity from them by identifying the following potential uncertainties arising
-from our data collection method.
+After collecting data, we began estimating the footfall or pedestrian activity
+from them by identifying the following potential uncertainties arising from our
+data collection method:
 
@@
         of interest. This can lead to a significant background noise at certain
-		locations. E.g. a phone shop or a bus stop located next to the study
-		area can increase the number of probe requests received by the sensor.
-		It is important to note that this method may not work effectively on
-		study locations with complex configurations such as the source of noise
-		and the area of study being located at the same distance from the
-		sensor. This aspect is explored in detail in the broader case study in
-		the following sections.
+        locations. For example, a phone shop or a bus stop located next to the
+        study area can artificially increase the number of probe requests
+        received by the sensor.  It is important to note that this method may
+        not work effectively on study locations with complex configurations such
+        as the source of noise and the area of study being located at the same
+        distance from the sensor. This aspect is explored in detail in the
+        broader case study in the following sections.
 \item 
-\textbf{MAC randomisation} - The mobile devices in recent years have been using
-		randomised 'local' MAC addresses for probe requests to protect the users
+\textbf{MAC randomisation} - mobile devices in recent years have been using
+        randomised `local' MAC addresses for probe requests to protect the users
         from being tracked. This makes it impossible to tell if the probe
-		requests are being sent by the same mobile device which is being
-		stationed next to the sensor. This along with the previous problem can
-		further increase the magnitude of error by several fold.
+        requests are being sent by the same mobile device. This along with the
+        previous problem can further increase the magnitude of error by several
+        fold.
 \item
-\textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely
+\textbf{Mobile ownership} - since the rate of mobile ownership can vary widely
 		across geography and demography, we cannot assume that every mobile
 		device translates to one pedestrian footfall. In addition to this, there
-		is a long term overall increase in mobile ownership which may lead to
+		is a long term overall increase in mobile ownership which may affect 
 		the number of probe requests collected overtime.
@@
 device and the sensor is the strength of the signal received by the sensor. The
-obvious approach here is to try and establish a relationship between the signal
-strength and distance first and use this to filter out the unwanted probe
-requests. This approach was found not to be feasible since the decay of signal
-strength with distance is not always constant. It varies with atmospheric
-conditions, presence of obstructions between the source and target, the nature
-of these obstructions and the strength (power level) of the source. This
-severely limits our ability in establishing a simple conversion between reported
-signal strength and distance. There is a need for a method which takes in to
-account these variables across various locations.
+obvious approach was to first try and establish a relationship between the
+signal strength and distance and to use this to filter out the unwanted probe
+requests. However this approach was found to not be feasible, since the decay of
+signal strength with respect to distance is not always constant. For instance,
+signal strength varies with atmospheric conditions, the presence of obstructions
+between the source and the target, the nature of these obstructions, and the
+strength (power level) of the source. This severely limits our ability to
+establish a simple conversion between reported signal strength and distance. As
+such, there was a need for a method which takes in to account all of these
+variables across the various locations.
 
-We assume that in configurations where a specific source of background
-noise is at a constant distance, there must be a distinct pattern in the number
-of probe requests reporting signal strength corresponding to that distance. For
-example, if there is a phone shop next to our sensor where hundreds of phones
-regularly send probe requests there should be a sharp rise of number of probe
-requests with reported signal strength corresponding to the distance between the
-sensor and the phone shop irrespective of the local conditions as shown in
-Figure \ref{methodology_schematic}. We could identify these breaks in the data
-using traditional one dimensional clustering algorithms such as `jenks natural
-breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since we are
-only looking for the break in the data and not for absolute values, the
-methodology should apply for all the variations due micro site conditions thus
+We assumed that, in configurations where a specific source of background noise
+was at a constant distance, there must be a distinct pattern in the number of
+probe requests reporting signal strength corresponding to that distance. For
+example, if there was a phone shop next to our sensor where hundreds of phones
+regularly sent probe requests, there should be a sharp rise in the of number of
+probe requests with reported signal strength corresponding to the distance
+between the sensor and the phone shop, irrespective of the local conditions as
+shown in Figure \ref{methodology_schematic}. We could identify these breaks in
+the data using traditional one dimensional clustering algorithms such as `jenks
+natural breaks', `k-means', `quantile' and `hierarchical clustering', etc. Since
+we were only looking for the break in the data and not for absolute values, the
+methodology should apply for all the variations due to micro site conditions
 reducing the overall noise in the collected data.
@@
 
-Since our primary unique identifier - MAC address, is being anonymised by new
-devices, we need to find other information present in the probe request for use
-as a unique identifier. Obvious approach here is to establish a factor of
-randomisation and adjust the counts for these probe requests based on this
-factor. We found this approach to be not feasible since the proportion of
-devices which randomise the MAC addresses increases over time. There is also a
-wide variation in the frequency at which the devices randomise the MAC addresses
-and the method used for the process. This lead us to look for a more
-generalisable approach which is independent of the device model.
+Since our primary unique identifier - MAC addresses are being anonymised by new
+devices, we needed to find other information present in the probe requests for
+use as a unique identifier. The obvious approach was to establish a factor of
+randomisation, and adjust the counts for the probe requests based on this
+factor. We found this approach to not be feasible since the proportion of
+devices which randomise the MAC addresses increased over time. There was also a
+wide variation in the frequency at which the devices randomised the MAC
+addresses and the method used for the process. This led us to look for a more
+generalisable approach which was independent of the device model.
 
-From our initial look at the data we found that OUI and the sequence
-number of the packet is the most promising information to achieve this.
-First we divide our dataset into sets of probe requests with randomised and
-non-randomised MAC addresses by looking at the second character of the vendor
-part of the MAC address and if it is E, A, 2 or 6, then these addresses are
-identified to be randomised. We keep the MAC address as the unique
-identifier for the non-randomised requests and further divide the randomised ones in to
-sub categories based on their OUI. We then  
-identify unique mobile devices from within these sets and assign a unique identifier to each device.
+From our initial look at the data, we found that OUI and the sequence number of
+the packet was the most promising information to achieve this.  First we divided
+our dataset into sets of probe requests with randomised and non-randomised MAC
+addresses by looking at the second character of the vendor part of the MAC
+address; if it was E, A, 2 or 6, then those addresses were identified to be
+randomised. We kept the MAC address as the unique identifier for the
+non-randomised requests and further divided the randomised ones in to
+sub-categories based on their OUI. We then  identified unique mobile devices
+from within those sets, and assigned a unique identifier to each device.
 
-The proposed algorithm creates a graph where the probe requests represented the
-nodes, and links are created between them based on the following rules:
+The proposed algorithm created a graph where the probe requests represented the
+nodes; links were created between them based on the following rules:
 
@@
 
-The nodes were then assigned a unique id based on the unique connected component
-they belong to as shown in Figure \ref{methodology_schematic}. This unique
-identifier is used in the place of MAC address for aggregation for the
-anonymised probe requests. Though the recycling of sequence number after 4096
-leads to multiple unique ids being reported from a single device from a sample
-experiment done on randomised probe requests sent by "Google" devices we found
-that the sequence number is reset for around 0.05\% of them. We hence assume
-this to be inconsequential.
+The nodes were then assigned a unique ID based on the unique connected component
+they belonged to as shown in Figure \ref{methodology_schematic}. This unique
+identifier was used in the place of MAC addresses for aggregation of the
+anonymised probe requests. Although the recycling of sequence numbers after 4096
+led to multiple unique IDs being reported from a single device, a sample
+consisting of all randomised probe requests sent by "Google" devices showed that
+only 0.05\% of the sample had their sequence number reset. This led assume this
+to be inconsequential.
 
@@
 
-Since mobile device ownership is an external uncertainty to our study and could
-arise from variety of spatio - temporal and demographic factors, we propose to
-solve this using a sample count done manually at each location.  We can then
-calculate an  adjustment factor or an offset for each location by comparing the
-sensor based counts and ground truth and in turn it can be used to adjust the
-data reliably to reflect the ground truth in absolute numbers for the future.
-This calibration can be carried out periodically at these locations to improve
-the quality of the estimation.
+Since proportions of mobile device ownership was an external uncertainty to our
+study and could arise from variety of spatio - temporal and demographic factors,
+we aimed to solve this by using a manual sample count at each location. We then
+calculated an adjustment factor, or an `offset' for each location by comparing
+the sensor-based counts and ground truth. In turn it was then used to adjust the
+data reliably to reflect the ground truth in absolute numbers.  This calibration
+can be carried out periodically at these locations to improve the quality of the
+estimation.

### 2018-12-07 review/response.md
intro and conclusions corrected
final before preparations

@@
 ## First review comments
+--------------------------------------------------------------------------------
 
@@
 ## 1 Editor
+--------------------------------------------------------------------------------
 
@@
 ## 2 Reviewer-1
+--------------------------------------------------------------------------------
 
@@
 
-> Thank you. We have included a citiation to the source, as well as one other
+> Thank you. We have included a citation to the source, as well as one other
 > recent addition to this fast-developing literature.
@@
 > have made reference to this discussion, while also making clear the
-> anonymisation is ensured by the system architecture that processess and
+> anonymisation is ensured by the system architecture that processes and
 > curates the data.  Specifically we rotated the random salt used to hash the
-> MAC addresses regulary to prevent any kind of tracking of the users across
+> MAC addresses regularly to prevent any kind of tracking of the users across
 > a longer period of time with which the hashing process is carried out.
@@
 concept and refer to the "Wi-Fi network interface" (or Wi-Fi interface) as
-the the element broadcasting a signal.
+the element broadcasting a signal.
 
@@
 
-> Thank you. This point has been boted and accommodated into the revised
+> Thank you. This point has been noted and accommodated into the revised
 > version.
@@
 > Thank you very much for the suggestion. We carried out some tests on a
-> sample data of 1910 probe requests corresponding to just google phones with
+> sample data of 1910 probe requests comprising of google phones with
 > randomised MAC addresses captured in the pilot study and found that only
@@
 > significant way considering the context. We have also included this in our
-> revised manuscript for the benefit of future research.
+> revised manuscript.
 
@@
 
-4. "privacy infringement" may not be the best word as infrigement relate to a
+4. "privacy infringement" may not be the best word as infringement relate to a
 law or a regulation. Privacy breach or leak may be a better choice.
@@
 ## 3 Reviewer-2
+--------------------------------------------------------------------------------
 
@@
 > footpath in front of it, we make sure that all the devices beyond a certain
-> distance will report similarly low signal strength. This threhold is
-> idenfied using the kmeans algorithm to find the natural break in the data.
+> distance will report similarly low signal strength. This threshold is
+> identified using the k-means algorithm to find the natural break in the data.
 > We have found that this successfully isolates a farther stationary source
 > of large number of probe requests such as a phone shop or a restaurant next
-> door, Bustops etc.  This doesn't work when the source of error is within
+> door, Bus stops etc.  This doesn't work when the source of error is within
 > the range of the footfall. As shown in our examples 2,4. Here our
@@
 > searching recycles the sequence number 1-2 times in a 30 mins interval. On
-> average this is expected to be much less than this. when the device doesn't
+> average this is expected to be much less than this. When the device doesn't
 > probe as frequently. We have included more details regarding the cycling of
-> sequence numbers and have mentioned method to overcomet this as mentioned
+> sequence numbers and have mentioned method to overcome this as mentioned
 > by other reviewer. The change in implementing the method is estimated to be
@@
 ## 4 Reviewer-3
+--------------------------------------------------------------------------------
 
@@
 >
-> There was no reason for selecting the time in the pilot study. the location
+> There was no reason for selecting the time in the pilot study. The location
 > was chosen for it it known to have high number of randomised probe requests
 > and has one of the highest footfall in London. For the main study the
-> counting was done such that each location has atleast 2/3 counts in the
+> counting was done such that each location has at least 2/3 counts in the
 > same time interval.
@@
 > computing unfeasible, it is neither big nor fast enough for a large scale
-> big data system. An opensource and free toolkit comprising of text
+> big data system. An open source and free toolkit comprising of text
 > processing tools and relational database management system was enough to
-> handle the compexity, size and speed of the data. The toolkit employed here
+> handle the complexity, size and speed of the data. The toolkit employed here
 > involves PostgreSQL, Bash, Unix tools for text processing (sed, awk, sort,

### 2019-01-21 review/review_02/response.md
almost done

## Reviewer 1
--------------------------------------------------------------------------------

I appreciate the authors’ effort to address my comments in the revised draft.
The manuscript has been greatly improved. I have a few minor comments:

 1. p9 l36-40: When determining the threshold for high/low signal strength, did
    you just set the number of clusters equals to 2 for kmeans and hierarchical
    clustering? If so, you should explicitly say so in the explanation.

> Thanks for the feedback. We have included this explicitly in the revised
> manuscript.

 2. Please enlarge and recreate the sub-figures in Figure 3. This figure is
    difficult to read in its current format.

> We agree. A more legible version of Figure 3 has been included in the revised
> manuscript.

 3. I would suggest adding the following paper to the literature review:
    Torrens, P. M. (2008) Wi-Fi Geographies. Annals of the Association of
    American Geographers, 98(1), 59-84.

> Thanks for pointing out this important literature that we missed to cite.  We
> have included this reference in the literature review.

## Reviewer 2
--------------------------------------------------------------------------------

The revised version of this paper is of better quality and the authors have
taken into account reviewers' comments.

 1. However, Figure 3 needs a slight improvement. In the final version, all the
    figures must be of good quality, with similar formatting (size).

> We agree. A more legible version of Figure 3 has been included in the revised
> manuscript.

 2. Please check the format of the references.

> Thanks for pointing this out. We have checked and fixed the format of the
> references

More specific comments can be found in the following,

 3. Page 8, Could you give more description about your real-time experimental
    data? (Like data size, the total number of records)

> Thank you for the feedback. Similar to the Oxford street case study, we have
> included more descriptive statistics on the data collected under the case
> study in Section 5 - Table 2.

 4. Page 10, “we observed that the best thresholds for time and sequence
    numbers in the clustering algorithm is around 16 and 60 respectively”,
    Could you give more detail information?

> Thank you for the feedback. These thresholds are arrived at through a trial
> and error. We have further clarified the process along with a more legible
> illustration which is shown in Figure 3.

 5. Page 14. “There are obvious extensions to understanding the ebbs and flows
    of activities in the 24-hour smart city.” Could you give some suggestions
    to use your method in these extensions?

> Thanks for the input. We have included few examples and cited recent
> research demonstrating them.

### 2019-01-21 paper/sections/main_study.tex
almost done

@@
 \begin{table}
-	\tbl{Locations where sensors were installed}
-	{\begin{tabular}{clll} 
+    \tbl{Locations where sensors were installes, volume and speed of probe requests collected by
+    the sendor and total pedestrians manually counted. The data occupies around 1.8 GB on disk 
+    when encoded in text format.}
+	{\begin{tabular}{clllccc} 
 		\toprule
-		 ID & Location & Type & Installation notes\\
+         ID & Location & Type & Installation notes & Probe Requests & Footfall\\
+         & & & & x10\textsuperscript{6} (per min) & No. (per min)\\
 		 \midrule
-		 1 & Camden High Street & Phone Shop & Bus stop in front\\
-		 2 & Central St.Giles Piazza & Restaurant & Seating area on both sides\\
-		 3 & Holborn Underground Station & Information Kiosk & Overlooks station entrance\\
-		 4 & Brunswick Center & Fast Food Restaurant & Has seating area on one side\\
-		 5 & The Strand & Tea Shop & Has phone shop next door \\
+         1 & Camden High Street & Phone Shop & Bus stop in front & 9.9 (297) & 3683 (33)\\
+         2 & Central St.Giles & Restaurant & Seating area on both sides & 3.9 (169) & 0346 (05)\\
+         3 & Holborn Station & Info. Kiosk & Overlooks station entrance & 4.3 (303) & 2956 (46)\\
+         4 & Brunswick Center & Fast Food & Has seating area on one side & 3.4 (210) & 0960 (12)\\
+         5 & The Strand & Tea Shop & Has phone shop next door & 8.4 (382) & 1969 (21)\\
 		 \bottomrule
@@
 sensors. The position of the sensor at these locations with respect to the
-context is shown the Figure \ref{main_signal_strength}. We can see that Location
-5 is the `cleanest' with one clear stationary source of noise (phone shop)
-while location 2 is the most complex due to the proximity of seating areas
-to the sensor.  The sensors were operational through out February and March,
-while manual counts were conducted in these locations in half hour sessions on at
-least two different days. For the purposes of comparing with ground truth, we
-considered the data from sensors which correspond to the 12 sets of
-available manual counts. The schedule of data collection is shown in Figure
-\ref{main_schedule}.
+context is shown the Figure \ref{main_signal_strength}. We can see that
+Location 5 is the `cleanest' with one clear stationary source of noise (phone
+shop) while location 2 is the most complex due to the proximity of seating
+areas to the sensor.  The sensors were operational through out February and
+March, while manual counts were conducted in these locations in half hour
+sessions on at least two different days. For the purposes of comparing with
+ground truth, we considered the data from sensors which correspond to the 12
+sets of available manual counts. The schedule of data collection is shown in
+Figure \ref{main_schedule}.
 
@@
 thresholds are shown in Table \ref{errors_table}.  Confirming our intuition, we
-see that the location 2 has the most MAPE followed by location 4 while rest of
-them have highly reduced MAPE.  It is significant that this method alone
-reduces our margin of error by a 50 - 100\% from the raw counts without any
-cleaning.  This makes the signal strength filtering quick and ideal method for
-practical applications which doesn't require absolute numbers such as creating
-large aggregated indexes to show long term trends.  We also see that the
-success of the signal strength filtering can be improved significantly by
-installing sensors such that the pedestrians and noise are at different
-distances from the sensor and are the field of measurement is distinct from the
-surroundings however noisy it might be.
+see that the location 2 has the most MAPE followed by location 4, while the rest
+of them have highly reduced MAPE.  It is significant that this method alone
+reduces our margin of error by 50 - 100\% from the raw counts without any
+cleaning.  This makes the signal strength filtering a quick and ideal method for
+practical applications, one which doesn't require absolute numbers such as
+creating large aggregated indexes to show long-term trends.  We also found that
+the success of the signal strength filtering can be improved significantly by
+installing sensors so that the pedestrians and source of noise are at different
+distances from the sensor. This ensures that the distribution of signal
+strengths within the field of measurement is distinct from that of the
+surroundings.
 
-We then run the sequence numbers based clustering process on the rest of the
+We then ran the sequence numbers based clustering process on the rest of the
 probe requests to reduce the MAPE by almost 50 - 100\% on all the sensors except
@@
 is the only one with large amount of pedestrians very close to the sensor.  This
-may be causing the over filtering caused by the previous process.  We finally
-run the calibration process where we calculate the adjustment factors from the
-ratio between the manual counts to the counts calculated for the sample period
-as shown in Table \ref{errors_table}.  We the use them to adjust the counts to
-achieve a MAPE ranging from 10 - 50\%. We can see that the sensors with people
-moving right next to them tend to under-count with our methodology while sensors
-with seating next to them tend to over-count significantly. However, using the
-filtering process, we can reduce the error to almost 10\% closer to that of the
-ground truth.
+may be the reason behind the over filtering observed in the previous process.
+We finally ran the calibration process where we calculated the adjustment
+factors from the ratio between the manual counts to the sensor based counts for
+the sample period as shown in Table \ref{errors_table}.  We used them to adjust
+the counts to achieve a MAPE ranging from 10 - 50\%. We observed that the
+sensors with people moving right next to them tend to under-count with our
+methodology, while sensors with seating next to them tend to over-count
+significantly. However, using the filtering process, we can reduce the error to
+almost 10\% closer to that of the ground truth.
 

### 2019-01-21 paper/sections/pilot_study.tex
almost done

@@
-begin{table}
+\begin{table}
 \tbl{Comparison of clustering algorithms with a sample of 40000 probe requests}
@@
 quantile, hierarchical clustering, bagged clustering, fisher and jenks natural
-breaks. The results are shown in Table \ref{classification-table}. We found that
-while hierarchical clustering and jenks gave us fairly low errors, they were too
-resource intensive for practical use with a larger dataset. We also found that
-k-means gave the quickest results with the lowest MAPE, closely followed by
-quantile algorithm. The cut-off point or threshold for the collected data with
-which we could classify as high and low was -71 dBm. We then removed all the
-probe requests which reported `low signal strength' and repeated the same
-aggregation process as before to produce footfall count. This process resulted
-in a footfall count with a net MAPE of 30\%. Although the results are
-encouraging we are still not completely confident that our filtering process is
-removing noise or has any correlation the configuration of sensor or position of
-the mobile devices. These concerns need to be addressed with a larger survey
-with multiple locations of varying orientations.
+breaks with the number of clusters set as 2. The results are shown in Table
+\ref{classification-table}. We found that while hierarchical clustering and
+jenks gave us fairly low errors, they were too resource intensive for practical
+use with a larger dataset. We also found that k-means gave the quickest results
+with the lowest MAPE, closely followed by quantile algorithm. The cut-off point
+or threshold for the collected data with which we could classify as high and
+low was -71 dBm. We then removed all the probe requests which reported `low
+signal strength' and repeated the same aggregation process as before to produce
+footfall count. This process resulted in a footfall count with a net MAPE of
+30\%. Although the results are encouraging we are still not completely
+confident that our filtering process is removing noise or has any correlation
+the configuration of sensor or position of the mobile devices. These concerns
+need to be addressed with a larger survey with multiple locations of varying
+orientations.
 
@@
 \vspace{10pt}
-        \includegraphics [width=\linewidth,trim=10 30 10 75,clip]
-            {images/pilot_clustering_params.png}
+\includegraphics [width=0.85\linewidth,trim=10 15 10 15,clip]
+    {images/pilot_clustering_params_rev.png}
 \caption{The clustering process was repeated with both increasing sequence
-            number threshold ($\alpha$) and time threshold ($\beta$), until we
-            arrived at the lowest parameters where the know device (black line) is
-            clustered as a single device.}
+    number threshold ($\alpha$) and time threshold ($\beta$), until we arrived
+    at the lowest parameters where the know device (black line) is clustered as
+    a single device.}
 \label{pilot_clustering_params}
@@
 	{\resizebox*{0.45\linewidth}{!}
-	{\includegraphics{images/pilot_clustering.jpeg}}}
+	{\includegraphics{images/pilot_clustering_rev1.jpeg}}}
 \hspace{20pt}
@@
 
-The next challenge was to identify the probe requests which are generated by the
-same device irrespective of the MAC randomisation process. We used the algorithm
-defined in Section \ref{methodology} and assigned a unique identifier or
-signature to each probe request, independent of their the MAC addresses. Since
-we didn't know the nature or frequency of the MAC address randomisation process,
-we used the surveyor's mobile device as a reference. As the surveyor's device
-was being actively used to count pedestrians and it's Wi-Fi module was kept
-active without establishing connection to any network, the device was known to
-be continuously probing for new networks. We also knew that the OUI of the
-device was 'Google' and the device was regularly randomising it's MAC address,
-thus providing us an excellent reference with which we could optimise the
-parameters for our clustering algorithm. Using this reference device we observed
-that the threshold for time $\alpha$ and the threshold for sequence numbers,
-$\beta$ are 16 seconds and 60 respectively via trial and error. This process is
-shown in Figure \ref{pilot_clustering_params}. This was undertaken on top the
-filtering done based on signal strength, and only for the probe requests with
-randomised MAC addresses. Figure \ref{pilot_results} shows the results of this
-clustering process on a small set of randomised probe requests. The probe
-requests with different randomised MAC address are shown by the coloured points
-and the lines joining them show that those probe requests were most likely be
-generated by the same device. We finally aggregated the probe requests as we did
-before but with the device signature rather than MAC addresses. This results in
-a footfall count with a MAPE of -18\% compared to the manual count. A comparison
-of minute by minute counts resulting from different filtering processes along
-with the ground truth is shown in Figure \ref{pilot_results} illustrating the
-promising effectiveness of the methods.
+The next challenge was to identify the probe requests which are generated by
+the same device irrespective of the MAC randomisation process. We used the
+algorithm defined in Section \ref{methodology} and assigned a unique identifier
+or signature to each probe request, independent of their the MAC addresses.
+Since we didn't know the nature or frequency of the MAC address randomisation
+process, we used the surveyor's mobile device as a reference. As the surveyor's
+device was being actively used to count pedestrians and it's Wi-Fi module was
+kept active without establishing connection to any network, the device was
+known to be continuously probing for new networks. We also knew that the OUI of
+the device was 'Google' and the device was regularly randomising it's MAC
+address, thus providing us an excellent reference with which we could optimise
+the parameters for our clustering algorithm. We then, by increasing the
+thresholds in steps of 1, found the minimum possible threshold for time and
+sequence numbers at which the algorithm clusters the reference device properly
+without over clustering the other probe requests.  This process is shown in
+Figure \ref{pilot_clustering_params}. We observed that the threshold for time
+$\alpha$ and the threshold for sequence numbers, $\beta$ are 16 seconds and 60
+respectively This was undertaken on top the filtering done based on signal
+strength, and only for the probe requests with randomised MAC addresses. Figure
+\ref{pilot_results} shows the results of this clustering process on a small set
+of randomised probe requests. The probe requests with different randomised MAC
+address are shown by the coloured points and the lines joining them show that
+those probe requests were most likely be generated by the same device. We
+finally aggregated the probe requests as we did before but with the device
+signature rather than MAC addresses. This results in a footfall count with a
+MAPE of -18\% compared to the manual count. A comparison of minute by minute
+counts resulting from different filtering processes along with the ground truth
+is shown in Figure \ref{pilot_results} illustrating the promising effectiveness
+of the methods.
 

### 2019-01-22 stages/first_submission/sections/methodology.tex
added stages of submission

The primary aim of this research is to enable us to collect a series of probe requests and process them into a usable pedestrian footfall count. 
We do this by using a Wi-Fi receiver to collect probe requests broadcast by mobile devices, filtering out the background noise and aggregating them based on the device that generated them.
In this section, we begin by looking at the characteristics of probe requests in detail, device a methodology to collect these probe requests in public areas, examine the systemic biases and uncertainties in the data collection method and device data processing methods to overcome these challenges.
Finally we compare the processed footfall counts to the ground truth recorded by primary surveys.

Probe requests are a special type of management packets broadcast by Wi-Fi enabled devices as part of the various functions such as scanning for available APs, quick geolocation by triangulation based known APs, etc.
These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or model of the devices though there is some variation on the frequency and the information transmitted through them.
In some cases, such as Android devices, these are broadcast even when the Wi-Fi functionality has been turned off by the user.
Thus these signals can be used to reliably identify the presence of Wi-Fi enabled mobile devices.
Being a first step of connection initiated by the mobile device, these packets have information regarding the characteristics of the mobile device itself. Some of the key information we can infer from these requests are,
\begin{enumerate}
	\item \textbf{Media Access Control (MAC) address} which is an unique identifier for the wireless hardware of the mobile device,
	\item \textbf{Sequence number} of the request for the mobile device to keep track of the responses,
	\item \textbf{Time stamp} at which the request was received by the AP,
	\item Total \textbf{length} of the request in number of bits, and 
	\item The \textbf{strength of the signal} which transmitted the request.
\end{enumerate}
The MAC address is the primary unique identifier for the mobile device.
It has two parts, first part is the Organisation Unique Identifier (OUI) which gives information about the manufacturer of the device and the second is unique to the device. In modern devices, to protect the users privacy, the MAC address can also be randomised (hence non unique) and is marked as such. Though sequence number and length of the packet are not strictly unique, we hypothesize that we can use them to estimate the number unique devices.

Data collection was done with the help of custom sensors built from modifying the hardware used in Smart street sensors \citep{sss2016} and updating them with custom software.
The sensor is essentially a Raspberry-Pi device with Wi-Fi and 3G modules.
It keeps the Wi-Fi module in `Monitor' mode and uses the open source software - WireShark \citep{wireshark2} to passively collect all packets sent to `broadcast', marked with type as management' and subtype as `probe requests'.
The MAC address in these probe requests is anonymised at the device level using a cryptographic hashing algorithm and transmitted through 3G connection to a central database via web-sockets protocol, where it is stored in a PostgreSQL database for further analysis.
A overall schematic of the data collection and storage process is shown in Figure \ref{datacollection_schematic}.
The ground truth on number of pedestrian footfall was recorded using a custom mobile application - Clicker \citep{bala2018clicker}

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{images/datacollection_schematic.jpeg}
	\caption{Schematic diagram showing the process of collecting and storing probe requests using the sensor} \label{datacollection_schematic}
\end{figure}

The next step after collecting data was to estimate the footfall or pedestrian activity from them. We identified the following potential uncertainties which arise from our collection methodology.
\begin{enumerate}
	\item \textbf{Background noise} - since the extent to which Wi-Fi signals travel differs subject to various factors such as interference and humidity, it is close to impossible to restrict our data collection to a finite area of interest. This can lead to a significant background noise at certain locations. E.g. a phone shop or a bus stop located next to the study area can increase the number of probe requests received by the sensor.
	\item \textbf{MAC randomisation} - The mobile devices in recent years have been using randomised 'local' MAC addresses for probe requests to protect the users from being tracked. This makes it impossible to tell if the probe requests are being sent by the same mobile device which is being stationed next to the sensor. This along with the previous problem can further increase the magnitude of error by several fold.
	\item \textbf{Mobile ownership} - Since the rate of mobile ownership can vary widely across geography and demography, we cannot assume that every mobile device translates to one pedestrian footfall. In addition to this, there is a long term overall increase in mobile ownership which may lead to the number of probe requests collected overtime. 
\end{enumerate}
We propose the following internal and external validation methods to tackle each of these uncertainties.

\begin{figure}
	\centering
	\subfigure[Distribution of signal strengths showing the filtering of background noise]{
		\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_signalstrength.jpeg}}}\hspace{20pt}
	\subfigure[Clustering probe requests as nodes in a graph using increasing sequence numbers]{
	\resizebox*{0.45\linewidth}{!}{\includegraphics{images/methodology_clustering.jpeg}}}
	\caption{Schematic diagrams explaining the methods for filtering by signal strength and clustering using sequence numbers} \label{methodology_schematic}
\end{figure}

\subsection{Filtering with Signal Strength}
One of the clues that we can use to estimate the distance between the mobile device and the sensor is the strength of the signal received by the sensor. 
The obvious approach here is to try and establish a relationship between the signal strength and distance first and use this to filter out the unwanted probe requests.
This approach was found not to be feasible since the decay of signal strength with distance is not always constant.
It varies with atmospheric conditions, presence of obstructions between the source and target, the nature of these obstructions and the strength (power level) of the source.
This severely limits our ability in establishing a simple conversion between reported signal strength and distance.
There is a need for a method which takes in to account these variables across various locations.

We hypothesise that in configurations where a specific source of background noise is at a constant distance, there must be a distinct pattern in the number of probe requests reporting signal strength corresponding to that distance.
For example, if there is a phone shop next to our sensor where hundreds of phones regularly send probe requests there should be a sharp rise of number of probe requests with reported signal strength corresponding to the distance between the sensor and the phone shop irrespective of the local conditions as shown in Figure \ref{methodology_schematic}.
We could identify these breaks in the data using traditional one dimensional classification algorithms such as `jenks natural breaks', `k-means', `quantile' and `hierarchical clustering', etc.
Since we are only looking for the break in the data and not for absolute values, the methodology should apply for all the variations due micro site conditions thus reducing the overall noise in the collected data.

\subsection{Clustering with sequence numbers}

Since our primary unique identifier - MAC address, is being anonymised by new devices, we need to find other information present in the probe request for use as a unique identifier.
Obvious approach here is to establish a factor of randomisation and adjust the counts for these probe requests based on this factor.
We found this approach to be not feasible since the proportion of devices which randomise the MAC addresses increases over time.
There is also a wide variation in the frequency at which the devices randomise the MAC addresses and the method used for the process.
This lead us to look for a more generalisable approach which is independent of the device model.

From our initial analysis we found that OUI, length of the packet and sequence number of the packet being the most promising information to achieve this.
First we divide our dataset into sets of probe requests with randomised and non-randomised MAC addresses and keep the MAC address as the unique identifier for the latter set. 
For randomised ones we further divide them in to sub categories based on their OUI and length of the packet.
Since the length tends to stay unique to specific models of devices we are left with the task of identifying the unique mobile devices from within these sets.

The proposed algorithm creates a graph where the probe requests represented the nodes, and links are created between them based on the following rules: 
\begin{itemize}
	\item A link could go only forward in time. 
	\item A link could go from low to high sequence numbers.
	\item A link could exist between nodes with a maximum time difference of $\alpha$ - time threshold.
	\item A link could exist between nodes with a maximum sequence number difference of $\beta$ - sequence threshold.
	\item A node could have only one incoming link and one outgoing link, which is the shortest of all such possible links in terms of both time and sequence number.
\end{itemize}
The nodes were then classified based on the unique connected component they belong to as shown in Figure \ref{methodology_schematic}.
This classification was assigned as the unique identifier for the anonymised probe requests in the place of MAC address.
Though the recycling of sequence number after 4000 leads to multiple classifications reported on single device, the magnitude of error is greatly reduced.

\subsection{Calibrating with Ground Truth}
Since mobile device ownership is an external uncertainty to our study and could arise from variety of spatio - temporal and demographic factors, we propose to solve this using external source of information.
We hypothesize that an adjustment factor could be arrived at for each location of data collection, comparing the sensor based counts and ground truth and it can be used to adjust the data reliably to reflect the ground truth in absolute numbers for the future.
This calibration can be carried over periodically and the frequency to improve the quality of the estimation.

### 2019-01-22 stages/second_submission/sections/abstract.tex
added stages of submission

The accurate measurement of human activity with high spatial and temporal
granularity is crucial for understanding the structure and function of the built
environment.  With increasing mobile ownership, the Wi-Fi `probe requests'
generated by mobile devices can act as a cheap, scalable and real-time source of
data for establishing such measures.  The two major challenges we face in using
these probe requests for estimating human activity are: filtering the noise
generated by the uncertain field of measurement, and clustering anonymised probe
requests generated by the same devices together without compromising the privacy
of the users.  In this paper we demonstrate that we can overcome these
challenges by using class intervals and a novel graph based technique for
filtering and clustering the probe requests which in turn, enables us to
reliably measure real-time pedestrian footfall at retail high streets.

### 2019-01-22 stages/second_submission/sections/conclusion.tex
added stages of submission

--- 2018-11-26 paper/sections/conclusion.tex
@@
-Sentient technologies make possible measurement of the human activities that are
-the life blood of the smart city. Yet the data that they harvest are frequently
-relevant only to sub-groups within society that avail themselves of particular
-goods and services – such as social media applications, transport modes or
-retail offers. In each of these cases, it is necessary to remember that the
-resulting data are a by-products of consumer transactions, and will as a
-consequence, only pertain to users of the relevant good or service. If the smart
-city is to be socially inclusive, it therefore follows that sentient data must
-represent entire populations, whether by design or by triangulation with
-external, population wide, sources. This usually poses a non-trivial task, since
-the ebbs and flows of smart device-enabled citizens rarely pertain to any
-clearly defined population in either administrative or functional terms Massam
-1973.
+Sentient technologies make measurement of the human activities that are the life
+blood of the smart city possible. Yet the data that they harvest are frequently
+relevant only to the sub-groups within society that avail themselves of
+particular goods and services – such as social media applications, transport
+modes or retail offers. In each of these cases, it is necessary to remember that
+the resulting data are by-products of consumer transactions, and will as a
+consequence, only pertain to users of the relevant goods or services. If the
+smart city is to be socially inclusive, it therefore follows that sentient data
+must represent entire populations, whether by design or by triangulation with
+external, population wide, sources. This is a non-trivial task, since the ebbs
+and flows of smart device-enabled citizens rarely pertain to any clearly defined
+population in either administrative or functional terms
+\citep{massam1975location}.
 
@@
 collection and processing of a novel consumer Big Dataset that enables valid
-measures of levels of footfall activity which has been scaled across a Great
-Britain wide network of sensors \citep{cdrc2018consumer}.  In both conceptual and technical
+measures of levels of footfall activity which has been scaled across a wide
+network of sensors \citep{cdrc2018consumer}.  In both conceptual and technical
 terms, it illustrates the ways in which passively collected consumer data can be
@@
 identifying or tracking the mobile devices themselves. External validation then
-entails reconciling adjusted counts with the footfall observed at sample
+entailed reconciling adjusted counts with the footfall observed at sample
 locations. This procedure makes it possible to generalise from locations at
@@
 this methodology for a large network of sensors to gather granular data on
-pedestrian footfall. A snapshot showing week’s worth of precise footfall in area
-around Charring cross, London is shown in Figure 7 to demonstrate the potential
-for such dataset. Projects such as SmartStreetSensors \citep{cdrc2018consumer}, may utilise
-this methodology to overcome the challenges introduced by the implementation of
-MAC address randomisation.
+pedestrian footfall. A snapshot showing a week’s worth of precise footfall in
+area around Charring cross, London is shown in Figure \ref{main_study_counts} in
+order to demonstrate the potential for such a dataset. Projects such as
+SmartStreetSensors \citep{cdrc2018consumer}, may utilise this methodology to
+overcome the challenges introduced by the implementation of MAC address
+randomisation.
 
@@
 individuals across the smart city using this approach.  This can be modelled
-using agent-based methods \citep{heppenstall2011agent}, however. In our own research
-we have also begun to link store time-lagged till receipts to footfall, and have
-used such data to better understand the dwell times that characterise such
-different retail uses as stores with window displays and fast food restaurants.
-Such analysis not only provides a more nuanced picture of movement through
-retail areas, but also enables valorisation of micro sites within retail
+using agent-based methods \citep{heppenstall2011agent}, however. In our own
+research we have also begun to link store time-lagged till receipts to footfall,
+and have used such data to better understand the dwell times that characterise
+such different retail uses as stores with window displays and fast food
+restaurants.  Such analysis not only provides a more nuanced picture of movement
+through retail areas, but also enables valorisation of micro sites within retail
 centres. In the UK, for example, this is of immediate practical importance in
@@
 the setting of retail unit rental values. There are obvious extensions to
-understanding the ebbs and flows of activities in the 24 hour smart city.
+understanding the ebbs and flows of activities in the 24-hour smart city.
 
@@
 to seek to differentiate the quality of different elements within footfall
-according to mission e.g. travel to adjacent workplace zones, leisure, etc and
+according to mission e.g. travel to adjacent workplace zones, leisure, etc., and
 personal characteristics such as spending power. In this respect, future

### 2019-01-22 stages/second_submission/sections/main_study.tex
added stages of submission

--- 2019-01-21 paper/sections/main_study.tex
@@
 \begin{table}
-    \tbl{Locations where sensors were installes, volume and speed of probe requests collected by
-    the sendor and total pedestrians manually counted. The data occupies around 1.8 GB on disk 
-    when encoded in text format.}
-	{\begin{tabular}{clllccc} 
+	\tbl{Locations where sensors were installed}
+	{\begin{tabular}{clll} 
 		\toprule
-         ID & Location & Type & Installation notes & Probe Requests & Footfall\\
-         & & & & x10\textsuperscript{6} (per min) & No. (per min)\\
+		 ID & Location & Type & Installation notes\\
 		 \midrule
-         1 & Camden High Street & Phone Shop & Bus stop in front & 9.9 (297) & 3683 (33)\\
-         2 & Central St.Giles & Restaurant & Seating area on both sides & 3.9 (169) & 0346 (05)\\
-         3 & Holborn Station & Info. Kiosk & Overlooks station entrance & 4.3 (303) & 2956 (46)\\
-         4 & Brunswick Center & Fast Food & Has seating area on one side & 3.4 (210) & 0960 (12)\\
-         5 & The Strand & Tea Shop & Has phone shop next door & 8.4 (382) & 1969 (21)\\
+		 1 & Camden High Street & Phone Shop & Bus stop in front\\
+		 2 & Central St.Giles Piazza & Restaurant & Seating area on both sides\\
+		 3 & Holborn Underground Station & Information Kiosk & Overlooks station entrance\\
+		 4 & Brunswick Center & Fast Food Restaurant & Has seating area on one side\\
+		 5 & The Strand & Tea Shop & Has phone shop next door \\
 		 \bottomrule

### 2019-01-22 stages/second_submission/sections/methodology.tex
added stages of submission

--- 2018-12-07 paper/sections/methodology.tex
@@
 consisting of all randomised probe requests sent by "Google" devices showed that
-only 0.05\% of the sample had their sequence number reset. This led assume this
+only 0.5\% of the sample had their sequence number reset. This led assume this
 to be inconsequential.

### 2019-01-22 stages/third_submission/sections/conclusion.tex
added stages of submission

--- 2019-01-22 stages/second_submission/sections/conclusion.tex
@@
 
-The vicissitudes of MAC randomisation, and the provisions of privacy legislation
-such as EU General Data Protection Regulations mitigate against tracking
-individuals across the smart city using this approach.  This can be modelled
-using agent-based methods \citep{heppenstall2011agent}, however. In our own
-research we have also begun to link store time-lagged till receipts to footfall,
-and have used such data to better understand the dwell times that characterise
-such different retail uses as stores with window displays and fast food
-restaurants.  Such analysis not only provides a more nuanced picture of movement
-through retail areas, but also enables valorisation of micro sites within retail
-centres. In the UK, for example, this is of immediate practical importance in
-evaluating business rates on properties, and has still wider implications for
-the setting of retail unit rental values. There are obvious extensions to
-understanding the ebbs and flows of activities in the 24-hour smart city.
+The vicissitudes of MAC randomisation, and the provisions of privacy
+legislation such as EU General Data Protection Regulations mitigate against
+tracking individuals across the smart city using this approach.  This can be
+modelled using agent-based methods \citep{heppenstall2011agent}, however. In
+our own research we have also begun to link store time-lagged till receipts to
+footfall, and have used such data to better understand the dwell times that
+characterise such different retail uses as stores with window displays and fast
+food restaurants.  Such analysis not only provides a more nuanced picture of
+movement through retail areas, but also enables valorisation of micro sites
+within retail centres. In the UK, for example, this is of immediate practical
+importance in evaluating business rates on properties, and has still wider
+implications for the setting of retail unit rental values. There are obvious
+extensions to understanding the ebbs and flows of activities in the 24-hour
+smart city such as understanding urban mobility \citep{gariazzo2019spatio} and
+conceptualising them with a people dimension \citep{nam2011conceptualizing}.
 

### 2019-01-22 stages/third_submission/sections/previous_work.tex
added stages of submission

--- 2018-12-07 paper/sections/previous_work.tex
@@
 technologies in providing high speed internet access to mobile devices such as
-smartphones, tablets and laptops in public and private spaces. This has
-resulted in multiple Wi-Fi networks being available at almost every location in
-dense urban environments. Traversing through this overlapping mesh of Wi-Fi
-networks, modern mobile devices with Wi-Fi network interfaces regularly
-broadcast a special type of signal known as `Probe Requests' in order to
-discover the Wi-Fi networks available to them. This helps these devices to
-connect and switch between the Wi-Fi networks seamlessly.
+smartphones, tablets and laptops in public and private spaces
+\citep{torrens2008wi}. This has resulted in multiple Wi-Fi networks being
+available at almost every location in dense urban environments. Traversing
+through this overlapping mesh of Wi-Fi networks, modern mobile devices with
+Wi-Fi network interfaces regularly broadcast a special type of signal known as
+`Probe Requests' in order to discover the Wi-Fi networks available to them.
+This helps these devices to connect and switch between the Wi-Fi networks
+seamlessly.
 
@@
 easily and quickly estimate trajectories of the mobile devices
-\citep{Sorensenlocation, musa2012tracking} which can be used similary to the
+\citep{musa2012tracking} which can be used similary to the
 GPS trajectories to understand individual travel patterns \citep{reki2007,

### 2019-02-21 paper/sections/main_study.tex
final submission

@@
 	\begin{center}
-		\includegraphics [width=0.90\linewidth] {images/main_schedule.jpeg}
+		\includegraphics [width=0.90\linewidth] {images/figure_5.jpeg}
 		\caption{Data collection schedule showing the days when sensors were active at their corresponding locations. The red squares show that manual counting of pedestrians was also done on that day.}
@@
 	\subfigure[Installation configuration of sensors at the survey locations (not to scale)]{
-		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/main_configs.jpeg}}}\hspace{20pt}
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 6 20 6,clip]{images/figure_6a.jpeg}}}\hspace{20pt}
 	\subfigure[Density distribution of signal strength reported in collected probe requests (lower values show higher signal strength)]{
-		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/main_signals.jpeg}}}
+		\resizebox*{0.46\linewidth}{!}{\includegraphics[trim=20 4 25 6,clip]{images/figure_6b.jpeg}}}
 	\caption{Distribution of signal strengths across locations} \label{main_signal_strength}
@@
 	\begin{center}
-		\includegraphics [width=\linewidth,trim=6 6 6 6,clip] {images/main_comparison.jpeg}
+		\includegraphics [width=\linewidth,trim=6 6 6 6,clip] {images/figure_7.jpeg}
 		\caption{Comparison of the filtering process with the ground truth in all the locations.}
@@
 	\begin{center}
-		\includegraphics [width=\linewidth] {images/main_footfall.jpeg}
+		\includegraphics [width=\linewidth] {images/figure_8.jpeg}
 		\caption{A week of pedestrian footfall at the Strand, London collected by the methodology. The counts are aggregated for 5 minute intervals.}

### 2019-02-21 paper/sections/methodology.tex
final submission

@@
 	\centering \includegraphics[width=\linewidth]
-		{images/datacollection_schematic.jpeg}
+		{images/figure_1.jpeg}
 	\caption 
@@
 		background noise] {\resizebox*{0.45\linewidth}{!}
-		{\includegraphics{images/methodology_signalstrength.jpeg}}}
+		{\includegraphics{images/figure_2a.jpeg}}}
 \hspace{20pt}
@@
 		sequence numbers] {\resizebox*{0.45\linewidth}{!}
-		{\includegraphics{images/methodology_clustering.jpeg}}}
+		{\includegraphics{images/figure_2b.jpeg}}}
 \caption {Schematic diagrams explaining the methods for filtering by signal

### 2019-02-21 paper/sections/pilot_study.tex
final submission

@@
 \begin{center}
-\vspace{10pt}
-\includegraphics [width=0.85\linewidth,trim=10 15 10 15,clip]
-    {images/pilot_clustering_params_rev.png}
+\includegraphics [width=\linewidth,trim=1 1.5 1 1.5,clip]
+    {images/figure_3.jpeg}
 \caption{The clustering process was repeated with both increasing sequence
@@
 	present in them.]
-	{\resizebox*{0.45\linewidth}{!}
-	{\includegraphics{images/pilot_clustering_rev1.jpeg}}}
+	{\resizebox*{0.475\linewidth}{!}
+	{\includegraphics{images/figure_4a.jpeg}}}
 \hspace{20pt}
@@
 	{\resizebox*{0.45\linewidth}{!}
-	{\includegraphics{images/pilot_comparison.jpeg}}}
+	{\includegraphics{images/figure_4b.jpeg}}}
 \caption{The results of the pilot study demonstrating the validity of the
